### 2014-02-02 paper.tex
first commit

%\documentclass[journal]{vgtc}                % final (journal style)
\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{vgtc}     % electronic version, journal

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

%% Please note that the use of figures other than the optional teaser is not permitted on the first page
%% of the journal version.  Figures should begin on the second page and be
%% in CMYK or Grey scale format, otherwise, colour shifting may occur
%% during the printing process.  Papers submitted with figures other than the optional teaser on the
%% first page will be refused.

%% These three lines bring in essential packages: ``mathptmx'' for Type 1
%% typefaces, ``graphicx'' for inclusion of EPS figures. and ``times''
%% for proper handling of the times font family.

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
\usepackage{xspace}
\usepackage{amssymb}

% \usepackage[onecolumn]{multicol}

%% algorithm packages (llins)
\usepackage{algorithm}
\usepackage{microtype}
\usepackage{algpseudocode}

%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.

%% This turns references into clickable hyperlinks.
\usepackage[bookmarks,backref=true,linkcolor=black]{hyperref} %,colorlinks
\hypersetup{
  pdfauthor = {},
  pdftitle = {},
  pdfsubject = {},
  pdfkeywords = {},
  colorlinks=true,
  linkcolor= black,
  citecolor= black,
  pageanchor=true,
  urlcolor = black,
  plainpages = false,
  linktocpage
}

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}

%% allow for this line if you want the electronic option to work properly
\vgtcinsertpkg

%% In preprint mode you may define your own headline.
%\preprinttext{To appear in an IEEE VGTC sponsored conference.}

%% Paper title.

\title{RCloud: Integrating Exploratory Visualization, Analysis and
  Deployment}

%% This is how authors are specified in the journal style

%% indicate IEEE Member or Student Member in form indicated below
\author{Stephen North and Carlos Scheidegger and Simon Urbanek and
  Gordon Woodhull}

\authorfooter{
%% insert punctuation at end of each item
}

%other entries to be set up for journal
% \shortauthortitle{Biv \MakeLowercase{\textit{et al.}}: Global Illumination for Fun and Profit}
%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}

%% Abstract section.
\abstract{
} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Some, Keywords, Here}

%% \CCScatlist{ % not used in journal version
%%  \CCScat{K.6.1}{Management of Computing and Information Systems}%
%% {Project and People Management}{Life Cycle};
%%  \CCScat{K.7.m}{The Computing Profession}{Miscellaneous}{Ethics}
%% }

%% Uncomment below to include a teaser figure.
\teaser{
  \centering
  %% \includegraphics[width=.9\linewidth]{figs/teaser.jpg}
  % \vspace{-.5em}
  \caption{We'll tease you.
  }
% \vspace{-.5em}
}

%% Uncomment below to disable the manuscript note
%\renewcommand{\manuscriptnotetxt}{}

%% Copyright space is enabled by default as required by guidelines.
%% It is disabled by the 'review' option or via the following command:
% \nocopyrightspace

%
% Useful Macros 
%

\newcommand{\eg}{e.g.\xspace} % e.g. and i.e. ARE NOT ITALIC!!
\newcommand{\ie}{i.e.\xspace}
\newcommand{\todo}[1]{\textcolor{red}{#1}}

\input{commentcommands.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% ALGORITHM %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\renewcommand{\algorithmicthen}{}

% aux. commands

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.

%% the only exception to this rule is the \firstsection command

\firstsection{Introduction}

\maketitle

\section{Related Work}

\section{Architecture}

\section{Case Studies}

\section{Lessons Learned}

\section{Discussion and Limitations}

\section{Conclusions}

\bibliographystyle{abbrv}

\bibliography{paper}
\end{document}

### 2014-02-13 paper.tex
abstract I wrote for an upcoming talk

@@
 %% Abstract section.
-\abstract{
+\abstract{ Consider the emerging role of a data science team within an
+  organization. Different data scientists and statisticians usually
+  work on loosely related problems, and transition fluidly from
+  exploratory data analysis to automated diagnostics and reports,
+  which need to be deployed for wider consumption on the
+  organization. There are two problems with the current
+  practice. First, there is a gap in this workflow: EDA is performed
+  with one set of tools, and automated reports and deployments with
+  another. Second, current EDA environments mostly assume a
+  single-developer perspective, while data scientist teams could
+  greatly benefit from easier sharing of scripts, beyond that which a
+  traditional version control system can provide. In this paper, we
+  will present RCloud, a system designed to enable collaborative data
+  analysis, visualization and deployment on the web. We will discuss
+  the design decisions, tradeoffs and limitations, and compare it to
+  current proposals.
 } % end of abstract

### 2014-02-14 paper.tex
bit of related work

@@
 
-\section{Architecture}
+Kandel's interview study points out the typical ``explore'',
+``model'', ``report'' cycle~\cite{Kandel:2012:EDA}. RCloud seeks to
+reduce the impedance mismatch.
+
+Heer and Agarwala identify a large number of design considerations for
+collaborative visual analytics~\cite{Heer:2008:DCF}. \emph{Starring},
+the system for signaling interest in notebooks, described in
+Section~\ref{sec:starring}, addresses social-psychological incentives,
+recommendation, and voting and ranking. RCloud's integrated deployment
+mechanism, described in Section~\ref{sec:deployment}, adresses cost of
+integration, content export, presentation and view sharing. Notebooks,
+and the integrated version control system for them, described in
+Section~\ref{sec:notebooks}, addresses modularity and granularity, and
+artifact histories.
+
+\section{System Architecture}
+
+\section{System Design}
+
+\subsection{Notebooks\label{sec:notebooks}}
+
+\subsection{Reputation and Interest: starring\label{sec:starring}}
+
+\subsection{Deployment of notebooks\label{sec:deployment}}
 

### 2014-02-18 paper.tex
Trying to make some headway.

@@
 \abstract{ Consider the emerging role of a data science team within an
-  organization. Different data scientists and statisticians usually
-  work on loosely related problems, and transition fluidly from
-  exploratory data analysis to automated diagnostics and reports,
-  which need to be deployed for wider consumption on the
-  organization. There are two problems with the current
-  practice. First, there is a gap in this workflow: EDA is performed
-  with one set of tools, and automated reports and deployments with
-  another. Second, current EDA environments mostly assume a
-  single-developer perspective, while data scientist teams could
-  greatly benefit from easier sharing of scripts, beyond that which a
-  traditional version control system can provide. In this paper, we
-  will present RCloud, a system designed to enable collaborative data
-  analysis, visualization and deployment on the web. We will discuss
-  the design decisions, tradeoffs and limitations, and compare it to
-  current proposals.
+  organization. Individual data scientists and statisticians usually
+  work on loosely related problems, and must devise effective ways to 
+  share their findings and to move results from exploratory data analysis
+  to automated diagnostics and reports deployed for wider consumption.
+  There are two problems with the current
+  practice. First, there are gaps in this workflow: EDA is performed
+  with one set of tools, and automated reports and deployments with another.
+  Second, EDA environments often assume a single-developer perspective,
+  while data scientist teams could greatly benefit from easier sharing of
+  scripts and data feeds, experiments, annotations, and automated
+  recommendations, which are well beyond what traditional version control
+  systems provide.  In this paper, we propose RCloud, a system that supports
+  collaborative data analysis, visualization and web deployment. We will
+  discuss the design decisions, tradeoffs and limitations, and compare RCloud
+  to other current proposals.
 } % end of abstract
@@
 ``model'', ``report'' cycle~\cite{Kandel:2012:EDA}. RCloud seeks to
-reduce the impedance mismatch.
+reduce the impedance mismatch. Kandel et al point out that larger teams
+are becoming more common in data analysis, that supporting
+collaboration is a difficult and important problem, and that
+sharing and versioning of data sources and artifacts was hindered
+by technology.  ``We found that analysts typically did not share scripts
+with each other. Scripts that were shared were disseminated similarly to
+intermediate data: either through shared drives or email.
+Analysts rarely stored their analytic code in source control.''
+Their work points to the need for better technology to support
+these aspects of data analysis in teams.
 
@@
 recommendation, and voting and ranking. RCloud's integrated deployment
-mechanism, described in Section~\ref{sec:deployment}, adresses cost of
+mechanism, described in Section~\ref{sec:deployment}, addresses cost of
 integration, content export, presentation and view sharing. Notebooks,
@@
 artifact histories.
+
+Our work has been inspired or has benefited from other modern technologies
+to improve the data analysis environments and processes, such as RStudio,
+R packages such as Markdown, knitr and Shiny, and iPython notebooks, to name a few.  % explain
 
@@
 
+% limitations
+Previous studies have pointed out frequent difficulties in the
+achieving production-level flexibility, scalability and maintainability
+with experimental code created by data analysts.  These properties cannot
+be ensured by any programming environment, but a reasonable approach in
+general is to provide suitable tools that make it easier for well-intentioned
+programmers to adopt such practices.
+
+% security issues
+
 \section{Conclusions}

### 2014-02-18 paper.tex
working

@@
 
-\title{RCloud: Integrating Exploratory Visualization, Analysis and
-  Deployment}
+\title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
 
@@
 Kandel's interview study points out the typical ``explore'',
-``model'', ``report'' cycle~\cite{Kandel:2012:EDA}. RCloud seeks to
-reduce the impedance mismatch. Kandel et al point out that larger teams
-are becoming more common in data analysis, that supporting
+``model'', ``report'' cycle~\cite{Kandel:2012:EDA}. There
+are many discontinuities in this cycle that cost time and
+effort to overcome.  RCloud seeks to reduce this impedance
+mismatch. Kandel et al also point out that larger teams are
+becoming more common in data analysis, that supporting
 collaboration is a difficult and important problem, and that
-sharing and versioning of data sources and artifacts was hindered
-by technology.  ``We found that analysts typically did not share scripts
-with each other. Scripts that were shared were disseminated similarly to
-intermediate data: either through shared drives or email.
-Analysts rarely stored their analytic code in source control.''
-Their work points to the need for better technology to support
-these aspects of data analysis in teams.
+sharing and versioning of data sources and artifacts is hindered
+by current technology in practice. ``We found that analysts typically
+did not share scripts with each other. Scripts that were shared were
+disseminated similarly to intermediate data: either through shared
+drives or email. Analysts rarely stored their analytic code in
+source control.'' Their work points to the opportunity for better
+technology to support collaboration and sharing by data analysis teams.
 
-Heer and Agarwala identify a large number of design considerations for
+Heer and Agarwala identify many design considerations for
 collaborative visual analytics~\cite{Heer:2008:DCF}. \emph{Starring},
@@
 and the integrated version control system for them, described in
-Section~\ref{sec:notebooks}, addresses modularity and granularity, and
+Section~\ref{sec:notebooks}, address modularity and granularity, and
 artifact histories.
 
-Our work has been inspired or has benefited from other modern technologies
-to improve the data analysis environments and processes, such as RStudio,
-R packages such as Markdown, knitr and Shiny, and iPython notebooks, to name a few.  % explain
+Our work has been inspired by and has benefited from other recent projects
+to improve data analysis environments and processes, including RStudio,
+R packages such as Markdown, knitr and Shiny, and iPython notebooks,
+to name a few. RStudio aims at providing an integrated development environment
+for R programming, with support for publishing code in packages. Markdown,
+knitR and Shiny provide R with sophisticated reporting capabilities, including
+interactive web interfaces. iPython \cite{Perez:2007:ISI:1251563.1251831}
+shares many of our goals...
 
@@
 % limitations
-Previous studies have pointed out frequent difficulties in the
-achieving production-level flexibility, scalability and maintainability
-with experimental code created by data analysts.  These properties cannot
-be ensured by any programming environment, but a reasonable approach in
-general is to provide suitable tools that make it easier for well-intentioned
-programmers to adopt such practices.
-
-% security issues
-
-\section{Conclusions}
+Previous studies have pointed out difficulties in achieving production-level
+flexibility, scalability and maintainability with experimental code created
+by data analysts. These properties cannot be ensured by any programming environment,
+but suitable tools can help well-intentioned analysts and programmers to create
+robust applications with less effort.
 

### 2014-03-08 paper.tex
a few new references, much more needed

@@
 
+"Coordination is done by meetings"
+
+"how do you trace an automated alert back to an EDA environment ?"
+
 \maketitle
@@
 
-Kandel's interview study points out the typical ``explore'',
-``model'', ``report'' cycle~\cite{Kandel:2012:EDA}. There
-are many discontinuities in this cycle that cost time and
-effort to overcome.  RCloud seeks to reduce this impedance
-mismatch. Kandel et al also point out that larger teams are
-becoming more common in data analysis, that supporting
-collaboration is a difficult and important problem, and that
-sharing and versioning of data sources and artifacts is hindered
-by current technology in practice. ``We found that analysts typically
-did not share scripts with each other. Scripts that were shared were
+Kandel et al. argue in their review of data wrangling work that data
+cleaning, wrangling and transformation is a major part of exploratory
+analysis and visualization~\cite{Kandel:2011:RDI}. Although RCloud
+does not by itself include modules specific to data cleaning and
+wrangling, we note from internal experience that these cleaning
+scripts themselves tend to change over time. RCloud addresses this
+issue by providing easy publishing of data-cleaning notebooks as web
+services, which reduces the total data-cleaning effort across an
+organization.
+
+Kandel et al.'s interview study points out the typical ``explore'',
+``model'', ``report'' cycle in enterprise data
+analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
+this cycle that cost time and effort to overcome.  RCloud seeks to
+reduce this impedance mismatch. They also point out that larger teams
+are becoming more common in data analysis, that supporting
+collaboration is a difficult and important problem, and that sharing
+and versioning of data sources and artifacts is hindered by current
+technology in practice. ``We found that analysts typically did not
+share scripts with each other. Scripts that were shared were
 disseminated similarly to intermediate data: either through shared
-drives or email. Analysts rarely stored their analytic code in
-source control.'' Their work points to the opportunity for better
-technology to support collaboration and sharing by data analysis teams.
+drives or email. Analysts rarely stored their analytic code in source
+control.'' Their work points to the opportunity for better technology
+to support collaboration and sharing by data analysis teams.
 
@@
 knitR and Shiny provide R with sophisticated reporting capabilities, including
-interactive web interfaces. iPython \cite{Perez:2007:ISI:1251563.1251831}
+interactive web interfaces. IPython \cite{Perez:2007:ISI:1251563.1251831}
 shares many of our goals...
@@
 \subsection{Deployment of notebooks\label{sec:deployment}}
+
+Notebooks as versioned subroutines, web services.
+
+Notebooks by default are visible by the entire organization: there
+exists a URL for every notebook in RCloud. This is deliberate. As
+pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS}, broad
+access to analysis outputs (in their case, in the form of NameVoyager)
+increases long-term engagement partly by the crossreferences in the
+web. Our main RCloud deployment is only visible inside an intranet,
+but we have nevertheless found anecdotal support for this theory by
+noticing links to RCloud notebooks in internal discussion fora and
+mailing lists.
 
@@
 
+Asynchronous collaborative visual analytics
+(ACVA)~\cite{Chen:2011:SEC}. This paper addresses visualization
+\emph{of} the ACVA process. It will be important when we talk about
+recommendation systems, and navigation of the set of notebooks, etc.
+
 \bibliographystyle{abbrv}

### 2014-03-12 paper.tex
working on refs
working

@@
 collaborative visual analytics~\cite{Heer:2008:DCF}. \emph{Starring},
-the system for signaling interest in notebooks, described in
+the means for signaling interest in notebooks, described in
 Section~\ref{sec:starring}, addresses social-psychological incentives,
@@
 
+Manyeyes \cite{Viegas:2007:MAS} was a landmark system for the integration
+of social media with visualization and data publishing. We build on this
+work by defining a rich interface for collaboration about code and for
+operating on metadata.
+
+The need for integrating statistics and visualization has been
+highlighted in previous studies and is widely understood by
+various technical communities \cite{Perer:2008:ISA}
+
+There has been considerable work on specific techniques such as
+social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV} that could benefit data scientists and deployment
+specialists if they were made available, together, in a broader environment.
+
+\cite{Santos:2009:VST}
+
+\cite{Mates:2011:CSA}
+
 Our work has been inspired by and has benefited from other recent projects
@@
 knitR and Shiny provide R with sophisticated reporting capabilities, including
-interactive web interfaces. IPython \cite{Perez:2007:ISI:1251563.1251831}
+interactive web interfaces. IPython \cite{Perez:2007:IAS}
 shares many of our goals...
+
+Reducing gap between implementors and deployers. Don't want a team of
+20 IT people to support 5 data scientists.
+
+% devops for data science.
 

### 2014-03-12 paper.tex
working

@@
 
+\cite{Keim:2008:VAS}
+
 Our work has been inspired by and has benefited from other recent projects
@@
 % limitations
-Previous studies have pointed out difficulties in achieving production-level
-flexibility, scalability and maintainability with experimental code created
-by data analysts. These properties cannot be ensured by any programming environment,
-but suitable tools can help well-intentioned analysts and programmers to create
-robust applications with less effort.
+Previous studies have pointed out difficulties in achieving the flexibility,
+scalability and maintainability expected of production software with experimental
+code created by data analysts.
+These properties cannot be ensured by any programming environment,
+but suitable tools can help well-motivated analysts and programmers to create
+robust applications with much less effort.
 

### 2014-03-12 paper.tex
working

@@
 There has been considerable work on specific techniques such as
-social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV} that could benefit data scientists and deployment
-specialists if they were made available, together, in a broader environment.
+social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
+to support collaborative or social 
+Our goal is to provide a productive environment in which such
+techniques may be integrated and made available to a community.
+Similarly, there are computational methods to support
+high performance execution in an incremental code development
+environment \cite{Guo:2010:TPI}.
 
@@
 to improve data analysis environments and processes, including RStudio,
-R packages such as Markdown, knitr and Shiny, and iPython notebooks,
+R packages such as Markdown, knitr and Shiny \cite{CRAN}, and iPython notebooks,
 to name a few. RStudio aims at providing an integrated development environment
@@
 
+discussion about binding times, running vs.caching.
+transparent vs. explicit operations done by programmers
+
 % limitations

### 2014-03-12 paper.tex
knitr cite
update

@@
 
+% DevOps for Visual Analysis and Data Science
+
 %% This is how authors are specified in the journal style
@@
   recommendations, which are well beyond what traditional version control
-  systems provide.  In this paper, we propose RCloud, a system that supports
+  systems provide.  \todo{sell idea, not system} In this paper, we propose RCloud, a system that supports
   collaborative data analysis, visualization and web deployment. We will
@@
 
-"Coordination is done by meetings"
+\maketitle
 
-"how do you trace an automated alert back to an EDA environment ?"
+\todo{Points we want to make."Coordination is done by meetings" "how do you trace an automated alert back to an EDA environment ?"}
 
-\maketitle
+% What's the area
+Consider the emerging role of a data science team within an
+organization.
+
+% How do people do it today?
+Individual data scientists and statisticians usually
+work on loosely related problems, and must devise effective ways to 
+share their findings and to move results from exploratory data analysis
+to automated diagnostics and reports deployed for wider consumption.
+
+% Why does this not work?
+There are two problems with the current
+practice. First, there are gaps in this workflow: EDA is performed
+with one set of tools, and automated reports and deployments with another.
+Second, EDA environments often assume a single-developer perspective,
+while data scientist teams could greatly benefit from easier sharing of
+scripts and data feeds, experiments, annotations, and automated
+recommendations, which are well beyond what traditional version control
+systems provide.
+
+% What are we going to do about this?
+\todo{Sell idea, not system} In this paper, we propose RCloud, a system that supports
+collaborative data analysis, visualization and web deployment. We will
+discuss the design decisions, tradeoffs and limitations, and compare RCloud
+to other current proposals.
+
+% What is the scope of our solution?
+
+% We want our solution to play well with the rest of the ecosystem
 
@@
 
+Design for cloud-friendly? This goes back to the point in the
+introduction about playing nice with the rest of the ecosystem.
+
 \section{System Design}
@@
 
+\subsection{Technologies: R, Python, HTML5, interactive notebooks, etc.}
+
+How do we do things that are not trivial to do with IPython (for example)
+
+dcplot. two-way communication between between backend session and
+frontend session.
+
 \section{Case Studies}
@@
 \section{Discussion and Limitations}
+
+What are the decisions and ideas that are central to this paper?
 

### 2014-03-14 paper.tex
yes

@@
 
-There has been considerable work on specific techniques such as
+There has been notable work on specific techniques such as
 social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
-to support collaborative or social 
-Our goal is to provide a productive environment in which such
-techniques may be integrated and made available to a community.
-Similarly, there are computational methods to support
-high performance execution in an incremental code development
-environment \cite{Guo:2010:TPI}.
+and crowdsourcing \cite{Fast:2014:ECS} to support collaborative
+or social development or analysis processes.
+Similarly, there are computational methods to support high
+performance execution in incremental code development
+environments \cite{Guo:2010:TPI}.
+Our goal is to define an environment in which many such
+techniques may be integrated and made available to a broad community.
 
@@
 interactive web interfaces. IPython \cite{Perez:2007:IAS}
-shares many of our goals...
+shares many of our goals, such as providing a comprehensive environment
+for analysis and programming, with sharable documents on the web.
 

### 2014-03-14 paper.tex
yes

@@
 Our work has been inspired by and has benefited from other recent projects
-to improve data analysis environments and processes, including RStudio,
-R packages such as Markdown, knitr and Shiny \cite{CRAN}, and iPython notebooks,
+to improve data analysis environments and processes,
+including RStudio \cite{RStudio:2013:SWA},
+R packages such as Markdown \cite{Allaire:2014:MMR},
+knitr \cite{Xie:2013:DDW}
+and Shiny \cite{RStudio:2013:SWA},
+and iPython notebooks \cite{Perez:2007:IAS}
 to name a few. RStudio aims at providing an integrated development environment

### 2014-03-14 paper.tex
yes

@@
 Consider the emerging role of a data science team within an
-organization.
+organization \cite{Keim:2008:VAS}.
 
@@
 to automated diagnostics and reports deployed for wider consumption.
+They generally rely on a patchwork of resources that are shared informally,
+or at best, through systems that attack some of the tasks or functions in
+the data analysis workcycle.
 
@@
 There are two problems with the current
-practice. First, there are gaps in this workflow: EDA is performed
-with one set of tools, and automated reports and deployments with another.
+practice. First, there are gaps in this workflow:
+Exploratory Data Analysis (EDA) is performed with one set of tools,
+and automated reports and deployments with another.
 Second, EDA environments often assume a single-developer perspective,
 while data scientist teams could greatly benefit from easier sharing of
-scripts and data feeds, experiments, annotations, and automated
-recommendations, which are well beyond what traditional version control
-systems provide.
+scripts, data sets, data feeds, experiments, annotations, and automated
+recommendations, which are well beyond what traditional version control systems provide.
 
 % What are we going to do about this?
-\todo{Sell idea, not system} In this paper, we propose RCloud, a system that supports
-collaborative data analysis, visualization and web deployment. We will
-discuss the design decisions, tradeoffs and limitations, and compare RCloud
-to other current proposals.
+\todo{Sell idea, not system} In this paper, we investigate the hypothesis
+that the practice of visual analytics can advance by adopting techniques
+from information retrieval and collaborative computer, formally supported
+within the computing environment. To investigate this, we propose RCloud,
+a system that supports collaborative data analysis, visualization and
+web deployment. We will discuss the design decisions, tradeoffs and
+limitations, and compare RCloud to other current proposals.
 

### 2014-03-14 paper.tex
arch. figure
yes

@@
 % What is the scope of our solution?
+Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
+for statistical computing and graphics. R is one of the most popular
+systems for ...
 
 % We want our solution to play well with the rest of the ecosystem
+R has developed a complex ecosystem of compatible packages, software
+tools and other resources. For example there are packages that make
+it convenient to execute map-reduce programs, generate web interfaces,
 
@@
 
+\begin{figure}
+\hspace{-0.08\linewidth}
+\includegraphics[width=1.1\linewidth]{fig/system/system.pdf}
+\caption{\label{fig:system}A diagram of RCloud's architecture}
+\end{figure}
+
 Design for cloud-friendly? This goes back to the point in the

### 2014-03-14 paper.tex
yes

@@
 
-There has been notable work on specific techniques such as
+There has been noteworthy work on specific techniques such as
 social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
@@
 
+VisMashup
 \cite{Santos:2009:VST}
 
+Crowdlabs
 \cite{Mates:2011:CSA}
 
-\cite{Keim:2008:VAS}
-
-Our work has been inspired by and has benefited from other recent projects
+Our work has been inspired by and benefited from other proposals
 to improve data analysis environments and processes,
@@
 
-Reducing gap between implementors and deployers. Don't want a team of
-20 IT people to support 5 data scientists.
+One overall goal is to reduce the gap between implementors and deployers
+of technology.  Don't want a team of 20 IT people to support 5 data scientists.
 

### 2014-03-14 paper.tex
worked on requirements

@@
 
+\todo{Define the requirements for the system}
+\begin{itemize}
+\item Technology transfer.
+Things developed by analysts need to become accessible in the production environment and should be supported in the environment, making it easy to push new features, experiments, etc.
+\item Dasboarding.
+There should be a mode in which people can operate the experiments without coding.
+At times code has to be written because some things cannot be done any other way but other people must not ever see code because their eyes will fall out.
+\item Version control as a primary concept in the repository. Need to control the visibility of 
+\item Modern interactive technologies.
+People want to interact with data to explore it.
+Scalable technologies like crossfilter exist.
+Want to support those in the context of an R environment.
+Need to support full 2-way communication between computing in the cloud and interaction on the web.
+\item Other requirements e.g.
+Kandel's user study.
+Not our contribution but things we must address, mentioned in related work section.
+Heer/Agarwala design considerations for collaborative visual analytics.
+\end{itemize}
+
 % What is the scope of our solution?

### 2014-03-18 paper.tex
fix typo
screenshot for teaser
vismashup and crowdlabs

@@
   \centering
+  \includegraphics[width=.3\linewidth]{fig/screenshots/wdcplotsmall.jpg}
   %% \includegraphics[width=.9\linewidth]{figs/teaser.jpg}
   % \vspace{-.5em}
-  \caption{We'll tease you.
+  \caption{Left: RCloud integration of HTML5 visualization libraries such
+    as D3~\cite{Bostock:2011:DDD} and Crossfilter~\cite{Square:2014:CFM}.
   }
@@
 Not our contribution but things we must address, mentioned in related work section.
-Heer/Agarwala design considerations for collaborative visual analytics.
+Heer/Agrawala design considerations for collaborative visual analytics.
 \end{itemize}
@@
 
-VisMashup
-\cite{Santos:2009:VST}
-
-Crowdlabs
-\cite{Mates:2011:CSA}
+VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
+are tools built on top of the VisTrails workflow management
+system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
+semantics for automatically deriving user interfaces from workflows,
+while Crowdlabs exposes these capabilities on a website feature
+workflow upload and remote execution. In our view, the impedance
+mismatch between a dataflow pipeline specification and the power of a
+general-purpose language is too great for the type of general
+exploratory work in data science teams. As a result, RCloud tries to
+provide a closer data-analysis to analysts accustomed to createing and
+executing R and Python code while keeping many of the attractive
+properties like automatic versioning and management.
 
@@
 \begin{figure}
-\hspace{-0.08\linewidth}
-\includegraphics[width=1.1\linewidth]{fig/system/system.pdf}
+\includegraphics[width=\linewidth]{fig/system/system.pdf}
 \caption{\label{fig:system}A diagram of RCloud's architecture}

### 2014-03-18 paper.tex
cloud-friendly pargraphs

@@
 \includegraphics[width=\linewidth]{fig/system/system.pdf}
-\caption{\label{fig:system}A diagram of RCloud's architecture}
+\caption{\label{fig:system}A diagram of RCloud's architecture. Dashed
+  lines represent features which not yet implemented.}
 \end{figure}
 
-Design for cloud-friendly? This goes back to the point in the
-introduction about playing nice with the rest of the ecosystem.
+The internal computing infrastructure of organizations has changed
+radically in the last fifteen years or so. Much of the move toward the
+``cloud'' brought with it an ecosystem of processes distributed over a
+network (many times virtually defined) that communicate over a
+high-level protocol of some sort. Messages are typically exchanged
+over HTTP. HTTP is attractive because web browsers and servers are
+ubiquituous and available in essentially all pieces hardware today,
+from tiny sensors in sensor networks to handheld devices to laptops to
+rack-mounted servers.
+
+As a result, in our view HTTP has become the lingua franca of IPC in
+an intranet. One of the design goals for RCloud was to provide an
+attractive environment for creators of data-analysis scripts, but at
+the same time to create a system that behaves as a first-class citizen
+in the ecosystem of computer services and networks with an
+organization.
+
+%% Design for cloud-friendly? This goes back to the point in the
+%% introduction about playing nice with the rest of the ecosystem.
 

### 2014-03-18 paper.tex
yes

@@
 
-\title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
-
-% DevOps for Visual Analysis and Data Science
+% \title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
+\title{RCloud: DevOps for Visual Analysis and Data Science}
 
@@
   work on loosely related problems, and must devise effective ways to 
-  share their findings and to move results from exploratory data analysis
-  to automated diagnostics and reports deployed for wider consumption.
+  share their findings and move results from exploratory data analysis
+  into automated diagnostics and reports deployed for wider consumption.
   There are two problems with the current
@@
   Second, EDA environments often assume a single-developer perspective,
-  while data scientist teams could greatly benefit from easier sharing of
-  scripts and data feeds, experiments, annotations, and automated
+  while data scientist teams would get much benefit from easier sharing
+  of scripts and data feeds, experiments, annotations, and automated
   recommendations, which are well beyond what traditional version control
@@
 Consider the emerging role of a data science team within an
-organization \cite{Keim:2008:VAS}.
+organization today \cite{Keim:2008:VAS}.
 
 % How do people do it today?
-Individual data scientists and statisticians usually
-work on loosely related problems, and must devise effective ways to 
-share their findings and to move results from exploratory data analysis
-to automated diagnostics and reports deployed for wider consumption.
-They generally rely on a patchwork of resources that are shared informally,
-or at best, through systems that attack some of the tasks or functions in
-the data analysis workcycle.
+Individual data scientists and statisticians usually work on loosely
+related problems, and must devise effective ways to share their findings
+and move results from exploratory data analysis to automated diagnostics
+and reports that are deployed for wider consumption.
+They generally rely on a patchwork of resources shared informally,
+or at best, through systems that attack individual tasks or functions
+in the data analysis workcycle.
 
@@
 There are two problems with the current
-practice. First, there are gaps in this workflow:
-Exploratory Data Analysis (EDA) is performed with one set of tools,
-and automated reports and deployments with another.
+practice. First, there are gaps in the workflow:
+Exploratory Data Analysis (EDA) is done with one set of tools,
+and automated reports and deployments with another. This makes
+transferring results from EDA to production slow, expensive, and error-prone.
 Second, EDA environments often assume a single-developer perspective,
-while data scientist teams could greatly benefit from easier sharing of
+while data scientist teams could greatly benefit from convenient sharing of
 scripts, data sets, data feeds, experiments, annotations, and automated
-recommendations, which are well beyond what traditional version control systems provide.
+analyses and recommendations. These features are well beyond what existing
+repositories and version control systems provide.
 
@@
 that the practice of visual analytics can advance by adopting techniques
-from information retrieval and collaborative computer, formally supported
-within the computing environment. To investigate this, we propose RCloud,
-a system that supports collaborative data analysis, visualization and
-web deployment. We will discuss the design decisions, tradeoffs and
-limitations, and compare RCloud to other current proposals.
+from information retrieval and collaborative work environments. To investigate
+these ideas, we designed and deployed a prototype environment called RCloud,
+that supports collaborative data analysis, visualization and web deployment.
+We will report on the design decisions, tradeoffs and limitations, and compare
+the concepts in RCloud to similar proposals.
 
 \todo{Define the requirements for the system}
-\begin{itemize}
-\item Technology transfer.
-Things developed by analysts need to become accessible in the production environment and should be supported in the environment, making it easy to push new features, experiments, etc.
-\item Dasboarding.
-There should be a mode in which people can operate the experiments without coding.
-At times code has to be written because some things cannot be done any other way but other people must not ever see code because their eyes will fall out.
-\item Version control as a primary concept in the repository. Need to control the visibility of 
+
+We begin by noting the main requirements for a data analysis environment for
+collaboration and transfer of results into production.
+
+begin{itemize}
+\item Technology transfer. Experiments in notebooks developed by analysts need to become accessible in the production environment, making it easy to push new features, prototypes, etc.
+\item Dashboarding. To support the above, there should be a mode in which non-experts
+can operate the experiments without coding.
+\item Version control as a primary concept in the repository. There needs to be a way to control the visibility of prototypes as they are being developed and tested.
 \item Modern interactive technologies.
@@
 code created by data analysts.
-These properties cannot be ensured by any programming environment,
-but suitable tools can help well-motivated analysts and programmers to create
+Although these properties cannot be enforced by any programming environment,
+suitable tools can help well-motivated analysts and programmers to create
 robust applications with much less effort.

### 2014-03-18 discussion.tex
split into multiple files

--- 2014-03-18 paper.tex
@@
-%\documentclass[journal]{vgtc}                % final (journal style)
-\documentclass[review,journal]{vgtc}         % review (journal style)
-%\documentclass[widereview]{vgtc}             % wide-spaced review
-%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
-%\documentclass[electronic,journal]{vgtc}     % electronic version, journal
-
-%% Uncomment one of the lines above depending on where your paper is
-%% in the conference process. ``review'' and ``widereview'' are for review
-%% submission, ``preprint'' is for pre-publication, and the final version
-%% doesn't use a specific qualifier. Further, ``electronic'' includes
-%% hyperreferences for more convenient online viewing.
-
-%% Please use one of the ``review'' options in combination with the
-%% assigned online id (see below) ONLY if your paper uses a double blind
-%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
-%% in the past.
-
-%% Please note that the use of figures other than the optional teaser is not permitted on the first page
-%% of the journal version.  Figures should begin on the second page and be
-%% in CMYK or Grey scale format, otherwise, colour shifting may occur
-%% during the printing process.  Papers submitted with figures other than the optional teaser on the
-%% first page will be refused.
-
-%% These three lines bring in essential packages: ``mathptmx'' for Type 1
-%% typefaces, ``graphicx'' for inclusion of EPS figures. and ``times''
-%% for proper handling of the times font family.
-
-\usepackage{mathptmx}
-\usepackage{graphicx}
-\usepackage{times}
-\usepackage{xspace}
-\usepackage{amssymb}
-
-% \usepackage[onecolumn]{multicol}
-
-%% algorithm packages (llins)
-\usepackage{algorithm}
-\usepackage{microtype}
-\usepackage{algpseudocode}
-
-%% We encourage the use of mathptmx for consistent usage of times font
-%% throughout the proceedings. However, if you encounter conflicts
-%% with other math-related packages, you may want to disable it.
-
-%% This turns references into clickable hyperlinks.
-\usepackage[bookmarks,backref=true,linkcolor=black]{hyperref} %,colorlinks
-\hypersetup{
-  pdfauthor = {},
-  pdftitle = {},
-  pdfsubject = {},
-  pdfkeywords = {},
-  colorlinks=true,
-  linkcolor= black,
-  citecolor= black,
-  pageanchor=true,
-  urlcolor = black,
-  plainpages = false,
-  linktocpage
-}
-
-%% If you are submitting a paper to a conference for review with a double
-%% blind reviewing process, please replace the value ``0'' below with your
-%% OnlineID. Otherwise, you may safely leave it at ``0''.
-\onlineid{}
-
-%% declare the category of your paper, only shown in review mode
-\vgtccategory{Research}
-
-%% allow for this line if you want the electronic option to work properly
-\vgtcinsertpkg
-
-%% In preprint mode you may define your own headline.
-%\preprinttext{To appear in an IEEE VGTC sponsored conference.}
-
-%% Paper title.
-
-\title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
-
-% DevOps for Visual Analysis and Data Science
-
-%% This is how authors are specified in the journal style
-
-%% indicate IEEE Member or Student Member in form indicated below
-\author{Stephen North and Carlos Scheidegger and Simon Urbanek and
-  Gordon Woodhull}
-
-\authorfooter{
-%% insert punctuation at end of each item
-}
-
-%other entries to be set up for journal
-% \shortauthortitle{Biv \MakeLowercase{\textit{et al.}}: Global Illumination for Fun and Profit}
-%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}
-
-%% Abstract section.
-\abstract{ Consider the emerging role of a data science team within an
-  organization. Individual data scientists and statisticians usually
-  work on loosely related problems, and must devise effective ways to 
-  share their findings and to move results from exploratory data analysis
-  to automated diagnostics and reports deployed for wider consumption.
-  There are two problems with the current
-  practice. First, there are gaps in this workflow: EDA is performed
-  with one set of tools, and automated reports and deployments with another.
-  Second, EDA environments often assume a single-developer perspective,
-  while data scientist teams could greatly benefit from easier sharing of
-  scripts and data feeds, experiments, annotations, and automated
-  recommendations, which are well beyond what traditional version control
-  systems provide.  \todo{sell idea, not system} In this paper, we propose RCloud, a system that supports
-  collaborative data analysis, visualization and web deployment. We will
-  discuss the design decisions, tradeoffs and limitations, and compare RCloud
-  to other current proposals.
-} % end of abstract
-
-%% Keywords that describe your work. Will show as 'Index Terms' in journal
-%% please capitalize first letter and insert punctuation after last keyword
-\keywords{Some, Keywords, Here}
-
-%% \CCScatlist{ % not used in journal version
-%%  \CCScat{K.6.1}{Management of Computing and Information Systems}%
-%% {Project and People Management}{Life Cycle};
-%%  \CCScat{K.7.m}{The Computing Profession}{Miscellaneous}{Ethics}
-%% }
-
-%% Uncomment below to include a teaser figure.
-\teaser{
-  \centering
-  \includegraphics[width=.3\linewidth]{fig/screenshots/wdcplotsmall.jpg}
-  %% \includegraphics[width=.9\linewidth]{figs/teaser.jpg}
-  % \vspace{-.5em}
-  \caption{Left: RCloud integration of HTML5 visualization libraries such
-    as D3~\cite{Bostock:2011:DDD} and Crossfilter~\cite{Square:2014:CFM}.
-  }
-% \vspace{-.5em}
-}
-
-%% Uncomment below to disable the manuscript note
-%\renewcommand{\manuscriptnotetxt}{}
-
-%% Copyright space is enabled by default as required by guidelines.
-%% It is disabled by the 'review' option or via the following command:
-% \nocopyrightspace
-
-%
-% Useful Macros 
-%
-
-\newcommand{\eg}{e.g.\xspace} % e.g. and i.e. ARE NOT ITALIC!!
-\newcommand{\ie}{i.e.\xspace}
-\newcommand{\todo}[1]{\textcolor{red}{#1}}
-
-\input{commentcommands.tex}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%% ALGORITHM %%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\renewcommand{\algorithmicthen}{}
-
-% aux. commands
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%% The ``\maketitle'' command must be the first command after the
-%% ``\begin{document}'' command. It prepares and prints the title block.
-
-%% the only exception to this rule is the \firstsection command
-
-\firstsection{Introduction}
-
-\maketitle
-
-\todo{Points we want to make."Coordination is done by meetings" "how do you trace an automated alert back to an EDA environment ?"}
-
-% What's the area
-Consider the emerging role of a data science team within an
-organization \cite{Keim:2008:VAS}.
-
-% How do people do it today?
-Individual data scientists and statisticians usually
-work on loosely related problems, and must devise effective ways to 
-share their findings and to move results from exploratory data analysis
-to automated diagnostics and reports deployed for wider consumption.
-They generally rely on a patchwork of resources that are shared informally,
-or at best, through systems that attack some of the tasks or functions in
-the data analysis workcycle.
-
-% Why does this not work?
-There are two problems with the current
-practice. First, there are gaps in this workflow:
-Exploratory Data Analysis (EDA) is performed with one set of tools,
-and automated reports and deployments with another.
-Second, EDA environments often assume a single-developer perspective,
-while data scientist teams could greatly benefit from easier sharing of
-scripts, data sets, data feeds, experiments, annotations, and automated
-recommendations, which are well beyond what traditional version control systems provide.
-
-% What are we going to do about this?
-\todo{Sell idea, not system} In this paper, we investigate the hypothesis
-that the practice of visual analytics can advance by adopting techniques
-from information retrieval and collaborative computer, formally supported
-within the computing environment. To investigate this, we propose RCloud,
-a system that supports collaborative data analysis, visualization and
-web deployment. We will discuss the design decisions, tradeoffs and
-limitations, and compare RCloud to other current proposals.
-
-\todo{Define the requirements for the system}
-\begin{itemize}
-\item Technology transfer.
-Things developed by analysts need to become accessible in the production environment and should be supported in the environment, making it easy to push new features, experiments, etc.
-\item Dasboarding.
-There should be a mode in which people can operate the experiments without coding.
-At times code has to be written because some things cannot be done any other way but other people must not ever see code because their eyes will fall out.
-\item Version control as a primary concept in the repository. Need to control the visibility of 
-\item Modern interactive technologies.
-People want to interact with data to explore it.
-Scalable technologies like crossfilter exist.
-Want to support those in the context of an R environment.
-Need to support full 2-way communication between computing in the cloud and interaction on the web.
-\item Other requirements e.g.
-Kandel's user study.
-Not our contribution but things we must address, mentioned in related work section.
-Heer/Agrawala design considerations for collaborative visual analytics.
-\end{itemize}
-
-% What is the scope of our solution?
-Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
-for statistical computing and graphics. R is one of the most popular
-systems for ...
-
-% We want our solution to play well with the rest of the ecosystem
-R has developed a complex ecosystem of compatible packages, software
-tools and other resources. For example there are packages that make
-it convenient to execute map-reduce programs, generate web interfaces,
-
-\section{Related Work}
-
-Kandel et al. argue in their review of data wrangling work that data
-cleaning, wrangling and transformation is a major part of exploratory
-analysis and visualization~\cite{Kandel:2011:RDI}. Although RCloud
-does not by itself include modules specific to data cleaning and
-wrangling, we note from internal experience that these cleaning
-scripts themselves tend to change over time. RCloud addresses this
-issue by providing easy publishing of data-cleaning notebooks as web
-services, which reduces the total data-cleaning effort across an
-organization.
-
-Kandel et al.'s interview study points out the typical ``explore'',
-``model'', ``report'' cycle in enterprise data
-analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
-this cycle that cost time and effort to overcome. RCloud seeks to
-reduce this impedance mismatch. They also point out that larger teams
-are becoming more common in data analysis, that supporting
-collaboration is a difficult and important problem, and that sharing
-and versioning of data sources and artifacts is hindered by current
-technology in practice. ``We found that analysts typically did not
-share scripts with each other. Scripts that were shared were
-disseminated similarly to intermediate data: either through shared
-drives or email. Analysts rarely stored their analytic code in source
-control.'' Their work points to the opportunity for better technology
-to support collaboration and sharing by data analysis teams.
-
-Heer and Agarwala identify many design considerations for
-collaborative visual analytics~\cite{Heer:2008:DCF}. \emph{Starring},
-the means for signaling interest in notebooks, described in
-Section~\ref{sec:starring}, addresses social-psychological incentives,
-recommendation, and voting and ranking. RCloud's integrated deployment
-mechanism, described in Section~\ref{sec:deployment}, addresses cost of
-integration, content export, presentation and view sharing. Notebooks,
-and the integrated version control system for them, described in
-Section~\ref{sec:notebooks}, address modularity and granularity, and
-artifact histories.
-
-Manyeyes \cite{Viegas:2007:MAS} was a landmark system for the integration
-of social media with visualization and data publishing. We build on this
-work by defining a rich interface for collaboration about code and for
-operating on metadata.
-
-The need for integrating statistics and visualization has been
-highlighted in previous studies and is widely understood by
-various technical communities \cite{Perer:2008:ISA}
-
-There has been noteworthy work on specific techniques such as
-social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
-and crowdsourcing \cite{Fast:2014:ECS} to support collaborative
-or social development or analysis processes.
-Similarly, there are computational methods to support high
-performance execution in incremental code development
-environments \cite{Guo:2010:TPI}.
-Our goal is to define an environment in which many such
-techniques may be integrated and made available to a broad community.
-
-VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
-are tools built on top of the VisTrails workflow management
-system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
-semantics for automatically deriving user interfaces from workflows,
-while Crowdlabs exposes these capabilities on a website feature
-workflow upload and remote execution. In our view, the impedance
-mismatch between a dataflow pipeline specification and the power of a
-general-purpose language is too great for the type of general
-exploratory work in data science teams. As a result, RCloud tries to
-provide a closer data-analysis to analysts accustomed to createing and
-executing R and Python code while keeping many of the attractive
-properties like automatic versioning and management.
-
-Our work has been inspired by and benefited from other proposals
-to improve data analysis environments and processes,
-including RStudio \cite{RStudio:2013:SWA},
-R packages such as Markdown \cite{Allaire:2014:MMR},
-knitr \cite{Xie:2013:DDW}
-and Shiny \cite{RStudio:2013:SWA},
-and iPython notebooks \cite{Perez:2007:IAS}
-to name a few. RStudio aims at providing an integrated development environment
-for R programming, with support for publishing code in packages. Markdown,
-knitR and Shiny provide R with sophisticated reporting capabilities, including
-interactive web interfaces. IPython \cite{Perez:2007:IAS}
-shares many of our goals, such as providing a comprehensive environment
-for analysis and programming, with sharable documents on the web.
-
-One overall goal is to reduce the gap between implementors and deployers
-of technology. Don't want a team of 20 IT people to support 5 data scientists.
-
-% devops for data science.
-
-\section{System Architecture}
-
-\begin{figure}
-\includegraphics[width=\linewidth]{fig/system/system.pdf}
-\caption{\label{fig:system}A diagram of RCloud's architecture}
-\end{figure}
-
-Design for cloud-friendly? This goes back to the point in the
-introduction about playing nice with the rest of the ecosystem.
-
-\section{System Design}
-
-\subsection{Notebooks\label{sec:notebooks}}
-
-\subsection{Reputation and Interest: starring\label{sec:starring}}
-
-\subsection{Deployment of notebooks\label{sec:deployment}}
-
-Notebooks as versioned subroutines, web services.
-
-Notebooks by default are visible by the entire organization: there
-exists a URL for every notebook in RCloud. This is deliberate. As
-pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS}, broad
-access to analysis outputs (in their case, in the form of NameVoyager)
-increases long-term engagement partly by the crossreferences in the
-web. Our main RCloud deployment is only visible inside an intranet,
-but we have nevertheless found anecdotal support for this theory by
-noticing links to RCloud notebooks in internal discussion fora and
-mailing lists.
-
-\subsection{Technologies: R, Python, HTML5, interactive notebooks, etc.}
-
-How do we do things that are not trivial to do with IPython (for example)
-
-dcplot. two-way communication between between backend session and
-frontend session.
-
-\section{Case Studies}
-
-\section{Lessons Learned}
-
-\section{Discussion and Limitations}
+\section{Lessons Learned, Discussion and Limitations}
 
@@
 recommendation systems, and navigation of the set of notebooks, etc.
-
-\bibliographystyle{abbrv}
-
-\bibliography{paper}
-\end{document}

### 2014-03-18 introduction.tex
split into multiple files

--- 2014-03-18 paper.tex
@@
-%\documentclass[journal]{vgtc}                % final (journal style)
-\documentclass[review,journal]{vgtc}         % review (journal style)
-%\documentclass[widereview]{vgtc}             % wide-spaced review
-%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
-%\documentclass[electronic,journal]{vgtc}     % electronic version, journal
-
-%% Uncomment one of the lines above depending on where your paper is
-%% in the conference process. ``review'' and ``widereview'' are for review
-%% submission, ``preprint'' is for pre-publication, and the final version
-%% doesn't use a specific qualifier. Further, ``electronic'' includes
-%% hyperreferences for more convenient online viewing.
-
-%% Please use one of the ``review'' options in combination with the
-%% assigned online id (see below) ONLY if your paper uses a double blind
-%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
-%% in the past.
-
-%% Please note that the use of figures other than the optional teaser is not permitted on the first page
-%% of the journal version.  Figures should begin on the second page and be
-%% in CMYK or Grey scale format, otherwise, colour shifting may occur
-%% during the printing process.  Papers submitted with figures other than the optional teaser on the
-%% first page will be refused.
-
-%% These three lines bring in essential packages: ``mathptmx'' for Type 1
-%% typefaces, ``graphicx'' for inclusion of EPS figures. and ``times''
-%% for proper handling of the times font family.
-
-\usepackage{mathptmx}
-\usepackage{graphicx}
-\usepackage{times}
-\usepackage{xspace}
-\usepackage{amssymb}
-
-% \usepackage[onecolumn]{multicol}
-
-%% algorithm packages (llins)
-\usepackage{algorithm}
-\usepackage{microtype}
-\usepackage{algpseudocode}
-
-%% We encourage the use of mathptmx for consistent usage of times font
-%% throughout the proceedings. However, if you encounter conflicts
-%% with other math-related packages, you may want to disable it.
-
-%% This turns references into clickable hyperlinks.
-\usepackage[bookmarks,backref=true,linkcolor=black]{hyperref} %,colorlinks
-\hypersetup{
-  pdfauthor = {},
-  pdftitle = {},
-  pdfsubject = {},
-  pdfkeywords = {},
-  colorlinks=true,
-  linkcolor= black,
-  citecolor= black,
-  pageanchor=true,
-  urlcolor = black,
-  plainpages = false,
-  linktocpage
-}
-
-%% If you are submitting a paper to a conference for review with a double
-%% blind reviewing process, please replace the value ``0'' below with your
-%% OnlineID. Otherwise, you may safely leave it at ``0''.
-\onlineid{}
-
-%% declare the category of your paper, only shown in review mode
-\vgtccategory{Research}
-
-%% allow for this line if you want the electronic option to work properly
-\vgtcinsertpkg
-
-%% In preprint mode you may define your own headline.
-%\preprinttext{To appear in an IEEE VGTC sponsored conference.}
-
-%% Paper title.
-
-% \title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
-\title{RCloud: DevOps for Visual Analysis and Data Science}
-
-%% This is how authors are specified in the journal style
-
-%% indicate IEEE Member or Student Member in form indicated below
-\author{Stephen North and Carlos Scheidegger and Simon Urbanek and
-  Gordon Woodhull}
-
-\authorfooter{
-%% insert punctuation at end of each item
-}
-
-%other entries to be set up for journal
-% \shortauthortitle{Biv \MakeLowercase{\textit{et al.}}: Global Illumination for Fun and Profit}
-%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}
-
-%% Abstract section.
-\abstract{ Consider the emerging role of a data science team within an
-  organization. Individual data scientists and statisticians usually
-  work on loosely related problems, and must devise effective ways to 
-  share their findings and move results from exploratory data analysis
-  into automated diagnostics and reports deployed for wider consumption.
-  There are two problems with the current
-  practice. First, there are gaps in this workflow: EDA is performed
-  with one set of tools, and automated reports and deployments with another.
-  Second, EDA environments often assume a single-developer perspective,
-  while data scientist teams would get much benefit from easier sharing
-  of scripts and data feeds, experiments, annotations, and automated
-  recommendations, which are well beyond what traditional version control
-  systems provide. \todo{sell idea, not system} In this paper, we propose RCloud, a system that supports
-  collaborative data analysis, visualization and web deployment. We will
-  discuss the design decisions, tradeoffs and limitations, and compare RCloud
-  to other current proposals.
-} % end of abstract
-
-%% Keywords that describe your work. Will show as 'Index Terms' in journal
-%% please capitalize first letter and insert punctuation after last keyword
-\keywords{Some, Keywords, Here}
-
-%% \CCScatlist{ % not used in journal version
-%%  \CCScat{K.6.1}{Management of Computing and Information Systems}%
-%% {Project and People Management}{Life Cycle};
-%%  \CCScat{K.7.m}{The Computing Profession}{Miscellaneous}{Ethics}
-%% }
-
-%% Uncomment below to include a teaser figure.
-\teaser{
-  \centering
-  \includegraphics[width=.3\linewidth]{fig/screenshots/wdcplotsmall.jpg}
-  %% \includegraphics[width=.9\linewidth]{figs/teaser.jpg}
-  % \vspace{-.5em}
-  \caption{Left: RCloud integration of HTML5 visualization libraries such
-    as D3~\cite{Bostock:2011:DDD} and Crossfilter~\cite{Square:2014:CFM}.
-  }
-% \vspace{-.5em}
-}
-
-%% Uncomment below to disable the manuscript note
-%\renewcommand{\manuscriptnotetxt}{}
-
-%% Copyright space is enabled by default as required by guidelines.
-%% It is disabled by the 'review' option or via the following command:
-% \nocopyrightspace
-
-%
-% Useful Macros 
-%
-
-\newcommand{\eg}{e.g.\xspace} % e.g. and i.e. ARE NOT ITALIC!!
-\newcommand{\ie}{i.e.\xspace}
-\newcommand{\todo}[1]{\textcolor{red}{#1}}
-
-\input{commentcommands.tex}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%% ALGORITHM %%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\renewcommand{\algorithmicthen}{}
-
-% aux. commands
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%% The ``\maketitle'' command must be the first command after the
-%% ``\begin{document}'' command. It prepares and prints the title block.
-
-%% the only exception to this rule is the \firstsection command
-
 \firstsection{Introduction}
@@
 
-begin{itemize}
+\begin{itemize}
 \item Technology transfer. Experiments in notebooks developed by analysts need to become accessible in the production environment, making it easy to push new features, prototypes, etc.
@@
 it convenient to execute map-reduce programs, generate web interfaces,
-
-\section{Related Work}
-
-Kandel et al. argue in their review of data wrangling work that data
-cleaning, wrangling and transformation is a major part of exploratory
-analysis and visualization~\cite{Kandel:2011:RDI}. Although RCloud
-does not by itself include modules specific to data cleaning and
-wrangling, we note from internal experience that these cleaning
-scripts themselves tend to change over time. RCloud addresses this
-issue by providing easy publishing of data-cleaning notebooks as web
-services, which reduces the total data-cleaning effort across an
-organization.
-
-Kandel et al.'s interview study points out the typical ``explore'',
-``model'', ``report'' cycle in enterprise data
-analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
-this cycle that cost time and effort to overcome. RCloud seeks to
-reduce this impedance mismatch. They also point out that larger teams
-are becoming more common in data analysis, that supporting
-collaboration is a difficult and important problem, and that sharing
-and versioning of data sources and artifacts is hindered by current
-technology in practice. ``We found that analysts typically did not
-share scripts with each other. Scripts that were shared were
-disseminated similarly to intermediate data: either through shared
-drives or email. Analysts rarely stored their analytic code in source
-control.'' Their work points to the opportunity for better technology
-to support collaboration and sharing by data analysis teams.
-
-Heer and Agarwala identify many design considerations for
-collaborative visual analytics~\cite{Heer:2008:DCF}. \emph{Starring},
-the means for signaling interest in notebooks, described in
-Section~\ref{sec:starring}, addresses social-psychological incentives,
-recommendation, and voting and ranking. RCloud's integrated deployment
-mechanism, described in Section~\ref{sec:deployment}, addresses cost of
-integration, content export, presentation and view sharing. Notebooks,
-and the integrated version control system for them, described in
-Section~\ref{sec:notebooks}, address modularity and granularity, and
-artifact histories.
-
-Manyeyes \cite{Viegas:2007:MAS} was a landmark system for the integration
-of social media with visualization and data publishing. We build on this
-work by defining a rich interface for collaboration about code and for
-operating on metadata.
-
-The need for integrating statistics and visualization has been
-highlighted in previous studies and is widely understood by
-various technical communities \cite{Perer:2008:ISA}
-
-There has been noteworthy work on specific techniques such as
-social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
-and crowdsourcing \cite{Fast:2014:ECS} to support collaborative
-or social development or analysis processes.
-Similarly, there are computational methods to support high
-performance execution in incremental code development
-environments \cite{Guo:2010:TPI}.
-Our goal is to define an environment in which many such
-techniques may be integrated and made available to a broad community.
-
-VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
-are tools built on top of the VisTrails workflow management
-system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
-semantics for automatically deriving user interfaces from workflows,
-while Crowdlabs exposes these capabilities on a website feature
-workflow upload and remote execution. In our view, the impedance
-mismatch between a dataflow pipeline specification and the power of a
-general-purpose language is too great for the type of general
-exploratory work in data science teams. As a result, RCloud tries to
-provide a closer data-analysis to analysts accustomed to createing and
-executing R and Python code while keeping many of the attractive
-properties like automatic versioning and management.
-
-Our work has been inspired by and benefited from other proposals
-to improve data analysis environments and processes,
-including RStudio \cite{RStudio:2013:SWA},
-R packages such as Markdown \cite{Allaire:2014:MMR},
-knitr \cite{Xie:2013:DDW}
-and Shiny \cite{RStudio:2013:SWA},
-and iPython notebooks \cite{Perez:2007:IAS}
-to name a few. RStudio aims at providing an integrated development environment
-for R programming, with support for publishing code in packages. Markdown,
-knitR and Shiny provide R with sophisticated reporting capabilities, including
-interactive web interfaces. IPython \cite{Perez:2007:IAS}
-shares many of our goals, such as providing a comprehensive environment
-for analysis and programming, with sharable documents on the web.
-
-One overall goal is to reduce the gap between implementors and deployers
-of technology. Don't want a team of 20 IT people to support 5 data scientists.
-
-% devops for data science.
-
-\section{System Architecture}
-
-\begin{figure}
-\includegraphics[width=\linewidth]{fig/system/system.pdf}
-\caption{\label{fig:system}A diagram of RCloud's architecture. Dashed
-  lines represent features which not yet implemented.}
-\end{figure}
-
-The internal computing infrastructure of organizations has changed
-radically in the last fifteen years or so. Much of the move toward the
-``cloud'' brought with it an ecosystem of processes distributed over a
-network (many times virtually defined) that communicate over a
-high-level protocol of some sort. Messages are typically exchanged
-over HTTP. HTTP is attractive because web browsers and servers are
-ubiquituous and available in essentially all pieces hardware today,
-from tiny sensors in sensor networks to handheld devices to laptops to
-rack-mounted servers.
-
-As a result, in our view HTTP has become the lingua franca of IPC in
-an intranet. One of the design goals for RCloud was to provide an
-attractive environment for creators of data-analysis scripts, but at
-the same time to create a system that behaves as a first-class citizen
-in the ecosystem of computer services and networks with an
-organization.
-
-%% Design for cloud-friendly? This goes back to the point in the
-%% introduction about playing nice with the rest of the ecosystem.
-
-\section{System Design}
-
-\subsection{Notebooks\label{sec:notebooks}}
-
-\subsection{Reputation and Interest: starring\label{sec:starring}}
-
-\subsection{Deployment of notebooks\label{sec:deployment}}
-
-Notebooks as versioned subroutines, web services.
-
-Notebooks by default are visible by the entire organization: there
-exists a URL for every notebook in RCloud. This is deliberate. As
-pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS}, broad
-access to analysis outputs (in their case, in the form of NameVoyager)
-increases long-term engagement partly by the crossreferences in the
-web. Our main RCloud deployment is only visible inside an intranet,
-but we have nevertheless found anecdotal support for this theory by
-noticing links to RCloud notebooks in internal discussion fora and
-mailing lists.
-
-\subsection{Technologies: R, Python, HTML5, interactive notebooks, etc.}
-
-How do we do things that are not trivial to do with IPython (for example)
-
-dcplot. two-way communication between between backend session and
-frontend session.
-
-\section{Case Studies}
-
-\section{Lessons Learned}
-
-\section{Discussion and Limitations}
-
-What are the decisions and ideas that are central to this paper?
-
-discussion about binding times, running vs.caching.
-transparent vs. explicit operations done by programmers
-
-% limitations
-Previous studies have pointed out difficulties in achieving the flexibility,
-scalability and maintainability expected of production software with experimental
-code created by data analysts.
-Although these properties cannot be enforced by any programming environment,
-suitable tools can help well-motivated analysts and programmers to create
-robust applications with much less effort.
-
-Asynchronous collaborative visual analytics
-(ACVA)~\cite{Chen:2011:SEC}. This paper addresses visualization
-\emph{of} the ACVA process. It will be important when we talk about
-recommendation systems, and navigation of the set of notebooks, etc.
-
-\bibliographystyle{abbrv}
-
-\bibliography{paper}
-\end{document}

### 2014-03-18 paper.tex
split into multiple files

--- 2014-03-18 paper.tex
@@
 
-\firstsection{Introduction}
-
-\maketitle
-
-\todo{Points we want to make."Coordination is done by meetings" "how do you trace an automated alert back to an EDA environment ?"}
-
-% What's the area
-Consider the emerging role of a data science team within an
-organization \cite{Keim:2008:VAS}.
-
-% How do people do it today?
-Individual data scientists and statisticians usually
-work on loosely related problems, and must devise effective ways to 
-share their findings and to move results from exploratory data analysis
-to automated diagnostics and reports deployed for wider consumption.
-They generally rely on a patchwork of resources that are shared informally,
-or at best, through systems that attack some of the tasks or functions in
-the data analysis workcycle.
-
-% Why does this not work?
-There are two problems with the current
-practice. First, there are gaps in this workflow:
-Exploratory Data Analysis (EDA) is performed with one set of tools,
-and automated reports and deployments with another.
-Second, EDA environments often assume a single-developer perspective,
-while data scientist teams could greatly benefit from easier sharing of
-scripts, data sets, data feeds, experiments, annotations, and automated
-recommendations, which are well beyond what traditional version control systems provide.
-
-% What are we going to do about this?
-\todo{Sell idea, not system} In this paper, we investigate the hypothesis
-that the practice of visual analytics can advance by adopting techniques
-from information retrieval and collaborative computer, formally supported
-within the computing environment. To investigate this, we propose RCloud,
-a system that supports collaborative data analysis, visualization and
-web deployment. We will discuss the design decisions, tradeoffs and
-limitations, and compare RCloud to other current proposals.
-
-\todo{Define the requirements for the system}
-\begin{itemize}
-\item Technology transfer.
-Things developed by analysts need to become accessible in the production environment and should be supported in the environment, making it easy to push new features, experiments, etc.
-\item Dasboarding.
-There should be a mode in which people can operate the experiments without coding.
-At times code has to be written because some things cannot be done any other way but other people must not ever see code because their eyes will fall out.
-\item Version control as a primary concept in the repository. Need to control the visibility of 
-\item Modern interactive technologies.
-People want to interact with data to explore it.
-Scalable technologies like crossfilter exist.
-Want to support those in the context of an R environment.
-Need to support full 2-way communication between computing in the cloud and interaction on the web.
-\item Other requirements e.g.
-Kandel's user study.
-Not our contribution but things we must address, mentioned in related work section.
-Heer/Agrawala design considerations for collaborative visual analytics.
-\end{itemize}
-
-% What is the scope of our solution?
-Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
-for statistical computing and graphics. R is one of the most popular
-systems for ...
-
-% We want our solution to play well with the rest of the ecosystem
-R has developed a complex ecosystem of compatible packages, software
-tools and other resources. For example there are packages that make
-it convenient to execute map-reduce programs, generate web interfaces,
-
-\section{Related Work}
-
-Kandel et al. argue in their review of data wrangling work that data
-cleaning, wrangling and transformation is a major part of exploratory
-analysis and visualization~\cite{Kandel:2011:RDI}. Although RCloud
-does not by itself include modules specific to data cleaning and
-wrangling, we note from internal experience that these cleaning
-scripts themselves tend to change over time. RCloud addresses this
-issue by providing easy publishing of data-cleaning notebooks as web
-services, which reduces the total data-cleaning effort across an
-organization.
-
-Kandel et al.'s interview study points out the typical ``explore'',
-``model'', ``report'' cycle in enterprise data
-analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
-this cycle that cost time and effort to overcome. RCloud seeks to
-reduce this impedance mismatch. They also point out that larger teams
-are becoming more common in data analysis, that supporting
-collaboration is a difficult and important problem, and that sharing
-and versioning of data sources and artifacts is hindered by current
-technology in practice. ``We found that analysts typically did not
-share scripts with each other. Scripts that were shared were
-disseminated similarly to intermediate data: either through shared
-drives or email. Analysts rarely stored their analytic code in source
-control.'' Their work points to the opportunity for better technology
-to support collaboration and sharing by data analysis teams.
-
-Heer and Agarwala identify many design considerations for
-collaborative visual analytics~\cite{Heer:2008:DCF}. \emph{Starring},
-the means for signaling interest in notebooks, described in
-Section~\ref{sec:starring}, addresses social-psychological incentives,
-recommendation, and voting and ranking. RCloud's integrated deployment
-mechanism, described in Section~\ref{sec:deployment}, addresses cost of
-integration, content export, presentation and view sharing. Notebooks,
-and the integrated version control system for them, described in
-Section~\ref{sec:notebooks}, address modularity and granularity, and
-artifact histories.
-
-Manyeyes \cite{Viegas:2007:MAS} was a landmark system for the integration
-of social media with visualization and data publishing. We build on this
-work by defining a rich interface for collaboration about code and for
-operating on metadata.
-
-The need for integrating statistics and visualization has been
-highlighted in previous studies and is widely understood by
-various technical communities \cite{Perer:2008:ISA}
-
-There has been noteworthy work on specific techniques such as
-social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
-and crowdsourcing \cite{Fast:2014:ECS} to support collaborative
-or social development or analysis processes.
-Similarly, there are computational methods to support high
-performance execution in incremental code development
-environments \cite{Guo:2010:TPI}.
-Our goal is to define an environment in which many such
-techniques may be integrated and made available to a broad community.
-
-VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
-are tools built on top of the VisTrails workflow management
-system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
-semantics for automatically deriving user interfaces from workflows,
-while Crowdlabs exposes these capabilities on a website feature
-workflow upload and remote execution. In our view, the impedance
-mismatch between a dataflow pipeline specification and the power of a
-general-purpose language is too great for the type of general
-exploratory work in data science teams. As a result, RCloud tries to
-provide a closer data-analysis to analysts accustomed to createing and
-executing R and Python code while keeping many of the attractive
-properties like automatic versioning and management.
-
-Our work has been inspired by and benefited from other proposals
-to improve data analysis environments and processes,
-including RStudio \cite{RStudio:2013:SWA},
-R packages such as Markdown \cite{Allaire:2014:MMR},
-knitr \cite{Xie:2013:DDW}
-and Shiny \cite{RStudio:2013:SWA},
-and iPython notebooks \cite{Perez:2007:IAS}
-to name a few. RStudio aims at providing an integrated development environment
-for R programming, with support for publishing code in packages. Markdown,
-knitR and Shiny provide R with sophisticated reporting capabilities, including
-interactive web interfaces. IPython \cite{Perez:2007:IAS}
-shares many of our goals, such as providing a comprehensive environment
-for analysis and programming, with sharable documents on the web.
-
-One overall goal is to reduce the gap between implementors and deployers
-of technology. Don't want a team of 20 IT people to support 5 data scientists.
-
-% devops for data science.
-
-\section{System Architecture}
-
-\begin{figure}
-\includegraphics[width=\linewidth]{fig/system/system.pdf}
-\caption{\label{fig:system}A diagram of RCloud's architecture}
-\end{figure}
-
-Design for cloud-friendly? This goes back to the point in the
-introduction about playing nice with the rest of the ecosystem.
-
-\section{System Design}
-
-\subsection{Notebooks\label{sec:notebooks}}
-
-\subsection{Reputation and Interest: starring\label{sec:starring}}
-
-\subsection{Deployment of notebooks\label{sec:deployment}}
-
-Notebooks as versioned subroutines, web services.
-
-Notebooks by default are visible by the entire organization: there
-exists a URL for every notebook in RCloud. This is deliberate. As
-pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS}, broad
-access to analysis outputs (in their case, in the form of NameVoyager)
-increases long-term engagement partly by the crossreferences in the
-web. Our main RCloud deployment is only visible inside an intranet,
-but we have nevertheless found anecdotal support for this theory by
-noticing links to RCloud notebooks in internal discussion fora and
-mailing lists.
-
-\subsection{Technologies: R, Python, HTML5, interactive notebooks, etc.}
-
-How do we do things that are not trivial to do with IPython (for example)
-
-dcplot. two-way communication between between backend session and
-frontend session.
-
-\section{Case Studies}
-
-\section{Lessons Learned}
-
-\section{Discussion and Limitations}
-
-What are the decisions and ideas that are central to this paper?
-
-discussion about binding times, running vs.caching.
-transparent vs. explicit operations done by programmers
-
-% limitations
-Previous studies have pointed out difficulties in achieving the flexibility,
-scalability and maintainability expected of production software with experimental
-code created by data analysts.
-These properties cannot be ensured by any programming environment,
-but suitable tools can help well-motivated analysts and programmers to create
-robust applications with much less effort.
-
-Asynchronous collaborative visual analytics
-(ACVA)~\cite{Chen:2011:SEC}. This paper addresses visualization
-\emph{of} the ACVA process. It will be important when we talk about
-recommendation systems, and navigation of the set of notebooks, etc.
+\input{introduction}
+\input{relatedwork}
+\input{system}
+\input{casestudies}
+\input{discussion}
+\input{conclusion}
 

### 2014-03-18 relatedwork.tex
split into multiple files

\section{Related Work}

Kandel et al. argue in their review of data wrangling work that data
cleaning, wrangling and transformation is a major part of exploratory
analysis and visualization~\cite{Kandel:2011:RDI}. Although RCloud
does not by itself include modules specific to data cleaning and
wrangling, we note from internal experience that these cleaning
scripts themselves tend to change over time. RCloud addresses this
issue by providing easy publishing of data-cleaning notebooks as web
services, which reduces the total data-cleaning effort across an
organization.

Kandel et al.'s interview study points out the typical ``explore'',
``model'', ``report'' cycle in enterprise data
analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
this cycle that cost time and effort to overcome. RCloud seeks to
reduce this impedance mismatch. They also point out that larger teams
are becoming more common in data analysis, that supporting
collaboration is a difficult and important problem, and that sharing
and versioning of data sources and artifacts is hindered by current
technology in practice. ``We found that analysts typically did not
share scripts with each other. Scripts that were shared were
disseminated similarly to intermediate data: either through shared
drives or email. Analysts rarely stored their analytic code in source
control.'' Their work points to the opportunity for better technology
to support collaboration and sharing by data analysis teams.

Heer and Agarwala identify many design considerations for
collaborative visual analytics~\cite{Heer:2008:DCF}. \emph{Starring},
the means for signaling interest in notebooks, described in
Section~\ref{sec:starring}, addresses social-psychological incentives,
recommendation, and voting and ranking. RCloud's integrated deployment
mechanism, described in Section~\ref{sec:deployment}, addresses cost of
integration, content export, presentation and view sharing. Notebooks,
and the integrated version control system for them, described in
Section~\ref{sec:notebooks}, address modularity and granularity, and
artifact histories.

Manyeyes \cite{Viegas:2007:MAS} was a landmark system for the integration
of social media with visualization and data publishing. We build on this
work by defining a rich interface for collaboration about code and for
operating on metadata.

The need for integrating statistics and visualization has been
highlighted in previous studies and is widely understood by
various technical communities \cite{Perer:2008:ISA}

There has been noteworthy work on specific techniques such as
social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
and crowdsourcing \cite{Fast:2014:ECS} to support collaborative
or social development or analysis processes.
Similarly, there are computational methods to support high
performance execution in incremental code development
environments \cite{Guo:2010:TPI}.
Our goal is to define an environment in which many such
techniques may be integrated and made available to a broad community.

VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
are tools built on top of the VisTrails workflow management
system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
semantics for automatically deriving user interfaces from workflows,
while Crowdlabs exposes these capabilities on a website feature
workflow upload and remote execution. In our view, the impedance
mismatch between a dataflow pipeline specification and the power of a
general-purpose language is too great for the type of general
exploratory work in data science teams. As a result, RCloud tries to
provide a closer data-analysis to analysts accustomed to createing and
executing R and Python code while keeping many of the attractive
properties like automatic versioning and management.

Our work has been inspired by and benefited from other proposals
to improve data analysis environments and processes,
including RStudio \cite{RStudio:2013:SWA},
R packages such as Markdown \cite{Allaire:2014:MMR},
knitr \cite{Xie:2013:DDW}
and Shiny \cite{RStudio:2013:SWA},
and iPython notebooks \cite{Perez:2007:IAS}
to name a few. RStudio aims at providing an integrated development environment
for R programming, with support for publishing code in packages. Markdown,
knitR and Shiny provide R with sophisticated reporting capabilities, including
interactive web interfaces. IPython \cite{Perez:2007:IAS}
shares many of our goals, such as providing a comprehensive environment
for analysis and programming, with sharable documents on the web.

One overall goal is to reduce the gap between implementors and deployers
of technology. Don't want a team of 20 IT people to support 5 data scientists.

% devops for data science.

### 2014-03-18 system.tex
split into multiple files

\section{The System}

% High-level architecture

\begin{figure}
\includegraphics[width=\linewidth]{fig/system/system.pdf}
\caption{\label{fig:system}A diagram of RCloud's architecture. Dashed
  lines represent features which not yet implemented.}
\end{figure}

The internal computing infrastructure of organizations has changed
radically in the last fifteen years or so. Much of the move toward the
``cloud'' brought with it an ecosystem of processes distributed over a
network (many times virtually defined) that communicate over a
high-level protocol of some sort. Messages are typically exchanged
over HTTP. HTTP is attractive because web browsers and servers are
ubiquituous and available in essentially all pieces hardware today,
from tiny sensors in sensor networks to handheld devices to laptops to
rack-mounted servers.

As a result, in our view HTTP has become the lingua franca of IPC in
an intranet. One of the design goals for RCloud was to provide an
attractive environment for creators of data-analysis scripts, but at
the same time to create a system that behaves as a first-class citizen
in the ecosystem of computer services and networks with an
organization.

%% Design for cloud-friendly? This goes back to the point in the
%% introduction about playing nice with the rest of the ecosystem.

\subsection{System Design}

We designed RCloud around the front-facing API, which has roughly one
entry point for each requirement.

Edit

View/Run/Share

Call

Browse/Search. ``More like this'', and a custom webpage solely for browsing.

\subsection{Notebooks\label{sec:notebooks}}

Notebooks as Github gists. 
%
We needed version control over a directory, gist provides that,
wrapped over an HTTP interface.
%
Mainly a tactical decision to save time.
%
Nevertheless, availability of notebook data as (barely) plain text had good side
effects, like ease to build search. 
%
Expect same with suggest.

\subsection{Reputation and Interest: starring\label{sec:starring}}

In RCloud, reputation and interest are a relationship between
\emph{notebooks} and \emph{users}, rather than a relationship between
user pairs. We decided on this approach because we expect typical
RCloud deployments to have relatively few users, but each user to have
 written relatively many notebooks.
In this case, assigning interest to users would not provide
``high-resolution'' data.

Explicit signaling of interest: starring.

Implicit signaling of interest: click-through count and execution
count.

These will be ingredients for the recommendation system.

\subsection{Deployment of notebooks\label{sec:deployment}}

Notebooks as versioned subroutines, web services.

Notebooks by default are visible by the entire organization: there
exists a URL for every notebook in RCloud. This is deliberate. As
pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS}, broad
access to analysis outputs (in their case, in the form of NameVoyager)
increases long-term engagement partly by the crossreferences in the
web. Our main RCloud deployment is only visible inside an intranet,
but we have nevertheless found anecdotal support for this theory by
noticing links to RCloud notebooks in internal discussion fora and
mailing lists.

\subsection{Technologies: R, Python, HTML5, interactive notebooks, etc.}

How do we do things that are not trivial to do with IPython (for example)

dcplot. two-way communication between between backend session and
frontend session.

### 2014-03-18 discussion.tex
yes

@@
 Previous studies have pointed out difficulties in achieving the flexibility,
-scalability and maintainability expected of production software with experimental
-code created by data analysts.
+scalability and maintainability expected of production software in experimental
+code written by data analysts.
 Although these properties cannot be enforced by any programming environment,
@@
 recommendation systems, and navigation of the set of notebooks, etc.
+
+Had to back off only one language sooner than we expected.
+
+Shared artifacts, artifact histories, 
+Discussion
+View sharing - bookmarking, everything available as a URL
+Content export - want to avoid, create ways of working where this is not necessary.
+Social-psychological incentives - starring
+Voting and ranking - starring
+
+Group management, size, diversity.  Lacking support,but clearly desirable.
+
+Has to play well in the ecosystem of the other tools and worked out as planned but there is a temptation to absorb
+
+What is relationship to a standard CMS with R/Python

### 2014-03-19 discussion.tex
yes

@@
 What are the decisions and ideas that are central to this paper?
-
-discussion about binding times, running vs.caching.
-transparent vs. explicit operations done by programmers
 
@@
 
+Returning to the design considerations outlined by Heer and Agrawala, experience with the RCloud prototype demonstrates the value of several
 Shared artifacts, artifact histories, 
@@
 
+discussion about binding times, running vs.caching.
+transparent vs. explicit operations done by programmers
+
 What is relationship to a standard CMS with R/Python
+
+Elaborate on to what extent can the users of the production notebooks use the tools like annotation etc. without coding or without being exposed to the analysts view. is this a gap in our design.

### 2014-03-21 system.tex
yes

@@
 The internal computing infrastructure of organizations has changed
-radically in the last fifteen years or so. Much of the move toward the
-``cloud'' brought with it an ecosystem of processes distributed over a
-network (many times virtually defined) that communicate over a
-high-level protocol of some sort. Messages are typically exchanged
-over HTTP. HTTP is attractive because web browsers and servers are
-ubiquituous and available in essentially all pieces hardware today,
-from tiny sensors in sensor networks to handheld devices to laptops to
-rack-mounted servers.
+radically in the last fifteen years. The shift from large
+servers toward scalable, lower-cost, distributed systems (``the cloud'')
+led to a software ecosystem of processes distributed over a
+network (often virtually defined) communicating via some
+high-level protocol, usually HTTP.  HTTP is dominant because web
+browsers and servers are ubiquituous and available in essentially
+all hardware devices, from tiny sensors, to handheld devices and
+laptops, to rack-mounted servers.
 
-As a result, in our view HTTP has become the lingua franca of IPC in
-an intranet. One of the design goals for RCloud was to provide an
-attractive environment for creators of data-analysis scripts, but at
-the same time to create a system that behaves as a first-class citizen
-in the ecosystem of computer services and networks with an
-organization.
+As a result, HTTP is the lingua franca of interprocess communication
+(IPC). One of the design goals for RCloud was to provide an attractive
+environment for creators of data-analysis scripts, that also behaves
+as a first-class citizen in the pre-existing ecosystem of computer
+services and networks with an organization.
 
@@
 
-We designed RCloud around the front-facing API, which has roughly one
-entry point for each requirement.
+We designed RCloud around the front-facing API, which provides roughly
+one entry point to correspond with each requirement.
 
@@
 RCloud deployments to have relatively few users, but each user to have
- written relatively many notebooks.
-In this case, assigning interest to users would not provide
-``high-resolution'' data.
+written relatively many notebooks. Under that assumption, assigning
+interest to users would not provide sufficiently ``high-resolution'' data.
 
-Explicit signaling of interest: starring.
+We incorporate both explicit and implicit indications of interest
+in notebooks. Explicit interest is signalled by ``starring,'' or
+clicking on a button that marks a notebook as interesting. 
+This makes explicit indication of interest a nearly trivial operation,
+always readily available, thus encouraging its use.
 
-Implicit signaling of interest: click-through count and execution
-count.
+Implicit signaling of interest is supported by keeping click-through
+\cite{Joachims:2005:AIC} and execution counts. (In addition to these
+classic web search techniques of collecting feedback, we have the
+prospect of applying static and dynamic code analysis to generate
+further fine-grained information about relationships, for example,
+which packages and data sets often appear together.)
 
-These will be ingredients for the recommendation system.
+\todo{explain} These will be ingredients for the recommendation system.
 
@@
 
-Notebooks by default are visible by the entire organization: there
-exists a URL for every notebook in RCloud. This is deliberate. As
-pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS}, broad
-access to analysis outputs (in their case, in the form of NameVoyager)
-increases long-term engagement partly by the crossreferences in the
-web. Our main RCloud deployment is only visible inside an intranet,
-but we have nevertheless found anecdotal support for this theory by
-noticing links to RCloud notebooks in internal discussion fora and
-mailing lists.
+There exists a URL for every notebook in RCloud, and notebooks by
+default are visible by the entire organization. This is deliberate.
+As pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS},
+broad access to analysis outputs (in their case, for NameVoyager)
+increases long-term engagement in part through crossreferences on
+the web. Although our prototype RCloud deployment is only visible
+inside a corporate intranet, we nevertheless found anecdotal support
+for this notion by discovering links to RCloud notebooks in internal
+discussion fora and mailing lists.
 
@@
 frontend session.
+
+A key engineering decision in RCloud was to rely on full two-way
+communication between human interface in a web client, and remote
+data and computing resources in the cloud.

### 2014-03-21 introduction.tex
yes

@@
 or at best, through systems that attack individual tasks or functions
-in the data analysis workcycle.
+in the data analysis workcycle. % might cite information visualization reference model or pirolli/card
 
@@
 and automated reports and deployments with another. This makes
-transferring results from EDA to production slow, expensive, and error-prone.
+transferring results between EDA and production (in both directions)
+slow, expensive, and error-prone. Coordination is done through
+meetings and may take months to implement.
+
 Second, EDA environments often assume a single-developer perspective,
-while data scientist teams could greatly benefit from convenient sharing of
-scripts, data sets, data feeds, experiments, annotations, and automated
-analyses and recommendations. These features are well beyond what existing
-repositories and version control systems provide.
+while data scientist teams could greatly benefit from convenient searching
+and sharing of scripts, data sets, data feeds, experiments, annotations,
+and automated analyses and recommendations. These capabilities have had
+tremendous impact on information retrieval in general but are well
+beyond what existing visual analytics environments provide, or what
+is found in today's repositories and version control systems.
 
@@
 for statistical computing and graphics. R is one of the most popular
-systems for ...
+systems for data analysis and is well-supported by 
+a complex ecosystem of compatible packages, software
+tools and other resources. Of importance to us, there are
+packages for high performance computing in the cloud,
+and for generating web graphics.
 
 % We want our solution to play well with the rest of the ecosystem
-R has developed a complex ecosystem of compatible packages, software
-tools and other resources. For example there are packages that make
-it convenient to execute map-reduce programs, generate web interfaces,

### 2014-03-25 casestudies.tex
merge

\section{Case Studies}

% IMPORTANT: what is unique about RCloud here

% From prototyping to dashboard

% Getting other information from the web

%

### 2014-03-25 introduction.tex
merge

@@
 
-\todo{Points we want to make."Coordination is done by meetings" "how do you trace an automated alert back to an EDA environment ?"}
+%% \todo{Points we want to make."Coordination is done by meetings" "how do you trace an automated alert back to an EDA environment ?"}
 
@@
 \begin{itemize}
-\item Technology transfer. Experiments in notebooks developed by analysts need to become accessible in the production environment, making it easy to push new features, prototypes, etc.
+\item Constant technology transfer. Experiments in notebooks developed
+by analysts need to become accessible in the production environment,
+making it easy to push new features, prototypes, etc.
 \item Dashboarding. To support the above, there should be a mode in which non-experts
 can operate the experiments without coding.
-\item Version control as a primary concept in the repository. There needs to be a way to control the visibility of prototypes as they are being developed and tested.
-\item Modern interactive technologies.
-People want to interact with data to explore it.
-Scalable technologies like crossfilter exist.
+\item Version control as a primary concept in the repository. There
+needs to be a way to control the visibility of prototypes as they are
+being developed and tested.
+\item Coexistence in the current data ecosystem. We perceive the
+isolation of current exploratory visualization and data analysis
+environments to be a serious threat to a wider adoption of modern
+techniques.
+\item Modern interactive technologies. current isolation
+notwithstanding, we find there is business demand for interactive
+visualization tools. The main challenge here is how to integrate
+these technologies so data scientists and statisticians can target
+them, \emph{and} business-facing users can use them.
+%% Scalable technologies like crossfilter exist.
 Want to support those in the context of an R environment.
 Need to support full 2-way communication between computing in the cloud and interaction on the web.
-\item Other requirements e.g.
+%% \item Other requirements e.g.
 Kandel's user study.
@@
 a complex ecosystem of compatible packages, software
-tools and other resources. Of importance to us, there are
+tools and other resources.
+%
+Of importance to us, there are existing
 packages for high performance computing in the cloud,
-and for generating web graphics.
+and for generating web graphics. % no, there aren't. we wrote those!
 

### 2014-03-25 discussion.tex
merge
working

@@
 \section{Lessons Learned, Discussion and Limitations}
+
+What are the consequences of introducing this technology into the
+data analysis ecosystem.
+
+Good and bad.
+
+We have observed that people start sharing ideas through the system,
+instead of hacking up prototypes and sending screenshots and Powerpoint
+they hack up experiments and send them. Makes this easier to collaborate
+about code but also makes it easy to collaborate about services that are
+deployed.
+
+The negative is that at every stage everything is deployed, so that if
+you break something
+
+Allows people to fork and try something else without restriction.
+Have not solved the problem of having a lot of bits and pieces lying around
+but have made it easier to get at them and to use algorithms to 
+help organize them.
+
+Recommendations - was done in Vistrails - within the same graph but
+there is clustering of workflows that could be applied to recommending
+across graphs.
+
+Because we developed the ideas and the system together
+
+Package dependencies and versions could be a problem
 
@@
 Previous studies have pointed out difficulties in achieving the flexibility,
-scalability and maintainability expected of production software in experimental
-code written by data analysts.
+scalability and maintainability expected of production software,
+with experimental code written by data analysts. In RCloud this problem
+is compounded by its flexible versioning, which may make it even
+more challenging to ensure that production services are stable.
+Versioning happens in several places: in scripts (that are under
+our control in RCloud through git), in libraries (that should be
+under the control of the R environment, though support for this
+is very weak), and in the external environment
+(such as operating system components and even the protocols
+spoken by external services, for which there is no realistic
+hope of formal control).  
+Also, because there is no explicit separation between the experimental
+and production environments in RCloud, it seems relatively easy for
+data scientists to inintentionally modify or break production services.
+
+Upgrading is troublesome, but what is better. When different things are being
+upgraded in flight how do you ensure the things you are calling behave as
+expected, and if someone upgrades 
+Would require support from the R environment which does not exist now.
+
+have only the right version of the script we are working on but
+can't do the same for the environment
+the environment includes remote services not under our control
+
 Although these properties cannot be enforced by any programming environment,

### 2014-03-25 introduction.tex
whatever

@@
 
+Main requirement is to satisfy these two sets of users: R coders and
+managers. They work differently and want different things. What should
+the infrastructure look like so they both coexist in peace? If this is
+such a fundamental problem, why have other tools not addressed it?
+Discuss Shiny (maybe in the disc. section).
+
 \begin{itemize}

### 2014-03-26 discussion.tex
working

--- 2014-03-18 discussion.tex
@@
 
-What are the decisions and ideas that are central to this paper?
+\subsection{Design Considerations}
 
-discussion about binding times, running vs.caching.
-transparent vs. explicit operations done by programmers
+Experience with deploying RCloud in a community of data analysts
+has provided some insight into how its capabilities were received
+that may be informative to researchers and designers of future systems
+supporting visual analytics.
+Returning to the design considerations outlined by Heer and Agrawala,
+our experience underscores the value of several of their proposals,
+and suggests areas for further exploration.
 
-% limitations
-Previous studies have pointed out difficulties in achieving the flexibility,
-scalability and maintainability expected of production software in experimental
-code written by data analysts.
-Although these properties cannot be enforced by any programming environment,
-suitable tools can help well-motivated analysts and programmers to create
-robust applications with much less effort.
+{\bf Shared artifacts and artifact histories} are a central feature.
+The prototype aims at making it convenient to collaborate on code
+and experiments with data. It was a success to observe that analysts
+readily shared ideas through RCloud. We observed them creating experiments
+in workbooks and sharing URLs to discuss data analysis and programming
+problems, instead of sending screenshots in Powerpoint.
+
+{\bf View sharing, bookmarking.} Likewise, everything is named
+and accessed as a URL.
+
+{\bf Discussion.} Annotation is supported in our human interface,
+and was also a central goal for our prototype.
+An interesting question is to decide to what extent the users of
+production notebooks also use tools like annotation etc. without coding
+or without being exposed to the analyst's view.\todo{Is this a gap in
+our design?}
+
+{\bf Content export} is not a capability we worked to support,
+but the R environment already has many packages for it.
+Our goal is to devise ways of working where explicit export is
+less necessary. However it is desirable to integrate results
+through web protocols such as JSON. We did assume that RCloud
+notebooks need to play well in the ecosystem of other tools
+and this worked out as planned, though it also leads to a temptation
+to absorb features almost without limit. (Should RCloud support
+templates for personal web pages? Should it incorporate emacs
+on the web, and an email client?)
+
+{\bf Social-psychological incentives} and {\bf voting and ranking}
+are supported through starring. This mechanism was employed often
+by our test audience. An obvious next step would be to enhance
+recommendations using relationships discovered by static and
+dynamic code analysis. This should be done both within and across
+collections of scripts (the latter being similar to work in Vistrails
+that enhanced recommendations by clustering multiple workflows).
+It may be valuable to data analysts to know which packages are often
+used together, or with a given type of data. It is probably helpful
+to know which data sources and packages often go together.
+
+{\bf Group management, size and diversity.} This area lacks sufficient
+support in RCloud, but is clearly desirable. We rely on conventional
+administrative and social processes, outside our platform. Automation
+of this capability in a large organization could rely on social media
+services
+
+\subsection{Limitations}
+
+Because we developed the ideas and the prototype together,
+we could not forsee some of the requirements that emerged
+after our audience started working with the prototype.
+
+It was unexpected that we had to had to back off so soon on
+the assumption that programming is being done in only one
+language (R). In retrospect, it should have been obvious that
+many data science problems are already being solved with
+cross-language approaches.
+
+A clear disdavantage is that some important aspects of the environment in
+RCloud are more difficult to manage than in conventional systems. Not only
+do we lack an explicit separation of the development and deployment
+environments, but every version of every prototype is shared by default.
+While this encourages collaboration on work in progress, errors such as
+misconfiguration, mismatches between versions of packages,or ordinary
+programming errors are also easily shared and could affect production
+websites unintentionally. We take the position that this cannot entirely
+be avoided, but the power of sharing is worth this price (just as that
+has proven true with distributed systems in general). Still, this 
+problem needs more attention. 
+Versioning occurs in several places: in scripts, in the R environment
+such as the installed R libraries and packages, and in the external
+environent such as the operating system and even protocols spoken by
+remote services. At one end of the scale, we have full control over the
+versioning of scripts via git, though there is still a need for conventions
+for example to name stable or working versions of scripts or versions known
+to work together. The R environment is under the control of its package
+manager. It is at least possible for RCloud to access information about
+package versions and configurations, but support for checking compatibility
+and maintaining multiple versions in the same environment is not strong.
+At the other end of the scale, there is not much reason to expect version
+control for the external environment. Most programmers rely on clean living
+and a conservative approach to upgrades.
+
+On top of this, RCloud allows analysts to fork experiments to try
+new ideas quickly. But we have not done much so far to address the
+probem of having a lot of bits and pieces of code and data lying
+around, though at least we have made it easier to get at them and
+to apply algorithms and metadata to organize them.
 
@@
 
-Had to back off only one language sooner than we expected.
+What is relationship to a standard CMS with R/Python
 
-Shared artifacts, artifact histories, 
-Discussion
-View sharing - bookmarking, everything available as a URL
-Content export - want to avoid, create ways of working where this is not necessary.
-Social-psychological incentives - starring
-Voting and ranking - starring
+Discussion about binding times, running vs.caching.
 
-Group management, size, diversity.  Lacking support,but clearly desirable.
-
-Has to play well in the ecosystem of the other tools and worked out as planned but there is a temptation to absorb
-
-What is relationship to a standard CMS with R/Python
+Transparent vs. explicit operations done by programmers

### 2014-03-26 discussion.tex
yes

@@
 Experience with deploying RCloud in a community of data analysts
-has provided some insight into how its capabilities were received
-that may be informative to researchers and designers of future systems
-supporting visual analytics.
+has provided some insight into how well the proposed requirements
+satisfied their needs. This may be informative to researchers and
+designers of future systems supporting visual analytics.
+
 Returning to the design considerations outlined by Heer and Agrawala,
@@
 {\bf Shared artifacts and artifact histories} are a central feature.
-The prototype aims at making it convenient to collaborate on code
+The prototype aimed at making it convenient to collaborate on code
 and experiments with data. It was a success to observe that analysts
 readily shared ideas through RCloud. We observed them creating experiments
-in workbooks and sharing URLs to discuss data analysis and programming
+in workbooks and sharing URLs to discuss analysis and programming
 problems, instead of sending screenshots in Powerpoint.
@@
 {\bf View sharing, bookmarking.} Likewise, everything is named
-and accessed as a URL.
+and accessed as a URL. This turned out to be an effective mechanism
+to support sharing and integration of analyses in notebooks
+with external processes and systems.
 
 {\bf Discussion.} Annotation is supported in our human interface,
-and was also a central goal for our prototype.
-An interesting question is to decide to what extent the users of
-production notebooks also use tools like annotation etc. without coding
-or without being exposed to the analyst's view.\todo{Is this a gap in
+and was also a central goal for our prototype. \todo{Elaborate.}
+An interesting question is to what extent users of production notebooks
+should be able to use tools like annotation without coding and
+without being exposed to the analyst's view.\todo{Is this a gap in
 our design?}
@@
 {\bf Content export} is not a capability we worked to support,
-but the R environment already has many packages for it.
-Our goal is to devise ways of working where explicit export is
+but the R environment already has many packages for this.
+Our goal was to devise ways of working where explicit export is
 less necessary. However it is desirable to integrate results
 through web protocols such as JSON. We did assume that RCloud
-notebooks need to play well in the ecosystem of other tools
-and this worked out as planned, though it also leads to a temptation
-to absorb features almost without limit. (Should RCloud support
-templates for personal web pages? Should it incorporate emacs
-on the web, and an email client?)
+notebooks need to play well in the ecosystem of other tools,
+and this worked out as intended, though we learned it also creates
+a temptation to absorb features almost without limit. (Should
+RCloud support templates for personal web pages? Should it
+incorporate emacs on the web, and an email client?)
 
@@
 used together, or with a given type of data. It is probably helpful
-to know which data sources and packages often go together.
+to know which data sources and packages often go together. In general,
+though, trivial or even passive mechanisms to collect data for
+recommendations are essential.
 
@@
 support in RCloud, but is clearly desirable. We rely on conventional
-administrative and social processes, outside our platform. Automation
-of this capability in a large organization could rely on social media
-services
+external administrative and social processes, outside our platform,
+to manage accounts and groups.  Automation of this capability in a
+large organization could rely on social media services
 
@@
 Transparent vs. explicit operations done by programmers
+
+Our environment assumes a high level of trust within the user community.
+Access control for information security is implemented by relying on the
+host operating system and web server. We admit this may require further

### 2014-03-27 discussion.tex
yes

@@
-\section{Lessons Learned, Discussion and Limitations}
+\section{Discussion, Lessons and Limitations}
 
@@
 has provided some insight into how well the proposed requirements
-satisfied their needs. This may be informative to researchers and
-designers of future systems supporting visual analytics.
+were met and matched their work. This may be informative to
+researchers and designers of future systems supporting visual analytics.
 
@@
 
-{\bf Shared artifacts and artifact histories} are a central feature.
-The prototype aimed at making it convenient to collaborate on code
-and experiments with data. It was a success to observe that analysts
-readily shared ideas through RCloud. We observed them creating experiments
+{\bf Shared artifacts and artifact histories} are a central feature
+of RCloud. The prototype aimed at making it convenient to collaborate
+on code and experiments with data. It was a success to observe that analysts
+readily shared ideas through RCloud. We noticed them creating experiments
 in workbooks and sharing URLs to discuss analysis and programming
-problems, instead of sending screenshots in Powerpoint.
+problems, instead of sending screenshots in PowerPoint.\todo{Talk about artifact histories?}
 
-{\bf View sharing, bookmarking.} Likewise, everything is named
-and accessed as a URL. This turned out to be an effective mechanism
-to support sharing and integration of analyses in notebooks
-with external processes and systems.
+{\bf View sharing, bookmarking.} Almost everything in RCloud is named
+and accessed as a URL. This proved to be an effective mechanism
+for sharing, and for analyses in notebooks to be integrated with
+external processes and systems.
 
 {\bf Discussion.} Annotation is supported in our human interface,
-and was also a central goal for our prototype. \todo{Elaborate.}
-An interesting question is to what extent users of production notebooks
-should be able to use tools like annotation without coding and
-without being exposed to the analyst's view.\todo{Is this a gap in
+and was another central goal of our prototype. \todo{Elaborate.}
+An interesting question is, to what extent should users of production
+notebooks also able to use annotation and similar tools without coding
+and without being exposed to the analyst's view?\todo{Is this a gap in
 our design?}
@@
 but the R environment already has many packages for this.
-Our goal was to devise ways of working where explicit export is
-less necessary. However it is desirable to integrate results
+Our goal was to devise ways of working where explicit export
+is less necessary. However it is desirable to integrate results
 through web protocols such as JSON. We did assume that RCloud
@@
 a temptation to absorb features almost without limit. (Should
-RCloud support templates for personal web pages? Should it
-incorporate emacs on the web, and an email client?)
+RCloud support templates for personal web pages and graphical
+editing? Should it incorporate an email client and emacs?
+Should it be able to push results to Facebook?)
 
@@
 by our test audience. An obvious next step would be to enhance
-recommendations using relationships discovered by static and
+recommendations via relationships discovered by static and
 dynamic code analysis. This should be done both within and across
-collections of scripts (the latter being similar to work in Vistrails
-that enhanced recommendations by clustering multiple workflows).
-It may be valuable to data analysts to know which packages are often
-used together, or with a given type of data. It is probably helpful
-to know which data sources and packages often go together. In general,
-though, trivial or even passive mechanisms to collect data for
-recommendations are essential.
+collections of scripts (the latter being similar to Vistrails'
+enhanced recommendations by clustering multiple workflows).
+It should be valuable to data analysts to know which packages are
+frequently used together, or appear with a given record type
+or data feed. In general, though, trivial or passive mechanisms
+to collect data for recommendations are essential.
 
 {\bf Group management, size and diversity.} This area lacks sufficient
-support in RCloud, but is clearly desirable. We rely on conventional
-external administrative and social processes, outside our platform,
-to manage accounts and groups.  Automation of this capability in a
-large organization could rely on social media services
+support in RCloud, but is clearly important to working in teams.
+We rely on existing administrative and social processes and conventions,
+outside RCloud, to manage accounts and groups. This capability
+could likely benefit from integrating social media to track
+identities and groups and to maintain communication channels.
 
@@
 Because we developed the ideas and the prototype together,
-we could not forsee some of the requirements that emerged
+we did not foresee some of the requirements that emerged
 after our audience started working with the prototype.
 
-It was unexpected that we had to had to back off so soon on
-the assumption that programming is being done in only one
-language (R). In retrospect, it should have been obvious that
-many data science problems are already being solved with
-cross-language approaches.
+We did not anticipate having to back off so soon on the
+assumption that programming is being done mainly in one
+language (R). In retrospect, it should have been obvious
+that many data science problems are already being solved
+with cross-language approaches.
 
-A clear disdavantage is that some important aspects of the environment in
-RCloud are more difficult to manage than in conventional systems. Not only
-do we lack an explicit separation of the development and deployment
-environments, but every version of every prototype is shared by default.
-While this encourages collaboration on work in progress, errors such as
-misconfiguration, mismatches between versions of packages,or ordinary
-programming errors are also easily shared and could affect production
-websites unintentionally. We take the position that this cannot entirely
-be avoided, but the power of sharing is worth this price (just as that
-has proven true with distributed systems in general). Still, this 
-problem needs more attention. 
+A disadvantage is that some important aspects of the environment
+are more difficult to manage in RCloud than in conventional systems.
+Not only does it lack an explicit separation of the development and
+deployment environments, but every version of every prototype is
+shared by default. While this encourages collaboration on work
+in progress, errors such as misconfiguration, mismatches between
+versions of packages, and ordinary programming errors are also
+easily shared and can affect production websites unintentionally.
+This cannot be avoided entirely, but the power of sharing is worth
+this price (as proven true with distributed systems in general).
+Still, this problem needs more attention if the environment is
+going to be scaled up to a wider number of users with fewer
+informal channels to coordinate work. \todo{need a transition}
 Versioning occurs in several places: in scripts, in the R environment
 such as the installed R libraries and packages, and in the external
-environent such as the operating system and even protocols spoken by
+environment such as the operating system and even protocols spoken by
 remote services. At one end of the scale, we have full control over the
@@
 new ideas quickly. But we have not done much so far to address the
-probem of having a lot of bits and pieces of code and data lying
+problem of having a lot of bits and pieces of code and data lying
 around, though at least we have made it easier to get at them and
@@
 
-Our environment assumes a high level of trust within the user community.
-Access control for information security is implemented by relying on the
-host operating system and web server. We admit this may require further
+Our environment assumes a high level of trust between its users.
+Access control for information security is deferred to the host
+operating system and web server. This decision should be re-examined,
+but more sophisticated approaches such as information flow analysis
+are an active research area and many subtle problems remain
+\cite{Moore:2011:SAF}. We are hopeful that providing a richer
+environment for collaboration and information sharing may aid
+the application of information security research.

### 2014-03-27 relatedwork.tex
yes

@@
 exploratory work in data science teams. As a result, RCloud tries to
-provide a closer data-analysis to analysts accustomed to createing and
+provide a closer data-analysis to analysts accustomed to creating and
 executing R and Python code while keeping many of the attractive
@@
 shares many of our goals, such as providing a comprehensive environment
-for analysis and programming, with sharable documents on the web.
+for analysis and programming, with shareable documents on the web.
 
-One overall goal is to reduce the gap between implementors and deployers
+One overall goal is to reduce the gap between implementers and deployers
 of technology. Don't want a team of 20 IT people to support 5 data scientists.

### 2014-03-27 system.tex
yes

@@
 high-level protocol, usually HTTP.  HTTP is dominant because web
-browsers and servers are ubiquituous and available in essentially
+browsers and servers are ubiquitous and available in essentially
 all hardware devices, from tiny sensors, to handheld devices and
@@
 We incorporate both explicit and implicit indications of interest
-in notebooks. Explicit interest is signalled by ``starring,'' or
+in notebooks. Explicit interest is signaled by ``starring,'' or
 clicking on a button that marks a notebook as interesting. 
@@
 broad access to analysis outputs (in their case, for NameVoyager)
-increases long-term engagement in part through crossreferences on
+increases long-term engagement in part through cross-references on
 the web. Although our prototype RCloud deployment is only visible

### 2014-03-27 discussion.tex
yes

@@
 has provided some insight into how well the proposed requirements
-were met and matched their work. This may be informative to
-researchers and designers of future systems supporting visual analytics.
+were met and were appropriate to the tasks in their work.
+This experience may be informative to researchers and designers
+of future systems supporting collaborative visual analytics.
 
-Returning to the design considerations outlined by Heer and Agrawala,
-our experience underscores the value of several of their proposals,
-and suggests areas for further exploration.
+Returning to the design considerations defined by Heer and Agrawala,
+our experience underscores the value of many of their proposals,
+and indicates areas for further exploration.
 
@@
 of RCloud. The prototype aimed at making it convenient to collaborate
-on code and experiments with data. It was a success to observe that analysts
-readily shared ideas through RCloud. We noticed them creating experiments
+on experimental workbooks and the analyses that result. We observed
+that analysts readily shared ideas through RCloud. 
+creating experiments
 in workbooks and sharing URLs to discuss analysis and programming
@@
 
-{\bf View sharing, bookmarking.} Almost everything in RCloud is named
-and accessed as a URL. This proved to be an effective mechanism
-for sharing, and for analyses in notebooks to be integrated with
+{\bf View sharing, bookmarking.} Most resources in RCloud are named
+and accessed as URLs. This proved to be an effective mechanism
+for sharing, and for analyses to be integrated with
 external processes and systems.

### 2014-03-27 discussion.tex
yes

@@
 Experience with deploying RCloud in a community of data analysts
-has provided some insight into how well the proposed requirements
-were met and were appropriate to the tasks in their work.
+has provided some insight about how the proposed requirements
+were met, and to what extent they were appropriate.
 This experience may be informative to researchers and designers
@@
 Returning to the design considerations defined by Heer and Agrawala,
-our experience underscores the value of many of their proposals,
+our experience underscores the importance of many of their proposals,
 and indicates areas for further exploration.
@@
 on experimental workbooks and the analyses that result. We observed
-that analysts readily shared ideas through RCloud. 
-creating experiments
-in workbooks and sharing URLs to discuss analysis and programming
-problems, instead of sending screenshots in PowerPoint.\todo{Talk about artifact histories?}
+that analysts readily shared ideas through RCloud. They used RCloud
+to discuss data analysis, programming problems, and to publish results
+by sharing URLs instead of instead of embedding screenshots in
+PowerPoint.\todo{Talk about artifact histories?}
 

### 2014-03-27 relatedwork.tex
yes

@@
 various technical communities \cite{Perer:2008:ISA}
+Lucas and Roth were early advocates of combining
+data exploration with presentation and publication \cite{Lucas:1996:EIV}.
 
@@
 One overall goal is to reduce the gap between implementers and deployers
-of technology. Don't want a team of 20 IT people to support 5 data scientists.
-
-% devops for data science.
+of technology in visual analytics. The fusion of development with production
+operations in software release management ("devops" \cite{Httermann:2012:DD}) is a trend in web services and related fields. There is certainly an economic
+benefit to making it possible for data scientists to expend a little more
+effort in creating experiments, to eliminate the need for a programming
+team to recreate their results in production.

### 2014-03-27 conclusion.tex
yes

\section{Conclusions and Future Work}

We presented the ideas that motivated the design of RCloud,
a prototype visual analytics environment to support
collaboration, searching, sharing, and publishing of
experiments, similar to DevOps development of web services.
Experience with the prototype provided evidence that
data science teams and the organizations in which they work
can benefit from having such features.

Future work will address cross-language development,
source code analysis, richer recommendation techniques,
and improvements to the HTML5 front end.

The code is available 
at \url{github.com/att/rcloud/}
under an MIT open source license.

### 2014-03-27 discussion.tex
yes

@@
 
-What is relationship to a standard CMS with R/Python
-
 Discussion about binding times, running vs.caching.
@@
 
-Our environment assumes a high level of trust between its users.
-Access control for information security is deferred to the host
-operating system and web server. This decision should be re-examined,
-but more sophisticated approaches such as information flow analysis
-are an active research area and many subtle problems remain
-\cite{Moore:2011:SAF}. We are hopeful that providing a richer
-environment for collaboration and information sharing may aid
-the application of information security research.
+Our environment assumes a high degree of trust between users.
+Access control for information security is delegated to the host
+operating system and web server. In practice, most operating
+systems and web servers rely either on coarse permission models,
+which tend to be ineffective, or on detailed access control
+lists, which tend to be cumbersome and complicated. 
+Information security in RCloud should be re-examined, but more
+sophisticated approaches such as information flow analysis are
+still an active research area, and many subtle problems remain
+\cite{Moore:2011:SAF}. We are hopeful that having a richer
+environment for collaboration and information sharing will encourage
+practical applications for new approaches to information security
+in visual analytics.

### 2014-03-27 introduction.tex
yes

@@
 
-Main requirement is to satisfy these two sets of users: R coders and
+Our main requirement is to satisfy these two sets of users: R coders and
 managers. They work differently and want different things. What should
@@
 making it easy to push new features, prototypes, etc.
-\item Dashboarding. To support the above, there should be a mode in which non-experts
-can operate the experiments without coding.
+\item Dashboarding. To support the above, there should be a mode
+in which non-experts can operate the experiments without coding.
+This may require some accomodation in code written by analysts,
+but our goal is to make that as simple as we can.
 \item Version control as a primary concept in the repository. There
-needs to be a way to control the visibility of prototypes as they are
-being developed and tested.
-\item Coexistence in the current data ecosystem. We perceive the
-isolation of current exploratory visualization and data analysis
-environments to be a serious threat to a wider adoption of modern
-techniques.
-\item Modern interactive technologies. current isolation
-notwithstanding, we find there is business demand for interactive
-visualization tools. The main challenge here is how to integrate
-these technologies so data scientists and statisticians can target
-them, \emph{and} business-facing users can use them.
+needs to be a way to control the visibility of prototypes as they
+are developed and tested.
+\item Coexistence in the current data ecosystem. The isolation of
+today's exploratory visualization and data analysis environments is
+a real obstacle to wider adoption of novel data analysis methods.
+\item Modern interactive technologies. Current isolation notwithstanding,
+there is demand for interactive visualization tools. In some respects,
+the prevalence of data journalism has raised expectations very high.
+The main challenge is how to integrate these technologies so data
+scientists and statisticians can program in them, \emph{and} business-facing
+workers can operate them.
 %% Scalable technologies like crossfilter exist.
@@
 for statistical computing and graphics. R is one of the most popular
-systems for data analysis and is well-supported by
-a complex ecosystem of compatible packages, software
-tools and other resources.
+data analysis systems and is well-supported by a complex ecosystem
+of compatible packages, software tools and other resources.
 %
-Of importance to us, there are existing
-packages for high performance computing in the cloud,
+Of importance to us, there are many existing R packages, including
+packages for high performance computing in the cloud, domain specific
+models,
 and for generating web graphics. % no, there aren't. we wrote those!
+% well, right about that
 

### 2014-03-27 relatedwork.tex
yes

@@
 
-Our work has been inspired by and benefited from other proposals
+Our work has been inspired by and benefited from other efforts
 to improve data analysis environments and processes,
 including RStudio \cite{RStudio:2013:SWA},
-R packages such as Markdown \cite{Allaire:2014:MMR},
+the R packages Markdown \cite{Allaire:2014:MMR},
 knitr \cite{Xie:2013:DDW}
@@
 to name a few. RStudio aims at providing an integrated development environment
-for R programming, with support for publishing code in packages. Markdown,
+for R, with support for publishing code in packages. Markdown,
 knitR and Shiny provide R with sophisticated reporting capabilities, including
@@
 shares many of our goals, such as providing a comprehensive environment
-for analysis and programming, with shareable documents on the web.
+for analysis and programming, and shareable documents on the web.
 
@@
 of technology in visual analytics. The fusion of development with production
-operations in software release management ("devops" \cite{Httermann:2012:DD}) is a trend in web services and related fields. There is certainly an economic
-benefit to making it possible for data scientists to expend a little more
-effort in creating experiments, to eliminate the need for a programming
-team to recreate their results in production.
+operations in software release management ("devops" \cite{Httermann:2012:DD})
+is a trend in web services and related fields. By making it possible
+for data scientists to expend a little more effort in the process of
+creating experiments, it may be possible to eliminate the need for
+programming teams to recreate their results in production with
+a concomitant penalty in time and expense.

### 2014-03-27 system.tex
yes

@@
 high-level protocol, usually HTTP.  HTTP is dominant because web
-browsers and servers are ubiquitous and available in essentially
+browsers and servers are ubiquitous and available in almost
 all hardware devices, from tiny sensors, to handheld devices and
@@
 
+What is relationship to a standard CMS with R/Python
+
 dcplot. two-way communication between between backend session and

### 2014-03-27 system.tex
yes

@@
 \cite{Joachims:2005:AIC} and execution counts. (In addition to these
-classic web search techniques of collecting feedback, we have the
-prospect of applying static and dynamic code analysis to generate
-further fine-grained information about relationships, for example,
-which packages and data sets often appear together.)
-
-\todo{explain} These will be ingredients for the recommendation system.
+classic techniques of collecting feedback from web search, we anticipate
+applying static and dynamic code analysis to generate fine-grained
+information about relationships, for example, which packages and data
+sets often appear together.)
 
@@
 A key engineering decision in RCloud was to rely on full two-way
-communication between human interface in a web client, and remote
-data and computing resources in the cloud.
+communication between the web client human interface, and data and
+computing resources in the cloud. It is easier to engineer a design
+in which the human interface is pushed one-way from the server
+and thereafter all interaction takes place within the client.
+Two-way communication is necessary, though, where the size of
+the data or computational performance makes it impractical to
+move all computation to the client.

### 2014-03-28 conclusion.tex
yes

@@
 
-We presented the ideas that motivated the design of RCloud,
-a prototype visual analytics environment to support
-collaboration, searching, sharing, and publishing of
-experiments, similar to DevOps development of web services.
-Experience with the prototype provided evidence that
+We described the motivation for RCloud, a prototype
+visual analytics environment that supports collaboration,
+searching, sharing, and publishing of experiments.
+This assists software development and release using
+practices similar to DevOps in web services.
+Experience with the Rcloud prototype provided evidence that
 data science teams and the organizations in which they work
-can benefit from having such features.
+benefit from having such capabilities.
 

### 2014-03-28 discussion.tex
yes

@@
 Experience with deploying RCloud in a community of data analysts
+(hackers in the parlance of Heer and Agrawala)
 has provided some insight about how the proposed requirements
@@
 Returning to the design considerations defined by Heer and Agrawala,
-our experience underscores the importance of many of their proposals,
-and indicates areas for further exploration.
+our experience underscores the importance of their proposals,
+and indicates areas for further exploration, especially in the
+following areas.
 
@@
 of RCloud. The prototype aimed at making it convenient to collaborate
-on experimental workbooks and the analyses that result. We observed
-that analysts readily shared ideas through RCloud. They used RCloud
+on experimental workbooks and the resulting analyses. We observed
+that hackers readily shared ideas through RCloud. They used RCloud
 to discuss data analysis, programming problems, and to publish results
 by sharing URLs instead of instead of embedding screenshots in
-PowerPoint.\todo{Talk about artifact histories?}
+PowerPoint. Artifact histories are provided through git. It is fair
+to criticize that it is not human-friendly, but we assume RCloud
+hackers are already familiar with it.
+
+{\bf Modularity and granularity.} The ability to segment work into
+independent units (modularity) is a key to working in teams. It is
+further an advantage if the units may be kept small, and team members
+can realize benefits that are at least proportional to their work
+on these units (granularity). RCloud's notebook and versioning
+capabilities allow work to be divided into units as fine as
+the underlying language allows, and versioning encourages making
+incremental changes at low cost and without disrupting the work
+of others.
 
 {\bf View sharing, bookmarking.} Most resources in RCloud are named
-and accessed as URLs. This proved to be an effective mechanism
-for sharing, and for analyses to be integrated with
-external processes and systems.
+and accessed as URLs. This proved to be an effective mechanism for
+sharing, and for analyses to be integrated with external processes
+and systems.
 
-{\bf Discussion.} Annotation is supported in our human interface,
-and was another central goal of our prototype. \todo{Elaborate.}
-An interesting question is, to what extent should users of production
-notebooks also able to use annotation and similar tools without coding
-and without being exposed to the analyst's view?\todo{Is this a gap in
-our design?}
+{\bf Discussion.} Annotation and commenting was another central goal
+of our prototype, and is provided in our human interface.
+An interesting question is, to what extent should application users
+be able to make annotations and operate similar tools in published
+notebooks without coding and being exposed to the hackers's view?
+Our prototype has not addressed this problem yet.
 
 {\bf Content export} is not a capability we worked to support,
-but the R environment already has many packages for this.
-Our goal was to devise ways of working where explicit export
-is less necessary. However it is desirable to integrate results
+but R already has many packages for this. Our goal was to devise
+ways of working where explicit export is less necessary.
+However it is desirable to integrate results
 through web protocols such as JSON. We did assume that RCloud
@@
 enhanced recommendations by clustering multiple workflows).
-It should be valuable to data analysts to know which packages are
+It should be valuable to hackers to know which packages are
 frequently used together, or appear with a given record type
@@
 we did not foresee some of the requirements that emerged
-after our audience started working with the prototype.
+after people started working with the prototype.
 
@@
 going to be scaled up to a wider number of users with fewer
-informal channels to coordinate work. \todo{need a transition}
-Versioning occurs in several places: in scripts, in the R environment
+informal channels to coordinate work. Effective sharing in RCloud
+will require control over versioning.
+Versioning must be managed in several places: in scripts, in the R environment
 such as the installed R libraries and packages, and in the external
@@
 
-On top of this, RCloud allows analysts to fork experiments to try
-new ideas quickly. But we have not done much so far to address the
+On top of this, RCloud allows hackers to fork experiments to try
+new ideas quickly. So far we have not done much more to address the
 problem of having a lot of bits and pieces of code and data lying

### 2014-03-28 introduction.tex
yes

@@
 beyond what existing visual analytics environments provide, or what
-is found in today's repositories and version control systems.
+is found in current repositories and version control systems.
 
@@
 there is demand for interactive visualization tools. In some respects,
-the prevalence of data journalism has raised expectations very high.
+the prevalence of data journalism has raised expectations very high,
+but handmade, one-off designs are too expensive to most teams to support.
 The main challenge is how to integrate these technologies so data
@@
 Heer/Agrawala design considerations for collaborative visual analytics.
+Commenting, annotation, modularity, granulatiry.
 \end{itemize}
@@
 for statistical computing and graphics. R is one of the most popular
-data analysis systems and is well-supported by a complex ecosystem
-of compatible packages, software tools and other resources.
+data analysis systems and is well-supported by a broad ecosystem
+of packages, tools and other resources.
 %
-Of importance to us, there are many existing R packages, including
-packages for high performance computing in the cloud, domain specific
-models,
-and for generating web graphics. % no, there aren't. we wrote those!
-% well, right about that
-
-% We want our solution to play well with the rest of the ecosystem
+There are many existing R packages, including packages for high performance
+computing in the cloud, domain specific data analysis, statistical computing,
+and web graphics (though the latter has to be adapted for our architecture).
+An additional requirement for our prototype is to work well with the rest
+of this ecosystem.

### 2014-03-28 discussion.tex
yes

@@
 
-{\bf Discussion.} Annotation and commenting was another central goal
-of our prototype, and is provided in our human interface.
+{\bf Discussion.} Annotation and commenting was another central goal.
+Commenting is supported through github, but our hackers found it
+awkward and did not exercise it as much as we expected.
+making it a bit awkward and
+not as widely used as we hoped.  More elaborate annotation was not
+implemented, and remains as future work.
 An interesting question is, to what extent should application users

### 2014-03-28 introduction.tex
update

--- 2014-03-25 introduction.tex
@@
 organization today \cite{Keim:2008:VAS}.
-
+%
 % How do people do it today?
@@
 % Why does this not work?
-There are two problems with the current
-practice. First, there are gaps in the workflow:
-Exploratory Data Analysis (EDA) is done with one set of tools,
-and automated reports and deployments with another. This makes
-transferring results between EDA and production (in both directions)
-slow, expensive, and error-prone. Coordination is done through
-meetings and may take months to implement.
+There are two problems with the current practice. First, there are
+gaps in the workflow: Exploratory Data Analysis (EDA) is done with one
+set of tools, and automated reports and deployments with another. This
+makes transferring results between EDA and production (in both
+directions) slow, expensive, and error-prone. Coordination is done
+through meetings and may take months to implement.
 
@@
 beyond what existing visual analytics environments provide, or what
-is found in today's repositories and version control systems.
+is found in today's repositories and version control
+systems. \todo{add cites to reco system work}
+
+In addition, take the recent trend in interactive visualization
+technologies. Let us consider prominent designers and developers of
+open source software, Jeffrey Heer, and Hadley Wickham. Heer has
+written (or helped write) Prefuse, Flare, Protovis, d3, and Vega, in
+that order. All but the first of these libraries target web
+technologies, and all but the second use open standards. Wickham has
+written ggplot, ggplot2, and ggvis. ggplot and
+ggplot2 generate static graphics, while ggvis uses Shiny and Vega to
+provide easy interactive visualizations in R, by opening a separate
+web server for each call. \todo{Add cites for all of these}
 
 % What are we going to do about this?
-\todo{Sell idea, not system} In this paper, we investigate the hypothesis
-that the practice of visual analytics can advance by adopting techniques
-from information retrieval and collaborative work environments. To investigate
-these ideas, we designed and deployed a prototype environment called RCloud,
-that supports collaborative data analysis, visualization and web deployment.
-We will report on the design decisions, tradeoffs and limitations, and compare
-the concepts in RCloud to similar proposals.
 
-\todo{Define the requirements for the system}
+%% \todo{Define the requirements for the system}
+%% We begin by noting the main requirements for a data analysis
+%% environment for collaboration and transfer of results into production.
 
-We begin by noting the main requirements for a data analysis environment for
-collaboration and transfer of results into production.
+%% Main requirement is to satisfy these two sets of users: R coders and
+%% managers. They work differently and want different things. What should
+%% the infrastructure look like so they both coexist in peace? If this is
+%% such a fundamental problem, why have other tools not addressed it?
+%% Discuss Shiny (maybe in the disc. section).
 
-Main requirement is to satisfy these two sets of users: R coders and
-managers. They work differently and want different things. What should
-the infrastructure look like so they both coexist in peace? If this is
-such a fundamental problem, why have other tools not addressed it?
-Discuss Shiny (maybe in the disc. section).
+We see the present state-of-the-art in environments for data science
+teams as tremendously powerful, but ultimately fragmented. Rather than
+design a completely new solution that risks even further
+fragmentation, we wish to find and remedy the shortcomings of data
+science as we see it practiced today. The following requirements
+summarize what we consider to be the main pain points:
 
-\begin{itemize}
-\item Constant technology transfer. Experiments in notebooks developed
-by analysts need to become accessible in the production environment,
-making it easy to push new features, prototypes, etc.
-\item Dashboarding. To support the above, there should be a mode in which non-experts
-can operate the experiments without coding.
-\item Version control as a primary concept in the repository. There
-needs to be a way to control the visibility of prototypes as they are
-being developed and tested.
-\item Coexistence in the current data ecosystem. We perceive the
+\begin{enumerate}
+
+\item Technology transfer. In our experience, development
+of new analyses and visualizations is done by \emph{hackers}
+and \emph{scripters} (where here are use Kandel et al.'s
+terminology~\cite{Kandel:2012:EDA}) who are, by definition, familiar
+to some extent with programming and software development.
+On the other hand, \emph{application
+users} are the ones extracting value from the data using the tools
+developed by hackers and scripters. Traditionally, the communication
+between these two world is done by IT staff. In an environment where
+business needs change possibly many times a week, manual IT support
+does not scale: there needs to exist
+infrastructure to connect the two worlds.
+%% The two classes of users
+%% have quite different perspectives but we
+%% infrastructure for continuous deployment of visualizations.
+
+\item Coexistence. In the current data science ecosystem, the
 isolation of current exploratory visualization and data analysis
-environments to be a serious threat to a wider adoption of modern
-techniques.
-\item Modern interactive technologies. current isolation
-notwithstanding, we find there is business demand for interactive
-visualization tools. The main challenge here is how to integrate
-these technologies so data scientists and statisticians can target
-them, \emph{and} business-facing users can use them.
+environments is a threat to wider adoption of modern techniques for
+both sides. We specifically see the richness of interactive
+visualization tools isolated from the richness of statistical
+programming environments.
+We wish to provide infrastructure such that
+developments on one side can more easily be employed on the other. In
+addition, this system itself should be easily accessible
+to the rest of the organization.
+
+\item Discoverability. From our experience in working with data
+analyst teams, we find that repeated work (``how can I connect to
+database X?'', ``how do you get clean data from column Y from feed
+Z?'') typically arises from lack of communication of solved
+subproblems. We wish to provide transparency about the work of
+team members in the organization.
+
+\end{enumerate}
+
+Technologies which fail to address these requirements
+are holding back the practice and impact of interactive visual
+analysis. In this paper, we discuss the design and deployment of
+RCloud, a prototype environment that supports collaborative data
+analysis, visualization and web deployment satisfying all requirements
+above. We will report on the design decisions, tradeoffs and
+limitations, and compare the concepts in RCloud to similar proposals.
+
+RCloud uses the R language~\cite{RCoreTeam:2013:R} as the main provider of
+statistical analysis tools. In fact, as we describe in
+Section~\todo{where?}, RCloud notebooks are sequences of R snippets
+interspersed with Markdown for easy generation of reports. In
+Section~\todo{where?}, we describe how RCloud enables
+interactive visualization widgets in this setting. In
+Section~\ref{sec:limitations}, we discuss some consequences of
+designing for a single language, and how this was received by
+our (small, but active) user community.
+
+%% Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
+%% for statistical computing and graphics. R is one of the most popular
+%% systems for data analysis and is well-supported by
+%% a complex ecosystem of compatible packages, software
+%% tools and other resources.
+%
+%% Of importance to us, there are existing
+%% packages for high performance computing in the cloud,
+%% and for generating web graphics. % no, there aren't. we wrote those!
+
+The facilities for \emph{sharing, viewing and embedding} RCloud
+notebooks address technology transfer, and are described in
+Section~\todo{where?}. RCloud notebooks are edited and shared as web
+pages. In order to support the tight integration of R and HTML5
+technologies, RCloud uses a full, bi-directional communication channel
+between the web browser and a running R process, described in
+Section~\todo{where?}.
+
+Because RCloud embraces technologies that are part of the computing
+environment (specifically, inter-process communication via HTTP and
+web services as remote procedure calls), we find it easier to
+integrate RCloud with satellite efforts in corporate environments. We
+discuss this architecture and consequences in
+Section~\ref{sec:system}.
+
+The requirement of continuous technology transfer creates challenges
+about the \emph{stability} of the many moving pieces: if notebooks are
+always updated and deployed automatically, how can the system handle
+changing requirements and public interfaces? We discuss our choice of
+transparent, integrated version control for RCloud notebooks in 
+Section~\ref{sec:notebooks}.
+
+%% Dashboards are a solution to requirement 1, using solutions of
+%% requirement 2.
+%% Requirement 2: \emph{how is accessibility supposed to happen?}
+
+%% Version control as a primary concept in the environment is a
+%% consequence of requirement 1. There
+%% needs to be a way to control the visibility of prototypes as they are
+%% being developed and tested.
+
+%% Tight integration between R process and web browser is a solution to
+%% requirement 2
+
+%% Current isolation notwithstanding, we find there is demand for interactive
+%% visualization tools. We do not believe that moving the \
+
+%% The main challenge here is how to integrate these
+%% technologies so data scientists and statisticians can target
+%% them, \emph{and} business-facing users can use them.
 %% Scalable technologies like crossfilter exist.
@@
 Heer/Agrawala design considerations for collaborative visual analytics.
-\end{itemize}
 
 % What is the scope of our solution?
-Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
-for statistical computing and graphics. R is one of the most popular
-systems for data analysis and is well-supported by
-a complex ecosystem of compatible packages, software
-tools and other resources.
-%
-Of importance to us, there are existing
-packages for high performance computing in the cloud,
-and for generating web graphics. % no, there aren't. we wrote those!
 

### 2014-03-28 paper.tex
merge
update

@@
 
-% \title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
-\title{RCloud: DevOps for Visual Analysis and Data Science}
+\title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
+%% \title{RCloud: DevOps for Visual Analysis and Data Science}
 
@@
   recommendations, which are well beyond what traditional version control
-  systems provide. \todo{sell idea, not system} In this paper, we propose RCloud, a system that supports
-  collaborative data analysis, visualization and web deployment. We will
+  systems provide. We contribute and justify the following three
+  requirements for systems built to support current data science teams
+  and users: \emph{technology transfer}, \emph{coexistence}, and
+  \emph{discoverability}. In addition, we contribute the design and
+  implementation of RCloud, a prototype system that supports the requirements
+  collaborative data analysis, visualization and web deployment. We
   discuss the design decisions, tradeoffs and limitations, and compare RCloud

### 2014-03-28 discussion.tex
yes

@@
 
+{\bf Multiple language support.}
 We did not anticipate having to back off so soon on the
@@
 
+{\bf Versioning.}
 A disadvantage is that some important aspects of the environment
@@
 
-Discussion about binding times, running vs.caching.
+{\bf Caching.}
+An important consideration is how and where to introduce caching
+in RCloud's distributed computation model. Caching can dramatically
+improve performance in a way that is otherwise transparent to
+application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
+Though caching is usually implicit, in some environments, such as
+Vistrails, programmers may also have some explicit control over cache
+management. This may be desirable to ensure the repeatability of
+computations that rely on volatile or unreliable data sources.
+For example, the stock ticker use case is one for which the data
+might temporarily become unavailable, so caching could improve
+reliability and consistency of results.  Alternatively, in a use case
+involving text analysis through a topic modeling technique such as
+Latent Dirichlet Allocation (LDA) that is relatively expensive,
+we might want caching to improve performance. Currently we would
+program this in RCloud by explicitly saving the analysis in a
+persistent database, but it would be desirable to implement this
+in a more general purpose associative cache. We envision cache
+management being implemented in a new middle layer not in RCloud's
+architecture today.
 
-Transparent vs. explicit operations done by programmers
-
+{\bf Security.}
+% We sprinkle the security dust on at the end!
 Our environment assumes a high degree of trust between users.

### 2014-03-28 discussion.tex
yes

@@
 {\bf Security.}
-% We sprinkle the security dust on at the end!
-Our environment assumes a high degree of trust between users.
+It is critical to consider security in an environment that
+
+RCloud uses an object capability model that prevents
+unauthenticated clients from making unauthorized accesses
+to RCloud objects \cite{
+
+Our back-end environment assumes a high degree of trust between users.
 Access control for information security is delegated to the host

### 2014-03-28 discussion.tex
yes

--- 2014-03-28 discussion.tex
@@
 {\bf Security.}
-% We sprinkle the security dust on at the end!
-Our environment assumes a high degree of trust between users.
+It is essential to provide security in an environment that encourages
+collaboration and data publishing.  RCloud uses an object capability model
+\cite{Miller:2006:RCT}
+recently added to the Rserve protocol \cite{Urbanek:2003:AFW}
+that prevents unauthenticated clients from making
+unauthorized calls to the RCloud runtime environment.
+
+Our back-end environment assumes a high degree of trust between users.
 Access control for information security is delegated to the host
@@
 which tend to be ineffective, or on detailed access control
-lists, which tend to be cumbersome and complicated. 
+lists, which tend to be cumbersome. 
 Information security in RCloud should be re-examined, but more

### 2014-03-28 discussion.tex
yes

@@
 {\bf Security.}
-It is essential to provide security in an environment that encourages
-collaboration and data publishing.  RCloud uses an object capability model
+It is essential to provide security in an environment for
+collaboration and data publishing. To provide some
+protection for RCloud workbooks in an organization's intranet,
+RCloud uses an object capability model
 \cite{Miller:2006:RCT}

### 2014-03-28 casestudies.tex
whatever
update
section ids
figure and case studies and discussion

@@
 
+\subsection{stock tickers\ref{sec:stockvis}}
+
+Basic stock ticker visualization via ggplot by Hadley Wickham.
+
+Easy to convert it into configurable entry point for dashboarding.
+
+Now consider an analyst that wants to understand the price dynamics of
+different stock attributes, different points in time, etc. For that,
+interactive visualization is very attractive. describe dcplot and
+notebook which uses it.
+
+Then describe abstraction of stock-data fetching, to talk about
+calling notebooks from other notebooks.
+
+then call.r?
+
+\subsection{text analysis\ref{sec:textvis}}
+
+Same.
+
 % IMPORTANT: what is unique about RCloud here
-
 % From prototyping to dashboard
-
 % Getting other information from the web
-
 %

### 2014-03-30 paper.tex
notebook figure and section

@@
   % \vspace{-.5em}
-  \caption{Left: RCloud integration of HTML5 visualization libraries such
-    as D3~\cite{Bostock:2011:DDD} and Crossfilter~\cite{Square:2014:CFM}.
+  \caption{RCloud provides infrastructure for data science teams so
+    that R \emph{hackers} and \emph{scripters}~\cite{Kandel:2012:EDA} can target modern
+    interactive visualization technologies such as
+    D3~\cite{Bostock:2011:DDD}, dc.js~\cite{DCJS:2014:DC} and
+    Crossfilter~\cite{Square:2014:CFM} (left), while at the same time
+    enabling seamless deployment and dashboarding for
+    \emph{application users} (right)
   }

### 2014-03-30 system.tex
update
notebook figure and section

@@
-\section{The System}
+\section{The System\label{sec:system}}
 
@@
 
-Notebooks as Github gists. 
+\begin{figure*}
+\includegraphics[width=\linewidth]{fig/notebook/notebook.pdf}
+\caption{\label{fig:notebook}An RCloud notebook is a sequence of
+  \emph{cells}, each a snippet of source in R or Markdown. Notebooks
+  are stored in GitHub \emph{gists}, which provide the version-control
+  primitives necessary for RCloud.}
+\end{figure*}
+
+The basic unit of operation in RCloud is a \emph{notebook}. RCloud
+notebooks hold a sequence of \emph{cells}, each of which represents a
+snippet of code or hypertext in Markdown. This is not a novel idea;
+executable documents structured in this way are present in a wide
+variety of present software systems, from Mathematica to IPython to
+Sage and others.
+
+One of the main differences between RCloud and other systems is the
+notion that notebooks are ``always deployed''. This has a variety of
+advantages which we discuss \todo{where?}. On the other hand, there
+are a variety of situations that might ask for a specific version of a
+notebook. Rather than have the notebook designer decide ahead of time
+which versions should be preserved, we choose to expose
+\emph{transparent} versioning. This is similar to models like
+Jankun-Kelly et al.'s p-set calculus \todo{cite} and VisTrails's
+version tree \todo{cite}, where every change in the state of the
+system is tracked.
+
+Implementing the entire version-control subsystem would be a large
+undertaking. Instead, we build the infrastructure on top of Github's
+\emph{gists}~\todo{cite}. Github offers a HTTP interface
+for creating simplified git repositories (the main limitation is the
+use of text-only files and a single directory). The GitHub web-service
+API provides much of the infrastructure we need for the versioning
+portion of the storage back end: access to previous versions,
+comments, starring, and forking.
+
+Using GitHub was mainly a tactical decision to reduce our
+time-to-prototype. At the same time, it had the fortuitous decision of
+forcing our backend to communicate with separate processes over HTTP
+and JSON. This had the serendipitous advantage of quickly enabling
+full search over RCloud notebooks (built using Apache SOLR).
 %
-We needed version control over a directory, gist provides that,
-wrapped over an HTTP interface.
+This perspective has implications for
+systems-based research and prototyping in visual analytics.
 %
-Mainly a tactical decision to save time.
+In order for small teams to efficiently explore ideas in the present
+software ecosystem, it pays off to use as many existing solutions as
+possible.
 %
-Nevertheless, availability of notebook data as (barely) plain text had good side
-effects, like ease to build search. 
-%
-Expect same with suggest.
+Even though it involved learning a variety of standards that are not
+directly related to visual analytics, our early decision to use the
+GitHub gist API has been quite fruitful. We expect similar payoffs to
+happen for related projects as well.
+
+%% %
+%% We needed version control over a directory, gist provides that,
+%% wrapped over an HTTP interface.
+%% %
+%% Mainly a tactical decision to save time.
+%% %
+%% Nevertheless, availability of notebook data as (barely) plain text had good side
+%% effects, like ease to build search.
+%% %
+%% Expect same with suggest.
 
@@
 
+\subsection{Executing R in a web browser}
+
+As we reiterated in 
+One of the design goals of the user experience in RCloud
+
 \subsection{Deployment of notebooks\label{sec:deployment}}
@@
 
-How do we do things that are not trivial to do with IPython (for example)
+How do we do things that are not trivial to do with IPython (for
+example)
 

### 2014-03-31 system.tex
more text

@@
   are stored in GitHub \emph{gists}, which provide the version-control
-  primitives necessary for RCloud.}
+  primitives needed in RCloud during editing (left) or viewing and
+  executing (right).}
 \end{figure*}
@@
 
-As we reiterated in 
-One of the design goals of the user experience in RCloud
+One our main goals is to provide ubiquitous access to the statistical
+tools of the R programming language in the wider environment in which data
+science teams participate.
+%
+Given the recent developments of interactive visualization and visual
+analytics in HTML5 and the web, we decided to create an
+\emph{interconnect} between the two environments.
+%
+Every RCloud session spawns a new R execution process in a remote
+server. The communication between the web browser and the
+
+%% \subsubsection{Object capabilities: R to Javascript RPC and access
+%%   control}
+
+HTTP and web services are convenient but much of the protocol is
+stateless. (For example, GET requests are required to not change the
+remote state, and results can be cached in transparent proxies along
+the network). In the case of close communication between a running
+R process and a web browser, we seek something that is more flexible
+than distinct URLs for every R process (with the resulting
+complication in providing access
+
+%% \subsection{Technologies: R, Python, HTML5, interactive notebooks, etc.}
+
+%% How do we do things that are not trivial to do with IPython (for
+%% example)
+
+%% What is relationship to a standard CMS with R/Python
+
+%% dcplot. two-way communication between between backend session and
+%% frontend session.
+
+%% A key engineering decision in RCloud was to rely on full two-way
+%% communication between the web client human interface, and data and
+%% computing resources in the cloud. It is easier to engineer a design
+%% in which the human interface is pushed one-way from the server
+%% and thereafter all interaction takes place within the client.
+%% Two-way communication is necessary, though, where the size of
+%% the data or computational performance makes it impractical to
+%% move all computation to the client.
 
@@
 discussion fora and mailing lists.
-
-\subsection{Technologies: R, Python, HTML5, interactive notebooks, etc.}
-
-How do we do things that are not trivial to do with IPython (for
-example)
-
-What is relationship to a standard CMS with R/Python
-
-dcplot. two-way communication between between backend session and
-frontend session.
-
-A key engineering decision in RCloud was to rely on full two-way
-communication between the web client human interface, and data and
-computing resources in the cloud. It is easier to engineer a design
-in which the human interface is pushed one-way from the server
-and thereafter all interaction takes place within the client.
-Two-way communication is necessary, though, where the size of
-the data or computational performance makes it impractical to
-move all computation to the client.

### 2014-03-31 casestudies.tex
more figures and description of stock price

--- 2014-03-25 casestudies.tex
@@
 
+In this section we present a pair of example applications of the
+capabilities provided by RCloud. The examples themselves are synthetic
+(unfortunately, we are unable to show production notebooks because of
+their business value), but the examples are representative of the
+type of progression we want the ecosystem to support.
+
+\subsection{Stock Price Analysis\label{sec:stockvis}}
+
+\begin{figure}
+\includegraphics[width=\linewidth]{fig/casestudy1/casestudy1.pdf}
+\caption{\label{sec:stockvis}Iterations of a stock ticker
+  visualization, based on an example suggested by Hadley Wickham.  A
+  simple, static visualization of the closing price of a single ticker
+  is progressively improved into a configurable visualization suitable
+  for dashboarding, then into an interactive visualization of the
+  volatility and volume of two separate stocks, then finally into API
+  calls for data access use in other RCloud notebooks. All notebooks
+  in this can be used directly as web pages. When a notebook
+  corresponding to a function call is used as a web page, its included
+  documentation is displayed.}
+\end{figure}
+
+Our first example is a sequence of visualizations of the performance
+of stock prices over a multi-year period. The first visualization uses
+ggplot2 and is due to Hadley Wickham. It uses data provided by Yahoo!
+Finance via their web service, and shows the stock price for a single
+symbol.
+
+The webpage that produces that visualization always includes a link
+back to the generating source code, from which a user can always fork
+the notebook and, in this case, add a configurable ticker based on the
+URL of the notebook. Notice how the notebook changes very little.
+
+From there, the notebook author (or a user) can decide to provide easy
+access to the data without producing a visualization. This is achieved
+by simply creating a notebook that defines a function. This notebook
+then becomes a \emph{subroutine} for other notebooks, and is
+version-controlled in the same way.
+
+Interactive version.
+
+%% This notebook can then be called in an \emph{interactive}
+%% visualization, which uses the Javascript
+
+%% Easy to convert it into configurable entry point for dashboarding.
+
+%% Now consider an analyst that wants to understand the price dynamics of
+%% different stock attributes, different points in time, etc. For that,
+%% interactive visualization is very attractive. describe dcplot and
+%% notebook which uses it.
+
+%% Then describe abstraction of stock-data fetching, to talk about
+%% calling notebooks from other notebooks.
+
+%% then call.r?
+
+\subsection{text analysis\ref{sec:textvis}}
+
+Same.
+
 % IMPORTANT: what is unique about RCloud here
-
 % From prototyping to dashboard
-
 % Getting other information from the web
-
 %

### 2014-03-31 relatedwork.tex
more figures and description of stock price

@@
+\begin{figure}
+\includegraphics[width=\linewidth]{fig/system/system.pdf}
+\caption{\label{fig:system}A diagram of RCloud's architecture. Dashed
+  lines represent features which not yet implemented.}
+\end{figure}
+
 \section{Related Work}
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=.8\linewidth]{fig/notebook/notebook.pdf}
+\caption{\label{fig:notebook}An RCloud notebook is a sequence of
+  \emph{cells}, each a snippet of source in R or Markdown. Notebooks
+  are stored in GitHub \emph{gists}, which provide the version-control
+  primitives needed in RCloud during editing (left) or viewing and
+  executing (right).}
+\end{figure*}
+
 Our work has been inspired by and benefited from other efforts

### 2014-03-31 system.tex
more figures and description of stock price

@@
 % High-level architecture
-
-\begin{figure}
-\includegraphics[width=\linewidth]{fig/system/system.pdf}
-\caption{\label{fig:system}A diagram of RCloud's architecture. Dashed
-  lines represent features which not yet implemented.}
-\end{figure}
 
@@
 \subsection{Notebooks\label{sec:notebooks}}
-
-\begin{figure*}
-\includegraphics[width=\linewidth]{fig/notebook/notebook.pdf}
-\caption{\label{fig:notebook}An RCloud notebook is a sequence of
-  \emph{cells}, each a snippet of source in R or Markdown. Notebooks
-  are stored in GitHub \emph{gists}, which provide the version-control
-  primitives needed in RCloud during editing (left) or viewing and
-  executing (right).}
-\end{figure*}
 

### 2014-03-31 casestudies.tex
yes

--- 2014-03-25 casestudies.tex
@@
 
+We present two example applications of the capabilities provided by RCloud
+based on synthetic data and applications.\footnote{Unfortunately, we could
+not show production notebooks because of concerns about revealing proprietary
+or confidential information.} The main pont of the examples is that they are
+representative of a progression through the visual analytics development
+and deployment process that we aim to support.
+
+\subsection{Stock Price Analysis\label{sec:stockvis}}
+
+\begin{figure}
+\includegraphics[width=\linewidth]{fig/casestudy1/casestudy1.pdf}
+\caption{\label{sec:stockvis}Iterations of a stock ticker
+  visualization, after an example by Hadley Wickham.  A
+  simple, static visualization of the closing price of a single ticker
+  is progressively developed into a configurable display suitable
+  for dashboarding, then into an interactive visualization of the
+  volatility and volume of two stocks, then finally into an API
+  for data access by other RCloud notebooks. All notebooks
+  in this example can be loaded as web pages. When a notebook
+  corresponding to a function call is displayed a web page,
+  its associated documentation is displayed.}
+\end{figure}
+
+Our first example is a sequence of visualizations of the performance
+of financial stocks over a multi-year period. The first visualization
+in this example uses ggplot2 and is due to Hadley Wickham. It reads
+data provided by Yahoo! Finance as a web service, and shows the price
+for a single trading symbol.
+
+The webpage that produces that visualization includes a link back to
+the generating source code, from which a user can fork the notebook
+and, in this case, add a configurable ticker based on the URL of
+the notebook. Notice how the notebook changes very little.
+
+From there, the notebook author (or another user) may decide to
+provide convenient access to the data apart from its visualization.
+This is achieved by creating a notebook that defines a function.
+This notebook then becomes a \emph{subroutine} for other notebooks,
+and is version-controlled like other notebooks.
+
+\todo{Interactive version.}
+
+%% This notebook can then be called in an \emph{interactive}
+%% visualization, which uses the Javascript
+
+%% Easy to convert it into configurable entry point for dashboarding.
+
+%% Now consider an analyst that wants to understand the price dynamics of
+%% different stock attributes, different points in time, etc. For that,
+%% interactive visualization is very attractive. describe dcplot and
+%% notebook which uses it.
+
+%% Then describe abstraction of stock-data fetching, to talk about
+%% calling notebooks from other notebooks.
+
+%% then call.r?
+
+\subsection{text analysis\ref{sec:textvis}}
+
+Same.
+
 % IMPORTANT: what is unique about RCloud here
-
 % From prototyping to dashboard
-
 % Getting other information from the web
-
 %

### 2014-03-31 conclusion.tex
yes

--- 2014-03-27 conclusion.tex
@@
 
-We presented the ideas that motivated the design of RCloud,
-a prototype visual analytics environment to support
-collaboration, searching, sharing, and publishing of
-experiments, similar to DevOps development of web services.
+We described the motivation introducing web technologies
+for searching, collaborating, recommending, publishing
+reusable notebooks as part of the visual analytics process.
+We implemented these ideas in the RCloud prototype.
+Rcloud assists software development and release using
+an approach similar to DevOps in web services.
 Experience with the prototype provided evidence that
-data science teams and the organizations in which they work
-can benefit from having such features.
+data science teams, and the organizations in which they work,
+benefit from having such capabilities.
 

### 2014-03-31 system.tex
yes

@@
 
-Edit
+\todo{Edit}
 
-View/Run/Share
+\todo{View/Run/Share}
 
-Call
+\todo{Call}
 
-Browse/Search. ``More like this'', and a custom webpage solely for browsing.
+\todor{Browse/Search}. ``More like this'', and a custom webpage solely for browsing.
 
@@
 
-The basic unit of operation in RCloud is a \emph{notebook}. RCloud
-notebooks hold a sequence of \emph{cells}, each of which represents a
+The basic unit of operation in RCloud is a \emph{notebook}. A notebook
+holds a sequence of \emph{cells}, each of which represents a
 snippet of code or hypertext in Markdown. This is not a novel idea;
 executable documents structured in this way are present in a wide
-variety of present software systems, from Mathematica to IPython to
-Sage and others.
+variety of other software systems, including Mathematica, IPython and Sage.
 
@@
 notebook. Rather than have the notebook designer decide ahead of time
-which versions should be preserved, we choose to expose
-\emph{transparent} versioning. This is similar to models like
-Jankun-Kelly et al.'s p-set calculus \todo{cite} and VisTrails's
-version tree \todo{cite}, where every change in the state of the
+which versions should be preserved, we embrace \emph{transparent} versioning.
+This is similar to models like Jankun-Kelly et al.'s p-set calculus \todo{cite}
+and VisTrails's version tree \todo{cite}, where every change in the state of the
 system is tracked.
@@
 Implementing the entire version-control subsystem would be a large
-undertaking. Instead, we build the infrastructure on top of Github's
+undertaking. Fortunately, we can build RCloud on top of Github's
 \emph{gists}~\todo{cite}. Github offers a HTTP interface
-for creating simplified git repositories (the main limitation is the
-use of text-only files and a single directory). The GitHub web-service
-API provides much of the infrastructure we need for the versioning
+for creating simplified git repositories (the main limitation being the
+use of text-only files in a single directory). The GitHub web-service
+API provides most of the semantics we need for the versioning
 portion of the storage back end: access to previous versions,
@@
 Using GitHub was mainly a tactical decision to reduce our
-time-to-prototype. At the same time, it had the fortuitous decision of
+time-to-prototype. At the same time, it had the beneficial effect of
 forcing our backend to communicate with separate processes over HTTP
 and JSON. This had the serendipitous advantage of quickly enabling
-full search over RCloud notebooks (built using Apache SOLR).
+full search over RCloud notebooks (using Apache SOLR).
 %
@@
 directly related to visual analytics, our early decision to use the
-GitHub gist API has been quite fruitful. We expect similar payoffs to
-happen for related projects as well.
+GitHub gist API has been quite fruitful. We expect related projects
+can gain similar benefits by adopting the same strategy.
 
@@
 This makes explicit indication of interest a nearly trivial operation,
-always readily available, thus encouraging its use.
+always readily available, to encourage its use.
 
@@
 the network). In the case of close communication between a running
-R process and a web browser, we seek something that is more flexible
+R process and a web browser, we need something that is more flexible
 than distinct URLs for every R process (with the resulting

### 2014-03-31 discussion.tex
figure and case studies and discussion
yes

@@
 \section{Discussion, Lessons and Limitations}
+\label{sec:limitations}
 
@@
 Experience with deploying RCloud in a community of data analysts
-(hackers in the parlance of Heer and Agrawala)
+(hackers in the classification of Heer and Agrawala)
 has provided some insight about how the proposed requirements
-were met, and to what extent they were appropriate.
-This experience may be informative to researchers and designers
+were met and to what extent they were appropriate.
+This experience may be informative to researchers and implementers
 of future systems supporting collaborative visual analytics.
 
-Returning to the design considerations defined by Heer and Agrawala,
-our experience underscores the importance of their proposals,
-and indicates areas for further exploration, especially in the
-following areas.
+Returning to Heer and Agrawala's design considerations, our experience
+both underscores the relevance of their proposals and indicates areas
+for further exploration. We briefly examine these aspects of the
+visual analytics environment.
 

### 2014-03-31 casestudies.tex
yes

@@
 
-We present two example applications of the capabilities provided by RCloud
-based on synthetic data and applications.\footnote{Unfortunately, we could
-not show production notebooks because of concerns about revealing proprietary
-or confidential information.} The main pont of the examples is that they are
-representative of a progression through the visual analytics development
-and deployment process that we aim to support.
+We present two example applications in RCloud based on synthetic data
+and applications.\footnote{Unfortunately, we could not show production
+notebooks that may contain proprietary or confidential information.}
+The main point of the examples is to show key steps in the visual
+analytics development and deployment process that we aimed to support.
 
@@
 \caption{\label{sec:stockvis}Iterations of a stock ticker
-  visualization, after an example by Hadley Wickham.  A
-  simple, static visualization of the closing price of a single ticker
+  visualization, based on an example by Hadley Wickham. A
+  simple, static visualization of the closing price of a single stock
   is progressively developed into a configurable display suitable
-  for dashboarding, then into an interactive visualization of the
-  volatility and volume of two stocks, then finally into an API
-  for data access by other RCloud notebooks. All notebooks
-  in this example can be loaded as web pages. When a notebook
+  for dashboarding, then into an interactive visualization for 
+  comparing the volatility and volume of two stocks, and finally
+  into an API for data access by other RCloud notebooks. The notebooks
+  in this example can all be loaded as web pages. When a notebook
   corresponding to a function call is displayed a web page,
@@
 and, in this case, add a configurable ticker based on the URL of
-the notebook. Notice how the notebook changes very little.
+the notebook. Notice that the notebook changes very little.
 
-From there, the notebook author (or another user) may decide to
-provide convenient access to the data apart from its visualization.
-This is achieved by creating a notebook that defines a function.
-This notebook then becomes a \emph{subroutine} for other notebooks,
+From there, the notebook author or a collaborator may decide to
+provide convenient access to the data, apart from its visualization.
+This is done by creating a notebook that defines a function.
+That notebook then becomes a \emph{subroutine} for other notebooks,
 and is version-controlled like other notebooks.

### 2014-03-31 discussion.tex
yes

@@
 Experience with deploying RCloud in a community of data analysts
-(hackers in the classification of Heer and Agrawala)
-has provided some insight about how the proposed requirements
-were met and to what extent they were appropriate.
+(``hackers'') has provided some insight about how the proposed
+requirements were met and to what extent they were appropriate.
 This experience may be informative to researchers and implementers

### 2014-03-31 introduction.tex
yes

@@
 directions) slow, expensive, and error-prone. Coordination is done
-through meetings and may take months to implement.
+through meetings and may take months to carry out.
 
@@
 is found in today's repositories and version control
-systems. \todo{add cites to reco system work}
+systems.\cite{Koren:2009:FIT}
 
-In addition, take the recent trend in interactive visualization
-technologies. Let us consider prominent designers and developers of
-open source software, Jeffrey Heer, and Hadley Wickham. Heer has
-written (or helped write) Prefuse, Flare, Protovis, d3, and Vega, in
-that order. All but the first of these libraries target web
-technologies, and all but the second use open standards. Wickham has
-written ggplot, ggplot2, and ggvis. ggplot and
-ggplot2 generate static graphics, while ggvis uses Shiny and Vega to
-provide easy interactive visualizations in R, by opening a separate
-web server for each call. \todo{Add cites for all of these}
+In addition, consider recent trends in interactive visualization
+technologies in the open source community. Projects led by Jeff Heer
+include Prefuse, Flare, Protovis, d3, and Vega, in that order.
+All but the first of these libraries take advantage of web technologies,
+and all but the second rely on open standards. Hadley Wickham is the
+author of ggplot, ggplot2, and ggvis. ggplot and ggplot2 generate
+static graphics, while ggvis provides convenient interactive visualizations
+in R, using the R package Shiny and the Vega abstract visualization grammar,
+opening a separate web server for each call.
 
@@
 
-We see the present state-of-the-art in environments for data science
-teams as tremendously powerful, but ultimately fragmented. Rather than
-design a completely new solution that risks even further
-fragmentation, we wish to find and remedy the shortcomings of data
-science as we see it practiced today. The following requirements
-summarize what we consider to be the main pain points:
+Today's state-of-the-art in environments for data science teams are extremely
+powerful, but also fragmented. Instead of creating a new solution that risks
+even further fragmentation, we wish to identify and remedy the shortcomings
+in current technology for the practice of data science. The following
+requirements summarize the main pain points:
 
@@
 of new analyses and visualizations is done by \emph{hackers}
-and \emph{scripters} (where here are use Kandel et al.'s
+and \emph{scripters} (using Kandel et al.'s
 terminology~\cite{Kandel:2012:EDA}) who are, by definition, familiar

### 2014-03-31 casestudies.tex
yes

@@
-\section{Case Studies}
+\section{Case Study}
 
-We present two example applications in RCloud based on synthetic data
-and applications.\footnote{Unfortunately, we could not show production
-notebooks that may contain proprietary or confidential information.}
-The main point of the examples is to show key steps in the visual
-analytics development and deployment process that we aimed to support.
-
-\subsection{Stock Price Analysis\label{sec:stockvis}}
+We present an example Stock Price Analysis application
+in RCloud.\footnote{Unfortunately,
+we could not include production notebooks containing proprietary or personal
+information.} The example, although simplified, shows key steps in the
+development and deployment process we aimed to support.
 
@@
 \includegraphics[width=\linewidth]{fig/casestudy1/casestudy1.pdf}
-\caption{\label{sec:stockvis}Iterations of a stock ticker
+\caption{\label{fig:stockvis}Iterations of a stock ticker
   visualization, based on an example by Hadley Wickham. A
@@
 
-Our first example is a sequence of visualizations of the performance
+The example shows a sequence of visualizations of the performance
 of financial stocks over a multi-year period. The first visualization
-in this example uses ggplot2 and is due to Hadley Wickham. It reads
-data provided by Yahoo! Finance as a web service, and shows the price
-for a single trading symbol.
+in this example uses ggplot2 and was written by Hadley Wickham.
+It reads data provided by Yahoo! Finance as a web service, and shows
+the price for a single trading symbol.
 
 The webpage that produces that visualization includes a link back to
-the generating source code, from which a user can fork the notebook
-and, in this case, add a configurable ticker based on the URL of
-the notebook. Notice that the notebook changes very little.
+the underlying source code. From this link, a user can fork the notebook.
+In the example,a configurable ticker is added, based on the URL of
+the notebook. The change to the notebook is very minor.
 
-From there, the notebook author or a collaborator may decide to
-provide convenient access to the data, apart from its visualization.
+Next, the notebook author or a collaborator may decide to extend
+the application to also provide convenient access to the data,
+apart from its visualization.
 This is done by creating a notebook that defines a function.
-That notebook then becomes a \emph{subroutine} for other notebooks,
-and is version-controlled like other notebooks.
+That notebook then becomes a \emph{subroutine} for other notebooks.
+It can be invoked from interactive notebooks, such as dashboards.
+The data access notebook is is version-controlled like other notebooks.
 
-\todo{Interactive version.}
+Finally, consider the situation where an analyst wants to understand
+the price dynamics of stocks with respects to various attributes and
+time ranges. For this, interactive visualization may be very helpful.
+In our example, the analyst creates an interactive tool with N views.
+\todo{Describe dcplot and notebook which uses it.}
 
-%% This notebook can then be called in an \emph{interactive}
-%% visualization, which uses the Javascript
-
-%% Easy to convert it into configurable entry point for dashboarding.
-
-%% Now consider an analyst that wants to understand the price dynamics of
-%% different stock attributes, different points in time, etc. For that,
-%% interactive visualization is very attractive. describe dcplot and
-%% notebook which uses it.
-
-%% Then describe abstraction of stock-data fetching, to talk about
-%% calling notebooks from other notebooks.
-
-%% then call.r?
-
-\subsection{text analysis\ref{sec:textvis}}
-
-Same.
+% \subsection{Text analysis\label{sec:textvis}}
+% Same.
 

### 2014-03-31 system.tex
yes

@@
 
-\todor{Browse/Search}. ``More like this'', and a custom webpage solely for browsing.
+\todo{Browse/Search}. ``More like this'', and a custom webpage solely for browsing.
 
@@
 which versions should be preserved, we embrace \emph{transparent} versioning.
-This is similar to models like Jankun-Kelly et al.'s p-set calculus \todo{cite}
-and VisTrails's version tree \todo{cite}, where every change in the state of the
-system is tracked.
+This is similar to models like Jankun-Kelly et al.'s p-set calculus \cite{Jankun-Kelly:2007:MFV}
+and VisTrails's version tree \cite{Callahan:2006:VVM}, where every change in the state of the system is tracked.
 
@@
 undertaking. Fortunately, we can build RCloud on top of Github's
-\emph{gists}~\todo{cite}. Github offers a HTTP interface
+\emph{gists}~\cite{Github:2014:GG}. Github offers a HTTP interface
 for creating simplified git repositories (the main limitation being the

### 2014-03-31 casestudies.tex
yes

@@
 
-We present an example Stock Price Analysis application
-in RCloud.\footnote{Unfortunately,
-we could not include production notebooks containing proprietary or personal
-information.} The example, although simplified, shows key steps in the
-development and deployment process we aimed to support.
+To illustrate some aspects of RCloud, we present an example application
+for stock price analysis.\footnote{Unfortunately we could not include
+production notebooks containing proprietary or personal information.}
+The example, although simplified, shows key steps in the development
+and deployment process we intend to support.
 
@@
 
-The example shows a sequence of visualizations of the performance
+It shows a sequence of visualizations of the performance
 of financial stocks over a multi-year period. The first visualization

### 2014-03-31 system.tex
yes

@@
 As a result, HTTP is the lingua franca of interprocess communication
-(IPC). One of the design goals for RCloud was to provide an attractive
+(IPC). One of the design goals for RCloud was to provide a productive
 environment for creators of data-analysis scripts, that also behaves
 as a first-class citizen in the pre-existing ecosystem of computer
-services and networks with an organization.
+services and networks in an organization.
 
@@
 One of the main differences between RCloud and other systems is the
-notion that notebooks are ``always deployed''. This has a variety of
-advantages which we discuss \todo{where?}. On the other hand, there
-are a variety of situations that might ask for a specific version of a
-notebook. Rather than have the notebook designer decide ahead of time
-which versions should be preserved, we embrace \emph{transparent} versioning.
+notion that notebooks are ``always deployed''. This improves the
+ability to share and modify experiments and compose the results
+without binding to specific versions of a notebook.
+On the other hand, there are a variety of situations where it might
+be necessary to name a specific version. Rather than force notebook
+designers to decide which versions need to be preserved,
+we embrace \emph{transparent} versioning.
 This is similar to models like Jankun-Kelly et al.'s p-set calculus \cite{Jankun-Kelly:2007:MFV}
@@
 
-Implementing the entire version-control subsystem would be a large
-undertaking. Fortunately, we can build RCloud on top of Github's
+Implementing an entire version-control subsystem from scratch would
+be a large undertaking. Fortunately, we can build RCloud on top of Github's
 \emph{gists}~\cite{Github:2014:GG}. Github offers a HTTP interface
@@
 time-to-prototype. At the same time, it had the beneficial effect of
-forcing our backend to communicate with separate processes over HTTP
+allowing our backend to communicate with separate processes over HTTP
 and JSON. This had the serendipitous advantage of quickly enabling
-full search over RCloud notebooks (using Apache SOLR).
+full search over RCloud notebooks, using Apache SOLR.
 %
 This perspective has implications for
-systems-based research and prototyping in visual analytics.
+systems-based research and prototyping in visual analytics:
+For small teams to efficiently explore novel ideas toward the
+improvement of a large, complex software ecosystem, it pays off
+to use existing solutions as far as possible.
 %
-In order for small teams to efficiently explore ideas in the present
-software ecosystem, it pays off to use as many existing solutions as
-possible.
-%
-Even though it involved learning a variety of standards that are not
-directly related to visual analytics, our early decision to use the
-GitHub gist API has been quite fruitful. We expect related projects
-can gain similar benefits by adopting the same strategy.
+Even though it involved learning standards that are not directly
+related to visual analytics, our early decision to use the GitHub
+gist API was succesful. We expect related projects would gain many
+of the same benefits by adopting a similar strategy.
 

### 2014-03-31 system.tex
stab at the web architecture / interactive / deployment system stuff

@@
 R process and a web browser, we need something that is more flexible
-than distinct URLs for every R process (with the resulting
-complication in providing access
+than distinct URLs for every R process, and which embraces
+statefulness of both calculations and insights in progress.
 
-%% \subsection{Technologies: R, Python, HTML5, interactive notebooks, etc.}
+\subsection{Interactive notebooks, etc.}
 
@@
 
-%% A key engineering decision in RCloud was to rely on full two-way
-%% communication between the web client human interface, and data and
-%% computing resources in the cloud. It is easier to engineer a design
-%% in which the human interface is pushed one-way from the server
-%% and thereafter all interaction takes place within the client.
-%% Two-way communication is necessary, though, where the size of
-%% the data or computational performance makes it impractical to
-%% move all computation to the client.
+A key engineering decision in RCloud was to rely on full two-way
+communication between the web client human interface, and data and
+computing resources in the cloud. It is easier to engineer a design
+in which the human interface is pushed one-way from the server
+and thereafter all interaction takes place within the client.
+Two-way communication is necessary, though, where the size of
+the data or computational performance makes it impractical to
+move all computation to the client.
+
+In addition, R is naturally better as a language for analysis, and
+JavaScript as a language for interactive visualization.  To draw on
+the strengths of both languages and both environments, the connection
+between the languages is not just procedural: it makes \emph{closures}
+and \emph{first-class functions} available across the network link.
+This provides great flexibility, so that for example, a chart built
+with dc.js or leaflet.js can call back to analysis functions in R
+without having to formalize the protocol between the processes.
 
@@
 
-Notebooks as versioned subroutines, web services.
+Because notebooks are continually published, any work is immediately
+available both as a subroutine and visual component. Close colleagues
+can start on the next stage of analysis, or delve into the data,
+even while the original author is polishing the algorithm or its
+presentation. The code is the page, and can either be shaped into
+a function of inputs and outputs, or the linear cells of the notebook
+can be reworked into a full-fledged HTML layout.
 

### 2014-03-31 introduction.tex
working

@@
 
-In addition, consider recent trends in interactive visualization
-technologies in the open source community. Projects led by Jeff Heer
+In addition, consider trends in interactive visualization software
+in the open source community. Projects led by Jeff Heer
 include Prefuse, Flare, Protovis, d3, and Vega, in that order.
@@
 in R, using the R package Shiny and the Vega abstract visualization grammar,
-opening a separate web server for each call.
+opening a separate web server for each call. These projects point to
+the trend toward visual analytics frameworks and systems that are
+increasingly interactive, composable, and accessible through the web.
 
@@
 
-Today's state-of-the-art in environments for data science teams are extremely
-powerful, but also fragmented. Instead of creating a new solution that risks
-even further fragmentation, we wish to identify and remedy the shortcomings
-in current technology for the practice of data science. The following
-requirements summarize the main pain points:
+Today's state-of-the-art environments for data science teams are unquestionably
+powerful, but also fragmented. Instead of creating a new solution that promotes
+even further fragmentation, we attempt to identify and remedy the shortcomings
+in the current practice of data science. The following requirements summarize
+the main pain points:
 
@@
 and \emph{scripters} (using Kandel et al.'s
-terminology~\cite{Kandel:2012:EDA}) who are, by definition, familiar
-to some extent with programming and software development.
+terminology~\cite{Kandel:2012:EDA}) who are, by definition,
+at least somewhat familiar with programming and software development.
 On the other hand, \emph{application
-users} are the ones extracting value from the data using the tools
-developed by hackers and scripters. Traditionally, the communication
-between these two world is done by IT staff. In an environment where
-business needs change possibly many times a week, manual IT support
-does not scale: there needs to exist
-infrastructure to connect the two worlds.
+users} ultimately derive value from the data by means of the tools
+developed by hackers and scripters. Traditionally, the connection
+between these two worlds is made by IT staff. In an environment where
+business needs can change every day, manual IT support does not scale:
+we need to create software infrastructure to connect the two worlds.
+(The path backward is equally important: when problems discovered in
+production require expert help, we would like the experts to be
+able to immediately apply their full range of analytic tools without
+moving the problem back to the EDA environment).
 %% The two classes of users
@@
 \item Coexistence. In the current data science ecosystem, the
-isolation of current exploratory visualization and data analysis
-environments is a threat to wider adoption of modern techniques for
-both sides. We specifically see the richness of interactive
-visualization tools isolated from the richness of statistical
-programming environments.
-We wish to provide infrastructure such that
-developments on one side can more easily be employed on the other. In
-addition, this system itself should be easily accessible
+isolation of exploratory visualization and data analysis
+environments is a hinderance to wider adoption of modern techniques
+on both sides. We specifically see the richness of interactive
+visualization tools as still being isolated from the richness
+of statistical programming environments.
+There is an opportunity to provide a framework so that developments
+on either side are more easily employed by the other.
+In addition, the integrated results should be more easily accessible
 to the rest of the organization.
@@
 \item Discoverability. From our experience in working with data
-analyst teams, we find that repeated work (``how can I connect to
-database X?'', ``how do you get clean data from column Y from feed
-Z?'') typically arises from lack of communication of solved
-subproblems. We wish to provide transparency about the work of
-team members in the organization.
+analyst teams, we find that repeated work (``How can I connect to
+database X?'', ``How do you get clean data from column Y from feed
+Z?'') arises from lack of communication about previously solved subproblems.
+We wish to provide transparency about the work of team members in the organization.
 
@@
 above. We will report on the design decisions, tradeoffs and
-limitations, and compare the concepts in RCloud to similar proposals.
+limitations, and compare the concepts in RCloud to related proposals.
 
-RCloud uses the R language~\cite{RCoreTeam:2013:R} as the main provider of
-statistical analysis tools. In fact, as we describe in
-Section~\todo{where?}, RCloud notebooks are sequences of R snippets
+RCloud uses the R language~\cite{RCoreTeam:2013:R} as the core set
+of statistical analysis tools. In fact, as we describe in
+Section~\ref{sec:notebooks}, RCloud notebooks are sequences of R snippets
 interspersed with Markdown for easy generation of reports. In
-Section~\todo{where?}, we describe how RCloud enables
+Section~\ref{sec:visualization}, we describe how RCloud enables
 interactive visualization widgets in this setting. In
@@
 Section~\todo{where?}. RCloud notebooks are edited and shared as web
-pages. In order to support the tight integration of R and HTML5
-technologies, RCloud uses a full, bi-directional communication channel
-between the web browser and a running R process, described in
-Section~\todo{where?}.
+pages. To support tight integration of R and HTML5, RCloud uses a
+full bi-directional communication channel between the web browser
+and a running R process, described in Section~\todo{where?}.
 
@@
 web services as remote procedure calls), we find it easier to
-integrate RCloud with satellite efforts in corporate environments. We
-discuss this architecture and consequences in
-Section~\ref{sec:system}.
+integrate RCloud with satellite efforts in corporate environments.
+We discuss this architecture and consequences in Section~\ref{sec:system}.
 
@@
 about the \emph{stability} of the many moving pieces: if notebooks are
-always updated and deployed automatically, how can the system handle
+updated and deployed automatically, how can the system handle
 changing requirements and public interfaces? We discuss our choice of

### 2014-03-31 system.tex
working

@@
 related to visual analytics, our early decision to use the GitHub
-gist API was succesful. We expect related projects would gain many
+gist API was successful. Future projects can gain many
 of the same benefits by adopting a similar strategy.

### 2014-03-31 casestudies.tex
try

@@
 To illustrate some aspects of RCloud, we present an example application
-for stock price analysis.\footnote{Unfortunately we could not include
-production notebooks containing proprietary or personal information.}
+for stock price analysis.\footnote{Unfortunately this report could not
+include production notebooks containing proprietary or personal information.}
 The example, although simplified, shows key steps in the development

### 2014-03-31 system.tex
try

@@
 
-\subsection{Interactive notebooks, etc.}
+\subsection{Interactive notebooks\label{sec:notebooks}}
 
@@
 
-In addition, R is naturally better as a language for analysis, and
-JavaScript as a language for interactive visualization.  To draw on
+In addition, R is well suited as a language for analysis, and
+JavaScript for interactive visualization. To draw on
 the strengths of both languages and both environments, the connection
 between the languages is not just procedural: it makes \emph{closures}
-and \emph{first-class functions} available across the network link.
-This provides great flexibility, so that for example, a chart built
+and \emph{first-class functions} available across the network.
+This provides considerable flexibility, so that for example, a chart built
 with dc.js or leaflet.js can call back to analysis functions in R
@@
 
-There exists a URL for every notebook in RCloud, and notebooks by
+Every notebook in RCloud is named by a URL, and notebooks by
 default are visible by the entire organization. This is deliberate.

### 2014-03-31 system.tex
UI overview

@@
 
-\todo{Edit}
-
-\todo{View/Run/Share}
-
-\todo{Call}
-
-\todo{Browse/Search}. ``More like this'', and a custom webpage solely for browsing.
+The central of the screen is for editing code and visualizations through
+the notebook interface. Supplementary ``asset'' files which are part of the
+notebook but not part of the notebook flow, are edited on the right.
+Controls at the top of the screen allow the user to quickly run, view, 
+and share the notebook. The call mechanism is the URL itself; arguments are
+read from the URL and the final cell is the result. On the left are the 
+browsing area for seeing other users' notebooks and for searching, as well
+as the help system.
 

### 2014-03-31 casestudies.tex
the easy vis language for analyst

@@
 In our example, the analyst creates an interactive tool with N views.
-\todo{Describe dcplot and notebook which uses it.}
+
+Through the tight integration of R and JavaScript functions and data,
+the analyst writes a function in R which passes the dataframe and the
+description of the charts to the dcplot library.  Simple R expressions
+are captured as trees to generate JavaScript expressions.  The terse
+chart description language, with intelligent defaults inspired by
+ggplot2, provides a simple yet powerful interface to the grouping
+and reduction functionality of crossfilter and the chart styles of
+dc.js and d3.js.
 

### 2014-04-01 casestudies.tex
yes

@@
 The example, although simplified, shows key steps in the development
-and deployment process we intend to support.
+and deployment process we support.
 
@@
 It can be invoked from interactive notebooks, such as dashboards.
-The data access notebook is is version-controlled like other notebooks.
+The data access notebook is version-controlled like other notebooks.
 
@@
 time ranges. For this, interactive visualization may be very helpful.
-In our example, the analyst creates an interactive tool with N views.
+In our example, the analyst creates an interactive tool with multiple
+linked views.
 
@@
 ggplot2, provides a simple yet powerful interface to the grouping
-and reduction functionality of crossfilter and the chart styles of
-dc.js and d3.js.
+and reduction functionality of the well-accepted charting libraries
+crossfilter, dc.js and d3.js.
 

### 2014-04-01 discussion.tex
curation para
yes

@@
 
+{\bf Curation.} Even without more formal group management, users are
+able to curate groups of related notebooks using the simple folder
+system.  This is particularly effective for bringing together
+training materials.
+
 \subsection{Limitations}
@@
 
-Asynchronous collaborative visual analytics
-(ACVA)~\cite{Chen:2011:SEC}. This paper addresses visualization
-\emph{of} the ACVA process. It will be important when we talk about
-recommendation systems, and navigation of the set of notebooks, etc.
+%Asynchronous collaborative visual analytics
+%(ACVA)~\cite{Chen:2011:SEC}. This paper addresses visualization
+%\emph{of} the ACVA process. It will be important when we talk about
+%recommendation systems, and navigation of the set of notebooks, etc.
 

### 2014-04-01 introduction.tex
yes

@@
 in R, using the R package Shiny and the Vega abstract visualization grammar,
-opening a separate web server for each call. These projects point to
-the trend toward visual analytics frameworks and systems that are
-increasingly interactive, composable, and accessible through the web.
+opening a separate web server for each call. These projects show
+a trend toward visual analytics frameworks and systems that are
+increasingly rich, interactive, composable, and accessible through the web.
 
@@
 we need to create software infrastructure to connect the two worlds.
-(The path backward is equally important: when problems discovered in
-production require expert help, we would like the experts to be
-able to immediately apply their full range of analytic tools without
-moving the problem back to the EDA environment).
 %% The two classes of users
@@
 of statistical programming environments.
+(The path backward from production to analysis can be 
+equally important: when problems discovered in
+production require expert help, we would like experts to be
+able to immediately apply the full range of analytic tools without
+extracting data to move it back to the EDA environment.)
 There is an opportunity to provide a framework so that developments
@@
 RCloud uses the R language~\cite{RCoreTeam:2013:R} as the core set
-of statistical analysis tools. In fact, as we describe in
-Section~\ref{sec:notebooks}, RCloud notebooks are sequences of R snippets
-interspersed with Markdown for easy generation of reports. In
-Section~\ref{sec:visualization}, we describe how RCloud enables
-interactive visualization widgets in this setting. In
-Section~\ref{sec:limitations}, we discuss some consequences of
-designing for a single language, and how this was received by
-our (small, but active) user community.
+of statistical analysis tools. As described in Section~\ref{sec:notebooks},
+RCloud notebooks are sequences of R snippets interspersed with Markdown
+for easy generation of reports. We also describe how RCloud enables
+interactive visualization widgets in this setting (Section~\ref{sec:executinginbrowser}).
 
@@
 The facilities for \emph{sharing, viewing and embedding} RCloud
-notebooks address technology transfer, and are described in
-Section~\todo{where?}. RCloud notebooks are edited and shared as web
-pages. To support tight integration of R and HTML5, RCloud uses a
-full bi-directional communication channel between the web browser
-and a running R process, described in Section~\todo{where?}.
-
+notebooks address technology transfer.
 Because RCloud embraces technologies that are part of the computing
 environment (specifically, inter-process communication via HTTP and
-web services as remote procedure calls), we find it easier to
+web services as remote procedure calls), it is not complicated to
 integrate RCloud with satellite efforts in corporate environments.
-We discuss this architecture and consequences in Section~\ref{sec:system}.
 
-The requirement of continuous technology transfer creates challenges
+The requirement of continuous technology transfer, though, creates challenges
 about the \emph{stability} of the many moving pieces: if notebooks are
@@
 transparent, integrated version control for RCloud notebooks in 
-Section~\ref{sec:notebooks}.
+Section~\ref{sec:deployment}.
+
+In Section~\ref{sec:limitations}, we discuss some consequences of
+designing for a single language, and how this was received by
+our (small, but active) user community.
 

### 2014-04-01 relatedwork.tex
yes

@@
 does not by itself include modules specific to data cleaning and
-wrangling, we note from internal experience that these cleaning
+wrangling, we note from practical experience that these cleaning
 scripts themselves tend to change over time. RCloud addresses this
@@
 this cycle that cost time and effort to overcome. RCloud seeks to
-reduce this impedance mismatch. They also point out that larger teams
+reduce this mismatch. They also point out that larger teams
 are becoming more common in data analysis, that supporting
@@
 Heer and Agarwala identify many design considerations for
-collaborative visual analytics~\cite{Heer:2008:DCF}. \emph{Starring},
-the means for signaling interest in notebooks, described in
+collaborative visual analytics~\cite{Heer:2008:DCF}.
+Notebooks, and the integrated version control system for them, described in
+Section~\ref{sec:notebooks}, address modularity and granularity, and
+artifact histories.
+\emph{Starring}, the means for signaling interest in notebooks, described in
 Section~\ref{sec:starring}, addresses social-psychological incentives,
@@
 mechanism, described in Section~\ref{sec:deployment}, addresses cost of
-integration, content export, presentation and view sharing. Notebooks,
-and the integrated version control system for them, described in
-Section~\ref{sec:notebooks}, address modularity and granularity, and
-artifact histories.
+integration, content export, presentation and view sharing.
 
@@
 exploratory work in data science teams. As a result, RCloud tries to
-provide a closer data-analysis to analysts accustomed to creating and
-executing R and Python code while keeping many of the attractive
+provide a closer match for analysts accustomed to creating and
+executing R and Python code, while retaining many attractive
 properties like automatic versioning and management.
@@
 of technology in visual analytics. The fusion of development with production
-operations in software release management ("devops" \cite{Httermann:2012:DD})
+operations in software release management (``DevOps'' \cite{Httermann:2012:DD})
 is a trend in web services and related fields. By making it possible

### 2014-04-01 system.tex
yes

@@
 
-The central of the screen is for editing code and visualizations through
+The center panel of the screen is for editing code and visualizations through
 the notebook interface. Supplementary ``asset'' files which are part of the
@@
 browsing area for seeing other users' notebooks and for searching, as well
-as the help system.
+as a help system.
 
@@
 systems-based research and prototyping in visual analytics:
-For small teams to efficiently explore novel ideas toward the
-improvement of a large, complex software ecosystem, it pays off
+for small teams to efficiently explore novel ideas toward the
+improvement of a large, complex software ecosystem, it pays
 to use existing solutions as far as possible.
@@
 classic techniques of collecting feedback from web search, we anticipate
-applying static and dynamic code analysis to generate fine-grained
+applying static and dynamic code analysis to infer fine-grained
 information about relationships, for example, which packages and data
@@
 
-\subsection{Executing R in a web browser}
+\subsection{Executing R in a web browser\label{sec:executinginbrowser}}
 
@@
 Every RCloud session spawns a new R execution process in a remote
-server. The communication between the web browser and the
+server.
 
@@
 than distinct URLs for every R process, and which embraces
-statefulness of both calculations and insights in progress.
+statefulness of both calculations and visualization parameters.
 
-\subsection{Interactive notebooks\label{sec:notebooks}}
+\subsection{Interactive notebooks\label{sec:interactivenotebooks}}
 
@@
 
-Because notebooks are continually published, any work is immediately
-available both as a subroutine and visual component. Close colleagues
-can start on the next stage of analysis, or delve into the data,
-even while the original author is polishing the algorithm or its
-presentation. The code is the page, and can either be shaped into
-a function of inputs and outputs, or the linear cells of the notebook
-can be reworked into a full-fledged HTML layout.
-
 Every notebook in RCloud is named by a URL, and notebooks by
@@
 discussion fora and mailing lists.
+
+Because notebooks are continually published, any work is immediately
+available both as a subroutine and a visual component. Close colleagues
+can start on the next stage of analysis, or delve into the data,
+even while the original author is polishing an algorithm or its
+presentation. The code is the page, and can either be shaped into
+a function of inputs and outputs, or the linear cells of the notebook
+can be reworked into a full-fledged HTML layout.

### 2014-04-01 introduction.tex
yes

@@
 of statistical programming environments.
-(The path backward from production to analysis can be 
-equally important: when problems discovered in
-production require expert help, we would like experts to be
-able to immediately apply the full range of analytic tools without
-extracting data to move it back to the EDA environment.)
+%(The path backward from production to analysis can be 
+%equally important: when problems discovered in
+%production require expert help, we would like experts to be
+%able to immediately apply the full range of analytic tools without
+%extracting data to move it back to the EDA environment.)
 There is an opportunity to provide a framework so that developments
@@
 
-RCloud uses the R language~\cite{RCoreTeam:2013:R} as the core set
-of statistical analysis tools. As described in Section~\ref{sec:notebooks},
-RCloud notebooks are sequences of R snippets interspersed with Markdown
-for easy generation of reports. We also describe how RCloud enables
-interactive visualization widgets in this setting (Section~\ref{sec:executinginbrowser}).
-
 %% Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
@@
 
-The facilities for \emph{sharing, viewing and embedding} RCloud
-notebooks address technology transfer.
+RCloud uses the R language~\cite{RCoreTeam:2013:R} as the core set
+of statistical analysis tools.  The facilities for \emph{sharing, viewing
+and embedding} RCloud notebooks address the {\bf technology transfer} requrement.
 Because RCloud embraces technologies that are part of the computing
 environment (specifically, inter-process communication via HTTP and
-web services as remote procedure calls), it is not complicated to
+web services as remote procedure calls, as explained
+in Section~\ref{sec:highlevelarchitecture}), it is straightforward to
 integrate RCloud with satellite efforts in corporate environments.
+
+{\bf Coexistence} between analysis and visualization is supported in
+its human interface architecture, described in~\ref{sec:humaninterface}. 
+As oulined in Section~\ref{sec:notebooks}, RCloud notebooks are sequences
+of R snippets interspersed with Markdown for easy generation of reports.
+We describe how RCloud enables interactive visualization widgets in
+this setting (Sections~\ref{sec:Rinbrowser} and~\ref{sec:interactivenotebooks}).
+Indexing, publishing and sharing of notebooks is also the main way
+Rcloud supports {\bf discoverability}.
 
@@
 
-In Section~\ref{sec:limitations}, we discuss some consequences of
-designing for a single language, and how this was received by
-our (small, but active) user community.
+Section~\ref{sec:casestudy} describes a case study illustrating some of these concepts.
+
+In Section~\ref{sec:limitations}, we report on experience toward meeting
+these requirements in RCloud, and limitations encountered.
 

### 2014-12-19 introduction.tex
notes

@@
 \maketitle
+
+\carlos{General plan: we take Kandel et al.'s interview study, and Heer and Agrawala design considerations, and make them integral to the explanation of RCloud, throughout the whole paper.}
+
+\carlos{We need to remake all the figures, pretty much. They need to have real notebooks, on each and every one of them.}
+
+\carlos{Update with new RCloud features. Multiple languages, full-text searching, shiny, ...}
 

### 2014-12-19 relatedwork.tex
notes

@@
 \caption{\label{fig:system}A diagram of RCloud's architecture. Dashed
-  lines represent features which not yet implemented.}
+  lines represent features which not yet implemented. \carlos{Must remove anything that isn't implemented}}
 \end{figure}
@@
 \section{Related Work}
+
+\carlos{There's no need to position ourselves defensively here; we're
+  addressing different needs, and the connection is already
+  tenuous. We're just hurting ourselves by comparing to wrangling and
+  then saying ``well, we don't have wrangling''}
 

### 2014-12-19 system.tex
yes
notes

@@
 
-% High-level architecture
+\subsection{High-level Architecture\label{sec:highlevelarchitecture}}
 
@@
 
-\subsection{System Design}
+\subsection{Human Interface\label{sec:humaninterface}}
 
@@
 
-\subsection{Executing R in a web browser\label{sec:executinginbrowser}}
+\subsection{Executing R in a web browser\label{sec:Rinbrowser}}
 
@@
 statefulness of both calculations and visualization parameters.
+
+\carlos{In-depth explanation of R<->web interconnect, and why it's
+  necessary or important. Main point: make programming in the R side
+  as close as possible to what R programming feels like, and make
+  programming in the Javascript side as close as possible to what
+  Javascript programming feels like. We do this because this part of
+  the system is targeted at programmers; Shiny targets web programming
+  at non-web-programmers. We don't attempt such a thing and land at a
+  different point in the spectrum.}
 

### 2014-12-19 paper.tex
yes
working
notes
abstract note

@@
 \usepackage{algpseudocode}
+
+%\usepackage{dblfloatfix}
+%\usepackage{fixltx2e}
+\usepackage{url}
 
@@
   \emph{discoverability}. In addition, we contribute the design and
-  implementation of RCloud, a prototype system that supports the requirements
+  implementation of RCloud, a system that supports the requirements
   collaborative data analysis, visualization and web deployment. We
@@
   to other current proposals.
+  \carlos{RCloud has been in production use for more than a year and has about one hundred active users. We will discuss the adoption issues and etc etc.}
 } % end of abstract
@@
 %% please capitalize first letter and insert punctuation after last keyword
-\keywords{Some, Keywords, Here}
+\keywords{visual analytics process, provenance, collaboration, visualization, computer-supported cooperative work}
 
@@
   \centering
-  \includegraphics[width=.3\linewidth]{fig/screenshots/wdcplotsmall.jpg}
+  \includegraphics[width=.28\linewidth]{fig/screenshots/wdcplotsmall.jpg}
   %% \includegraphics[width=.9\linewidth]{figs/teaser.jpg}
+  \hspace{1em}
+  \includegraphics[width=.6\linewidth]{fig/casestudy1/casestudy1.graffle/image8.png}
   % \vspace{-.5em}
@@
     enabling seamless deployment and dashboarding for
-    \emph{application users} (right)
+    \emph{application users} (right).
+    \carlos{Replace this with something using either a Shiny demo or in-process nanocubes?}
   }

### 2015-01-02 paper.tex
new teaser skeleton

@@
 
-\title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
-%% \title{RCloud: DevOps for Visual Analysis and Data Science}
+\title{RCloud: Integrated Visualization, Analysis, Development, and Deployment}
 
@@
   \centering
-  \includegraphics[width=.28\linewidth]{fig/screenshots/wdcplotsmall.jpg}
-  %% \includegraphics[width=.9\linewidth]{figs/teaser.jpg}
-  \hspace{1em}
-  \includegraphics[width=.6\linewidth]{fig/casestudy1/casestudy1.graffle/image8.png}
+  \includegraphics[width=.95\linewidth]{fig/teaser/teaser.pdf}
   % \vspace{-.5em}
-  \caption{RCloud provides infrastructure for data science teams so
-    that R \emph{hackers} and \emph{scripters}~\cite{Kandel:2012:EDA} can target modern
-    interactive visualization technologies such as
-    D3~\cite{Bostock:2011:DDD}, dc.js~\cite{DCJS:2014:DC} and
-    Crossfilter~\cite{Square:2014:CFM} (left), while at the same time
-    enabling seamless deployment and dashboarding for
-    \emph{application users} (right).
-    \carlos{Replace this with something using either a Shiny demo or in-process nanocubes?}
+  \caption{
+    \carlos{Come up with caption for this. The story is: hackers/scripters work in the middle (and bottom-right), rest of organization has automatic access to their work on left and top-right.}
   }

### 2015-01-15 casestudies.tex
yes

@@
-\section{Case Study}
+\section{Case Study\label{sec:casestudy}}
 
-To illustrate some aspects of RCloud, we present an example application
+To illustrate some aspects of RCloud, we present an example notebook
 for stock price analysis.\footnote{Unfortunately this report could not
-include production notebooks containing proprietary or personal information.}
+include production notebooks containing proprietary information.}
 The example, although simplified, shows key steps in the development
-and deployment process we support.
+and deployment of an application.
 
@@
   is progressively developed into a configurable display suitable
-  for dashboarding, then into an interactive visualization for 
-  comparing the volatility and volume of two stocks, and finally
+  for a dashboard, then into an interactive visualization that
+  compares the volatility and volume of two stocks, and finally
   into an API for data access by other RCloud notebooks. The notebooks
@@
 It shows a sequence of visualizations of the performance
-of financial stocks over a multi-year period. The first visualization
+of financial stocks over several years. The first visualization
 in this example uses ggplot2 and was written by Hadley Wickham.
@@
 
-The webpage that produces that visualization includes a link back to
+The webpage that produces that visualization includes a link to
 the underlying source code. From this link, a user can fork the notebook.
 In the example,a configurable ticker is added, based on the URL of
-the notebook. The change to the notebook is very minor.
+the notebook. The change to the notebook is minor.
 
@@
 Finally, consider the situation where an analyst wants to understand
-the price dynamics of stocks with respects to various attributes and
+the price dynamics of stocks with respects to other attributes and
 time ranges. For this, interactive visualization may be very helpful.
@@
 
-Through the tight integration of R and JavaScript functions and data,
-the analyst writes a function in R which passes the dataframe and the
-description of the charts to the dcplot library.  Simple R expressions
-are captured as trees to generate JavaScript expressions.  The terse
-chart description language, with intelligent defaults inspired by
+Demonstrating the integration of R and JavaScript, here the analyst
+writes an R function that passes the dataframe and description of
+the charts to the dcplot library.  Simple R expressions
+are captured as trees to generate JavaScript expressions.
+The terse chart description language, with sensible defaults inspired by
 ggplot2, provides a simple yet powerful interface to the grouping

### 2015-01-23 conclusion.tex
yes
stuff

@@
 
-We described the motivation introducing web technologies
+We presented a case for introducing web technologies
 for searching, collaborating, recommending, publishing
-reusable notebooks as part of the visual analytics process.
+reusable notebooks to the visual analytics process.
 We implemented these ideas in the RCloud prototype.
-Rcloud assists software development and release using
+RCloud assists software development and release using
 an approach similar to DevOps in web services.
@@
 
-Future work will address cross-language development,
-source code analysis, richer recommendation techniques,
-and improvements to the HTML5 front end.
+Reasonable next steps are to support cross-language development,
+source code analysis, richer recommendation techniques, and
+improvements to the HTML5 human interface.
 

### 2015-01-23 discussion.tex
notes
stuff

@@
 
+\carlos{All this stuff needs to appear in the paper much earlier.}
+
 Experience with deploying RCloud in a community of data analysts
-(``hackers'') has provided some insight about how the proposed
-requirements were met and to what extent they were appropriate.
-This experience may be informative to researchers and implementers
-of future systems supporting collaborative visual analytics.
+(``hackers'') has provided some insight about whether the proposed
+requirements were appropriate, and how well they were met.
+\stephen{The term {\it hackers} may be too tainted for us to use.}
+% I know the following is true but also feel we need to let the work speak for itself. SCN
+%This experience may be informative to researchers and implementers
+%of future systems supporting collaborative visual analytics.
 
 Returning to Heer and Agrawala's design considerations, our experience
-both underscores the relevance of their proposals and indicates areas
+underscores the relevance of their proposals, and indicates areas
 for further exploration. We briefly examine these aspects of the

### 2015-01-23 introduction.tex
stuff

@@
 related problems, and must devise effective ways to share their findings
-and move results from exploratory data analysis to automated diagnostics
-and reports that are deployed for wider consumption.
+and move results from exploratory data analysis to automated reports and
+diagnostics that are deployed for wider consumption.
 They generally rely on a patchwork of resources shared informally,
@@
 set of tools, and automated reports and deployments with another. This
-makes transferring results between EDA and production (in both
-directions) slow, expensive, and error-prone. Coordination is done
-through meetings and may take months to carry out.
+makes transferring results between EDA and production slow, expensive,
+and error-prone, in both directions. Coordination is done through meetings
+and may take months to carry out.
 
@@
 and automated analyses and recommendations. These capabilities have had
-tremendous impact on information retrieval in general but are well
+extraordinary impact on information retrieval in general, but are well
 beyond what existing visual analytics environments provide, or what

### 2015-01-23 paper.tex
yes
stuff

@@
 %% Abstract section.
-\abstract{ Consider the emerging role of a data science team within an
-  organization. Individual data scientists and statisticians usually
+\abstract{ Consider the emerging role of data science teams within
+  organizations. Individual data scientists and statisticians usually
   work on loosely related problems, and must devise effective ways to
@@
     \carlos{Come up with caption for this. The story is: hackers/scripters work in the middle (and bottom-right), rest of organization has automatic access to their work on left and top-right.}
+    \stephen{An overview of RCloud.
+a) Workbooks of published experiments and production applications are visible and searchable by all authorized users in an organization b) Data scientists create, publish and anotate workbooks with support for sharing and recommendation c) Workbooks can be parameterized and deployed as production services}
   }

### 2015-01-27 introduction.tex
working

--- 2014-04-01 introduction.tex
@@
 
-%% \todo{Points we want to make."Coordination is done by meetings" "how do you trace an automated alert back to an EDA environment ?"}
+Consider the role of a data science team within an organization today.
+Analysts work in loose or tight collaboration with colleagues.
+Project teams vary in size, from just a few people to dozens or
+more, even within one project's duration. Tasks are broad, including problem
+identification, data wrangling, visualization, modeling and analysis,
+interpreting, explaining and advocating the results, and often helping
+clients to realize the benefits of the work in ongoing processes.
+There are many details to these tasks.
+For example, data wrangling can involve finding data, figuring out its
+syntax and semantics, assessing data quality, performing normalization and
+data quality remediation, and making the data available for other tasks
+to follow.
 
-% What's the area
-Consider the emerging role of a data science team within an
-organization today \cite{Keim:2008:VAS}.
-%
-% How do people do it today?
-Individual data scientists and statisticians usually work on loosely
-related problems, and must devise effective ways to share their findings
-and move results from exploratory data analysis to automated diagnostics
-and reports that are deployed for wider consumption.
-They generally rely on a patchwork of resources shared informally,
-or at best, through systems that attack individual tasks or functions
-in the data analysis workcycle. % might cite information visualization reference model or pirolli/card
+Visual analytics in practice depends greatly on communication and collaboration.
+At almost every step, detailed knowledge about data, code and tasks
+is shared by collaborators. 
+Further, data scientists are increasingly asked to work more closely
+with business people who are domain specialists and less technically oriented.
+Thus, data scientists and developers are being asked to become
+very broad, and integrate work across the spectrum. (Hiring ``full stack
+developers'' to perform integration seems to be a trend, but does not achieve
+the potential benefits of specialization and division of labor in larger
+teams, and is not a practice to be encouraged.)
+This pushes the boundaries for communication and collaboration even further.
+Unfortunately, processes and technology that support this work are fragmented.
+Knowledge is shared through informal conversations, email online messages,
+meetings, phone calls, within code itself, in wikis, in project documents and
+by other means. Results of experiments are often shared first by copying static
+output, later, by holding meetings that lead to moving or even rewriting
+entire programs.
 
-% Why does this not work?
-There are two problems with the current practice. First, there are
-gaps in the workflow: Exploratory Data Analysis (EDA) is done with one
-set of tools, and automated reports and deployments with another. This
-makes transferring results between EDA and production (in both
-directions) slow, expensive, and error-prone. Coordination is done
-through meetings and may take months to carry out.
+This disjointed process inhibits the use of tools to bridge
+the many gaps and to help people find resources.
+For example, when code is shared, we still not may know where to find examples
+that show how to use it. stephen{Why did I shift to first person?}
+When we find a useful data source, we don't know what packages or
+other data sources usually go along with it, or what problems with the data
+others have already discovered and corrected, or where to obtain online feeds
+to get updates of the data.
 
-Second, EDA environments often assume a single-developer perspective,
-while data scientist teams could greatly benefit from convenient searching
-and sharing of scripts, data sets, data feeds, experiments, annotations,
-and automated analyses and recommendations. These capabilities have had
-tremendous impact on information retrieval in general but are well
-beyond what existing visual analytics environments provide, or what
-is found in today's repositories and version control
-systems.\cite{Koren:2009:FIT}
+The current situation affects how well data science teams work. It limits
+the ability to share work and to rapidly deploy and refine results.
+We have seen this in our lab environment.
+Other authors report similar findings.
 
-In addition, consider trends in interactive visualization software
-in the open source community.  Projects led by Jeff Heer
-include Prefuse, Flare, Protovis, d3, and Vega, in that order.
-All but the first of these libraries take advantage of web technologies,
-and all but the second rely on open standards. Hadley Wickham is the
-author of ggplot, ggplot2, and ggvis. ggplot and ggplot2 generate
-static graphics, while ggvis provides convenient interactive visualizations
-in R, using the R package Shiny and the Vega abstract visualization grammar,
-opening a separate web server for each call. These projects show
-a trend toward visual analytics frameworks and systems that are
-increasingly rich, interactive, composable, and accessible through the web.
+Some symptoms of the current situation are:
+\begin{itemize}
+\item It is hard to find data, metadata, and previously hard-won knowledge about it.
+\item Most coordination is done through meetings, whose content is not related to other artifacts and may not even be stored.
+\item Production deployment requires porting code to a different environment or even completely rewriting it, thwarting continuous release.
+\item Many tools adopt a standalone or single developer perspective,
+not suited to collaboration and web deployment. 
+\item Analysts find insufficient support for tracing events or issues from production back to the EDA process
+\end{itemize}
 
-% What are we going to do about this?
+In view of the opportunity to improve this situation,
+we created a software environment that supports the end to end
+visual analytics process for individuals and teams.
+This environment knits together some familiar tools, and it
+provides new features to find data and code; create experimental workbooks;
+run experiments; annotate and deploy experiments as end-user websites or
+as reusable, callable services; and to share, search and recommend these artifacts. The artifacts are stored in a version
+control system that provides a common workspace, as well as needed control
+and isolation between stable and experimental versions of code and other
+resources.
 
-%% \todo{Define the requirements for the system}
-%% We begin by noting the main requirements for a data analysis
-%% environment for collaboration and transfer of results into production.
+Over the past two years, we prototyped this environment and deployed it
+in a production community of data scientists, business analysts and other colleagues.
+The platform today has over 200 active accounts, and several dozen people have forked
+and customized code in the system.
 
-%% Main requirement is to satisfy these two sets of users: R coders and
-%% managers. They work differently and want different things. What should
-%% the infrastructure look like so they both coexist in peace? If this is
-%% such a fundamental problem, why have other tools not addressed it?
-%% Discuss Shiny (maybe in the disc. section).
+A key point is that, for the most part, the improved capabilities are
+available by default, without much explicit intervention on the part of
+data scientists and other customers. Visual analytics experiments are
+conveniently shared and turned into production websites, without moving
+or porting code. All the published artifacts in the system can be
+searched immediately. Recommendation is as easy as clicking a star
+on a useful workbook.
 
-Today's state-of-the-art environments for data science teams are unquestionably
-powerful, but also fragmented. Instead of creating a new solution that promotes
-even further fragmentation, we attempt to identify and remedy the shortcomings
-in the current practice of data science. The following requirements summarize
-the main pain points:
+The high level architecture of the prototype system
+is shown in figure~\ref{fig:high-level-architecture}.
+We chose R as the foundation for our prototype because it is already
+the dominant statistical computing language in our lab.
+R also has many useful packages for data analysis and visualization,
+and the core system and its packages are open source that can be modified to
+support research.
+
+There is much previous work that inspired or influenced ours. Some
+software prototypes that we learned from include:
+
+{\bf Vistrails} as an important prototype of a scientific workflow
+and provenance management system. The prototype helped the community
+to understand the value of capturing aspects of the processes that
+surround data analysis experiments and tools, including detailed
+history, collaboration, and deployment.
+
+{\bf iPython notebooks (Jupyter)} are the closest work we are aware of.
+Of particular note,
+
+{\bf RStudio, RShiny} and other packages based on R for creating and
+publishing web content such as knitR, Rpubs, rCharts, and gg2v.
+
+{\bf bl.ocks, jsfiddle, plotly} and other web services for sharing code
+and demos.
+
+{ManyEyes} is the best-known foundational work showing the potential for
+integrating data visualization with social media.
+
+Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
+data from lab experiments and sharing it within research teams \cite{Rubacha:ELN:2011}. In a sense, we hope to adapt and extend this concept to the work of
+visual analytics teams.
+
+From a more principled or theoretical perspective, visual analytics goals
+or objectives for emerging systems, described in previous work, closely match
+the goals of our prototype.  Adopting the terminology of X Y and Z, we noted
+the following.
 
@@
 
-\item Technology transfer. In our experience, development
-of new analyses and visualizations is done by \emph{hackers}
-and \emph{scripters} (using Kandel et al.'s
-terminology~\cite{Kandel:2012:EDA}) who are, by definition,
-at least somewhat familiar with programming and software development.
-On the other hand, \emph{application
-users} ultimately derive value from the data by means of the tools
-developed by hackers and scripters. Traditionally, the connection
-between these two worlds is made by IT staff. In an environment where
-business needs can change every day, manual IT support does not scale:
-we need to create software infrastructure to connect the two worlds.
-%% The two classes of users
-%% have quite different perspectives but we
-%% infrastructure for continuous deployment of visualizations.
+\item Technology transfer.
+In most organizations, development of analyses and visualizations 
+s done by \emph{hackers} and \emph{scripters} (using Kandel et al.'s
+terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
+gain the benefits of the tools developed by hackers and scripters.
+Generally, the connection between these two worlds is made by IT staff,
+who port or even rewrite code so it can run as a stable production service.
+In an environment where business needs can change rapidly, this
+process does not scale. Our objective is to connect the two worlds
+of EDA and production.
 
@@
 isolation of exploratory visualization and data analysis
-environments is a hinderance to wider adoption of modern techniques
-on both sides. We specifically see the richness of interactive
-visualization tools as still being isolated from the richness
-of statistical programming environments.
-%(The path backward from production to analysis can be 
-%equally important: when problems discovered in
-%production require expert help, we would like experts to be
-%able to immediately apply the full range of analytic tools without
-%extracting data to move it back to the EDA environment.)
+environments hinders wider adoption of modern techniques.
+The richness of interactive visualization tools is still somewhat
+separated from the power of statistical programming environments.
 There is an opportunity to provide a framework so that developments
-on either side are more easily employed by the other.
-In addition, the integrated results should be more easily accessible
-to the rest of the organization.
+on each side are more easily adopted by the other, and made available
+in production services.
 
-\item Discoverability. From our experience in working with data
-analyst teams, we find that repeated work (``How can I connect to
-database X?'', ``How do you get clean data from column Y from feed
-Z?'') arises from lack of communication about previously solved subproblems.
-We wish to provide transparency about the work of team members in the organization.
+\item Discoverability. A chronic complaint of data analysis teams is
+repeated work (``How can I connect to database X?'', ``How do you get
+clean data from column Y from feed Z?'')  This work arises from lack of
+communication about previously solved subproblems.  Our goal is encourage
+and support transparency of work between team members.
 
@@
 
-Technologies which fail to address these requirements
-are holding back the practice and impact of interactive visual
-analysis. In this paper, we discuss the design and deployment of
-RCloud, a prototype environment that supports collaborative data
-analysis, visualization and web deployment satisfying all requirements
-above. We will report on the design decisions, tradeoffs and
-limitations, and compare the concepts in RCloud to related proposals.
-
-%% Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
-%% for statistical computing and graphics. R is one of the most popular
-%% systems for data analysis and is well-supported by
-%% a complex ecosystem of compatible packages, software
-%% tools and other resources.
-%
-%% Of importance to us, there are existing
-%% packages for high performance computing in the cloud,
-%% and for generating web graphics. % no, there aren't. we wrote those!
-
-RCloud uses the R language~\cite{RCoreTeam:2013:R} as the core set
-of statistical analysis tools.  The facilities for \emph{sharing, viewing
-and embedding} RCloud notebooks address the {\bf technology transfer} requrement.
-Because RCloud embraces technologies that are part of the computing
-environment (specifically, inter-process communication via HTTP and
-web services as remote procedure calls, as explained
-in Section~\ref{sec:highlevelarchitecture}), it is straightforward to
-integrate RCloud with satellite efforts in corporate environments.
-
-{\bf Coexistence} between analysis and visualization is supported in
-its human interface architecture, described in~\ref{sec:humaninterface}. 
-As oulined in Section~\ref{sec:notebooks}, RCloud notebooks are sequences
-of R snippets interspersed with Markdown for easy generation of reports.
-We describe how RCloud enables interactive visualization widgets in
-this setting (Sections~\ref{sec:Rinbrowser} and~\ref{sec:interactivenotebooks}).
-Indexing, publishing and sharing of notebooks is also the main way
-Rcloud supports {\bf discoverability}.
-
-The requirement of continuous technology transfer, though, creates challenges
-about the \emph{stability} of the many moving pieces: if notebooks are
-updated and deployed automatically, how can the system handle
-changing requirements and public interfaces? We discuss our choice of
-transparent, integrated version control for RCloud notebooks in 
-Section~\ref{sec:deployment}.
-
-Section~\ref{sec:casestudy} describes a case study illustrating some of these concepts.
-
-In Section~\ref{sec:limitations}, we report on experience toward meeting
-these requirements in RCloud, and limitations encountered.
-
-%% Dashboards are a solution to requirement 1, using solutions of
-%% requirement 2.
-%% Requirement 2: \emph{how is accessibility supposed to happen?}
-
-%% Version control as a primary concept in the environment is a
-%% consequence of requirement 1. There
-%% needs to be a way to control the visibility of prototypes as they are
-%% being developed and tested.
-
-%% Tight integration between R process and web browser is a solution to
-%% requirement 2
-
-%% Current isolation notwithstanding, we find there is demand for interactive
-%% visualization tools. We do not believe that moving the \
-
-%% The main challenge here is how to integrate these
-%% technologies so data scientists and statisticians can target
-%% them, \emph{and} business-facing users can use them.
-%% Scalable technologies like crossfilter exist.
-Want to support those in the context of an R environment.
-Need to support full 2-way communication between computing in the cloud and interaction on the web.
-%% \item Other requirements e.g.
-Kandel's user study.
-Not our contribution but things we must address, mentioned in related work section.
-Heer/Agrawala design considerations for collaborative visual analytics.
-
-% What is the scope of our solution?
-
-% We want our solution to play well with the rest of the ecosystem
+\stephen{We could now tell them what we're going to tell them here, but I'm
+not sure that's a good approach - unless we feel the reader is having
+trouble following the path we are laying out, it seems to disrupt the
+flow of the discussion.}

### 2015-01-28 conclusion.tex
yes

--- 2014-03-31 conclusion.tex
@@
 
-We described the motivation introducing web technologies
-for searching, collaborating, recommending, publishing
-reusable notebooks as part of the visual analytics process.
+We presented a case for a visual analytics environment that supports
+the entire work process, from exploratory data analysis
+through development and production deployment. One might think of
+this as ``DevOps for data science.'' The environment supports
+collaboration and communication, with capabilities for
+creating and sharing experimental workbooks, version control,
+searching, recommending, annotating, and deploying experiments
+as web sites and reusable services.
+
 We implemented these ideas in the RCloud prototype.
-Rcloud assists software development and release using
-an approach similar to DevOps in web services.
-Experience with the prototype provided evidence that
-data science teams, and the organizations in which they work,
+Experience with the prototype provided evidence that data
+science teams, and the organizations in which they work,
 benefit from having such capabilities.
 
-Future work will address cross-language development,
-source code analysis, richer recommendation techniques,
-and improvements to the HTML5 front end.
+Reasonable next steps are to support cross-language development,
+richer source code analysis and recommendation techniques, and
+improvements to the HTML5 human interface.
 

### 2015-01-30 conclusion.tex
yes

--- 2014-03-31 conclusion.tex
@@
 
-We described the motivation introducing web technologies
-for searching, collaborating, recommending, publishing
-reusable notebooks as part of the visual analytics process.
+We presented a case for an environment that supports the entire
+visual analytics work process, from data acquisition and exploratory
+data analysis through development and production deployment. One might
+think of this approach as ``DevOps for data science.'' The environment
+supports collaboration and communication with capabilities for
+creating and sharing experimental workbooks, version control,
+searching, recommending, annotating, and deploying experiments
+as web sites and as reusable services.
+
 We implemented these ideas in the RCloud prototype.
-Rcloud assists software development and release using
-an approach similar to DevOps in web services.
-Experience with the prototype provided evidence that
-data science teams, and the organizations in which they work,
+Experience with the prototype provides evidence that data
+science teams, and the organizations in which they work,
 benefit from having such capabilities.
 
-Future work will address cross-language development,
-source code analysis, richer recommendation techniques,
-and improvements to the HTML5 front end.
+Next steps are to support cross-language development, to incorporate
+richer source code analysis and recommendation techniques, and
+to make improvements in the HTML5 human interface.
 

### 2015-01-30 discussion.tex
yes

@@
 protection for RCloud workbooks in an organization's intranet,
-RCloud uses an object capability model
-\cite{Miller:2006:RCT}
+RCloud uses an object capability model \cite{Miller:2006:RCT}
 recently added to the Rserve protocol \cite{Urbanek:2003:AFW}
@@
 sophisticated approaches such as information flow analysis are
-still an active research area, and many subtle problems remain
-\cite{Moore:2011:SAF}. We are hopeful that having a richer
+still an active research area, and many problems have
+been identified \cite{Moore:2011:SAF}. Having a richer
 environment for collaboration and information sharing will encourage

### 2015-01-30 introduction.tex
yes

@@
 \maketitle
+
+Over fifty years ago, John Tukey published ``The Future of Data Analysis''
+in which he defined a remarkably prescient agenda for the future of
+data analysis and statistical computing. His vision has been realized
+through environments like S and R, that have encouraged the adoption of
+statistical graphics, and robust techniques for practical data modeling.
+The software people use today has fulfilled his vision for the practice
+of data analysis by experts.
+
+The field has now moved beyond the boundaries of what Tukey originally
+foresaw. Both computational and networking capabilities today far exceed
+what could have been imagined fifty years ago. Cheap, highly parallel
+virtualized systems allow analysts to solve problems at a scale that
+was once difficult to conceive. But platforms today also enable collaboration
+and support complex tasks and work processes in ways that were also
+hard to imagine in the past. We turn our attention to this opportunity.
 

### 2015-01-30 system.tex
more notes
yes

@@
 
+\stephen{Let's work from front to back: from human interface to the
+services and frameworks that support it.}
+
+\stephen{I don't think we need to even make this argument, it seems so obvious.}
 The internal computing infrastructure of organizations has changed
@@
 
-We designed RCloud around the front-facing API, which provides roughly
-one entry point to correspond with each requirement.
+%\stephen{What??}
+%We designed RCloud around the front-facing API, which provides roughly
+%one entry point to correspond with each requirement.
 
-The center panel of the screen is for editing code and visualizations through
-the notebook interface. Supplementary ``asset'' files which are part of the
-notebook but not part of the notebook flow, are edited on the right.
-Controls at the top of the screen allow the user to quickly run, view, 
-and share the notebook. The call mechanism is the URL itself; arguments are
-read from the URL and the final cell is the result. On the left are the 
-browsing area for seeing other users' notebooks and for searching, as well
-as a help system.
+Figure~\ref{fig:something} shows the RCloud developer interface.
+The center panel is a notebook for code and visualizations.
+Code is edited in this panel, and visualizations are rendered.
+Supplementary ``asset'' files, which are part of the notebook,
+though not part of its executable flow, are edited on the right.
+Controls at the top of the screen allow the user to run, view, and
+share the notebook. The call mechanism is the URL itself; arguments are
+read from the URL and the final cell is the result. On the left is a
+browsing area for searching and seeing other users' notebooks, and a
+help system.
 
@@
 
-The basic unit of operation in RCloud is a \emph{notebook}. A notebook
-holds a sequence of \emph{cells}, each of which represents a
+The main unit of computation in RCloud is a \emph{notebook}.
+A notebook holds a sequence of \emph{cells}, each of which contains a
 snippet of code or hypertext in Markdown. This is not a novel idea;
-executable documents structured in this way are present in a wide
-variety of other software systems, including Mathematica, IPython and Sage.
+executable documents structured this way are present in many
+other systems, including Mathematica, IPython and Sage.
 
-One of the main differences between RCloud and other systems is the
-notion that notebooks are ``always deployed''. This improves the
-ability to share and modify experiments and compose the results
-without binding to specific versions of a notebook.
-On the other hand, there are a variety of situations where it might
-be necessary to name a specific version. Rather than force notebook
-designers to decide which versions need to be preserved,
-we embrace \emph{transparent} versioning.
+One of the main contributions of RCloud is the notion that notebooks
+are ``always deployed''. This improves the ability to share and modify
+experiments and compose the results without binding to specific versions
+of a notebook.
+On the other hand, there are situations where it might be important to
+be able to name a specific version. We do not expect designers to 
+decide which versions need to be preserved, but we embrace
+\emph{transparent} versioning.
 This is similar to models like Jankun-Kelly et al.'s p-set calculus \cite{Jankun-Kelly:2007:MFV}
@@
 
-Implementing an entire version-control subsystem from scratch would
-be a large undertaking. Fortunately, we can build RCloud on top of Github's
-\emph{gists}~\cite{Github:2014:GG}. Github offers a HTTP interface
-for creating simplified git repositories (the main limitation being the
-use of text-only files in a single directory). The GitHub web-service
+To implement this, it is fortunate that we can build RCloud on top of
+Github's \emph{gists}~\cite{Github:2014:GG}. Github offers a HTTP interface
+for creating simplified git repositories, the main limitation being the
+restriction to text-only files in a single directory. The GitHub web-service
 API provides most of the semantics we need for the versioning
@@
 
-Using GitHub was mainly a tactical decision to reduce our
-time-to-prototype. At the same time, it had the beneficial effect of
-allowing our backend to communicate with separate processes over HTTP
-and JSON. This had the serendipitous advantage of quickly enabling
-full search over RCloud notebooks, using Apache SOLR.
+Using GitHub for storage and versioning also has the advantage of allowing
+us to use other Github capabilities that can be invoked by HTTP and JSON.
+Particularly, we can easily provide full search, using Apache SOLR
+\stephen{whatever that is!}
 %
@@
 
-%% %
-%% We needed version control over a directory, gist provides that,
-%% wrapped over an HTTP interface.
-%% %
-%% Mainly a tactical decision to save time.
-%% %
-%% Nevertheless, availability of notebook data as (barely) plain text had good side
-%% effects, like ease to build search.
-%% %
-%% Expect same with suggest.
-
 \subsection{Reputation and Interest: starring\label{sec:starring}}
@@
 \emph{notebooks} and \emph{users}, rather than a relationship between
-user pairs. We decided on this approach because we expect typical
-RCloud deployments to have relatively few users, but each user to have
-written relatively many notebooks. Under that assumption, assigning
+user pairs. We chose this approach because we expect initial 
+RCloud deployments to have relatively few users, but some users will
+relatively many notebooks. Under that assumption, assigning
 interest to users would not provide sufficiently ``high-resolution'' data.
@@
 
+\carlos{compare against https://www.opencpu.org, compare against
+  IPython notebooks. Two-way communication? Actually describe what it
+  is. IPython manipulate notebooks? Understand and contrast}
+
 \subsection{Interactive notebooks\label{sec:interactivenotebooks}}

### 2015-01-30 conclusion.tex
yes

@@
 
-We presented a case for an environment that supports the entire
-visual analytics work process, from data acquisition and exploratory
-data analysis through development and production deployment. One might
-think of this approach as ``DevOps for data science.'' The environment
-supports collaboration and communication with capabilities for
-creating and sharing experimental workbooks, version control,
-searching, recommending, annotating, and deploying experiments
-as web sites and as reusable services.
+We presented a case for an environment that supports the visual
+analytics process, from data acquisition and exploratory
+data analysis, through development and deployment. One might
+think of this approach as ``DevOps for data science.''
+The environment supports collaboration and communication
+with capabilities for creating and sharing experimental workbooks,
+version control, searching, recommending, annotating, and publishing
+experiments as web sites and as reusable services.
 
 We implemented these ideas in the RCloud prototype.
+
 Experience with the prototype provides evidence that data
@@
 
-Next steps are to support cross-language development, to incorporate
-richer source code analysis and recommendation techniques, and
-to make improvements in the HTML5 human interface.
+Some possible next steps are to support cross-language development,
+to incorporate richer source code analysis and recommendation
+techniques, to provide fine-grained security, and to improve
+the usability of the HTML5 human interface.
 

### 2015-01-30 introduction.tex
yes

--- 2015-01-27 introduction.tex
@@
 
-Consider the role of a data science team within an organization today.
+Over fifty years ago, John Tukey published ``The Future of Data Analysis''
+in which he defined a remarkably prescient agenda for the future of
+data analysis and statistical computing. His vision has been realized
+through environments like S and R, that fostered the adoption of
+statistical graphics and robust techniques for practical data modeling.
+The software that we use today has fulfilled his brilliant vision for
+the practice of computational data analysis by individual experts.
+
+The field has now moved beyond the boundaries of what Tukey originally
+foresaw. Processing and networking capabilities today far exceed
+what was only imagined fifty years ago. Cheap, highly parallel
+virtualized systems allow analysts to routinely solve problems
+involving millions or even billions of data values. Computer
+systems today also enable collaboration and the automation of
+complex tasks and processes in ways that were hardly even considered
+fifty years ago. We turn our attention to this opportunity to
+use computation to support data science teams and the visual
+analytics process.
+
+As a starting point, consider the role of a data science team within
+an organization today.
 Analysts work in loose or tight collaboration with colleagues.
 Project teams vary in size, from just a few people to dozens or
-more, even within one project's duration. Tasks are broad, including problem
-identification, data wrangling, visualization, modeling and analysis,
-interpreting, explaining and advocating the results, and often helping
-clients to realize the benefits of the work in ongoing processes.
-There are many details to these tasks.
-For example, data wrangling can involve finding data, figuring out its
-syntax and semantics, assessing data quality, performing normalization and
-data quality remediation, and making the data available for other tasks
-to follow.
+more, even within one project's duration. Assignments are often
+broad, including problem identification, data wrangling, modeling and analysis,
+visualization, summarizing and interpreting results, and explaining and advocating
+them within the organization, helping clients to realize the benefits of the work.
+Eventually, the knowledge or working prototypes created by data science teams
+may be transferred to other organizations to employ them in production.
+There are many details to all these tasks.  For example, data wrangling can involve
+finding data, understanding its syntax and semantics, assessing data quality,
+performing normalization and data quality remediation, and making the data available
+for the other tasks that will follow.
 
@@
 
-\stephen{We could now tell them what we're going to tell them here, but I'm
-not sure that's a good approach - unless we feel the reader is having
-trouble following the path we are laying out, it seems to disrupt the
-flow of the discussion.}
+\stephen{We could tell them what we're going to tell them here, but I'm
+not sure that's good - unless we feel the reader will have trouble following
+the path we are setting, it seems to disrupt the flow of the discussion.}

### 2015-01-31 system.tex
yes

@@
 
-\stephen{Let's work from front to back: from human interface to the
+\stephen{Let's work from front to back, from human interface to the
 services and frameworks that support it.}
@@
 Code is edited in this panel, and visualizations are rendered.
-Supplementary ``asset'' files, which are part of the notebook,
-though not part of its executable flow, are edited on the right.
+An inventory of supplementary ``asset'' files, that are part of
+the notebook, though not in its executable flow, are edited on the right.
 Controls at the top of the screen allow the user to run, view, and
-share the notebook. The call mechanism is the URL itself; arguments are
-read from the URL and the final cell is the result. On the left is a
-browsing area for searching and seeing other users' notebooks, and a
-help system.
+share the notebook. The result of code execution is the final cell of
+the notebook.  On the left is a browsing area for searching and viewing
+other users' notebooks, and a help system.
+
+The call mechanism is the URL itself; arguments are read from the URL
+and the final cell is rendered as the result.  \stephen{That needs
+more explanation.}
 
@@
 snippet of code or hypertext in Markdown. This is not a novel idea;
-executable documents structured this way are present in many
+executable documents structured this way are a feature of many
 other systems, including Mathematica, IPython and Sage.
+\stephen{Code is executed when...?}
 
-One of the main contributions of RCloud is the notion that notebooks
-are ``always deployed''. This improves the ability to share and modify
-experiments and compose the results without binding to specific versions
-of a notebook.
-On the other hand, there are situations where it might be important to
-be able to name a specific version. We do not expect designers to 
-decide which versions need to be preserved, but we embrace
-\emph{transparent} versioning.
+One of the main contributions of RCloud is the idea that notebooks
+are ``always deployed''. In other words, the most recent version of
+a notebook is immediately available.\stephen{What if it is broken?}
+This makes it convenient to share and modify experiments and compose
+results without binding to a specific version of a notebook.
+On the other hand, there are situations where it might be important
+to name specific versions. We do not expect designers to decide which
+versions need to be preserved, but we embrace \emph{transparent} versioning.
 This is similar to models like Jankun-Kelly et al.'s p-set calculus \cite{Jankun-Kelly:2007:MFV}
@@
 
-To implement this, it is fortunate that we can build RCloud on top of
-Github's \emph{gists}~\cite{Github:2014:GG}. Github offers a HTTP interface
-for creating simplified git repositories, the main limitation being the
-restriction to text-only files in a single directory. The GitHub web-service
-API provides most of the semantics we need for the versioning
-portion of the storage back end: access to previous versions,
-comments, starring, and forking.
+To implement this, we built RCloud on top of Github's \emph{gists}~\cite{Github:2014:GG}.
+Github offers a HTTP interface for creating simplified git repositories, the main limitation
+being a restriction to text-only files in a single directory. The GitHub web-service
+API provides most of the semantics we need for the versioning portion of the storage back end:
+access to previous versions, comments, starring, and forking.
 
-Using GitHub for storage and versioning also has the advantage of allowing
-us to use other Github capabilities that can be invoked by HTTP and JSON.
-Particularly, we can easily provide full search, using Apache SOLR
-\stephen{whatever that is!}
+Using GitHub for storage and versioning also exposes other capabilities
+that can be invoked with JSON and HTTP.
+Particularly, we can provide full text search using Apache SOLR.
+SOLR is scalable, can index in near-realtime, supports multiple character sets,
+indexes several common types of documents, and has schemas and faceted search.
 %
-This perspective has implications for
-systems-based research and prototyping in visual analytics:
-for small teams to efficiently explore novel ideas toward the
-improvement of a large, complex software ecosystem, it pays
-to use existing solutions as far as possible.
-%
-Even though it involved learning standards that are not directly
-related to visual analytics, our early decision to use the GitHub
-gist API was successful. Future projects can gain many
-of the same benefits by adopting a similar strategy.
+Integrating existing services and software components, especially open source,
+instead of implementing custom code is a trend in the software industry, and
+is very positive for systems research and prototyping in visual analytics.
+By adopting existing technology, small teams with limited resoures can explore
+novel ideas toward the improvement of large, complex software ecosystems.
+In our case, even though this approach involved learning standards not directly
+related to visual analytics, our early decision to use the GitHub gist API
+turned out to be successful. Future projects could gain many of the same benefits
+by adopting this strategy. In fact, it will be essential for technology adoption
+and transfer, because it is almost impossible for any one tool or platform to
+``own'' or support the entire visual analytics process.
+\stephen{Are we preaching too much?}
 
 \subsection{Reputation and Interest: starring\label{sec:starring}}
+
+Information retrieval based on collecting usage and recommendations
+is a cornerstone of modern web services. We would like to help data
+scientists to find workbooks (and therefore items in their contents
+such as code, data, and colleagues with specific expertise) with
+the benefit of such information.
 
@@
 user pairs. We chose this approach because we expect initial 
-RCloud deployments to have relatively few users, but some users will
-relatively many notebooks. Under that assumption, assigning
-interest to users would not provide sufficiently ``high-resolution'' data.
+RCloud deployments to have relatively few users, but some users to
+create many notebooks. Under that assumption, assigning interest
+to users would not provide sufficiently ``high-resolution'' data.
 
 We incorporate both explicit and implicit indications of interest
-in notebooks. Explicit interest is signaled by ``starring,'' or
+in notebooks. Explicit interest is indicated by ``starring,'' or
 clicking on a button that marks a notebook as interesting.
@@
 
-Implicit signaling of interest is supported by keeping click-through
+Implicit signaling of interest is supported by keeping click-through counts
 \cite{Joachims:2005:AIC} and execution counts. (In addition to these
-classic techniques of collecting feedback from web search, we anticipate
+standard techniques of collecting feedback from web search, we anticipate
 applying static and dynamic code analysis to infer fine-grained
@@
 even while the original author is polishing an algorithm or its
-presentation. The code is the page, and can either be shaped into
+presentation. \stephen{We glossed over the things that can go wrong.
+We should at least discuss that briefly.}
+The code is the page, and can either be shaped into
 a function of inputs and outputs, or the linear cells of the notebook

### 2015-02-17 quotes.tex
Added quotes OCRd from Carlos Kindle capture.

% part1.txt
Shellman: Finally, prototyping our products so that internal
customers can use them early on has been crucial for
our success. It doesn't even have to be something
fancy. For instance, our recommendations preview
tool doesn't have a particularly interesting
visualization, but its enough to show the result of our
algorithms. Now we can shoot off a URL to internal
customers and it allows them to sit at their desk and
understand the behavior of our product and
experiment with it, and provide feedback way before
we're talking about getting it into production. This
has been super helpful and has been a really great
way to get people excited about what we're working
on. Building and maintaining those relationships are
just like anything else-- you always have to be
working on them. Nurturing those relationships are
part of the job, and if you leave them unattended,
they might not be there later.

% part2.txt
Gutierrez: How do you share the knowledge you are
building with others?
Shellman: We’re obsessed with Conﬂuence from
Atlassian, and the Data Lab has a very active
Conﬂuence space. Recommendo is fully documented
on Conﬂuence, so anyone in the company could learn
how it’s built, where to ﬁnd it, and how to use it. We
also share exploratory analyses and reports on
Conﬂuence so that we can still exchange knowledge
even if the work didn’t make it into a larger project.

% part3.txt
Foreman: I lead the data science team at
MailChimp, and I like to get my hands dirty too.
Some of the big pieces of my day involve working
with my team to take stock of current projects and
figure out where to go next, doing my own work and
prototyping things-- I do projects just like my peers
do and then also my talking with other teams,
talking with management, and planning for the
future.
On our team, we’ve got different folks facing different
% part4.txt
kinds of projects. We’ve got one person who really
owns compliance and looks at our compliance
processes. We’ve got another data scientist who
focuses on more of the user experience side of the
house and understanding our customers. I help them
and others as is needed while I do the three things
mentioned earlier-- build data products, be a
translator, and have conversations with the data
science community.

% part5.txt
Lenaghan: First and foremost, it is very important
to be self-critical: always question your assumptions
and be paranoid about your outputs. That is the easy
part, In terms of skills that people should have if they
really want to succeed in the data science ﬁeld, it is
essential to have good software engineering skills. So
even though we may hire people who come in with
very little programming experience, we work very
hard to instill in them very quickly the importance of
engineering, engineering practices, and a lot of good
agile programming practices, This is helpful to them
and us, as these can all be applied almost one-to-one
to data science right now.
If you look at dev ops right now, they have things
such as continuous integration, continuous build,
automated testing, and test harnesses-- a1l of which
map very well from the dev ops world to the data ops
(a phrase I stole from Red Monk) world very easily. I
think this is a very powerful notion. It is important to
have testing frameworks for all of your data, so that if
you make a code change, you can go back and test all
of your data. % Having an engineering mindset is

% part6.txt
In addition to Bitly’s extraordinary data sets, I was
given all the resources that I needed. These were
resources that I’d only heard about but had never
seen. I went from asking what a Hadoop cluster was,
as my school didn't have one, to being able to work
with one. Now I could play with one to figure out how
to make my work better and how to change my
algorithms so they were cleaner and ran faster on
Hadoop. The whole transition was eye opening. Not
only that, I could also look at other people’s code and
play with their code and data sets as well. It was
fantastic and I was convinced that working with data
was my thing.

### 2015-02-18 introduction.tex
yes

--- 2015-01-27 introduction.tex
@@
 
-Consider the role of a data science team within an organization today.
-Analysts work in loose or tight collaboration with colleagues.
+Over fifty years ago, John Tukey published ``The Future of Data Analysis''
+in which he defined a remarkably prescient agenda for the future of
+data analysis and statistical computing. His vision has been realized
+through environments like S and R, that started  a revolution by
+integrating statistical graphics, computation, and robust techniques
+for practical data modeling. Every day, many thousands of data scientists
+and others specialists benefit from this brilliant vision.
+
+The field has now moved beyond the boundaries of what Tukey originally
+foresaw. Processing and networking capabilities today far exceed
+what was barely imagined fifty years ago. Inexpensive, scalable
+virtualized systems allow analysts to routinely solve problems
+involving millions or even billions of data values. Networking
+also enables both collaboration and the automation of complex
+tasks and processes in ways that were not considered even ten years ago.
+We turn our attention to this opportunity to use computation
+to support, not only individuals, but entire teams and their
+work within larger organizations.
+
+As a starting point, consider the role of a data science team within
+a business or technical organization today.
+Analysts work in loose or tight collaboration with colleagues and
+domain experts.
 Project teams vary in size, from just a few people to dozens or
-more, even within one project's duration. Tasks are broad, including problem
-identification, data wrangling, visualization, modeling and analysis,
-interpreting, explaining and advocating the results, and often helping
-clients to realize the benefits of the work in ongoing processes.
-There are many details to these tasks.
-For example, data wrangling can involve finding data, figuring out its
-syntax and semantics, assessing data quality, performing normalization and
-data quality remediation, and making the data available for other tasks
-to follow.
+more, even within one project's duration. Assignments are often
+broad, and include tasks such as problem identification,
+data wrangling, modeling, analysis, visualization, summarization
+and interpretation of results, and explaining them and recommending
+decisions, helping clients to realize the benefits of the analysis.
+Eventually, knowledge or working prototypes created by data science teams
+may be transferred to other organizations to employ them in production.
+There are many details to these tasks. For example, data wrangling can involve
+finding data, understanding its syntax and semantics, assessing data quality,
+performing normalization and data quality remediation, and making the data
+available to other tasks that will follow.
 
-Visual analytics in practice depends greatly on communication and collaboration.
+Visual analytics depends greatly on communication and collaboration.
 At almost every step, detailed knowledge about data, code and tasks
@@
 Thus, data scientists and developers are being asked to become
-very broad, and integrate work across the spectrum. (Hiring ``full stack
-developers'' to perform integration seems to be a trend, but does not achieve
-the potential benefits of specialization and division of labor in larger
-teams, and is not a practice to be encouraged.)
-This pushes the boundaries for communication and collaboration even further.
-Unfortunately, processes and technology that support this work are fragmented.
+very broad, and integrate work across the spectrum.
+
+For example, interview comments found in {\it Data Scientists at Work}
+by Sebastian Gutierrez illuminate the current practice of visual analytics.
+There is a common theme, that data science or visual analytics is 
+being done in teams that more share their work, both data and code,
+more freely than before. We particularly want to point out the remarks
+by Jonathan Lenaghan, who anticipates an environment for continuous
+integration and publication of data analysis.
+
+\begin{addmargin}[1em]{1em}  % left and right margin change
+\small
+\input{quotes}
+\end{addmargin}
+
+% (Hiring ``full stack developers'' to perform integration seems to be a trend, but does not achieve the potential benefits of specialization and division of labor in larger teams, and is not a practice to be encouraged.)
+
+Unfortunately, processes and technology that support visual analytics today
+work are fragmented.
 Knowledge is shared through informal conversations, email online messages,

### 2015-02-18 quotes.tex
yes

@@
 % part1.txt
-Shellman: Finally, prototyping our products so that internal
+Erin Shellman: Finally, prototyping our products so that internal
 customers can use them early on has been crucial for
@@
 % part3.txt
-Foreman: I lead the data science team at
+John Foreman: I lead the data science team at
 MailChimp, and I like to get my hands dirty too.
@@
 % part5.txt
-Lenaghan: First and foremost, it is very important
+Jonathan Lenaghan: First and foremost, it is very important
 to be self-critical: always question your assumptions
@@
 part, In terms of skills that people should have if they
-really want to succeed in the data science ﬁeld, it is
+really want to succeed in the data science field, it is
 essential to have good software engineering skills. So
@@
 % part6.txt
-In addition to Bitly’s extraordinary data sets, I was
-given all the resources that I needed. These were
+Anna Smith: In addition to Bitly’s extraordinary data sets,
+I was given all the resources that I needed. These were
 resources that I’d only heard about but had never

### 2015-02-19 discussion.tex
Yes

@@
 
-Returning to Heer and Agrawala's design considerations, our experience
-underscores the relevance of their proposals, and indicates areas
-for further exploration. We briefly examine these aspects of the
-visual analytics environment.
+Reconsidering Heer and Agrawala's design considerations, our
+experience confirms the relevance of their proposals, and indicates
+areas for further exploration. We briefly examine these aspects of
+the visual analytics environment.
 
-{\bf Shared artifacts and artifact histories} are a central feature
-of RCloud. The prototype aimed at making it convenient to collaborate
-on experimental workbooks and the resulting analyses. We observed
-that hackers readily shared ideas through RCloud. They used RCloud
-to discuss data analysis, programming problems, and to publish results
-by sharing URLs instead of instead of embedding screenshots in
-PowerPoint. Artifact histories are provided through git. It is fair
-to criticize that it is not human-friendly, but we assume RCloud
-hackers are already familiar with it.
+{\bf Shared artifacts and artifact histories} are key features
+of RCloud. We observed that hackers readily used its capabilities
+to share code and experiments, discuss data analysis, help one another
+with programming problems, and to publish results by sharing URLs
+instead of embedding screenshots in PowerPoint. Artifact histories are
+maintained in git. We admit that git is not human-friendly, but many
+hackers are familiar with it. Better graphical tools to explore and
+control provenance would be an advantage.
 
 {\bf Modularity and granularity.} The ability to segment work into
-independent units (modularity) is a key to working in teams. It is
-further an advantage if the units may be kept small, and team members
-can realize benefits that are at least proportional to their work
-on these units (granularity). RCloud's notebook and versioning
-capabilities allow work to be divided into units as fine as
-the underlying language allows, and versioning encourages making
-incremental changes at low cost and without disrupting the work
-of others.
+independent modules is a key to working in teams. It is a further
+advantage if the units can be kept small, and team members can
+realize benefits that are at least proportional to their work
+on these units, which is referred to as granularity. RCloud's notebook
+and versioning capabilities allow work to be divided into units as fine
+as the underlying language allows, and versioning encourages making
+incremental changes at low cost and without disrupting the work of others.
 
@@
 and accessed as URLs. This proved to be an effective mechanism for
-sharing, and for analyses to be integrated with external processes
-and systems.
+sharing, and their ubiquity allows analyses to be integrated into
+external processes and systems.
 
-{\bf Discussion.} Annotation and commenting was another central goal.
+{\bf Discussion.} Annotation and commenting was another important goal.
 Commenting is supported through github, but our hackers found it
 awkward and did not exercise it as much as we expected.
-making it a bit awkward and
-not as widely used as we hoped.  More elaborate annotation was not
-implemented, and remains as future work.
+More elaborate annotation was not implemented, and remains as future work.
 An interesting question is, to what extent should application users
-be able to make annotations and operate similar tools in published
-notebooks without coding and being exposed to the hackers's view?
-Our prototype has not addressed this problem yet.
+be able to make annotations in published notebooks without coding and
+being exposed to the hackers's view?
+The RCloud prototype has not addressed this problem yet.
 
@@
 editing? Should it incorporate an email client and emacs?
-Should it be able to push results to Facebook?)
+Should it be able to push content to Facebook?)
 
@@
 are supported through starring. This mechanism was employed often
-by our test audience. An obvious next step would be to enhance
+by our test audience. A potential next step would be to enhance
 recommendations via relationships discovered by static and
-dynamic code analysis. This should be done both within and across
+dynamic code analysis. This could be done both within and across
 collections of scripts (the latter being similar to Vistrails'
 enhanced recommendations by clustering multiple workflows).
-It should be valuable to hackers to know which packages are
-frequently used together, or appear with a given record type
-or data feed. In general, though, trivial or passive mechanisms
-to collect data for recommendations are essential.
+It should be highly valuable to hackers to know which packages
+are frequently used together, or appear with a given record type
+or data feed. Such knowledge is important, and difficult to acquire today.
+In general, trivial or passive mechanisms to collect data for recommendations
+are essential.
 
@@
 able to curate groups of related notebooks using the simple folder
-system.  This is particularly effective for bringing together
+system. This proved to be particularly effective for organizing
 training materials.
@@
 We did not anticipate having to back off so soon on the
-assumption that programming is being done mainly in one
-language (R). In retrospect, it should have been obvious
-that many data science problems are already being solved
-with cross-language approaches.
+assumption that programming is done mainly in one language (R).
+In retrospect, we should have realized that many data science problems
+are already being solved with cross-language approaches.
 
@@
 are more difficult to manage in RCloud than in conventional systems.
-Not only does it lack an explicit separation of the development and
-deployment environments, but every version of every prototype is
-shared by default. While this encourages collaboration on work
-in progress, errors such as misconfiguration, mismatches between
+Not only does RCloud intentionally lack an explicit separation of the
+development and deployment environments, but every version of every
+prototype is shared by default. While this encourages collaboration on
+work in progress, errors such as misconfiguration, mismatches between
 versions of packages, and ordinary programming errors are also
@@
 this price (as proven true with distributed systems in general).
-Still, this problem needs more attention if the environment is
-going to be scaled up to a wider number of users with fewer
-informal channels to coordinate work. Effective sharing in RCloud
-will require control over versioning.
+Still, this problem needs more attention if the RCloud environment is
+going to be scaled up to more users with fewer informal channels to
+coordinate work.
+
+Effective sharing in RCloud will require more control over versioning.
 Versioning must be managed in several places: in scripts, in the R environment
@@
 control for the external environment. Most programmers rely on clean living
-and a conservative approach to upgrades.
+and a conservative approach to global upgrades.
 
@@
 computations that rely on volatile or unreliable data sources.
-For example, the stock ticker use case is one for which the data
+For example, data fetched from a remote service (like historical stock data)
 might temporarily become unavailable, so caching could improve
@@
 sophisticated approaches such as information flow analysis are
-still an active research area, and many problems have
-been identified \cite{Moore:2011:SAF}. Having a richer
-environment for collaboration and information sharing will encourage
-practical applications for new approaches to information security
-in visual analytics.
+still an active research area, and many unsolved problems have
+been identified \cite{Moore:2011:SAF}. Having a richer environment for
+collaboration and information sharing will encourage more research
+and the adoption of better solutions to this problem.

### 2015-02-19 introduction.tex
Yes

@@
 Over fifty years ago, John Tukey published ``The Future of Data Analysis''
-in which he defined a remarkably prescient agenda for the future of
+in which he defined a remarkably farsighted agenda for the future of
 data analysis and statistical computing. His vision has been realized
@@
 for practical data modeling. Every day, many thousands of data scientists
-and others specialists benefit from this brilliant vision.
+and others specialists benefit from this astounding vision.
 
@@
 broad, and include tasks such as problem identification,
-data wrangling, modeling, analysis, visualization, summarization
-and interpretation of results, and explaining them and recommending
-decisions, helping clients to realize the benefits of the analysis.
+data wrangling, modeling, analysis, visualization, summarization,
+presentation and interpretation of results, and recommending
+actions to help clients to realize the benefits of the analysis.
 Eventually, knowledge or working prototypes created by data science teams
@@
 Further, data scientists are increasingly asked to work more closely
-with business people who are domain specialists and less technically oriented.
+with business or domain specialists who may be less technically oriented.
 Thus, data scientists and developers are being asked to become
@@
 
-For example, interview comments found in {\it Data Scientists at Work}
-by Sebastian Gutierrez illuminate the current practice of visual analytics.
-There is a common theme, that data science or visual analytics is 
-being done in teams that more share their work, both data and code,
-more freely than before. We particularly want to point out the remarks
-by Jonathan Lenaghan, who anticipates an environment for continuous
+The following interview comments from
+{\it Data Scientists at Work}
+by Sebastian Gutierrez illuminate the current practice of teamwork
+and sharing in visual analytics teams.
+We particularly noted the remarks by
+Jonathan Lenaghan, who anticipates an environment for continuous
 integration and publication of data analysis.
@@
 \small
+\stephen{Can we get away with \verb"\smaller"? Ha ha.}
 \input{quotes}
@@
 
-This disjointed process inhibits the use of tools to bridge
+This disjointed, inconsistent process inhibits the use of tools to bridge
 the many gaps and to help people find resources.
-For example, when code is shared, we still not may know where to find examples
-that show how to use it. stephen{Why did I shift to first person?}
+For example, even when code is shared, we still not may know where to
+find examples that show how to use it.
 When we find a useful data source, we don't know what packages or
-other data sources usually go along with it, or what problems with the data
+other data sources usually go along with it, what problems with the data
 others have already discovered and corrected, or where to obtain online feeds
-to get updates of the data.
+to get updates.
 
-The current situation affects how well data science teams work. It limits
-the ability to share work and to rapidly deploy and refine results.
+This situation definitely affects how well data science teams work.
+It limits our ability to share work and to rapidly deploy and refine results.
 We have seen this in our lab environment.
@@
 \begin{itemize}
-\item It is hard to find data, metadata, and previously hard-won knowledge about it.
+\item It is hard to find data, metadata, and knowledge about them.
 \item Most coordination is done through meetings, whose content is not related to other artifacts and may not even be stored.
@@
 not suited to collaboration and web deployment. 
-\item Analysts find insufficient support for tracing events or issues from production back to the EDA process
+\item Analysts find insufficient support for tracing events or issues from production back to the EDA process.
 \end{itemize}
@@
 visual analytics process for individuals and teams.
-This environment knits together some familiar tools, and it
+This environment knits together some familiar tools, and 
 provides new features to find data and code; create experimental workbooks;
@@
 The platform today has over 200 active accounts, and several dozen people have forked
-and customized code in the system.
+and customized code in the system. Most of the active users are members of
+the statistics research department of our lab, but some others are data
+scientists and other collaborators in business units. \stephen{I hope this is right.}
 
@@
 
-There is much previous work that inspired or influenced ours. Some
-software prototypes that we learned from include:
+There is much previous work that inspired or influenced ours.
+Indeed, many studies have addressed the requirements for future visual
+analytics environments.
+Previous work that we learned from includes:
 
-{\bf Vistrails} as an important prototype of a scientific workflow
-and provenance management system. The prototype helped the community
+{\bf Vistrails} as a landmark prototype of a scientific workflow
+and provenance management system. This prototype helped the community
 to understand the value of capturing aspects of the processes that
@@
 {\bf iPython notebooks (Jupyter)} are the closest work we are aware of.
-Of particular note,
+The goal of this project is to provide shared notebooks and a remote
+execution environment for interpreted scripts. \stephen{Are there 
+significant differences involving versioning, provenance, annotation
+and other broad process support?}
 
@@
 publishing web content such as knitR, Rpubs, rCharts, and gg2v.
+These tools are not collaborative, but they are aimed at better
+graphics and web applications.
 
@@
 
-{ManyEyes} is the best-known foundational work showing the potential for
+{ManyEyes} is the best-known original work showing the potential for
 integrating data visualization with social media.
@@
 Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
-data from lab experiments and sharing it within research teams \cite{Rubacha:ELN:2011}. In a sense, we hope to adapt and extend this concept to the work of
+and sharing data from scientitic lab experiments\cite{Rubacha:ELN:2011}.
+In a sense, we hope to adapt and extend this concept to the work of
 visual analytics teams.
@@
 In most organizations, development of analyses and visualizations 
-s done by \emph{hackers} and \emph{scripters} (using Kandel et al.'s
+s done by \emph{hackers} and \emph{scripters} (adopting Kandel et al.'s
 terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
@@
 In an environment where business needs can change rapidly, this
-process does not scale. Our objective is to connect the two worlds
+process does not scale. Our objective is to merge the worlds
 of EDA and production.
@@
 isolation of exploratory visualization and data analysis
-environments hinders wider adoption of modern techniques.
+environments hinders wider adoption of modern techniques from each.
 The richness of interactive visualization tools is still somewhat
@@
 \end{enumerate}
-
-\stephen{We could tell them what we're going to tell them here, but I'm
-not sure that's good - unless we feel the reader will have trouble following
-the path we are setting, it seems to disrupt the flow of the discussion.}

### 2015-02-19 quotes.tex
Yes

@@
 % part1.txt
-Erin Shellman: Finally, prototyping our products so that internal
+Erin Shellman [Nordstrom]: Finally, prototyping our products so that internal
 customers can use them early on has been crucial for
@@
 % part5.txt
-Jonathan Lenaghan: First and foremost, it is very important
+Jonathan Lenaghan [PlaceIQ]: First and foremost, it is very important
 to be self-critical: always question your assumptions
@@
 % part6.txt
-Anna Smith: In addition to Bitly’s extraordinary data sets,
+Anna Smith: In addition to Bitly's extraordinary data sets,
 I was given all the resources that I needed. These were

### 2015-03-11 conclusion.tex
yes

@@
 analytics process, from data acquisition and exploratory
-data analysis, through development and deployment. One might
-think of this approach as ``DevOps for data science.''
+data analysis, through development and deployment.
+This approach can be considered ``DevOps for data science''
+or reproducible data science.
 The environment supports collaboration and communication
@@
 We implemented these ideas in the RCloud prototype.
-
 Experience with the prototype provides evidence that data
-science teams, and the organizations in which they work,
+science teams and the organizations in which they work
 benefit from having such capabilities.
@@
 Some possible next steps are to support cross-language development,
-to incorporate richer source code analysis and recommendation
-techniques, to provide fine-grained security, and to improve
-the usability of the HTML5 human interface.
+to incorporate richer recommendation techniques, to provide fine-grained
+information security, and to improve the usability of the human interface.
 

### 2015-03-20 conclusion.tex
Renamed some files

@@
 
-We presented a case for an environment that supports the visual
+We presented a case for environments that support the visual
 analytics process, from data acquisition and exploratory
 data analysis, through development and deployment.
-This approach can be considered ``DevOps for data science''
-or reproducible data science.
 The environment supports collaboration and communication
-with capabilities for creating and sharing experimental workbooks,
-version control, searching, recommending, annotating, and publishing
+through shared experimental workbooks with version control,
+searching, recommending, annotating, and publishing
 experiments as web sites and as reusable services.
@@
 
-Some possible next steps are to support cross-language development,
+This work prototype a step toward practical ``DevOps for data
+science'' and offers reproducible experiments.
+Possible next steps are to support cross-language development,
 to incorporate richer recommendation techniques, to provide fine-grained

### 2015-03-20 discussion.tex
Renamed some files

--- 2015-01-23 discussion.tex
@@
 
-{\bf Curation.} Even without more formal group management, users are
-able to curate groups of related notebooks using the simple folder
-system.  This is particularly effective for bringing together
-training materials.
+{\bf Curation.} Even without formal group management, users are
+able to curate groups of related notebooks using the folder
+system. This is particularly effective for collecting and
+distributing training materials.
 
@@
 
-Because we developed the ideas and the prototype together,
-we did not foresee some of the requirements that emerged
-after people started working with the prototype.
+Because we developed these ideas while creating a prototype,
+we did not foresee some of the requirements that emerged after
+people started working with it.
 
 {\bf Multiple language support.}
-We did not anticipate having to back off so soon on the
-assumption that programming is being done mainly in one
-language (R). In retrospect, it should have been obvious
-that many data science problems are already being solved
-with cross-language approaches.
+We did not anticipate having to back off so soon on the assumption that
+programming is done mainly in R.  In retrospect, it is clear that many
+data science problems are solved with cross-language approaches
+(for example, scikit-learn in Python, weka in Java).
 
 {\bf Versioning.}
-A disadvantage is that some important aspects of the environment
-are more difficult to manage in RCloud than in conventional systems.
-Not only does it lack an explicit separation of the development and
-deployment environments, but every version of every prototype is
-shared by default. While this encourages collaboration on work
-in progress, errors such as misconfiguration, mismatches between
-versions of packages, and ordinary programming errors are also
-easily shared and can affect production websites unintentionally.
-This cannot be avoided entirely, but the power of sharing is worth
+Some important aspects of the environment are more difficult to manage
+in RCloud than in conventional systems.  RCloud does not explicitly
+separate the development and deployment environments. Actually, every
+version of every prototype is shared by default. Although this encourages
+collaboration on work in progress, it also quickly exposes errors such as
+misconfiguration, mismatches between versions of packages, and programming
+errors that can affect production websites unintentionally.
+This cannot be avoided entirely, but the power of sharing is often worth
 this price (as proven true with distributed systems in general).
-Still, this problem needs more attention if the environment is
+Still, effective control over versioning when sharing 
+problem needs more attention if the environment is
 going to be scaled up to a wider number of users with fewer
-informal channels to coordinate work. Effective sharing in RCloud
-will require control over versioning.
+informal channels to coordinate work.
 Versioning must be managed in several places: in scripts, in the R environment
@@
 remote services. At one end of the scale, we have full control over the
-versioning of scripts via git, though there is still a need for conventions
-for example to name stable or working versions of scripts or versions known
-to work together. The R environment is under the control of its package
-manager. It is at least possible for RCloud to access information about
+versioning of scripts via git, along with conventions to name stable or
+working versions of scripts or versions known to work together.
+The R environment is under the control of its package manager.
+It is at least possible for RCloud to access information about
 package versions and configurations, but support for checking compatibility
@@
 problem of having a lot of bits and pieces of code and data lying
-around, though at least we have made it easier to get at them and
-to apply algorithms and metadata to organize them.
+around, though we have provided a structure in which it should be
+easier to find them and to apply algorithms and metadata to organize them.
+\stephen{Why should the reader believe that?}
 
@@
 might temporarily become unavailable, so caching could improve
-reliability and consistency of results.  Alternatively, in a use case
-involving text analysis through a topic modeling technique such as
-Latent Dirichlet Allocation (LDA) that is relatively expensive,
-we might want caching to improve performance. Currently we would
-program this in RCloud by explicitly saving the analysis in a
-persistent database, but it would be desirable to implement this
-in a more general purpose associative cache. We envision cache
-management being implemented in a new middle layer not in RCloud's
-architecture today.
+reliability and consistency of results.  Alternatively, in situations
+involving relatively expensive analysis (for example, analysis of large
+text corpora, clustering of multivariate time series, deep learning, etc.)
+caching may be essential to provide adequate performance. Currently we would
+program this in RCloud by explicitly saving the analysis in a persistent database,
+but it would be desirable to implement this in a general purpose associative
+cache, not in application-dependent libraries.  We envision cache management
+being implemented in a new middle layer not in RCloud's architecture today.
 
@@
 lists, which tend to be cumbersome. 
-Information security in RCloud should be re-examined, but more
-sophisticated approaches such as information flow analysis are
-still an active research area, and many subtle problems remain
+Information security in RCloud should be improved, but more
+sophisticated approaches (such as information flow analysis,
+and query languages that ensure differential privacy) are
+active research topics, and many difficult problems remain
 \cite{Moore:2011:SAF}. We are hopeful that having a richer
 environment for collaboration and information sharing will encourage
-practical applications for new approaches to information security
-in visual analytics.
+new approaches to information security in visual analytics.

### 2015-03-20 conclusion.tex
working

@@
 
-We implemented these ideas in the RCloud prototype.
-Experience with the prototype provides evidence that data
+Building from the extensive previous literature identifying
+requirements for future visual analytics systems, our goal was
+to gain experience and further insight by implementing and
+deploying a prototype with a community of working data scientists.
+Experience with the RCloud prototype provides evidence that data
 science teams and the organizations in which they work
-benefit from having such capabilities.
+benefit from having such capabilities that support a
+broad view of the visual analytics process.
 
-This work prototype a step toward practical ``DevOps for data
-science'' and offers reproducible experiments.
+Rcloud proposes a step toward practical ``DevOps for data
+science'' with reproducible experiments.
 Possible next steps are to support cross-language development,

### 2015-03-20 discussion.tex
working

@@
 \section{Discussion, Lessons and Limitations}
-\label{sec:limitations}
+\label{sec:discussion}
 
@@
 
-\carlos{All this stuff needs to appear in the paper much earlier.}
+Experience with deploying RCloud in a community of data analysts
+(``hackers'') gave us some insight into whether the proposed
+requirements are appropriate, and how well they were met.
 
-Experience with deploying RCloud in a community of data analysts
-(``hackers'') has provided some insight about whether the proposed
-requirements were appropriate, and how well they were met.
-\stephen{The term {\it hackers} may be too tainted for us to use.}
-% I know the following is true but also feel we need to let the work speak for itself. SCN
-%This experience may be informative to researchers and implementers
-%of future systems supporting collaborative visual analytics.
-
-Returning to Heer and Agrawala's design considerations, our experience
-underscores the relevance of their proposals, and indicates areas
-for further exploration. We briefly examine these aspects of the
-visual analytics environment.
+Our experience underscored the relevance of Heer and Agrawala's design
+considerations, and indicated areas for further exploration. We adopt
+their taxonomy for the following discussion.
 
 {\bf Shared artifacts and artifact histories} are a central feature
-of RCloud. The prototype aimed at making it convenient to collaborate
-on experimental workbooks and the resulting analyses. We observed
-that hackers readily shared ideas through RCloud. They used RCloud
-to discuss data analysis, programming problems, and to publish results
-by sharing URLs instead of instead of embedding screenshots in
-PowerPoint. Artifact histories are provided through git. It is fair
-to criticize that it is not human-friendly, but we assume RCloud
-hackers are already familiar with it.
+of RCloud, through its shared workbooks of experiments and analyses.
+We observed that hackers readily shared work through RCloud. They
+used it to discuss ideas and results in data analysis, to ask for help
+with programming problems, and to publish results by sharing URLs.
+Artifact histories are provided through a version management system (git).
+\stephen{Do analysts type raw git commands?}
 
-{\bf Modularity and granularity.} The ability to segment work into
-independent units (modularity) is a key to working in teams. It is
-further an advantage if the units may be kept small, and team members
-can realize benefits that are at least proportional to their work
-on these units (granularity). RCloud's notebook and versioning
-capabilities allow work to be divided into units as fine as
-the underlying language allows, and versioning encourages making
-incremental changes at low cost and without disrupting the work
-of others.
+{\bf Modularity and granularity.} The ability to partition work into
+independent units (modularity) is a key to working productively in teams.
+It is good if the units can be kept small, so team members can realize
+benefits at least proportional to the work on the units (granularity).
+RCloud's notebook and versioning capabilities allow work to be divided
+into units as fine as the underlying language allows, and versioning
+encourages making incremental changes at low cost and without disrupting
+the work of others.
 
@@
 and accessed as URLs. This proved to be an effective mechanism for
-sharing, and for analyses to be integrated with external processes
-and systems.
+sharing and for integrating analyses with external processes and systems.
+It is particularly advantageous for work to be shared as URLs that
+provide access to code and experiments, instead of by pasting static
+screenshots into documents and presentations.
 
@@
 awkward and did not exercise it as much as we expected.
-making it a bit awkward and
-not as widely used as we hoped.  More elaborate annotation was not
-implemented, and remains as future work.
 An interesting question is, to what extent should application users
-be able to make annotations and operate similar tools in published
-notebooks without coding and being exposed to the hackers's view?
-Our prototype has not addressed this problem yet.
+be able to make annotations in published notebooks without coding
+and being exposed to the hackers's view? The design of
+more elaborate, integrated annotation remains as future work.
 
-{\bf Content export} is not a capability we worked to support,
-but R already has many packages for this. Our goal was to devise
-ways of working where explicit export is less necessary.
-However it is desirable to integrate results
-through web protocols such as JSON. We did assume that RCloud
-notebooks need to play well in the ecosystem of other tools,
-and this worked out as intended, though we learned it also creates
-a temptation to absorb features almost without limit. (Should
-RCloud support templates for personal web pages and graphical
+{\bf Content export} is not a capability we aimed at supporting,
+and R already has many packages for this. Instead, our goal was to
+encourage ways of working without needing as much explicit export.
+We did assume that RCloud notebooks should play well in the ecosystem
+of other tools, and this worked as intended, though we learned it also
+creates a temptation to absorb features almost without limit. (Should
+RCloud provide templates for personal web pages and graphical
 editing? Should it incorporate an email client and emacs?
-Should it be able to push results to Facebook?)
+Should it push results to Facebook?) Experience shows that interfaces
+to external systems and other features almost always win over architectural
+purity in the real world.
 
@@
 by our test audience. An obvious next step would be to enhance
-recommendations via relationships discovered by static and
-dynamic code analysis. This should be done both within and across
+recommendations using relationships discovered by static and
+dynamic code analysis. This may be considered both within and across
 collections of scripts (the latter being similar to Vistrails'
 enhanced recommendations by clustering multiple workflows).
-It should be valuable to hackers to know which packages are
-frequently used together, or appear with a given record type
-or data feed. In general, though, trivial or passive mechanisms
-to collect data for recommendations are essential.
+It seems valuable to know which packages are frequently used together,
+or appear in proximity to a certain record types or data feeds.
+In general, trivial or passive mechanisms to collect data for
+recommendations are essential.
 
-{\bf Group management, size and diversity.} This area lacks sufficient
+{\bf Group management, size and diversity.} This area needs better
 support in RCloud, but is clearly important to working in teams.
-We rely on existing administrative and social processes and conventions,
-outside RCloud, to manage accounts and groups. This capability
-could likely benefit from integrating social media to track
-identities and groups and to maintain communication channels.
+We rely on external administrative processes and social conventions
+to manage accounts and groups. Rcloud could benefit greatly from
+integrating social media to track identities and groups and to maintain
+communication channels, instead of having its own isolated solution.
 
-{\bf Curation.} Even without formal group management, users are
-able to curate groups of related notebooks using the folder
-system. This is particularly effective for collecting and
-distributing training materials.
+{\bf Curation.} Even without formal group management, users may
+curate groups of related notebooks using file system folders.
+We found this particularly effective in collecting and distributing
+training materials.
 
 \subsection{Limitations}
-
-Because we developed these ideas while creating a prototype,
+Because we developed our ideas while simultaneously creating a prototype,
 we did not foresee some of the requirements that emerged after
-people started working with it.
+people started working with RCloud.
 
 {\bf Multiple language support.}
-We did not anticipate having to back off so soon on the assumption that
-programming is done mainly in R.  In retrospect, it is clear that many
-data science problems are solved with cross-language approaches
-(for example, scikit-learn in Python, weka in Java).
+We did not forsee the demand to relax the assumption that programming
+is done mainly in R. In retrospect, it is clear that many data science
+problems are solved with cross-language approaches (for example, scikit-learn
+in Python, weka in Java).\stephen{Isn't this fixed in some way now?}
 
@@
 in RCloud than in conventional systems. RCloud does not explicitly
-separate the development and deployment environments. Actually, every
-version of every prototype is shared by default. Although this encourages
-collaboration on work in progress, it also quickly exposes errors such as
+separate the development and deployment environments. More than that,
+ever version of every prototype is shared by default. Although this encourages
+collaboration on work in progress it also quickly exposes errors such as
 misconfiguration, mismatches between versions of packages, and programming
-errors that can affect production websites unintentionally.
-This cannot be avoided entirely, but the power of sharing is often worth
-this price (as proven true with distributed systems in general).
-Still, effective control over versioning when sharing 
-problem needs more attention if the environment is
-going to be scaled up to a wider number of users with fewer
+errors that can unintentionally affect production websites. Such errors
+are difficult to completely avoid, but the power of convenient sharing is often
+worth it (as proven true with distributed systems in general).
+Still, effective control over versioning when sharing problem needs more
+attention if Rcloud would be scaled up to a wider number of users lacking
 informal channels to coordinate work.
-Versioning must be managed in several places: in scripts, in the R environment
-such as the installed R libraries and packages, and in the external
-environment such as the operating system and even protocols spoken by
-remote services. At one end of the scale, we have full control over the
-versioning of scripts via git, along with conventions to name stable or
-working versions of scripts or versions known to work together.
-The R environment is under the control of its package manager.
+Versioning needs to be managed in several places: in scripts;
+in the R environment such as the installed libraries and packages;
+and in the external environment such as the operating system and
+protocols spoken by remote services. At one end of the scale, we have
+full control over the versioning of scripts via git, along with conventions
+to name stable or working versions of scripts, and versions known to work together.
+The R environment itself is under the control of its package manager
+that has rules to ensure reasonable consistency.
 It is at least possible for RCloud to access information about
@@
 control for the external environment. Most programmers rely on clean living
-and a conservative approach to upgrades.
+and a careful approach to system upgrades.
 
-On top of this, RCloud allows hackers to fork experiments to try
-new ideas quickly. So far we have not done much more to address the
-problem of having a lot of bits and pieces of code and data lying
-around, though we have provided a structure in which it should be
-easier to find them and to apply algorithms and metadata to organize them.
-\stephen{Why should the reader believe that?}
+On top of this, RCloud allows or even encourages hackers to fork experiments
+to try new ideas quickly. So far we have not done much more to address the
+problem of having a lot of bits and pieces of code and data lying around,
+though we have provided a framework in which it should be easier to find them
+and to apply algorithms and metadata to organize them.
+\stephen{I think we need to defend that a bit more.}
 
@@
 reliability and consistency of results. Alternatively, in situations
-involving relatively expensive analysis (for example, analysis of large
-text corpora, clustering of multivariate time series, deep learning, etc.)
-caching may be essential to provide adequate performance. Currently we would
-program this in RCloud by explicitly saving the analysis in a persistent database,
-but it would be desirable to implement this in a general purpose associative
-cache, not in application-dependent libraries.  We envision cache management
-being implemented in a new middle layer not in RCloud's architecture today.
+involving relatively expensive computation (for example, analysis of large
+text corpora, clustering multivariate time series or deep learning algorithms)
+caching may be essential to adequate performance. Currently we would
+program this in RCloud by explicitly saving an analysis in a persistent database,
+but it seems better to implement this capability in a general purpose associative
+cache, instead of application-specific libraries. We envision cache management
+being implemented in a new middle layer to be added in the future.
 

### 2015-03-20 discussion.tex
working

@@
 
-\subsection{Design Considerations}
+\subsection{Reflection on Design Considerations}
 
@@
 We rely on external administrative processes and social conventions
-to manage accounts and groups. Rcloud could benefit greatly from
+to manage accounts and groups. RCloud could benefit greatly from
 integrating social media to track identities and groups and to maintain
@@
 {\bf Multiple language support.}
-We did not forsee the demand to relax the assumption that programming
+We did not foresee the demand to relax the assumption that programming
 is done mainly in R. In retrospect, it is clear that many data science
@@
 Still, effective control over versioning when sharing problem needs more
-attention if Rcloud would be scaled up to a wider number of users lacking
+attention if RCloud would be scaled up to a wider number of users lacking
 informal channels to coordinate work.

### 2015-03-20 introduction.tex
yes
working

@@
 \small
-\stephen{Can we get away with \verb"\smaller"? Ha ha.}
 \input{quotes}
@@
 Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
-and sharing data from scientitic lab experiments\cite{Rubacha:ELN:2011}.
+and sharing data from scientific lab experiments\cite{Rubacha:ELN:2011}.
 In a sense, we hope to adapt and extend this concept to the work of

### 2015-03-20 relatedwork.tex
working

--- 2014-04-01 relatedwork.tex
@@
 \section{Related Work}
-
-Kandel et al. argue in their review of data wrangling work that data
-cleaning, wrangling and transformation is a major part of exploratory
-analysis and visualization~\cite{Kandel:2011:RDI}. Although RCloud
-does not by itself include modules specific to data cleaning and
-wrangling, we note from practical experience that these cleaning
-scripts themselves tend to change over time. RCloud addresses this
-issue by providing easy publishing of data-cleaning notebooks as web
-services, which reduces the total data-cleaning effort across an
-organization.
 
@@
 this cycle that cost time and effort to overcome. RCloud seeks to
-reduce this mismatch. They also point out that larger teams
+reduce this mismatch. Kandel et al also point out that larger teams
 are becoming more common in data analysis, that supporting
-collaboration is a difficult and important problem, and that sharing
+collaboration is difficult and important, and that sharing
 and versioning of data sources and artifacts is hindered by current
-technology in practice. ``We found that analysts typically did not
+technology. ``We found that analysts typically did not
 share scripts with each other. Scripts that were shared were
@@
 control.'' Their work points to the opportunity for better technology
-to support collaboration and sharing by data analysis teams.
+to support collaboration and sharing in data analysis teams.
+
+An earlier study by Kandel et al argues that data wrangling
+(cleaning, parsing and transformation)is a major part of exploratory
+analysis and visualization~\cite{Kandel:2011:RDI}. We view this
+as attacking a different semantic level than ours, but points
+to the need for an environment that enables better sharing
+of the knowledge, tools and processes to do this. Anecdotally
+we find much frustration among practitioners that this knowledge
+is difficult to find and often not recorded or provided in a reusable
+form even within one organization.
 
@@
 collaborative visual analytics~\cite{Heer:2008:DCF}.
-Notebooks, and the integrated version control system for them, described in
-Section~\ref{sec:notebooks}, address modularity and granularity, and
-artifact histories.
+RCloud notebooks, and the integrated version control system for them,
+described in Section~\ref{sec:notebooks}, address modularity and granularity,
+and artifact histories.
 \emph{Starring}, the means for signaling interest in notebooks, described in
@@
 recommendation, and voting and ranking. RCloud's integrated deployment
-mechanism, described in Section~\ref{sec:deployment}, addresses cost of
+mechanism, described in Section~\ref{sec:deployment}, addresses the cost of
 integration, content export, presentation and view sharing.
@@
 of social media with visualization and data publishing. We build on this
-work by defining a rich interface for collaboration about code and for
+work by defining a rich interface for collaboration about code, and for
 operating on metadata.
@@
 environments \cite{Guo:2010:TPI}.
-Our goal is to define an environment in which many such
+Our present goal is to define an environment in which many such
 techniques may be integrated and made available to a broad community.
@@
 provide a closer match for analysts accustomed to creating and
-executing R and Python code, while retaining many attractive
+executing R and Python code, while retaining attractive
 properties like automatic versioning and management.
@@
 for R, with support for publishing code in packages. Markdown,
-knitR and Shiny provide R with sophisticated reporting capabilities, including
+knitR and Shiny augment R with sophisticated reporting capabilities, including
 interactive web interfaces. IPython \cite{Perez:2007:IAS}
 shares many of our goals, such as providing a comprehensive environment
-for analysis and programming, and shareable documents on the web.
+for analysis and programming, and share-able documents on the web.
+\stephen{Do we now need to introduce one of those remarks where we
+explain that we are not the same as IPython.}
 
@@
 of technology in visual analytics. The fusion of development with production
-operations in software release management (``DevOps'' \cite{Httermann:2012:DD})
-is a trend in web services and related fields. By making it possible
-for data scientists to expend a little more effort in the process of
-creating experiments, it may be possible to eliminate the need for
-programming teams to recreate their results in production with
-a concomitant penalty in time and expense.
+operations in software release management (``DevOps'' \cite{Httermann:2012:DD}
+or ``continuous integration'' \cite{Fowler:2006:Continuous}) is a trend in web
+services and related fields.
+By making it convenient for data scientists to expend just a little more effort
+in the process of creating experiments, we may be able to eliminate the need
+for programming teams to recreate their work in production which has a high
+penalty in time, expense and sometimes correctness.

### 2015-03-20 introduction.tex
fix bib things

@@
 Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
-and sharing data from scientific lab experiments\cite{Rubacha:ELN:2011}.
+and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
 In a sense, we hope to adapt and extend this concept to the work of

### 2015-03-20 introduction.tex
Did something with comments from that book by Guiterrez

@@
 
-The following interview comments from
-{\it Data Scientists at Work}
-by Sebastian Gutierrez illuminate the current practice of teamwork
-and sharing in visual analytics teams.
-We particularly noted the remarks by
-Jonathan Lenaghan, who anticipates an environment for continuous
-integration and publication of data analysis.
+Interviews in {\it Data Scientists at Work} by Sebastian Gutierrez illuminate
+the teamwork and sharing in the current practice of visual analytics in industry.
+We particularly appreciated the remarks by Jonathan Lenaghan about the potential
+for continuous integration and publication.
 
-\begin{addmargin}[1em]{1em}  % left and right margin change
-\small
-\input{quotes}
-\end{addmargin}
+% \begin{addmargin}[1em]{1em}  % left and right margin change
+% part1.txt
+Erin Shellman [of Nordstrom] described the benefits of
+continuous deployment of experiments, and sharing knowledge
+through software and data analysis artifacts.
+``Finally, prototyping our products so that internal
+customers can use them early on has been crucial for
+our success. It doesn't even have to be something
+fancy. For instance, our recommendations preview
+tool doesn't have a particularly interesting
+visualization, but its enough to show the result of our
+algorithms. Now we can shoot off a URL to internal
+customers and it allows them to sit at their desk and
+understand the behavior of our product and
+experiment with it, and provide feedback way before
+we're talking about getting it into production. This
+has been super helpful and has been a really great
+way to get people excited about what we're working
+on...
+%Building and maintaining those relationships are
+%just like anything else-- you always have to be
+%working on them. Nurturing those relationships are
+%part of the job, and if you leave them unattended,
+%they might not be there later.
+%
+% part2.txt
+%Gutierrez: How do you share the knowledge you are
+%building with others?
+We’re obsessed with Confluence from
+Atlassian, and the Data Lab has a very active
+Confluence space. Recommendo is fully documented
+on Confluence, so anyone in the company could learn
+how it’s built, where to find it, and how to use it.
+We also share exploratory analyses and reports on
+Confluence so that we can still exchange knowledge
+even if the work didn’t make it into a larger project.''
+
+% part3.txt
+John Foreman commented on the scope of the work in
+data science teams.
+``I lead the data science team at
+MailChimp, and I like to get my hands dirty too.
+Some of the big pieces of my day involve working
+with my team to take stock of current projects and
+figure out where to go next, doing my own work and
+prototyping things-- I do projects just like my peers
+do and then also my talking with other teams,
+talking with management, and planning for the
+future.
+On our team, we’ve got different folks facing different
+% part4.txt
+kinds of projects. We’ve got one person who really
+owns compliance and looks at our compliance
+processes. We’ve got another data scientist who
+focuses on more of the user experience side of the
+house and understanding our customers. I help them
+and others as is needed while I do the three things
+mentioned earlier-- build data products, be a
+translator, and have conversations with the data
+science community.''
+
+% part5.txt
+Jonathan Lenaghan from PlaceIQ discussed to the potential
+for continous deployment in data science.
+``If you look at dev ops right now, they have things
+such as continuous integration, continuous build,
+automated testing, and test harnesses-- a1l of which
+map very well from the dev ops world to the data ops
+(a phrase I stole from Red Monk) world very easily. I
+think this is a very powerful notion. It is important to
+have testing frameworks for all of your data, so that if
+you make a code change, you can go back and test all
+of your data.''
+
+% part6.txt
+Anna Smith from Bitly spoke about the benefits of access to
+a rich environment of shared data and code.
+``In addition to Bitly's extraordinary data sets,
+I was given all the resources that I needed. These were
+resources that I’d only heard about but had never
+seen. I went from asking what a Hadoop cluster was,
+as my school didn't have one, to being able to work
+with one. Now I could play with one to figure out how
+to make my work better and how to change my
+algorithms so they were cleaner and ran faster on
+Hadoop. The whole transition was eye opening. Not
+only that, I could also look at other people’s code and
+play with their code and data sets as well. It was
+fantastic and I was convinced that working with data was my thing.''
+% \end{addmargin}
 

### 2015-03-20 introduction.tex
fix weird ligatures in quote document

@@
 Confluence so that we can still exchange knowledge
-even if the work didn’t make it into a larger project.''
+even if the work didn't make it into a larger project.''
 
@@
 future.
-On our team, we’ve got different folks facing different
+On our team, we've got different folks facing different
 % part4.txt
-kinds of projects. We’ve got one person who really
+kinds of projects. We've got one person who really
 owns compliance and looks at our compliance
-processes. We’ve got another data scientist who
+processes. We've got another data scientist who
 focuses on more of the user experience side of the
@@
 Jonathan Lenaghan from PlaceIQ discussed to the potential
-for continous deployment in data science.
+for continuous deployment in data science.
 ``If you look at dev ops right now, they have things

### 2015-03-20 conclusion.tex
fix

--- 2015-03-20 conclusion.tex
@@
 
-We presented a case for environments that support the visual
-analytics process, from data acquisition and exploratory
-data analysis, through development and deployment.
-The environment supports collaboration and communication
-through shared experimental workbooks with version control,
-searching, recommending, annotating, and publishing
+We presented a case for a data analytics environment that
+supports the visual analytics process, from data acquisition
+and exploratory data analysis, through development and deployment.
+Building on the extensive previous work on requirements for
+visual analytics systems, we designed and deployed a prototype
+environment, RCloud.  RCloud supports collaboration and communication
+through shared experimental workbooks with version
+control, searching, recommending, annotating, and publishing
 experiments as web sites and as reusable services.
 
-We implemented these ideas in the RCloud prototype.
-Experience with the prototype provides evidence that data
-science teams and the organizations in which they work
-benefit from having such capabilities.
+Our purpose was to gain experience and insight by implementing and
+deploying a prototype with a community of working data scientists.
+Experience with the RCloud prototype provides evidence that data
+science teams and the organizations in which they work benefit from
+having capabilities that support the broad visual analytics process.
 
-This work prototype a step toward practical ``DevOps for data
-science'' and offers reproducible experiments.
+Rcloud is a step toward practical ``DevOps for data science'' and
+reproducible, publishable experiments.
+
 Possible next steps are to support cross-language development,
@@
 
-The code is available 
+RCloud code is available 
 at \url{github.com/att/rcloud/}

### 2015-03-20 conclusion.tex
fixed up conclusions

@@
 We presented a case for a data analytics environment that
-supports the visual analytics process, from data acquisition
-and exploratory data analysis, through development and deployment.
-Building on the extensive previous work on requirements for
-visual analytics systems, we designed and deployed a prototype
-environment, RCloud.  RCloud supports collaboration and communication
+supports the visual analytics process for work by teams.
+This inclides data acquisition, exploratory data analysis,
+code development and deployment.
+
+Building on previous work that defines requirements for
+visual analytics, we designed a prototype environment, RCloud.
+RCloud supports collaboration and communication
 through shared experimental workbooks with version
@@
 
-Our purpose was to gain experience and insight by implementing and
-deploying a prototype with a community of working data scientists.
+We deployed RCloud in a community of working data scientists.
 Experience with the RCloud prototype provides evidence that data
 science teams and the organizations in which they work benefit from
-having capabilities that support the broad visual analytics process.
+having capabilities that support collaboration and integrate
+the entire visual analytics process.
 
@@
 
-RCloud code is available 
-at \url{github.com/att/rcloud/}
+RCloud code is available at \url{github.com/att/rcloud/}
 under an MIT open source license.

### 2015-03-21 system.tex
i was working

@@
 
-\stephen{Let's work from front to back, from human interface to the
-services and frameworks that support it.}
+% Stephen: Let's work from front to back, from human interface to the
+% services and frameworks that support it.
 
-\stephen{I don't think we need to even make this argument, it seems so obvious.}
 The internal computing infrastructure of organizations has changed
@@
 led to a software ecosystem of processes distributed over a
-network (often virtually defined) communicating via some
-high-level protocol, usually HTTP. HTTP is dominant because web
-browsers and servers are ubiquitous and available in almost
+network, usually communicating via HTTP. HTTP is dominant because
+web browsers and servers are ubiquitous and available in almost
 all hardware devices, from tiny sensors, to handheld devices and
@@
 The call mechanism is the URL itself; arguments are read from the URL
-and the final cell is rendered as the result.  \stephen{That needs
-more explanation.}
+and the final cell is rendered as the result.
 

### 2015-03-22 interviews.tex
cram in a ton of quotes - completely unedited and without narrative

\section{Interview Study\label{sec:interviews}}

To evaluate the effectiveness of RCloud, we interviewed 13 current and past users of RCloud. Of these, 9 are adata analysts, and 4 build tools for data analysts and business needs.

\gordon{Add user statistics here: so many users use the Research instance a lot, so many use it an intermediate amount, so many just tried it out. Simon to work up numbers on this. Also will get stats on number of notebooks, forks, stars, from Prateek's notebook.}

\subsection{Sharing of results}
Rick: ``I like being able to look at other people's notebooks.''

Clint: ``I can share it with people and copy it pretty readily. So if my supervisor wants to see what I've done or QA it, I can just send her a link.'' ``I think the best part about it is how easily you can share code... you can find a working example, rather than wearing out Google and finding questionable examples that may or may not work.''

Corey: ``being able to share notebooks or codes, or collaborate on something, easily being able to share that information. That's an obvious advantage and nice to have.''

Raif: ``I've been sharing my notebooks in order for other people to see what I've done. It's very convenient for that purpose, I would say'' ``we haven't been editing them together, but we've been just sharing, and then I've been looking at these things that other people do. So there's no collaborative aspect of it, like developing the code, but there's like sharing aspects for other people to see and for me to see other people's notebooks'' ``the fact that we have all the notebooks there, searchable, it just saves me from replicating what other people have done'' ``there already is some person who has done something similar, and then you're able to just edit that, and that's saved a lot of work time for me''

Shelly: Is unable to use RCloud for organizational reasons but likes the ``concept of being able to create notebooks and share them'' ``wiki is not the best way for communication results - it's kind of like writing a blog post with very limited functionality. can't input equations easily.'' ``have to save every picture and post it as an image'' ``I can't share all of the code because it would just get crowded and wouldn't look right on a wiki''

Bart: Also unable to use for org/tech reasons. ``So if I could make a folder on RCloud and have Python notebooks and also Pig notebooks there, and execute them from RCloud, that would be much better than my current thing, because that would free me from manual documentation and version control and also telling people where my code was. It would be just, hey, go look on RCloud, here's the stuff.''

Horace: ``It's not for myself, but for other people who use RCloud vs RStudio, they will always think that they want to prototype on RStudio. Maybe because it's more familiar for them. So, they work on RStudio, on a small data set, and they fix up everything, make sure everything is right, and then they place it on RCloud for sharing.''

Taylor: ``I use it more for sharing code with other people, and for doing tutorials for iotools or hmr or explaining'' ``a lot of times I'll share the code, because I'm doing so much package development. I want people to see how the package works, so I clearly want them to see the code… I kind of write it just like I would write GitHub Markdown, where you have little code snippets and text, but RCloud lets me actually run the snippets” with results''

\subsection{Forking}
Rick: ``I fork my own notebooks because I'm going off and doing some other analogous project, so I've sort of got interesting content that I've already done in a previous analysis that I want to start from and then tweak to match a new set of data.''

Clint: ``I use [forking] on my own work quite a bit, and I use it on Research a ton. Honestly, it's one of the handiest functions in it, because instead of having to find it, copy, paste it, you just hit Fork, you rename it, and it's done. It's pretty amazing.''

Raif: ``I've been forking other people's notebooks myself, because essentially they have done some work, but then I want to run it on a different part of data, or I want to change some parts, I don't want to see this column, I want a different column, things like that''

Shelly: ``gives the audience a way to try their own analysis, like maybe Shelly should have used this parameter and they can just change the parameters''

Horace: on working with other group using his notebooks, ``I basically point them to a few different parameters that they need to fix.'' ``I'll teach people to intercept the result in the middle and see what you should see, and what's not being returned, that sort of thing'' ``If the computation went wrong in the middle, and something blew up, or you're just returning NULL or empty data, where should we check those things?'' ``We should, but a lot of the time we just insert print statements here and there and check values. Having debugging code inside a functioning program is actually helpful for others to understand what you're doing?''

\subsection{Automatic source control}
Rick: ``when I've finished something, there's a nice clean record of it.''

Clint: ``instead of looking back and saying I've got a billion files here in this subdirectory and I hope I've got them backed up, if they're on RCloud I know they are'' ``I don't have to worry about it if I drop my computer in the bathtub, everythings gone, because I'm not great about uploading my own for versioning, right?''

Bart: ``I like the fact that it has a built-in editor, so if you need to, like, fixing a typo in a link, or an extra line break or any of this other nonsense, you can just switch to the edit view, pull up your asset, type something, it's automatically saved, committed, everything. You don't have to go back to your source code, change it, commit it to the repo, you pull the repo to your distribution version, none of that's necessary. A lot of web development is fixing stupid little bugs, and you can do that kind of instantly without a lot of overhead, which is nice.''

\subsection{Web integration}
Horace ``that also means the other guys who want to do analytics on the data, they can first pull the data, do the analytics on it, and then feed the viewer the data. So if anyone from the stats group wants to insert some code in between the data pulling procedure - now it's just database to GUI, and you can always insert something in between. We were trying to make such an example so that they don't have to rewrite the GUI part and they know how to get the data source, once you get the data into RCloud, then you have a dataframe to work with, and then they know how to produce another dataframe. We give them the specification of the dataframes, we give them the data input, they do whatever they want, they can reuse the data connection part and the GUI part. We can keep changing these components and that's a very nice thing.''

Jyothi: ``Having an open session with R certainly helps too, because both of my applications need to be doing something with the data, so having an open session where I can run R command or R functions without having to invoke an API or send out a request and then wait for the request to come back to do the results.  So that kind of seamless integration is extremely helpful in writing the application.''

Jyothi: ``another thing I really like about RCloud is your publish feature, so that non-RCloud users, the anonymous login, non-RCloud users can access the notebooks without having to login to RCloud''

accidental feature: ``some of my jobs are huge and I like to just call them and the fact that the user can just shut down his browser or shut down his computer and the job still runs in the background is actually a very neat thing. So what, um, the way I have coded is, whenever I have a huge job to run, I just display to the user, you go do whatever you want, we'll send you an email when the job is done, um that's another feature of RCloud that's very helpful, that the job will continue to run in the background''

\subsection{Proliferation of notebook cells}
Rick: ``Making new cells and taking some stuff in it, then having it do something, then having to make a new cell, and I don't know, it just gets in my way, so if I'm doing really straightforward simple stuff, I'll just get into R and do it myself.'' ``I'm always scrolling up and down the screen'' ``every time I type more stuff, the notebook gets longer and longer and it's harder to deal with'' ``I've got to do a lot of scrolling around to get to where I'm going, whereas if I'm doing it in vi, I can search for things in my typing, and get to places I want to be without having to scroll through it and try to read it'' ``times when I just want to type in a couple of quick and get some results that are going to tell me what to do next, and they're not necessarily archival in any sense'' ``being literate about anything typically takes a lot of rewriting and going back over things... I just want to explore a few things and then I'll know what I want to write.'' ``I'll often type in expressions that allow me to check that I'm the right track, so I know that if the computation worked out, then the sum of the residuals is gonna be zero. So I'll type in sum(resid) and if it prints out zero, then I'm all happy, but I don't want that in my notebook.'' ``it tempts me to type a big long thing and then run the whole thing, as opposed to, type a few little pieces that I'm thinking of on the inner parts and then put them together to make the big thing''

Ivan has the complaint that too many cells are created to use the notebook as a console.  This has a lot to do with the extra real estate that gets taken up by the cells controls, blank space, etc of cells versus a straight command line.

KC and Ivan both complained that when there are long results or plots, it causes the code to scroll off the screen.

Kenny: ``It saves everything I do like everything is gold, but most of it is junk not meant to be saved'' ``the console is purely disposable''

Raif: ``cells are really useful'' but ``whatever you type on the command line, it becomes a part of your notebook, and I find it a little bit annoying, because I really type a lot of things to check my data, and I don't want that to become a part of the notebook'' ``Like you're kind of checking out what your data is, or you make a plot of the data. Things that should not really become part of a notebook, but things that help you understand your data better.'' ``for example imagine I run a cell whose output is huge, right, and now that output would block the whole window, and in order to find the next cell, I have to scroll down, and find where that cell starts, right? So, I'm losing the continuity of my code to the output''

Horace: ``Well, the nice thing is if you have the whole notebook, you can run it step by step, and try to mess around in between, interactively.''

Rick on auditing (possible solution): ``you didn't see your mistakes, they were in there but you never really saw them because a mistake never led to an answer at the end'' ``the real result of this analysis was these three plots, so go back and figure out everything that I did that was involved in creating those three plots. So that I could start from the raw data and create those three plots.''

\subsection{Long-running cells}
Kenny: Problem of long-running cells.  ``The long tail.''  First cell takes seconds to run, next cell minutes, next cell 5 hours.  In this scenario, which is common for Kenny, the ``Run whole notebook'' button is dangerous. This is also a huge issue for demoing code.  Suggests something like knitr's caching of results. When converting his work to notebooks, Kenny ends up with a lot of comments that say ``this cell takes a long time to run.'' He ends up ``littering the notebook with little switches that comment out'' the slow parts and ``load the object instead'', ``save the object to disk''.

\subsection{Misc. notebook interface annoyance}
Rick: ``if I was typing into one of the cells near that top, I had to think of `I'm editing this cell and then I have to execute it', and if was doing the one at the bottom I could type it but then it would automatically execute, but then it became a standard cell and I have to edit it. It was just another set of modes that I didn't like''

Clint: ``it took me a couple of weeks of looking at it to become comfortable because I saw the cells and it just threw me for a total loop. I mean, it's a good idea because it's essentially like I can have this part and this part and this part. I can run it all at once if I want or I can break these into sections for either debugging or staging purposes which, I really like it now, but when I first saw it, [it was] very confusing... I think I prefer it now, because you can cut your code up.''

Corey: In the previous version, it switched between code and results. Now you see both - ``and there are times when that's good, but there's certainly times when it'd be nice to - you know, you have 10 or 15 cells and you just want to get back and there's all this output and you're scrolling all over the place so, if there were a way to control both, that'd be great. There are times when I do want to see the output, but there are times when I've seen enough of the same thing.''

Jyothi: ``It's good that I can see the results but if I'm working and I need more room and I want to hide the results, I can't.” Could we have “an option to close the results window?”

Ivan: RStudio's layout is more helpful because
- the charts are kept in another pane and stay in one place while doing analysis
- having ``long term'' code (which would be in assets) above the command prompt means that both can be sized very wide to accomodate long lines

Kenny: Problem with screen real estate: side panes take up too much space. Doesn't need the notebook tree pane.

KC thought that having a scratch pad that doesn't persist between different notebooks is 
``useless''.

\subsection{Other workflows}
Rick: ``I find that the style of editing is clunky. I would never edit like that if I'm doing ordinary work. I usually use vi.'' ``I mean it's nice to have it saved, but you know, there's sort of this trade-off between it making it easier for me to present something or to save something, and my easy of typing and correcting and things in a plain old editor window.'' ``I just have the discipline of having a file that describes what I'm doing with the commands that I'm using so that I can go back and recreate it or pass it on to someone else, so it's a little bit like having an RCloud notebook, it's not necessarily executable but most of the commands that I've typed are in there.''

Kenny: the UI doesn't work for his workflow, which is pasting snippets from a file into the console in the R GUI.  he tends to type type type into the file, paste paste paste to the console, rearrange so the good code is at the top of the file, then save the file and call it quits once he's got the right code at the top of the file.

Taylor: ``The big one is I don't know why would I want to use RCloud over my current setup. If it's just me, I like my text editor and terminal. There's nothing that I want that those two don't give me'' ``In R, when I'm doing development work, I copy and paste in like 5-10 lines of code, so when something breaks, I get an error message on that one line, and I can up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell, so the only way to get that same functionality is if every line's in one cell. Because if it runs like lines 1,2,3, then 4 fails, I have to either run the whole cell again, or have line 4 in its own cell. I can't rerun just one line.'' ``There's nothing that I want that's not in the terminal. I don't think you should try to support all the things I want to do in the terminal.''

Byers: installing random packages is part of exploratory data analysis

\subsection{Version management difficulties}
Rick: ``I don't need something that keeps track of every mistake I've made or every direction I've tried.” ``namespaces are already crowded, so trying to remember the names of notebooks is hard, much less the names of states within each notebook.''

\subsection{A sea of notebooks}

Byers: Problem with forking a notebook and correcting it, is that the original notebook still exists with the error.  We need ways to see that there is a more recent fork of notebook and to fetch in the changes from it.

Taylor: ``I'll make a notebook of a little tutorial of, here's the basic functions and what they do, and then I'll send them out, and then people will give me feedback on them, or have some new feature, or I'll realize something's wrong, and then I'll go back and change it and update that page, but then what happens is people have in the meantime, forked it, and then they have a version that… and usually what it'll be is actually that I'll have to change a package as well, so their old fork stops working, and then they complain, and then, oh no, you have to go get the original [notebook]. Yeah, that's the sort of.. This probably bites me more than other people, because I'm doing development so things are changing as they're looking at the code. Like, iotools, a lot of the functions change names or change syntax.''

Taylor: ``That's just a problem because people will have old notebooks that don't work anymore. It's annoying when it's a notebook that I still have and the format has changed just a bit. More than that is when people have, now that I know that you can't even delete notebooks, they'll have a fork of something that doesn't exist anymore on my thing.'' ``Now I'm really gun-shy about sharing notebooks because, I don't want to like, it's like, do I want to support this forever?''

Taylor: ``A good example is, someone was complaining about doing forking in an RCloud notebook. They were trying to use a parallel package. And I said, no no I think it works, and I did like a one-cell thing that just did some sweep kind of thing to show that it was actually working in parallel, and then I just deleted it, thinking it was gone. And then it saved something in a directory that I then deleted, because it was just a temporary directory. Three months later, I don't know why they were running it, they said `oh it doesn't work'. I think that's the biggest issue I'm having right now, all these forks''

Taylor: ``there's no way for both of us to have ownership of a notebook, so the only way is to fork it back and forth, and so we have dozens of old copies. [We end up] deleting all the old ones, but people still have links to them, because they don't actually disappear.'' ``The UI makes it easy to go back to the old stale versions, because you can say `forked from' `forked from' `forked from', but there's actually no way to go the way that we want people to go, which is, like, up the tree.''

Rick: ``I don't necessarily need to see everybody's notebook that uses RCloud'' ``every time I do something new, I get a new notebook and so now my notebooks are maybe 50 or 60. That's enough to think about just on my own, but if I've got everybody's 50 or 60 sitting on my display, I find that it's more than I want to know about.''

\subsection{Parameterization}
Clint: ``For the things that I'll do multiple times, I'll try to parameterize it, and I'm working now on putting some packages together that will operate in some fashion outside of RCloud, straight from the command line''

Taylor: ``like the top of it would be, `This is a report of the volume of all of our feeds for this month', and someone would want to look at it for the next month or the previous month, so they'd fork it to change the month''

Jyothi: ``I think that if a notebook is web-enabled [parameterized] then it gives the person who is running the notebook, I mean they can always fork the notebooks and make changes, but I feel that if the owner of the notebook web-enables it, then whoever is running it, it's easier to run, instead of forking it, if they can set options, it's probably more efficient, and also they can fork it if they want to.''

\subsection{Web snafus}
Disconnection, slowness

Corey: ``“What frustrates me is that if I do leave this window for what feels like 5 or 7 minutes, it disconnects and I have to reconnect and run the whole notebook again.''

Kenny: it really sucks when you get disconnected

Raif: ``being web-based, it's obviously slower than something you would be running on your machine''

Byers: even the little lag of sending a command to a server and receiving the result is intolerable

\subsection{Other}
Byers: There is also a big problem with ``bitrot'' - notebooks often stop working because the user changed the structure of their data, changed a filename or a database.  We need ``organizational protocols'' to catch up with the technology.  There is a ``hysteresis between the technical and the social'' and we have not yet adapted.

Taylor: `` It would be a file inside a directory that only I had read access to and so someone else couldn't run it... That was usually what it was, it was a temporary file on something that was read-only to everyone else. But I've gotten careful about that.''

Bart: ``One UI notebook and a bunch of other notebooks that do various tasks and are called with API calls. The problem is, if you clone all the notebooks, then you have to go update every single notebook, because all the IDs are different now. There's no sort of relative path type stuff that you can do. You can do it with static assets, but you can't do it with Ajax calls''

### 2015-03-22 interviews.tex
organize based on feature/complaint grid

@@
 It's very convenient for that purpose, I would say'' ``we haven't been editing them together, but we've been just sharing, and then I've been looking at these things that other people do.
-So there's no collaborative aspect of it, like developing the code, but there's like sharing aspects for other people to see and for me to see other people's notebooks'' ``the fact that we have all the notebooks there, searchable, it just saves me from replicating what other people have done'' ``there already is some person who has done something similar, and then you're able to just edit that, and that's saved a lot of work time for me''
+So there's no collaborative aspect of it, like developing the code, but there's like sharing aspects for other people to see and for me to see other people's notebooks'' ``there already is some person who has done something similar, and then you're able to just edit that, and that's saved a lot of work time for me''
+
+Taylor: ``I use it more for sharing code with other people, and for doing tutorials for iotools or hmr or explaining'' ``a lot of times I'll share the code, because I'm doing so much package development.
+I want people to see how the package works, so I clearly want them to see the code… I kind of write it just like I would write GitHub Markdown, where you have little code snippets and text, but RCloud lets me actually run the snippets” with results''
 
@@
 It would be just, hey, go look on RCloud, here's the stuff.''
-
-Horace: ``It's not for myself, but for other people who use RCloud vs RStudio, they will always think that they want to prototype on RStudio.
-Maybe because it's more familiar for them.
-So, they work on RStudio, on a small data set, and they fix up everything, make sure everything is right, and then they place it on RCloud for sharing.''
-
-Taylor: ``I use it more for sharing code with other people, and for doing tutorials for iotools or hmr or explaining'' ``a lot of times I'll share the code, because I'm doing so much package development.
-I want people to see how the package works, so I clearly want them to see the code… I kind of write it just like I would write GitHub Markdown, where you have little code snippets and text, but RCloud lets me actually run the snippets” with results''
 
@@
 
-Shelly: ``gives the audience a way to try their own analysis, like maybe Shelly should have used this parameter and they can just change the parameters''
-
 Horace: on working with other group using his notebooks, ``I basically point them to a few different parameters that they need to fix.'' ``I'll teach people to intercept the result in the middle and see what you should see, and what's not being returned, that sort of thing'' ``If the computation went wrong in the middle, and something blew up, or you're just returning NULL or empty data, where should we check those things?'' ``We should, but a lot of the time we just insert print statements here and there and check values.
 Having debugging code inside a functioning program is actually helpful for others to understand what you're doing?''
+
+Shelly: ``gives the audience a way to try their own analysis, like maybe Shelly should have used this parameter and they can just change the parameters''
 
@@
 
-Clint: ``instead of looking back and saying I've got a billion files here in this subdirectory and I hope I've got them backed up, if they're on RCloud I know they are'' ``I don't have to worry about it if I drop my computer in the bathtub, everythings gone, because I'm not great about uploading my own for versioning, right?''
+Clint: ``instead of looking back and saying I've got a billion files here in this subdirectory and I hope I've got them backed up, if they're on RCloud I know they are'' ``I don't have to worry about it if I drop my computer in the bathtub, everythings gone, because I'm not great about uploading my own for versioning''
 
@@
 
+\subsection{Search}
+Raif: ``the fact that we have all the notebooks there, searchable, it just saves me from replicating what other people have done'' ``One way of thinking of it is like Google Code, you know you just search for something and you find something useful and you just use it.
+It’s a repository of code''
+
+Rick: ``And the same kind of thing when you search through notebooks, usually I know somebody’s notebooks that I want to search through, like I’ll search through Simon Urbanek’s first, then I’ll go off to somebody else’s, because I know that the kind of thing I’m looking for something more obscure than I’m likely to find in some random person’s thing''
+
 \subsection{Web integration}
 Horace ``that also means the other guys who want to do analytics on the data, they can first pull the data, do the analytics on it, and then feed the viewer the data.
-So if anyone from the stats group wants to insert some code in between the data pulling procedure - now it's just database to GUI, and you can always insert something in between.
+So if anyone from the stats group wants to insert some code in between the data pulling procedure, [they] can always insert something in between.
 We were trying to make such an example so that they don't have to rewrite the GUI part and they know how to get the data source, once you get the data into RCloud, then you have a dataframe to work with, and then they know how to produce another dataframe.
@@
 So what, um, the way I have coded is, whenever I have a huge job to run, I just display to the user, you go do whatever you want, we'll send you an email when the job is done, um that's another feature of RCloud that's very helpful, that the job will continue to run in the background''
+
+\subsection{Other workflows}
+Rick: ``I find that the style of editing is clunky.
+I would never edit like that if I'm doing ordinary work.
+I usually use vi.'' ``I mean it's nice to have it saved, but you know, there's sort of this trade-off between it making it easier for me to present something or to save something, and my easy of typing and correcting and things in a plain old editor window.'' ``I just have the discipline of having a file that describes what I'm doing with the commands that I'm using so that I can go back and recreate it or pass it on to someone else, so it's a little bit like having an RCloud notebook, it's not necessarily executable but most of the commands that I've typed are in there.''
+
+Kenny: the UI doesn't work for his workflow, which is pasting snippets from a file into the console in the R GUI.
+he tends to type type type into the file, paste paste paste to the console, rearrange so the good code is at the top of the file, then save the file and call it quits once he's got the right code at the top of the file.
+
+Taylor: ``The big one is I don't know why would I want to use RCloud over my current setup.
+If it's just me, I like my text editor and terminal.
+There's nothing that I want that those two don't give me'' ``In R, when I'm doing development work, I copy and paste in like 5-10 lines of code, so when something breaks, I get an error message on that one line, and I can up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell, so the only way to get that same functionality is if every line's in one cell.
+Because if it runs like lines 1,2,3, then 4 fails, I have to either run the whole cell again, or have line 4 in its own cell.
+I can't rerun just one line.'' ``There's nothing that I want that's not in the terminal.
+I don't think you should try to support all the things I want to do in the terminal.''
+
+Horace: ``It's not for myself, but for other people who use RCloud vs RStudio, they will always think that they want to prototype on RStudio.
+Maybe because it's more familiar for them.
+So, they work on RStudio, on a small data set, and they fix up everything, make sure everything is right, and then they place it on RCloud for sharing.''
+
+Byers: installing random packages is part of exploratory data analysis
 
@@
 
-Raif: ``cells are really useful'' but ``whatever you type on the command line, it becomes a part of your notebook, and I find it a little bit annoying, because I really type a lot of things to check my data, and I don't want that to become a part of the notebook'' ``Like you're kind of checking out what your data is, or you make a plot of the data.
-Things that should not really become part of a notebook, but things that help you understand your data better.'' ``for example imagine I run a cell whose output is huge, right, and now that output would block the whole window, and in order to find the next cell, I have to scroll down, and find where that cell starts, right?
-So, I'm losing the continuity of my code to the output''
+Raif: ``whatever you type on the command line, it becomes a part of your notebook, and I find it a little bit annoying, because I really type a lot of things to check my data, and I don't want that to become a part of the notebook'' ``Like you're kind of checking out what your data is, or you make a plot of the data.
+Things that should not really become part of a notebook, but things that help you understand your data better.''
 
@@
 So that I could start from the raw data and create those three plots.''
-
-\subsection{Long-running cells}
-Kenny: Problem of long-running cells.
-``The long tail.''  First cell takes seconds to run, next cell minutes, next cell 5 hours.
-In this scenario, which is common for Kenny, the ``Run whole notebook'' button is dangerous.
-This is also a huge issue for demoing code.
-Suggests something like knitr's caching of results.
-When converting his work to notebooks, Kenny ends up with a lot of comments that say ``this cell takes a long time to run.'' He ends up ``littering the notebook with little switches that comment out'' the slow parts and ``load the object instead'', ``save the object to disk''.
 
@@
 
+Raif: ``cells are really useful'' but ``you want to see your output in a different window, on a separate part of the window'' ``for example imagine I run a cell whose output is huge, right, and now that output would block the whole window, and in order to find the next cell, I have to scroll down, and find where that cell starts, right?
+So, I'm losing the continuity of my code to the output''
+
 Kenny: Problem with screen real estate: side panes take up too much space. Doesn't need the notebook tree pane.
 
-KC thought that having a scratch pad that doesn't persist between different notebooks is 
-``useless''.
+KC thought that having a scratch pad that doesn't persist between different notebooks is ``useless''.
 
-\subsection{Other workflows}
-Rick: ``I find that the style of editing is clunky.
-I would never edit like that if I'm doing ordinary work.
-I usually use vi.'' ``I mean it's nice to have it saved, but you know, there's sort of this trade-off between it making it easier for me to present something or to save something, and my easy of typing and correcting and things in a plain old editor window.'' ``I just have the discipline of having a file that describes what I'm doing with the commands that I'm using so that I can go back and recreate it or pass it on to someone else, so it's a little bit like having an RCloud notebook, it's not necessarily executable but most of the commands that I've typed are in there.''
-
-Kenny: the UI doesn't work for his workflow, which is pasting snippets from a file into the console in the R GUI.
-he tends to type type type into the file, paste paste paste to the console, rearrange so the good code is at the top of the file, then save the file and call it quits once he's got the right code at the top of the file.
-
-Taylor: ``The big one is I don't know why would I want to use RCloud over my current setup.
-If it's just me, I like my text editor and terminal.
-There's nothing that I want that those two don't give me'' ``In R, when I'm doing development work, I copy and paste in like 5-10 lines of code, so when something breaks, I get an error message on that one line, and I can up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell, so the only way to get that same functionality is if every line's in one cell.
-Because if it runs like lines 1,2,3, then 4 fails, I have to either run the whole cell again, or have line 4 in its own cell.
-I can't rerun just one line.'' ``There's nothing that I want that's not in the terminal.
-I don't think you should try to support all the things I want to do in the terminal.''
-
-Byers: installing random packages is part of exploratory data analysis
-
-\subsection{Version management difficulties}
-Rick: ``I don't need something that keeps track of every mistake I've made or every direction I've tried.” ``namespaces are already crowded, so trying to remember the names of notebooks is hard, much less the names of states within each notebook.''
+Kenny: Problem of long-running cells.
+``The long tail.''  First cell takes seconds to run, next cell minutes, next cell 5 hours.
+In this scenario, which is common for Kenny, the ``Run whole notebook'' button is dangerous.
+This is also a huge issue for demoing code.
+Suggests something like knitr's caching of results.
+When converting his work to notebooks, Kenny ends up with a lot of comments that say ``this cell takes a long time to run.'' He ends up ``littering the notebook with little switches that comment out'' the slow parts and ``load the object instead'', ``save the object to disk''.
 
@@
 
-\subsection{Parameterization}
-Clint: ``For the things that I'll do multiple times, I'll try to parameterize it, and I'm working now on putting some packages together that will operate in some fashion outside of RCloud, straight from the command line''
-
-Taylor: ``like the top of it would be, `This is a report of the volume of all of our feeds for this month', and someone would want to look at it for the next month or the previous month, so they'd fork it to change the month''
-
-Jyothi: ``I think that if a notebook is web-enabled [parameterized] then it gives the person who is running the notebook, I mean they can always fork the notebooks and make changes, but I feel that if the owner of the notebook web-enables it, then whoever is running it, it's easier to run, instead of forking it, if they can set options, it's probably more efficient, and also they can fork it if they want to.''
+\subsection{Versioning not useful}
+Rick: ``I don't need something that keeps track of every mistake I've made or every direction I've tried.” ``namespaces are already crowded, so trying to remember the names of notebooks is hard, much less the names of states within each notebook.''
 
 \subsection{Web snafus}
-Disconnection, slowness
-
 Corey: ``“What frustrates me is that if I do leave this window for what feels like 5 or 7 minutes, it disconnects and I have to reconnect and run the whole notebook again.''
@@
 
-\subsection{Other}
+\subsection{Parameterization}
+Clint: ``For the things that I'll do multiple times, I'll try to parameterize it, and I'm working now on putting some packages together that will operate in some fashion outside of RCloud, straight from the command line''
+
+Taylor: ``like the top of it would be, `This is a report of the volume of all of our feeds for this month', and someone would want to look at it for the next month or the previous month, so they'd fork it to change the month''
+
+Jyothi: ``I think that if a notebook is web-enabled [parameterized] then it gives the person who is running the notebook, I mean they can always fork the notebooks and make changes, but I feel that if the owner of the notebook web-enables it, then whoever is running it, it's easier to run, instead of forking it, if they can set options, it's probably more efficient, and also they can fork it if they want to.''
+
+\subsection{Other observations}
 Byers: There is also a big problem with ``bitrot'' - notebooks often stop working because the user changed the structure of their data, changed a filename or a database.

### 2015-03-23 interviews.tex
start interview prose and wrap lines

@@
 
-To evaluate the effectiveness of RCloud, we interviewed 13 current and past users of RCloud. Of these, 9 are adata analysts, and 4 build tools for data analysts and business needs.
+To evaluate the effectiveness of RCloud, we interviewed 13 current and past
+users of RCloud. Of these, 9 are adata analysts, and 4 build tools for data
+analysts and business needs.
 
-\gordon{Add user statistics here: so many users use the Research instance a lot, so many use it an intermediate amount, so many just tried it out.
-Simon to work up numbers on this.
-Also will get stats on number of notebooks, forks, stars, from Prateek's notebook.}
+\gordon{Add user statistics here: number of notebooks, forks, stars, users
+who have done one of these things. Number of users who use RCloud a lot,
+a moderate amount, or just tried it out.}
 
 \subsection{Sharing of results}
+Sharing of notebooks is the core feature of RCloud, and a popular one. Almost
+everyone we spoke to praised this feature. By default, all notebooks are
+publicly visible in RCloud, and notebooks can be found through the notebook tree
+or search.
+
+However, the method for sharing that users mentioned most is by sending links
+through email. Clint says, ``I can share it with people and copy it pretty
+readily. So if my supervisor wants to see what I've done or QA it, I can just
+send her a link.''
+
+``I think the best part about it is how easily you can share code... you can
+find a working example, rather than wearing out Google and finding questionable
+examples that may or may not work.''
+
+Corey: ``being able to share notebooks or codes, or collaborate on something,
+easily being able to share that information. That's an obvious advantage and
+nice to have.''
+
+Raif: ``I've been sharing my notebooks in order for other people to see what
+I've done. It's very convenient for that purpose, I would say'' ``we haven't
+been editing them together, but we've been just sharing, and then I've been
+looking at these things that other people do. So there's no collaborative aspect
+of it, like developing the code, but there's like sharing aspects for other
+people to see and for me to see other people's notebooks'' ``there already is
+some person who has done something similar, and then you're able to just edit
+that, and that's saved a lot of work time for me''
+
+Taylor: ``I use it more for sharing code with other people, and for doing
+tutorials for iotools or hmr or explaining'' ``a lot of times I'll share the
+code, because I'm doing so much package development. I want people to see how
+the package works, so I clearly want them to see the code… I kind of write it
+just like I would write GitHub Markdown, where you have little code snippets and
+text, but RCloud lets me actually run the snippets'' with results.
+
+Shelly: Is unable to use RCloud for organizational reasons but likes the
+``concept of being able to create notebooks and share them'' ``wiki is not the
+best way for communication results - it's kind of like writing a blog post with
+very limited functionality. can't input equations easily.'' ``have to save every
+picture and post it as an image'' ``I can't share all of the code because it
+would just get crowded and wouldn't look right on a wiki''
+
+Bart: Also unable to use for org/tech reasons. ``So if I could make a folder on
+RCloud and have Python notebooks and also Pig notebooks there, and execute them
+from RCloud, that would be much better than my current thing, because that would
+free me from manual documentation and version control and also telling people
+where my code was. It would be just, hey, go look on RCloud, here's the stuff.''
+
 Rick: ``I like being able to look at other people's notebooks.''
 
-Clint: ``I can share it with people and copy it pretty readily.
-So if my supervisor wants to see what I've done or QA it, I can just send her a link.'' ``I think the best part about it is how easily you can share code...
-you can find a working example, rather than wearing out Google and finding questionable examples that may or may not work.''
+\subsection{Forking}
+Rick: ``I fork my own notebooks because I'm going off and doing some other
+analogous project, so I've sort of got interesting content that I've already
+done in a previous analysis that I want to start from and then tweak to match a
+new set of data.''
 
-Corey: ``being able to share notebooks or codes, or collaborate on something, easily being able to share that information. That's an obvious advantage and nice to have.''
+Clint: ``I use [forking] on my own work quite a bit, and I use it on Research a
+ton. Honestly, it's one of the handiest functions in it, because instead of
+having to find it, copy, paste it, you just hit Fork, you rename it, and it's
+done. It's pretty amazing.''
 
-Raif: ``I've been sharing my notebooks in order for other people to see what I've done.
-It's very convenient for that purpose, I would say'' ``we haven't been editing them together, but we've been just sharing, and then I've been looking at these things that other people do.
-So there's no collaborative aspect of it, like developing the code, but there's like sharing aspects for other people to see and for me to see other people's notebooks'' ``there already is some person who has done something similar, and then you're able to just edit that, and that's saved a lot of work time for me''
+Raif: ``I've been forking other people's notebooks myself, because essentially
+they have done some work, but then I want to run it on a different part of data,
+or I want to change some parts, I don't want to see this column, I want a
+different column, things like that''
 
-Taylor: ``I use it more for sharing code with other people, and for doing tutorials for iotools or hmr or explaining'' ``a lot of times I'll share the code, because I'm doing so much package development.
-I want people to see how the package works, so I clearly want them to see the code… I kind of write it just like I would write GitHub Markdown, where you have little code snippets and text, but RCloud lets me actually run the snippets” with results''
+Horace: on working with other group using his notebooks, ``I basically point
+them to a few different parameters that they need to fix.'' ``I'll teach people
+to intercept the result in the middle and see what you should see, and what's
+not being returned, that sort of thing'' ``If the computation went wrong in the
+middle, and something blew up, or you're just returning NULL or empty data,
+where should we check those things?'' ``We should, but a lot of the time we just
+insert print statements here and there and check values. Having debugging code
+inside a functioning program is actually helpful for others to understand what
+you're doing?''
 
-Shelly: Is unable to use RCloud for organizational reasons but likes the ``concept of being able to create notebooks and share them'' ``wiki is not the best way for communication results - it's kind of like writing a blog post with very limited functionality.
-can't input equations easily.'' ``have to save every picture and post it as an image'' ``I can't share all of the code because it would just get crowded and wouldn't look right on a wiki''
-
-Bart: Also unable to use for org/tech reasons.
-``So if I could make a folder on RCloud and have Python notebooks and also Pig notebooks there, and execute them from RCloud, that would be much better than my current thing, because that would free me from manual documentation and version control and also telling people where my code was.
-It would be just, hey, go look on RCloud, here's the stuff.''
-
-\subsection{Forking}
-Rick: ``I fork my own notebooks because I'm going off and doing some other analogous project, so I've sort of got interesting content that I've already done in a previous analysis that I want to start from and then tweak to match a new set of data.''
-
-Clint: ``I use [forking] on my own work quite a bit, and I use it on Research a ton.
-Honestly, it's one of the handiest functions in it, because instead of having to find it, copy, paste it, you just hit Fork, you rename it, and it's done.
-It's pretty amazing.''
-
-Raif: ``I've been forking other people's notebooks myself, because essentially they have done some work, but then I want to run it on a different part of data, or I want to change some parts, I don't want to see this column, I want a different column, things like that''
-
-Horace: on working with other group using his notebooks, ``I basically point them to a few different parameters that they need to fix.'' ``I'll teach people to intercept the result in the middle and see what you should see, and what's not being returned, that sort of thing'' ``If the computation went wrong in the middle, and something blew up, or you're just returning NULL or empty data, where should we check those things?'' ``We should, but a lot of the time we just insert print statements here and there and check values.
-Having debugging code inside a functioning program is actually helpful for others to understand what you're doing?''
-
-Shelly: ``gives the audience a way to try their own analysis, like maybe Shelly should have used this parameter and they can just change the parameters''
+Shelly: ``gives the audience a way to try their own analysis, like maybe Shelly
+should have used this parameter and they can just change the parameters''
 
@@
 
-Clint: ``instead of looking back and saying I've got a billion files here in this subdirectory and I hope I've got them backed up, if they're on RCloud I know they are'' ``I don't have to worry about it if I drop my computer in the bathtub, everythings gone, because I'm not great about uploading my own for versioning''
+Clint: ``instead of looking back and saying I've got a billion files here in
+this subdirectory and I hope I've got them backed up, if they're on RCloud I
+know they are'' ``I don't have to worry about it if I drop my computer in the
+bathtub, everythings gone, because I'm not great about uploading my own for
+versioning''
 
-Bart: ``I like the fact that it has a built-in editor, so if you need to, like, fixing a typo in a link, or an extra line break or any of this other nonsense, you can just switch to the edit view, pull up your asset, type something, it's automatically saved, committed, everything.
-You don't have to go back to your source code, change it, commit it to the repo, you pull the repo to your distribution version, none of that's necessary.
-A lot of web development is fixing stupid little bugs, and you can do that kind of instantly without a lot of overhead, which is nice.''
+Bart: ``I like the fact that it has a built-in editor, so if you need to, like,
+fixing a typo in a link, or an extra line break or any of this other nonsense,
+you can just switch to the edit view, pull up your asset, type something, it's
+automatically saved, committed, everything. You don't have to go back to your
+source code, change it, commit it to the repo, you pull the repo to your
+distribution version, none of that's necessary. A lot of web development is
+fixing stupid little bugs, and you can do that kind of instantly without a lot
+of overhead, which is nice.''
 
 \subsection{Search}
-Raif: ``the fact that we have all the notebooks there, searchable, it just saves me from replicating what other people have done'' ``One way of thinking of it is like Google Code, you know you just search for something and you find something useful and you just use it.
-It’s a repository of code''
+Raif: ``the fact that we have all the notebooks there, searchable, it just saves
+me from replicating what other people have done'' ``One way of thinking of it is
+like Google Code, you know you just search for something and you find something
+useful and you just use it. It’s a repository of code''
 
-Rick: ``And the same kind of thing when you search through notebooks, usually I know somebody’s notebooks that I want to search through, like I’ll search through Simon Urbanek’s first, then I’ll go off to somebody else’s, because I know that the kind of thing I’m looking for something more obscure than I’m likely to find in some random person’s thing''
+Rick: ``And the same kind of thing when you search through notebooks, usually I
+know somebody’s notebooks that I want to search through, like I’ll search
+through Simon Urbanek’s first, then I’ll go off to somebody else’s, because I
+know that the kind of thing I’m looking for something more obscure than I’m
+likely to find in some random person’s thing''
 
 \subsection{Web integration}
-Horace ``that also means the other guys who want to do analytics on the data, they can first pull the data, do the analytics on it, and then feed the viewer the data.
-So if anyone from the stats group wants to insert some code in between the data pulling procedure, [they] can always insert something in between.
-We were trying to make such an example so that they don't have to rewrite the GUI part and they know how to get the data source, once you get the data into RCloud, then you have a dataframe to work with, and then they know how to produce another dataframe.
-We give them the specification of the dataframes, we give them the data input, they do whatever they want, they can reuse the data connection part and the GUI part.
-We can keep changing these components and that's a very nice thing.''
+Horace ``that also means the other guys who want to do analytics on the data,
+they can first pull the data, do the analytics on it, and then feed the viewer
+the data. So if anyone from the stats group wants to insert some code in between
+the data pulling procedure, [they] can always insert something in between. We
+were trying to make such an example so that they don't have to rewrite the GUI
+part and they know how to get the data source, once you get the data into
+RCloud, then you have a dataframe to work with, and then they know how to
+produce another dataframe. We give them the specification of the dataframes, we
+give them the data input, they do whatever they want, they can reuse the data
+connection part and the GUI part. We can keep changing these components and
+that's a very nice thing.''
 
-Jyothi: ``Having an open session with R certainly helps too, because both of my applications need to be doing something with the data, so having an open session where I can run R command or R functions without having to invoke an API or send out a request and then wait for the request to come back to do the results.
-So that kind of seamless integration is extremely helpful in writing the application.''
+Jyothi: ``Having an open session with R certainly helps too, because both of my
+applications need to be doing something with the data, so having an open session
+where I can run R command or R functions without having to invoke an API or send
+out a request and then wait for the request to come back to do the results.  So
+that kind of seamless integration is extremely helpful in writing the
+application.''
 
-Jyothi: ``another thing I really like about RCloud is your publish feature, so that non-RCloud users, the anonymous login, non-RCloud users can access the notebooks without having to login to RCloud''
+Jyothi: ``another thing I really like about RCloud is your publish feature, so
+that non-RCloud users, the anonymous login, non-RCloud users can access the
+notebooks without having to login to RCloud''
 
-accidental feature: ``some of my jobs are huge and I like to just call them and the fact that the user can just shut down his browser or shut down his computer and the job still runs in the background is actually a very neat thing.
-So what, um, the way I have coded is, whenever I have a huge job to run, I just display to the user, you go do whatever you want, we'll send you an email when the job is done, um that's another feature of RCloud that's very helpful, that the job will continue to run in the background''
+accidental feature: ``some of my jobs are huge and I like to just call them and
+the fact that the user can just shut down his browser or shut down his computer
+and the job still runs in the background is actually a very neat thing. So what,
+um, the way I have coded is, whenever I have a huge job to run, I just display
+to the user, you go do whatever you want, we'll send you an email when the job
+is done, um that's another feature of RCloud that's very helpful, that the job
+will continue to run in the background''
 
 \subsection{Other workflows}
-Rick: ``I find that the style of editing is clunky.
-I would never edit like that if I'm doing ordinary work.
-I usually use vi.'' ``I mean it's nice to have it saved, but you know, there's sort of this trade-off between it making it easier for me to present something or to save something, and my easy of typing and correcting and things in a plain old editor window.'' ``I just have the discipline of having a file that describes what I'm doing with the commands that I'm using so that I can go back and recreate it or pass it on to someone else, so it's a little bit like having an RCloud notebook, it's not necessarily executable but most of the commands that I've typed are in there.''
+Rick: ``I find that the style of editing is clunky. I would never edit like that
+if I'm doing ordinary work. I usually use vi.'' ``I mean it's nice to have it
+saved, but you know, there's sort of this trade-off between it making it easier
+for me to present something or to save something, and my easy of typing and
+correcting and things in a plain old editor window.'' ``I just have the
+discipline of having a file that describes what I'm doing with the commands that
+I'm using so that I can go back and recreate it or pass it on to someone else,
+so it's a little bit like having an RCloud notebook, it's not necessarily
+executable but most of the commands that I've typed are in there.''
 
-Kenny: the UI doesn't work for his workflow, which is pasting snippets from a file into the console in the R GUI.
-he tends to type type type into the file, paste paste paste to the console, rearrange so the good code is at the top of the file, then save the file and call it quits once he's got the right code at the top of the file.
+Kenny: the UI doesn't work for his workflow, which is pasting snippets from a
+file into the console in the R GUI.  he tends to type type type into the file,
+paste paste paste to the console, rearrange so the good code is at the top of
+the file, then save the file and call it quits once he's got the right code at
+the top of the file.
 
-Taylor: ``The big one is I don't know why would I want to use RCloud over my current setup.
-If it's just me, I like my text editor and terminal.
-There's nothing that I want that those two don't give me'' ``In R, when I'm doing development work, I copy and paste in like 5-10 lines of code, so when something breaks, I get an error message on that one line, and I can up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell, so the only way to get that same functionality is if every line's in one cell.
-Because if it runs like lines 1,2,3, then 4 fails, I have to either run the whole cell again, or have line 4 in its own cell.
-I can't rerun just one line.'' ``There's nothing that I want that's not in the terminal.
-I don't think you should try to support all the things I want to do in the terminal.''
+Taylor: ``The big one is I don't know why would I want to use RCloud over my
+current setup. If it's just me, I like my text editor and terminal. There's
+nothing that I want that those two don't give me'' ``In R, when I'm doing
+development work, I copy and paste in like 5-10 lines of code, so when something
+breaks, I get an error message on that one line, and I can up-arrow and change
+it and fix it, whereas in RCloud I have to run a whole cell, so the only way to
+get that same functionality is if every line's in one cell. Because if it runs
+like lines 1,2,3, then 4 fails, I have to either run the whole cell again, or
+have line 4 in its own cell. I can't rerun just one line.'' ``There's nothing
+that I want that's not in the terminal. I don't think you should try to support
+all the things I want to do in the terminal.''
 
-Horace: ``It's not for myself, but for other people who use RCloud vs RStudio, they will always think that they want to prototype on RStudio.
-Maybe because it's more familiar for them.
-So, they work on RStudio, on a small data set, and they fix up everything, make sure everything is right, and then they place it on RCloud for sharing.''
+Horace: ``It's not for myself, but for other people who use RCloud vs RStudio,
+they will always think that they want to prototype on RStudio. Maybe because
+it's more familiar for them. So, they work on RStudio, on a small data set, and
+they fix up everything, make sure everything is right, and then they place it on
+RCloud for sharing.''
 
@@
 \subsection{Proliferation of notebook cells}
-Rick: ``Making new cells and taking some stuff in it, then having it do something, then having to make a new cell, and I don't know, it just gets in my way, so if I'm doing really straightforward simple stuff, I'll just get into R and do it myself.'' ``I'm always scrolling up and down the screen'' ``every time I type more stuff, the notebook gets longer and longer and it's harder to deal with'' ``I've got to do a lot of scrolling around to get to where I'm going, whereas if I'm doing it in vi, I can search for things in my typing, and get to places I want to be without having to scroll through it and try to read it'' ``times when I just want to type in a couple of quick and get some results that are going to tell me what to do next, and they're not necessarily archival in any sense'' ``being literate about anything typically takes a lot of rewriting and going back over things...
-I just want to explore a few things and then I'll know what I want to write.'' ``I'll often type in expressions that allow me to check that I'm the right track, so I know that if the computation worked out, then the sum of the residuals is gonna be zero.
-So I'll type in sum(resid) and if it prints out zero, then I'm all happy, but I don't want that in my notebook.'' ``it tempts me to type a big long thing and then run the whole thing, as opposed to, type a few little pieces that I'm thinking of on the inner parts and then put them together to make the big thing''
+Rick: ``Making new cells and taking some stuff in it, then having it do
+something, then having to make a new cell, and I don't know, it just gets in my
+way, so if I'm doing really straightforward simple stuff, I'll just get into R
+and do it myself.'' ``I'm always scrolling up and down the screen'' ``every time
+I type more stuff, the notebook gets longer and longer and it's harder to deal
+with'' ``I've got to do a lot of scrolling around to get to where I'm going,
+whereas if I'm doing it in vi, I can search for things in my typing, and get to
+places I want to be without having to scroll through it and try to read it''
+``times when I just want to type in a couple of quick and get some results that
+are going to tell me what to do next, and they're not necessarily archival in
+any sense'' ``being literate about anything typically takes a lot of rewriting
+and going back over things... I just want to explore a few things and then I'll
+know what I want to write.'' ``I'll often type in expressions that allow me to
+check that I'm the right track, so I know that if the computation worked out,
+then the sum of the residuals is gonna be zero. So I'll type in sum(resid) and
+if it prints out zero, then I'm all happy, but I don't want that in my
+notebook.'' ``it tempts me to type a big long thing and then run the whole
+thing, as opposed to, type a few little pieces that I'm thinking of on the inner
+parts and then put them together to make the big thing''
 
-Ivan has the complaint that too many cells are created to use the notebook as a console.
-This has a lot to do with the extra real estate that gets taken up by the cells controls, blank space, etc of cells versus a straight command line.
+Ivan has the complaint that too many cells are created to use the notebook as a
+console.  This has a lot to do with the extra real estate that gets taken up by
+the cells controls, blank space, etc of cells versus a straight command line.
 
-KC and Ivan both complained that when there are long results or plots, it causes the code to scroll off the screen.
+KC and Ivan both complained that when there are long results or plots, it causes
+the code to scroll off the screen.
 
-Kenny: ``It saves everything I do like everything is gold, but most of it is junk not meant to be saved'' ``the console is purely disposable''
+Kenny: ``It saves everything I do like everything is gold, but most of it is
+junk not meant to be saved'' ``the console is purely disposable''
 
-Raif: ``whatever you type on the command line, it becomes a part of your notebook, and I find it a little bit annoying, because I really type a lot of things to check my data, and I don't want that to become a part of the notebook'' ``Like you're kind of checking out what your data is, or you make a plot of the data.
-Things that should not really become part of a notebook, but things that help you understand your data better.''
+Raif: ``whatever you type on the command line, it becomes a part of your
+notebook, and I find it a little bit annoying, because I really type a lot of
+things to check my data, and I don't want that to become a part of the
+notebook'' ``Like you're kind of checking out what your data is, or you make a
+plot of the data. Things that should not really become part of a notebook, but
+things that help you understand your data better.''
 
-Horace: ``Well, the nice thing is if you have the whole notebook, you can run it step by step, and try to mess around in between, interactively.''
+Horace: ``Well, the nice thing is if you have the whole notebook, you can run it
+step by step, and try to mess around in between, interactively.''
 
-Rick on auditing (possible solution): ``you didn't see your mistakes, they were in there but you never really saw them because a mistake never led to an answer at the end'' ``the real result of this analysis was these three plots, so go back and figure out everything that I did that was involved in creating those three plots.
-So that I could start from the raw data and create those three plots.''
+Rick on auditing (possible solution): ``you didn't see your mistakes, they were
+in there but you never really saw them because a mistake never led to an answer
+at the end'' ``the real result of this analysis was these three plots, so go
+back and figure out everything that I did that was involved in creating those
+three plots. So that I could start from the raw data and create those three
+plots.''
 
 \subsection{Misc. notebook interface annoyance}
-Rick: ``if I was typing into one of the cells near that top, I had to think of `I'm editing this cell and then I have to execute it', and if was doing the one at the bottom I could type it but then it would automatically execute, but then it became a standard cell and I have to edit it.
-It was just another set of modes that I didn't like''
+Rick: ``if I was typing into one of the cells near that top, I had to think of
+`I'm editing this cell and then I have to execute it', and if was doing the one
+at the bottom I could type it but then it would automatically execute, but then
+it became a standard cell and I have to edit it. It was just another set of
+modes that I didn't like''
 
-Clint: ``it took me a couple of weeks of looking at it to become comfortable because I saw the cells and it just threw me for a total loop.
-I mean, it's a good idea because it's essentially like I can have this part and this part and this part.
-I can run it all at once if I want or I can break these into sections for either debugging or staging purposes which, I really like it now, but when I first saw it, [it was] very confusing...
-I think I prefer it now, because you can cut your code up.''
+Clint: ``it took me a couple of weeks of looking at it to become comfortable
+because I saw the cells and it just threw me for a total loop. I mean, it's a
+good idea because it's essentially like I can have this part and this part and
+this part. I can run it all at once if I want or I can break these into sections
+for either debugging or staging purposes which, I really like it now, but when I
+first saw it, [it was] very confusing... I think I prefer it now, because you
+can cut your code up.''
 
-Corey: In the previous version, it switched between code and results.
-Now you see both - ``and there are times when that's good, but there's certainly times when it'd be nice to - you know, you have 10 or 15 cells and you just want to get back and there's all this output and you're scrolling all over the place so, if there were a way to control both, that'd be great.
-There are times when I do want to see the output, but there are times when I've seen enough of the same thing.''
+Corey: In the previous version, it switched between code and results. Now you
+see both - ``and there are times when that's good, but there's certainly times
+when it'd be nice to - you know, you have 10 or 15 cells and you just want to
+get back and there's all this output and you're scrolling all over the place so,
+if there were a way to control both, that'd be great. There are times when I do
+want to see the output, but there are times when I've seen enough of the same
+thing.''
 
-Jyothi: ``It's good that I can see the results but if I'm working and I need more room and I want to hide the results, I can't.” Could we have “an option to close the results window?”
+Jyothi: ``It's good that I can see the results but if I'm working and I need
+more room and I want to hide the results, I can't.'' Could we have ``an option
+to close the results window?''
 
-Ivan: RStudio's layout is more helpful because
-- the charts are kept in another pane and stay in one place while doing analysis
-- having ``long term'' code (which would be in assets) above the command prompt means that both can be sized very wide to accomodate long lines
+Ivan: RStudio's layout is more helpful because - the charts are kept in another
+pane and stay in one place while doing analysis - having ``long term'' code
+(which would be in assets) above the command prompt means that both can be sized
+very wide to accomodate long lines
 
-Raif: ``cells are really useful'' but ``you want to see your output in a different window, on a separate part of the window'' ``for example imagine I run a cell whose output is huge, right, and now that output would block the whole window, and in order to find the next cell, I have to scroll down, and find where that cell starts, right?
-So, I'm losing the continuity of my code to the output''
+Raif: ``cells are really useful'' but ``you want to see your output in a
+different window, on a separate part of the window'' ``for example imagine I run
+a cell whose output is huge, right, and now that output would block the whole
+window, and in order to find the next cell, I have to scroll down, and find
+where that cell starts, right? So, I'm losing the continuity of my code to the
+output''
 
-Kenny: Problem with screen real estate: side panes take up too much space. Doesn't need the notebook tree pane.
+Kenny: Problem with screen real estate: side panes take up too much
+space. Doesn't need the notebook tree pane.
 
-KC thought that having a scratch pad that doesn't persist between different notebooks is ``useless''.
+KC thought that having a scratch pad that doesn't persist between different
+notebooks is ``useless''.
 
-Kenny: Problem of long-running cells.
-``The long tail.''  First cell takes seconds to run, next cell minutes, next cell 5 hours.
-In this scenario, which is common for Kenny, the ``Run whole notebook'' button is dangerous.
-This is also a huge issue for demoing code.
-Suggests something like knitr's caching of results.
-When converting his work to notebooks, Kenny ends up with a lot of comments that say ``this cell takes a long time to run.'' He ends up ``littering the notebook with little switches that comment out'' the slow parts and ``load the object instead'', ``save the object to disk''.
+Kenny: Problem of long-running cells.  ``The long tail.''  First cell takes
+seconds to run, next cell minutes, next cell 5 hours.  In this scenario, which
+is common for Kenny, the ``Run whole notebook'' button is dangerous. This is
+also a huge issue for demoing code.  Suggests something like knitr's caching of
+results. When converting his work to notebooks, Kenny ends up with a lot of
+comments that say ``this cell takes a long time to run.'' He ends up ``littering
+the notebook with little switches that comment out'' the slow parts and ``load
+the object instead'', ``save the object to disk''.
 
@@
 
-Byers: Problem with forking a notebook and correcting it, is that the original notebook still exists with the error.
-We need ways to see that there is a more recent fork of notebook and to fetch in the changes from it.
+Byers: Problem with forking a notebook and correcting it, is that the original
+notebook still exists with the error.  We need ways to see that there is a more
+recent fork of notebook and to fetch in the changes from it.
 
-Taylor: ``I'll make a notebook of a little tutorial of, here's the basic functions and what they do, and then I'll send them out, and then people will give me feedback on them, or have some new feature, or I'll realize something's wrong, and then I'll go back and change it and update that page, but then what happens is people have in the meantime, forked it, and then they have a version that… and usually what it'll be is actually that I'll have to change a package as well, so their old fork stops working, and then they complain, and then, oh no, you have to go get the original [notebook].
-Yeah, that's the sort of..
-This probably bites me more than other people, because I'm doing development so things are changing as they're looking at the code.
-Like, iotools, a lot of the functions change names or change syntax.''
+Taylor: ``I'll make a notebook of a little tutorial of, here's the basic
+functions and what they do, and then I'll send them out, and then people will
+give me feedback on them, or have some new feature, or I'll realize something's
+wrong, and then I'll go back and change it and update that page, but then what
+happens is people have in the meantime, forked it, and then they have a version
+that… and usually what it'll be is actually that I'll have to change a package
+as well, so their old fork stops working, and then they complain, and then, oh
+no, you have to go get the original [notebook]. Yeah, that's the sort of.. This
+probably bites me more than other people, because I'm doing development so
+things are changing as they're looking at the code. Like, iotools, a lot of the
+functions change names or change syntax.''
 
-Taylor: ``That's just a problem because people will have old notebooks that don't work anymore.
-It's annoying when it's a notebook that I still have and the format has changed just a bit.
-More than that is when people have, now that I know that you can't even delete notebooks, they'll have a fork of something that doesn't exist anymore on my thing.'' ``Now I'm really gun-shy about sharing notebooks because, I don't want to like, it's like, do I want to support this forever?''
+Taylor: ``That's just a problem because people will have old notebooks that
+don't work anymore. It's annoying when it's a notebook that I still have and the
+format has changed just a bit. More than that is when people have, now that I
+know that you can't even delete notebooks, they'll have a fork of something that
+doesn't exist anymore on my thing.'' ``Now I'm really gun-shy about sharing
+notebooks because, I don't want to like, it's like, do I want to support this
+forever?''
 
-Taylor: ``A good example is, someone was complaining about doing forking in an RCloud notebook.
-They were trying to use a parallel package.
-And I said, no no I think it works, and I did like a one-cell thing that just did some sweep kind of thing to show that it was actually working in parallel, and then I just deleted it, thinking it was gone.
-And then it saved something in a directory that I then deleted, because it was just a temporary directory.
-Three months later, I don't know why they were running it, they said `oh it doesn't work'.
-I think that's the biggest issue I'm having right now, all these forks''
+Taylor: ``A good example is, someone was complaining about doing forking in an
+RCloud notebook. They were trying to use a parallel package. And I said, no no I
+think it works, and I did like a one-cell thing that just did some sweep kind of
+thing to show that it was actually working in parallel, and then I just deleted
+it, thinking it was gone. And then it saved something in a directory that I then
+deleted, because it was just a temporary directory. Three months later, I don't
+know why they were running it, they said `oh it doesn't work'. I think that's
+the biggest issue I'm having right now, all these forks''
 
-Taylor: ``there's no way for both of us to have ownership of a notebook, so the only way is to fork it back and forth, and so we have dozens of old copies.
-[We end up] deleting all the old ones, but people still have links to them, because they don't actually disappear.'' ``The UI makes it easy to go back to the old stale versions, because you can say `forked from' `forked from' `forked from', but there's actually no way to go the way that we want people to go, which is, like, up the tree.''
+Taylor: ``there's no way for both of us to have ownership of a notebook, so the
+only way is to fork it back and forth, and so we have dozens of old copies. [We
+  end up] deleting all the old ones, but people still have links to them,
+because they don't actually disappear.'' ``The UI makes it easy to go back to
+the old stale versions, because you can say `forked from' `forked from' `forked
+from', but there's actually no way to go the way that we want people to go,
+which is, like, up the tree.''
 
-Rick: ``I don't necessarily need to see everybody's notebook that uses RCloud'' ``every time I do something new, I get a new notebook and so now my notebooks are maybe 50 or 60.
-That's enough to think about just on my own, but if I've got everybody's 50 or 60 sitting on my display, I find that it's more than I want to know about.''
+Rick: ``I don't necessarily need to see everybody's notebook that uses RCloud''
+``every time I do something new, I get a new notebook and so now my notebooks
+are maybe 50 or 60. That's enough to think about just on my own, but if I've got
+everybody's 50 or 60 sitting on my display, I find that it's more than I want to
+know about.''
 
 \subsection{Versioning not useful}
-Rick: ``I don't need something that keeps track of every mistake I've made or every direction I've tried.” ``namespaces are already crowded, so trying to remember the names of notebooks is hard, much less the names of states within each notebook.''
+Rick: ``I don't need something that keeps track of every mistake I've made or
+every direction I've tried.'' ``namespaces are already crowded, so trying to
+remember the names of notebooks is hard, much less the names of states within
+each notebook.''
 
 \subsection{Web snafus}
-Corey: ``“What frustrates me is that if I do leave this window for what feels like 5 or 7 minutes, it disconnects and I have to reconnect and run the whole notebook again.''
+Corey: ``What frustrates me is that if I do leave this window for what feels
+like 5 or 7 minutes, it disconnects and I have to reconnect and run the whole
+notebook again.''
 
@@
 
-Raif: ``being web-based, it's obviously slower than something you would be running on your machine''
+Raif: ``being web-based, it's obviously slower than something you would be
+running on your machine''
 
-Byers: even the little lag of sending a command to a server and receiving the result is intolerable
+Byers: even the little lag of sending a command to a server and receiving the
+result is intolerable
 
 \subsection{Parameterization}
-Clint: ``For the things that I'll do multiple times, I'll try to parameterize it, and I'm working now on putting some packages together that will operate in some fashion outside of RCloud, straight from the command line''
+Clint: ``For the things that I'll do multiple times, I'll try to parameterize
+it, and I'm working now on putting some packages together that will operate in
+some fashion outside of RCloud, straight from the command line''
 
-Taylor: ``like the top of it would be, `This is a report of the volume of all of our feeds for this month', and someone would want to look at it for the next month or the previous month, so they'd fork it to change the month''
+Taylor: ``like the top of it would be, `This is a report of the volume of all of
+our feeds for this month', and someone would want to look at it for the next
+month or the previous month, so they'd fork it to change the month''
 
-Jyothi: ``I think that if a notebook is web-enabled [parameterized] then it gives the person who is running the notebook, I mean they can always fork the notebooks and make changes, but I feel that if the owner of the notebook web-enables it, then whoever is running it, it's easier to run, instead of forking it, if they can set options, it's probably more efficient, and also they can fork it if they want to.''
+Jyothi: ``I think that if a notebook is web-enabled [parameterized] then it
+gives the person who is running the notebook, I mean they can always fork the
+notebooks and make changes, but I feel that if the owner of the notebook
+web-enables it, then whoever is running it, it's easier to run, instead of
+forking it, if they can set options, it's probably more efficient, and also they
+can fork it if they want to.''
 
 \subsection{Other observations}
-Byers: There is also a big problem with ``bitrot'' - notebooks often stop working because the user changed the structure of their data, changed a filename or a database.
-We need ``organizational protocols'' to catch up with the technology.
-There is a ``hysteresis between the technical and the social'' and we have not yet adapted.
+Byers: There is also a big problem with ``bitrot'' - notebooks often stop
+working because the user changed the structure of their data, changed a filename
+or a database.  We need ``organizational protocols'' to catch up with the
+technology.  There is a ``hysteresis between the technical and the social'' and
+we have not yet adapted.
 
-Taylor: `` It would be a file inside a directory that only I had read access to and so someone else couldn't run it...
-That was usually what it was, it was a temporary file on something that was read-only to everyone else.
-But I've gotten careful about that.''
+Taylor: `` It would be a file inside a directory that only I had read access to
+and so someone else couldn't run it... That was usually what it was, it was a
+temporary file on something that was read-only to everyone else. But I've gotten
+careful about that.''
 
-Bart: ``One UI notebook and a bunch of other notebooks that do various tasks and are called with API calls.
-The problem is, if you clone all the notebooks, then you have to go update every single notebook, because all the IDs are different now.
-There's no sort of relative path type stuff that you can do.
-You can do it with static assets, but you can't do it with Ajax calls''
+Bart: ``One UI notebook and a bunch of other notebooks that do various tasks and
+are called with API calls. The problem is, if you clone all the notebooks, then
+you have to go update every single notebook, because all the IDs are different
+now. There's no sort of relative path type stuff that you can do. You can do it
+with static assets, but you can't do it with Ajax calls''

### 2015-03-23 system.tex
better distinguish colors; comment; capitalize GitHub

@@
 
-Figure~\ref{fig:something} shows the RCloud developer interface.
-The center panel is a notebook for code and visualizations.
-Code is edited in this panel, and visualizations are rendered.
-An inventory of supplementary ``asset'' files, that are part of
-the notebook, though not in its executable flow, are edited on the right.
-Controls at the top of the screen allow the user to run, view, and
-share the notebook. The result of code execution is the final cell of
-the notebook.  On the left is a browsing area for searching and viewing
-other users' notebooks, and a help system.
+Figure~\ref{fig:something} shows the RCloud developer interface.  The center
+panel is a notebook for code and visualizations.  Code is edited in this panel,
+and visualizations are rendered.  An inventory of supplementary ``asset'' files,
+that are part of the notebook, though not in its executable flow, are edited on
+the right.  Controls at the top of the screen allow the user to run, view, and
+share the notebook. \gordon{Why are we talking about the callable interface
+here?} The result of code execution is the final cell of the notebook.  On the
+left is a browsing area for searching and viewing other users' notebooks, and a
+help system.
 
@@
 
-To implement this, we built RCloud on top of Github's \emph{gists}~\cite{Github:2014:GG}.
-Github offers a HTTP interface for creating simplified git repositories, the main limitation
+To implement this, we built RCloud on top of GitHub's \emph{gists}~\cite{GitHub:2014:GG}.
+GitHub offers a HTTP interface for creating simplified git repositories, the main limitation
 being a restriction to text-only files in a single directory. The GitHub web-service

### 2015-03-23 interviews.tex
a little bit more narrative
long ways to go

@@
 \subsection{Sharing of results}
-Sharing of notebooks is the core feature of RCloud, and a popular one. Almost
-everyone we spoke to praised this feature. By default, all notebooks are
-publicly visible in RCloud, and notebooks can be found through the notebook tree
-or search.
+Sharing of notebooks is the core feature of RCloud, and a popular one. All
+active users we spoke to praised this feature.
 
-However, the method for sharing that users mentioned most is by sending links
-through email. Clint says, ``I can share it with people and copy it pretty
-readily. So if my supervisor wants to see what I've done or QA it, I can just
-send her a link.''
+Raif says ``I've been sharing my notebooks in order for other people to see what
+I've done. It's very convenient for that purpose, I would say.'' ``we haven't
+been editing them together, but we've been just sharing, and I've been looking
+at things that other people do. So there's no collaborative aspect of it, like
+developing the code, but there's sharing aspects for other people to see and for
+me to see other people's notebooks''
 
-``I think the best part about it is how easily you can share code... you can
-find a working example, rather than wearing out Google and finding questionable
-examples that may or may not work.''
+By default, all notebooks are publicly visible in RCloud, and notebooks can be
+found through the notebook tree or search. However, the method for sharing that
+users mentioned most is by sending links through email. Clint says, ``I can
+share [my notebook] with people and copy it pretty readily. So if my supervisor
+wants to see what I've done or QA it, I can just send her a link.''
 
-Corey: ``being able to share notebooks or codes, or collaborate on something,
-easily being able to share that information. That's an obvious advantage and
-nice to have.''
+Besides providing a way to present work to others, notebook sharing can also
+provide a starting point. Clint says, ``I think the best part about it is how
+easily you can share code... you can find a working example, rather than wearing
+out Google and finding questionable examples that may or may not work.'' Raif
+notes, if ``some person has done something similar, then you're able to just
+edit that, and that's saved a lot of work time for me''
 
-Raif: ``I've been sharing my notebooks in order for other people to see what
-I've done. It's very convenient for that purpose, I would say'' ``we haven't
-been editing them together, but we've been just sharing, and then I've been
-looking at these things that other people do. So there's no collaborative aspect
-of it, like developing the code, but there's like sharing aspects for other
-people to see and for me to see other people's notebooks'' ``there already is
-some person who has done something similar, and then you're able to just edit
-that, and that's saved a lot of work time for me''
-
-Taylor: ``I use it more for sharing code with other people, and for doing
-tutorials for iotools or hmr or explaining'' ``a lot of times I'll share the
-code, because I'm doing so much package development. I want people to see how
-the package works, so I clearly want them to see the code… I kind of write it
-just like I would write GitHub Markdown, where you have little code snippets and
-text, but RCloud lets me actually run the snippets'' with results.
+Taylor develops packages for analysts: ``I use it more for sharing code with
+other people, and for doing tutorials for iotools or hmr or explaining'' issues with data sources. ``A lot
+of times I'll share the code because I'm doing so much package development. I
+want people to see how the package works, so I clearly want them to see the
+code... I kind of write it just like I would write GitHub Markdown, where you have
+little code snippets and text, but RCloud lets me actually run the snippets''
+and display the results.
 
@@
 
-Rick: ``I like being able to look at other people's notebooks.''
+%% Corey: ``being able to share notebooks or codes, or collaborate on something,
+%% easily being able to share that information. That's an obvious advantage and
+%% nice to have.''
+
+%% Rick: ``I like being able to look at other people's notebooks.''
 
@@
 happens is people have in the meantime, forked it, and then they have a version
-that… and usually what it'll be is actually that I'll have to change a package
+that... and usually what it'll be is actually that I'll have to change a package
 as well, so their old fork stops working, and then they complain, and then, oh
-no, you have to go get the original [notebook]. Yeah, that's the sort of.. This
+no, you have to go get the original [notebook]. Yeah, that's the sort of... This
 probably bites me more than other people, because I'm doing development so

### 2015-03-23 discussion.tex
various snarky comments

@@
 \label{sec:discussion}
+
+\gordon{This section now overlaps with the Interview Study quite a bit - we can
+  use that to motivate this, but that would mean not immediately responding to
+  the complaints raised there. On the other hand, trying to merge these two
+  sections would be a beast. The good news is, we anticipated most of the
+  problems that people are facing now. :-/}
 
@@
 encourages making incremental changes at low cost and without disrupting
-the work of others.
+the work of others. \gordon{But this design choice means that developing
+larger, multi-page apps is less well supported.}
 
@@
 to external systems and other features almost always win over architectural
-purity in the real world.
+purity in the real world. \gordon{If you don't export, doesn't that mean
+you have to support every possible operation downstream? Due to popular demand,
+we did add plot export.}
 
-{\bf Social-psychological incentives} and {\bf voting and ranking}
-are supported through starring. This mechanism was employed often
-by our test audience. An obvious next step would be to enhance
-recommendations using relationships discovered by static and
-dynamic code analysis. This may be considered both within and across
-collections of scripts (the latter being similar to Vistrails'
-enhanced recommendations by clustering multiple workflows).
-It seems valuable to know which packages are frequently used together,
-or appear in proximity to a certain record types or data feeds.
-In general, trivial or passive mechanisms to collect data for
-recommendations are essential.
+{\bf Social-psychological incentives} and {\bf voting and ranking} are supported
+through starring. This mechanism was employed often by our test
+audience. \gordon{actually, the notebook tree quickly became so unwieldy that
+no one uses this.} An obvious next step would be to enhance recommendations using
+relationships discovered by static and dynamic code analysis. This may be
+considered both within and across collections of scripts (the latter being
+similar to Vistrails' enhanced recommendations by clustering multiple
+workflows).  It seems valuable to know which packages are frequently used
+together, or appear in proximity to a certain record types or data feeds.  In
+general, trivial or passive mechanisms to collect data for recommendations are
+essential.
 
@@
 {\bf Curation.} Even without formal group management, users may
-curate groups of related notebooks using file system folders.
+curate groups of related notebooks using notebook tree folders.
 We found this particularly effective in collecting and distributing
@@
 in Python, weka in Java).\stephen{Isn't this fixed in some way now?}
+\gordon{Should we have a section on ways that we've responded to popular demand?}
 
@@
 new approaches to information security in visual analytics.
+\gordon{This sounds like we think we are helping by pushing the
+system to the breaking point, is that the idea?}

### 2015-03-24 paper.tex
working
yes
Renamed some files
cram in a ton of quotes - completely unedited and without narrative
better distinguish colors; comment; capitalize GitHub
abstract, teaser

--- 2015-01-02 paper.tex
@@
 \usepackage{url}
+\usepackage{scrextend}   % for addmargin in introduction.tex
+\usepackage[usenames,dvipsnames]{xcolor}
 
@@
 
-\title{RCloud: Integrated Visualization, Analysis, Development, and Deployment}
+\title{Social Visual Analytics: DevOps for Data Scientists}
 
@@
 %% Abstract section.
-\abstract{ Consider the emerging role of a data science team within an
-  organization. Individual data scientists and statisticians usually
-  work on loosely related problems, and must devise effective ways to
-  share their findings and move results from exploratory data analysis
-  into automated diagnostics and reports deployed for wider consumption.
+\abstract{ Consider the emerging role of data science teams embedded in
+  larger organizations. Individual analysts work on loosely related problems,
+  and must share their findings with each other and the organization at large,
+  moving results from exploratory data analyses (EDA)
+  into automated visualizations, diagnostics and reports deployed for wider consumption.
   There are two problems with the current
@@
   with one set of tools, and automated reports and deployments with another.
-  Second, EDA environments often assume a single-developer perspective,
+  Second, these environments often assume a single-developer perspective,
   while data scientist teams would get much benefit from easier sharing
@@
   \emph{discoverability}. In addition, we contribute the design and
-  implementation of RCloud, a system that supports the requirements
-  collaborative data analysis, visualization and web deployment. We
-  discuss the design decisions, tradeoffs and limitations, and compare RCloud
+  implementation of RCloud, a system that supports the requirements of
+  collaborative data analysis, visualization and web deployment.
+  The biggest deployment of RCloud has been in active use for more than a year,
+  and has about fifty active users. We report on interviews with some of these users, and
+  discuss the design decisions, tradeoffs and limitations, comparing RCloud
   to other current proposals.
-  \carlos{RCloud has been in production use for more than a year and has about one hundred active users. We will discuss the adoption issues and etc etc.}
 } % end of abstract
@@
   \caption{
-    \carlos{Come up with caption for this. The story is: hackers/scripters work in the middle (and bottom-right), rest of organization has automatic access to their work on left and top-right.}
+    An overview of features in RCloud.
+RCloud was built to support an environment in which a large number of loosely-related problems in data and visual analytics are solved by a small number of \emph{hackers} and \emph{scripters}.
+In such a shared environment, \emph{discoverability} is also a problem.
+RCloud supports search, annotation, recommendation, and commenting for all notebooks, enabling an overview mode where users can \emph{browse} popular and recent analyses.
+When problems and data sources change constantly, deployment becomes a time sink; RCloud supports \emph{transparent and automatic} deployment of analyses as web pages and web services, allowing a seamless transition from exploratory work to production code.
   }
@@
 \input{casestudies}
+\input{interviews}
 \input{discussion}

### 2015-03-24 discussion.tex
table

@@
 \label{sec:discussion}
+
+\begin{table*}
+  \centering
+  \begin{tabular}{l|cccccc}
+           & Versioning/Forking & Collaboration & Deployment & Multilanguage & Integrated Reports & Integrated Analysis \\
+    \hline
+RCloud     &       v            &       v       &     v      &      v        &           v        &           v         \\
+RStudio    &                    &               &            &               &           v        &           v         \\
+JSFiddle   &       v            &       v       &            &               &                    &                     \\
+bl.ocks    &       v            &       v       &            &               &           v        &                     \\
+shiny      &                    &               &     v      &               &           v        &           v         \\
+Jupyter    &                    &               &            &      v        &           v        &           v         \\
+Tableau    &                    &               &            &               &                    &                     
+  \end{tabular}
+  \caption{Comparison of system features.}
+\end{table*}
 

### 2015-03-25 introduction.tex
intro work

@@
 
-Over fifty years ago, John Tukey published ``The Future of Data Analysis''
-in which he defined a remarkably farsighted agenda for the future of
-data analysis and statistical computing. His vision has been realized
-through environments like S and R, that started a revolution by
-integrating statistical graphics, computation, and robust techniques
-for practical data modeling. Every day, many thousands of data scientists
-and others specialists benefit from this astounding vision.
+Almost a
+half-century ago, Tukey foresaw
+much of what is now commonplace in data analysis~\cite{TukeyFDA}.
+Powerful, interactive environments for analysis and programming were
+the goal, together with an unflinching (and, at the time, somewhat
+heretical) insistence in keeping humans as a central part of the
+discovery process. His now-famous quip that ``the picture-examining eye is the
+best finder we have of the wholly unanticipated'' has come to define
+much of visual analytics and exploratory visualization~\cite{TukeyEDA}.
 
-The field has now moved beyond the boundaries of what Tukey originally
-foresaw. Processing and networking capabilities today far exceed
-what was barely imagined fifty years ago. Inexpensive, scalable
-virtualized systems allow analysts to routinely solve problems
-involving millions or even billions of data values. Networking
-also enables both collaboration and the automation of complex
-tasks and processes in ways that were not considered even ten years ago.
-We turn our attention to this opportunity to use computation
+In some way, we have moved far beyond what Tukey imagined:
+computing and networking capabilities today far exceed
+what was barely imaginable then. 
+We argue, on the other hand, that we \emph{develop} our data analysis
+solutions, and context in which they become part of the broader issues
+being solved in organizations, has not changed as much: the S language
+was developed essentially alongside Unix~\cite{TheSSystem}, after all.
+In this paper, we turn our attention to this opportunity to use computation
 to support, not only individuals, but entire teams and their
@@
 
-As a starting point, consider the role of a data science team within
+Consider the role of a data science team within
 a business or technical organization today.
-Analysts work in loose or tight collaboration with colleagues and
-domain experts.
 Project teams vary in size, from just a few people to dozens or
@@
 actions to help clients to realize the benefits of the analysis.
-Eventually, knowledge or working prototypes created by data science teams
-may be transferred to other organizations to employ them in production.
+Eventually, knowledge or working prototypes created by these teams
+are transferred to other organizations to employ them in production.
 There are many details to these tasks. For example, data wrangling can involve

### 2015-03-25 introduction.tex
refine usership stats

--- 2015-03-20 introduction.tex
@@
 
-Over the past two years, we prototyped this environment and deployed it
-in a production community of data scientists, business analysts and other colleagues.
-The platform today has over 200 active accounts, and several dozen people have forked
-and customized code in the system. Most of the active users are members of
-the statistics research department of our lab, but some others are data
-scientists and other collaborators in business units. \stephen{I hope this is right.}
+Over the past two years, we prototyped this environment and deployed it in a
+production community of data scientists, business analysts and other colleagues.
+The platform today has over 300 active accounts; about 20 people use it more
+than twice a week, and about 50 people use it more than once a week.  Most of
+the active users are members of AT&T Labs, but some others are data scientists
+and other collaborators in business units. \gordon{This doesn't measure how
+intensely people use it...}
 

### 2015-03-25 introduction.tex
summarize interviews

--- 2015-03-25 introduction.tex
@@
 
-Interviews in {\it Data Scientists at Work} by Sebastian Gutierrez illuminate
-the teamwork and sharing in the current practice of visual analytics in industry.
-We particularly appreciated the remarks by Jonathan Lenaghan about the potential
-for continuous integration and publication.
+%% Interviews in {\it Data Scientists at Work} by Sebastian Gutierrez illuminate
+%% the teamwork and sharing in the current practice of visual analytics in industry.
+%% We particularly appreciated the remarks by Jonathan Lenaghan about the potential
+%% for continuous integration and publication.
 
-% \begin{addmargin}[1em]{1em}  % left and right margin change
-% part1.txt
-Erin Shellman [of Nordstrom] described the benefits of
-continuous deployment of experiments, and sharing knowledge
-through software and data analysis artifacts.
-``Finally, prototyping our products so that internal
+%% % \begin{addmargin}[1em]{1em}  % left and right margin change
+%% % part1.txt
+%% Erin Shellman [of Nordstrom] described the benefits of
+%% continuous deployment of experiments, and sharing knowledge
+%% through software and data analysis artifacts.
+%% ``Finally, prototyping our products so that internal
+%% customers can use them early on has been crucial for
+%% our success. It doesn't even have to be something
+%% fancy. For instance, our recommendations preview
+%% tool doesn't have a particularly interesting
+%% visualization, but its enough to show the result of our
+%% algorithms. Now we can shoot off a URL to internal
+%% customers and it allows them to sit at their desk and
+%% understand the behavior of our product and
+%% experiment with it, and provide feedback way before
+%% we're talking about getting it into production. This
+%% has been super helpful and has been a really great
+%% way to get people excited about what we're working
+%% on...
+%% %Building and maintaining those relationships are
+%% %just like anything else-- you always have to be
+%% %working on them. Nurturing those relationships are
+%% %part of the job, and if you leave them unattended,
+%% %they might not be there later.
+%% %
+%% % part2.txt
+%% %Gutierrez: How do you share the knowledge you are
+%% %building with others?
+%% We’re obsessed with Confluence from
+%% Atlassian, and the Data Lab has a very active
+%% Confluence space. Recommendo is fully documented
+%% on Confluence, so anyone in the company could learn
+%% how it’s built, where to find it, and how to use it.
+%% We also share exploratory analyses and reports on
+%% Confluence so that we can still exchange knowledge
+%% even if the work didn't make it into a larger project.''
+
+%% % part3.txt
+%% John Foreman commented on the scope of the work in
+%% data science teams.
+%% ``I lead the data science team at
+%% MailChimp, and I like to get my hands dirty too.
+%% Some of the big pieces of my day involve working
+%% with my team to take stock of current projects and
+%% figure out where to go next, doing my own work and
+%% prototyping things-- I do projects just like my peers
+%% do and then also my talking with other teams,
+%% talking with management, and planning for the
+%% future.
+%% On our team, we've got different folks facing different
+%% % part4.txt
+%% kinds of projects. We've got one person who really
+%% owns compliance and looks at our compliance
+%% processes. We've got another data scientist who
+%% focuses on more of the user experience side of the
+%% house and understanding our customers. I help them
+%% and others as is needed while I do the three things
+%% mentioned earlier-- build data products, be a
+%% translator, and have conversations with the data
+%% science community.''
+
+%% % part5.txt
+%% Jonathan Lenaghan from PlaceIQ discussed to the potential
+%% for continuous deployment in data science.
+%% ``If you look at dev ops right now, they have things
+%% such as continuous integration, continuous build,
+%% automated testing, and test harnesses-- a1l of which
+%% map very well from the dev ops world to the data ops
+%% (a phrase I stole from Red Monk) world very easily. I
+%% think this is a very powerful notion. It is important to
+%% have testing frameworks for all of your data, so that if
+%% you make a code change, you can go back and test all
+%% of your data.''
+
+%% % part6.txt
+%% Anna Smith from Bitly spoke about the benefits of access to
+%% a rich environment of shared data and code.
+%% ``In addition to Bitly's extraordinary data sets,
+%% I was given all the resources that I needed. These were
+%% resources that I’d only heard about but had never
+%% seen. I went from asking what a Hadoop cluster was,
+%% as my school didn't have one, to being able to work
+%% with one. Now I could play with one to figure out how
+%% to make my work better and how to change my
+%% algorithms so they were cleaner and ran faster on
+%% Hadoop. The whole transition was eye opening. Not
+%% only that, I could also look at other people’s code and
+%% play with their code and data sets as well. It was
+%% fantastic and I was convinced that working with data was my thing.''
+%% % \end{addmargin}
+
+%% % (Hiring ``full stack developers'' to perform integration seems to be a trend, but does not achieve the potential benefits of specialization and division of labor in larger teams, and is not a practice to be encouraged.)
+
+Unfortunately, the processes and technologies supporting data analysis visual analytics are fragmented.
+Knowledge is shared through informal conversations, emails,
+meetings, phone calls, source code itself, in wikis, in project documents and
+by other means. Results of experiments are often shared first by copying static
+output (usually a simple screen shot or image output). By the time
+decisions are made based on the results of that screenshot, the data
+and the processes might have changed enough to bring to question the
+original decision.
+
+%% This disjointed, inconsistent process inhibits the use of tools to bridge
+%% the many gaps and to help people find resources.
+%% For example, even when code is shared, we still not may know where to
+%% find examples that show how to use it.
+%% When we find a useful data source, we don't know what packages or
+%% other data sources usually go along with it, what problems with the data
+%% others have already discovered and corrected, or where to obtain online feeds
+%% to get updates.
+
+This situation can affect how data science teams; as analysts and
+toolbuilders, we have experienced this first-hand in our environment, and
+others report similar findings. Gutierrez collected interviews with data
+scientists~\cite{Gutierrez}, which include the following:
+
+\begin{itemize}
+\item (upon being given access to other code and data analysis, in
+order to learn about Hadoop) ``[...] Not only that, I could also look at other people’s code and
+play with their code and data sets as well''
+\item (discussing the value of sharing prototypes rather than static
+data) ``[...] Finally, prototyping our products so that internal
 customers can use them early on has been crucial for
-our success. It doesn't even have to be something
-fancy. For instance, our recommendations preview
-tool doesn't have a particularly interesting
-visualization, but its enough to show the result of our
-algorithms. Now we can shoot off a URL to internal
-customers and it allows them to sit at their desk and
-understand the behavior of our product and
-experiment with it, and provide feedback way before
-we're talking about getting it into production. This
-has been super helpful and has been a really great
-way to get people excited about what we're working
-on...
-%Building and maintaining those relationships are
-%just like anything else-- you always have to be
-%working on them. Nurturing those relationships are
-%part of the job, and if you leave them unattended,
-%they might not be there later.
-%
-% part2.txt
-%Gutierrez: How do you share the knowledge you are
-%building with others?
-We’re obsessed with Confluence from
-Atlassian, and the Data Lab has a very active
-Confluence space. Recommendo is fully documented
-on Confluence, so anyone in the company could learn
-how it’s built, where to find it, and how to use it.
-We also share exploratory analyses and reports on
-Confluence so that we can still exchange knowledge
-even if the work didn't make it into a larger project.''
+our success. [...] Now we can shoot off a URL to internal
+customers and it allows them to [...] provide feedback way before
+we're talking about getting it into production.''
+\item (on sharing more than just a finished product) ``We also share
+exploratory analyses and reports [...] so that we can still exchange
+knowledge even if the work didn't make it into a larger project.''
+\item (on changing analyses and processes) ``It is important to
+have testing frameworks for all of your data, so [...] you can go back and test all
+of your data. 
+\end{itemize}
 
-% part3.txt
-John Foreman commented on the scope of the work in
-data science teams.
-``I lead the data science team at
-MailChimp, and I like to get my hands dirty too.
-Some of the big pieces of my day involve working
-with my team to take stock of current projects and
-figure out where to go next, doing my own work and
-prototyping things-- I do projects just like my peers
-do and then also my talking with other teams,
-talking with management, and planning for the
-future.
-On our team, we've got different folks facing different
-% part4.txt
-kinds of projects. We've got one person who really
-owns compliance and looks at our compliance
-processes. We've got another data scientist who
-focuses on more of the user experience side of the
-house and understanding our customers. I help them
-and others as is needed while I do the three things
-mentioned earlier-- build data products, be a
-translator, and have conversations with the data
-science community.''
-
-% part5.txt
-Jonathan Lenaghan from PlaceIQ discussed to the potential
-for continuous deployment in data science.
-``If you look at dev ops right now, they have things
-such as continuous integration, continuous build,
-automated testing, and test harnesses-- a1l of which
-map very well from the dev ops world to the data ops
-(a phrase I stole from Red Monk) world very easily. I
-think this is a very powerful notion. It is important to
-have testing frameworks for all of your data, so that if
-you make a code change, you can go back and test all
-of your data.''
-
-% part6.txt
-Anna Smith from Bitly spoke about the benefits of access to
-a rich environment of shared data and code.
-``In addition to Bitly's extraordinary data sets,
-I was given all the resources that I needed. These were
-resources that I’d only heard about but had never
-seen. I went from asking what a Hadoop cluster was,
-as my school didn't have one, to being able to work
-with one. Now I could play with one to figure out how
-to make my work better and how to change my
-algorithms so they were cleaner and ran faster on
-Hadoop. The whole transition was eye opening. Not
-only that, I could also look at other people’s code and
-play with their code and data sets as well. It was
-fantastic and I was convinced that working with data was my thing.''
-% \end{addmargin}
-
-% (Hiring ``full stack developers'' to perform integration seems to be a trend, but does not achieve the potential benefits of specialization and division of labor in larger teams, and is not a practice to be encouraged.)
-
-Unfortunately, processes and technology that support visual analytics today
-work are fragmented.
-Knowledge is shared through informal conversations, email online messages,
-meetings, phone calls, within code itself, in wikis, in project documents and
-by other means. Results of experiments are often shared first by copying static
-output, later, by holding meetings that lead to moving or even rewriting
-entire programs.
-
-This disjointed, inconsistent process inhibits the use of tools to bridge
-the many gaps and to help people find resources.
-For example, even when code is shared, we still not may know where to
-find examples that show how to use it.
-When we find a useful data source, we don't know what packages or
-other data sources usually go along with it, what problems with the data
-others have already discovered and corrected, or where to obtain online feeds
-to get updates.
-
-This situation definitely affects how well data science teams work.
-It limits our ability to share work and to rapidly deploy and refine results.
-We have seen this in our lab environment.
-Other authors report similar findings.
-
-Some symptoms of the current situation are:
+Currently, these concerns need to be addressed \emph{alongside} the
+visual analytics environment. Some symptoms of the current situation are:
 \begin{itemize}

### 2015-03-25 interviews.tex

@@
 To evaluate the effectiveness of RCloud, we interviewed 13 current and past
-users of RCloud. Of these, 9 are adata analysts, and 4 build tools for data
+users of RCloud. Of these, 9 are data analysts, and 4 build tools for data
 analysts and business needs.
@@
 Taylor develops packages for analysts: ``I use it more for sharing code with
-other people, and for doing tutorials for iotools or hmr or explaining'' issues with data sources. ``A lot
-of times I'll share the code because I'm doing so much package development. I
-want people to see how the package works, so I clearly want them to see the
-code... I kind of write it just like I would write GitHub Markdown, where you have
-little code snippets and text, but RCloud lets me actually run the snippets''
-and display the results.
+other people, and for doing tutorials for iotools or hmr'' or explaining issues
+with data sources. ``A lot of times I'll share the code because I'm doing so
+much package development. I want people to see how the package works, so I
+clearly want them to see the code... I kind of write it just like I would write
+GitHub Markdown, where you have little code snippets and text, but RCloud lets
+me actually run the snippets'' and display the results.
 
-Shelly: Is unable to use RCloud for organizational reasons but likes the
-``concept of being able to create notebooks and share them'' ``wiki is not the
+Some users who tried out RCloud but are not able to use it for organizational
+reasons, already miss the functionality. Shelly likes
+``the concept of being able to create notebooks and share them.'' A ``wiki is not the
 best way for communication results - it's kind of like writing a blog post with
-very limited functionality. can't input equations easily.'' ``have to save every
-picture and post it as an image'' ``I can't share all of the code because it
-would just get crowded and wouldn't look right on a wiki''
-
-Bart: Also unable to use for org/tech reasons. ``So if I could make a folder on
+very limited functionality... I have to save every
+picture and post it as an image'' and ``I can't share all of the code because it
+would just get crowded and wouldn't look right on a wiki.'' Bart says, ``if I could
+make a folder on
 RCloud and have Python notebooks and also Pig notebooks there, and execute them
@@
 \subsection{Forking}
-Rick: ``I fork my own notebooks because I'm going off and doing some other
+Forking is another popular feature. 
+Clint: ``I use [forking] on my own work quite a bit, and I use it on Research a
+ton. Honestly, it's one of the handiest functions, because instead of
+having to find it, copy, paste it, you just hit Fork, you rename it, and it's
+done. It's pretty amazing.''
+
+Although we always intended forking to be used to improve others' code, we
+we initially didn't support forking one's own notebooks, and this has proved very useful. Rick says ``I fork my own notebooks because I'm going off and doing some other
 analogous project, so I've sort of got interesting content that I've already
@@
 
-Clint: ``I use [forking] on my own work quite a bit, and I use it on Research a
-ton. Honestly, it's one of the handiest functions in it, because instead of
-having to find it, copy, paste it, you just hit Fork, you rename it, and it's
-done. It's pretty amazing.''
-
-Raif: ``I've been forking other people's notebooks myself, because essentially
+A somewhat more problematic and extremely common use of forking is just to change parameters. Raif: ``I've been forking other people's notebooks myself, because essentially
 they have done some work, but then I want to run it on a different part of data,
 or I want to change some parts, I don't want to see this column, I want a
-different column, things like that''
+different column, things like that.'' Horace: on working with other group using his notebooks, ``I basically point them to a few different parameters that they need to fix.''
 
-Horace: on working with other group using his notebooks, ``I basically point
-them to a few different parameters that they need to fix.'' ``I'll teach people
+But forking also provides a way for others to troubleshoot when something goes wrong.
+Horace continues, ``I'll teach people
 to intercept the result in the middle and see what you should see, and what's
-not being returned, that sort of thing'' ``If the computation went wrong in the
+not being returned, that sort of thing... If the computation went wrong in the
 middle, and something blew up, or you're just returning NULL or empty data,
-where should we check those things?'' ``We should, but a lot of the time we just
-insert print statements here and there and check values. Having debugging code
-inside a functioning program is actually helpful for others to understand what
-you're doing?''
+where should we check those things?''
 
-Shelly: ``gives the audience a way to try their own analysis, like maybe Shelly
-should have used this parameter and they can just change the parameters''
+%% ``a lot of the time we just
+%% insert print statements here and there and check values. Having debugging code
+%% inside a functioning program is helpful for others to understand what
+%% you're doing.''
+
+%% Shelly: ``gives the audience a way to try their own analysis, like maybe Shelly
+%% should have used this parameter and they can just change the parameters''
 
@@
 
-Bart: ``I like the fact that it has a built-in editor, so if you need to, like,
-fixing a typo in a link, or an extra line break or any of this other nonsense,
+Bart: ``I like the fact that it has a built-in editor, so if you need to
+fix a typo in a link, or an extra line break or any of this other nonsense,
 you can just switch to the edit view, pull up your asset, type something, it's
 automatically saved, committed, everything. You don't have to go back to your
-source code, change it, commit it to the repo, you pull the repo to your
+source code, change it, commit it to the repo, pull the repo to your
 distribution version, none of that's necessary. A lot of web development is
-fixing stupid little bugs, and you can do that kind of instantly without a lot
+fixing stupid little bugs, and you can do that instantly without a lot
 of overhead, which is nice.''
 
-\subsection{Search}
-Raif: ``the fact that we have all the notebooks there, searchable, it just saves
+\subsection{Search versus Browsing}
+Raif: ``the fact that we have all the notebooks there, searchable, it saves
 me from replicating what other people have done'' ``One way of thinking of it is
@@
 
-Rick: ``And the same kind of thing when you search through notebooks, usually I
+But global search doesn't work for everyone, and may need to be refined. Other users
+prefer to just browse the notebooks of the experts they know. Rick says, ``usually I
 know somebody’s notebooks that I want to search through, like I’ll search

### 2015-03-25 relatedwork.tex
intro and related work

@@
 \section{Related Work}
+
+There is much previous work that inspired or influenced ours.
+Indeed, many studies have addressed the requirements for future visual
+analytics environments.
+Previous work that we learned from includes:
+
+{\bf Vistrails} as a landmark prototype of a scientific workflow
+and provenance management system. This prototype helped the community
+to understand the value of capturing aspects of the processes that
+surround data analysis experiments and tools, including detailed
+history, collaboration, and deployment.
+
+{\bf iPython notebooks (Jupyter)} are the closest work we are aware of.
+The goal of this project is to provide shared notebooks and a remote
+execution environment for interpreted scripts. \stephen{Are there 
+significant differences involving versioning, provenance, annotation
+and other broad process support?}
+
+{\bf RStudio, RShiny} and other packages based on R for creating and
+publishing web content such as knitR, Rpubs, rCharts, and gg2v.
+These tools are not collaborative, but they are aimed at better
+graphics and web applications.
+
+{\bf bl.ocks, jsfiddle, plotly} and other web services for sharing code
+and demos.
+
+{ManyEyes} is the best-known original work showing the potential for
+integrating data visualization with social media.
+
+Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
+and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
+In a sense, we hope to adapt and extend this concept to the work of
+visual analytics teams.
 

### 2015-03-25 introduction.tex
summarized interviews
backslash
intro and related work

@@
 
-Almost a
-half-century ago, Tukey foresaw
+More than a half-century ago, Tukey foresaw
 much of what is now commonplace in data analysis~\cite{TukeyFDA}.
@@
 was developed essentially alongside Unix~\cite{TheSSystem}, after all.
-In this paper, we turn our attention to this opportunity to use computation
+We turn our attention to this opportunity to use computation
 to support, not only individuals, but entire teams and their
-work within larger organizations.
+work within larger organizations. In this paper, we contribute the
+design of RCloud, together with an interview study conducted over the
+course of its development and deployment at AT\&T Labs, where RCloud has
+been in use for about two years.
 
@@
 others report similar findings. Gutierrez collected interviews with data
-scientists~\cite{Gutierrez}, which include the following:
+scientists~\cite{Gutierrez:2014:DSA}, which include the following:
 
@@
 
-Over the past two years, we prototyped this environment and deployed it in a
-production community of data scientists, business analysts and other colleagues.
-The platform today has over 300 active accounts; about 20 people use it more
-than twice a week, and about 50 people use it more than once a week.  Most of
-the active users are members of AT&T Labs, but some others are data scientists
-and other collaborators in business units. \gordon{This doesn't measure how
-intensely people use it...}
-
 A key point is that, for the most part, the improved capabilities are
@@
 support research.
-
-There is much previous work that inspired or influenced ours.
-Indeed, many studies have addressed the requirements for future visual
-analytics environments.
-Previous work that we learned from includes:
-
-{\bf Vistrails} as a landmark prototype of a scientific workflow
-and provenance management system. This prototype helped the community
-to understand the value of capturing aspects of the processes that
-surround data analysis experiments and tools, including detailed
-history, collaboration, and deployment.
-
-{\bf iPython notebooks (Jupyter)} are the closest work we are aware of.
-The goal of this project is to provide shared notebooks and a remote
-execution environment for interpreted scripts. \stephen{Are there 
-significant differences involving versioning, provenance, annotation
-and other broad process support?}
-
-{\bf RStudio, RShiny} and other packages based on R for creating and
-publishing web content such as knitR, Rpubs, rCharts, and gg2v.
-These tools are not collaborative, but they are aimed at better
-graphics and web applications.
-
-{\bf bl.ocks, jsfiddle, plotly} and other web services for sharing code
-and demos.
-
-{ManyEyes} is the best-known original work showing the potential for
-integrating data visualization with social media.
-
-Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
-and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
-In a sense, we hope to adapt and extend this concept to the work of
-visual analytics teams.
 
@@
 \end{enumerate}
+
+Over the past two years, we prototyped RCloud and deployed it to a
+community of data scientists, business analysts and other colleagues.
+The platform today has over 300 active accounts; about 20 people use
+it regularly, and about 50 people use it more than once a week.  Most of
+the active users are members of AT\&T Labs, but some are data scientists
+and other collaborators in other business units. \gordon{This doesn't measure how
+intensely people use it...}
+
+\carlos{we need a segue into related work}.

### 2015-03-26 conclusion.tex
various snarky comments
junk

@@
 
-We presented a case for a data analytics environment that
-supports the visual analytics process for work by teams.
-This inclides data acquisition, exploratory data analysis,
+\stephen{Is this the kind of bad conclusion that contains
+a lot of material that should have been explained at the
+beginning of the paper to move it along faster and tell
+people why they should be interested in the work? What
+would be our conclusion?  These capabilties are important,
+but they are really hard to program, and adoption is slow
+because you're meddling with people's work process and
+toolset, which is the heart of their environment and
+not some little toy you provide out on the fringes.
+Infratsructure is hard.}
+
+We presented a case for an environment that supports the
+visual analytics process for work by teams. The process
+includes data acquisition, exploratory data analysis,
 code development and deployment.
 
-Building on previous work that defines requirements for
+Building on previous work to define requirements for
 visual analytics, we designed a prototype environment, RCloud.
@@
 
-We deployed RCloud in a community of working data scientists.
+We also deployed RCloud in a community of working data scientists.
 Experience with the RCloud prototype provides evidence that data
@@
 
-Rcloud is a step toward practical ``DevOps for data science'' and
+RCloud is a step toward practical ``DevOps for data science'' and
 reproducible, publishable experiments.

### 2015-03-27 discussion.tex
yes

@@
     \hline
-RCloud     &       v            &       v       &     v      &      v        &           v        &           v         \\
-RStudio    &                    &               &            &               &           v        &           v         \\
-JSFiddle   &       v            &       v       &            &               &                    &                     \\
-bl.ocks    &       v            &       v       &            &               &           v        &                     \\
-shiny      &                    &               &     v      &               &           v        &           v         \\
-Jupyter    &                    &               &            &      v        &           v        &           v         \\
-Tableau    &                    &               &            &               &                    &                     
+RCloud     &       x            &       x       &     x      &      x        &           x        &           x         \\
+RStudio    &                    &               &            &               &           x        &           x         \\
+JSFiddle   &       x            &       x       &            &               &                    &                     \\
+bl.ocks    &       x            &       x       &            &               &           x        &                     \\
+shiny      &                    &               &     x      &               &           x        &           x         \\
+Jupyter    &                    &               &            &      x        &           x        &           x         \\
+Tableau    &                     &      x        &    x       &               &          x         &                     
   \end{tabular}
@@
 \end{table*}
+% SN says Tableau: versioning=no  collaboration=yes  deployment=yes multilanguage=no  reports=yes  analysis=no
 

### 2015-03-27 casestudies.tex
case studies skeleton

@@
-\section{Case Study\label{sec:casestudy}}
+\section{Case Study\label{sec:casestudy}: topic visualization in RCloud}
 
-To illustrate some aspects of RCloud, we present an example notebook
-for stock price analysis.\footnote{Unfortunately this report could not
-include production notebooks containing proprietary information.}
-The example, although simplified, shows key steps in the development
-and deployment of an application.
+In this section, we present a simple, but real-world example
+application currently deployed using RCloud.
 
-\begin{figure}
-\includegraphics[width=\linewidth]{fig/casestudy1/casestudy1.pdf}
-\caption{\label{fig:stockvis}Iterations of a stock ticker
-  visualization, based on an example by Hadley Wickham. A
-  simple, static visualization of the closing price of a single stock
-  is progressively developed into a configurable display suitable
-  for a dashboard, then into an interactive visualization that
-  compares the volatility and volume of two stocks, and finally
-  into an API for data access by other RCloud notebooks. The notebooks
-  in this example can all be loaded as web pages. When a notebook
-  corresponding to a function call is displayed a web page,
-  its associated documentation is displayed.}
-\end{figure}
+This example application is an implementation of LDAVis, a
+recently-published method for visualizing large collections of
+text. It was specifically designed to enable non-expert users to
+explore collections of short text documents using topic modeling and
+visualization. Topic modeling~\cite{Blei:2003:LDA} is a standard
+technique for text summarization which, although powerful, requires a
+certain amount of manual intervention and
+interpretation~\cite{Siever:2014:LAM}.
 
-It shows a sequence of visualizations of the performance
-of financial stocks over several years. The first visualization
-in this example uses ggplot2 and was written by Hadley Wickham.
-It reads data provided by Yahoo! Finance as a web service, and shows
-the price for a single trading symbol.
+\begin{figure*}
+  \includegraphics[width=\linewidth]{fig/casestudytext/casestudytext.pdf}
+  \caption{\label{fig:textvis}An example application developed and deployed in RCloud.}
+\end{figure*}
 
-The webpage that produces that visualization includes a link to
-the underlying source code. From this link, a user can fork the notebook.
-In the example,a configurable ticker is added, based on the URL of
-the notebook. The change to the notebook is minor.
+LDAVis itself was developed by two members of the technical staff at
+AT\&T Labs, and originally was developed entirely in RStudio's
+Shiny~\cite{RStudio:2013:SWA}, a framework for developing R
+applications for the web. The main reason for moving the application
+to an RCloud development was precisely ease of deployment. Shiny
+provides ease of development, but when the application is finished,
+one needs to find a publicly available web server that can host the
+Shiny application; user management needs to be handled, and so
+on. Using RCloud, deployment happens transparently. This is a big win
+in terms of efficiency.
 
-Next, the notebook author or a collaborator may decide to extend
-the application to also provide convenient access to the data,
-apart from its visualization.
-This is done by creating a notebook that defines a function.
-That notebook then becomes a \emph{subroutine} for other notebooks.
-It can be invoked from interactive notebooks, such as dashboards.
-The data access notebook is version-controlled like other notebooks.
+Describe the application and the technologies. Runs code from an R
+package and custom analysis code; files are uploaded from the web
+client, results are provided in JSON format and visualized in the web
+client using d3~\cite{Bostock:2011:DDD}.
 
-Finally, consider the situation where an analyst wants to understand
-the price dynamics of stocks with respects to other attributes and
-time ranges. For this, interactive visualization may be very helpful.
-In our example, the analyst creates an interactive tool with multiple
-linked views.
+%% Text analysis in
 
-Demonstrating the integration of R and JavaScript, here the analyst
-writes an R function that passes the dataframe and description of
-the charts to the dcplot library.  Simple R expressions
-are captured as trees to generate JavaScript expressions.
-The terse chart description language, with sensible defaults inspired by
-ggplot2, provides a simple yet powerful interface to the grouping
-and reduction functionality of the well-accepted charting libraries
-crossfilter, dc.js and d3.js.
+%% To illustrate some aspects of RCloud, we present an example notebook
+%% for stock price analysis.\footnote{Unfortunately this report could not
+%% include production notebooks containing proprietary information.}
+%% The example, although simplified, shows key steps in the development
+%% and deployment of an application.
+
+%% \begin{figure}
+%% \includegraphics[width=\linewidth]{fig/casestudy1/casestudy1.pdf}
+%% \caption{\label{fig:stockvis}Iterations of a stock ticker
+%%   visualization, based on an example by Hadley Wickham. A
+%%   simple, static visualization of the closing price of a single stock
+%%   is progressively developed into a configurable display suitable
+%%   for a dashboard, then into an interactive visualization that
+%%   compares the volatility and volume of two stocks, and finally
+%%   into an API for data access by other RCloud notebooks. The notebooks
+%%   in this example can all be loaded as web pages. When a notebook
+%%   corresponding to a function call is displayed a web page,
+%%   its associated documentation is displayed.}
+%% \end{figure}
+
+%% It shows a sequence of visualizations of the performance
+%% of financial stocks over several years. The first visualization
+%% in this example uses ggplot2 and was written by Hadley Wickham.
+%% It reads data provided by Yahoo! Finance as a web service, and shows
+%% the price for a single trading symbol.
+
+%% The webpage that produces that visualization includes a link to
+%% the underlying source code. From this link, a user can fork the notebook.
+%% In the example,a configurable ticker is added, based on the URL of
+%% the notebook. The change to the notebook is minor.
+
+%% Next, the notebook author or a collaborator may decide to extend
+%% the application to also provide convenient access to the data,
+%% apart from its visualization.
+%% This is done by creating a notebook that defines a function.
+%% That notebook then becomes a \emph{subroutine} for other notebooks.
+%% It can be invoked from interactive notebooks, such as dashboards.
+%% The data access notebook is version-controlled like other notebooks.
+
+%% Finally, consider the situation where an analyst wants to understand
+%% the price dynamics of stocks with respects to other attributes and
+%% time ranges. For this, interactive visualization may be very helpful.
+%% In our example, the analyst creates an interactive tool with multiple
+%% linked views.
+
+%% Demonstrating the integration of R and JavaScript, here the analyst
+%% writes an R function that passes the dataframe and description of
+%% the charts to the dcplot library.  Simple R expressions
+%% are captured as trees to generate JavaScript expressions.
+%% The terse chart description language, with sensible defaults inspired by
+%% ggplot2, provides a simple yet powerful interface to the grouping
+%% and reduction functionality of the well-accepted charting libraries
+%% crossfilter, dc.js and d3.js.
 

### 2015-03-27 relatedwork.tex
fig 2
working

@@
 
+\stephen{In addition to the immense foundation of work on systems
+for statistical computing,}
 There is much previous work that inspired or influenced ours.
-Indeed, many studies have addressed the requirements for future visual
+Indeed, many studies have addressed requirements for future visual
 analytics environments.
@@
 
-{\bf Vistrails} as a landmark prototype of a scientific workflow
-and provenance management system. This prototype helped the community
-to understand the value of capturing aspects of the processes that
+{\bf VisTrails} as a landmark prototype of a scientific workflow
+and provenance management system. This prototype demonstrated
+the value of capturing aspects of the processes that
 surround data analysis experiments and tools, including detailed
@@
 The goal of this project is to provide shared notebooks and a remote
-execution environment for interpreted scripts. \stephen{Are there 
-significant differences involving versioning, provenance, annotation
-and other broad process support?}
+execution environment for interpreted scripts. 
+\stephen{Are there significant differences involving versioning,
+provenance, annotation and other broad process support?}
 
@@
 and demos.
-
-{ManyEyes} is the best-known original work showing the potential for
-integrating data visualization with social media.
 
@@
 drives or email. Analysts rarely stored their analytic code in source
-control.'' Their work points to the opportunity for better technology
-to support collaboration and sharing in data analysis teams.
+control.'' Their study highlights the opportunity for better ways
+of supporting collaboration and sharing in data analysis teams.
 
@@
 analysis and visualization~\cite{Kandel:2011:RDI}. We view this
-as attacking a different semantic level than ours, but points
-to the need for an environment that enables better sharing
+as attacking a different semantic level than ours, but also
+showing the need for an environment that enables better sharing
 of the knowledge, tools and processes to do this. Anecdotally
 we find much frustration among practitioners that this knowledge
-is difficult to find and often not recorded or provided in a reusable
-form even within one organization.
+is difficult to find and often isnot recorded or available in a
+reusable form even within the same organization.
 
@@
 
-Manyeyes \cite{Viegas:2007:MAS} was a landmark system for the integration
+{\bf Manyeyes} \cite{Viegas:2007:MAS} was a landmark system for the integration
 of social media with visualization and data publishing. We build on this
@@
 environments \cite{Guo:2010:TPI}.
-Our present goal is to define an environment in which many such
+The goal of RCloud is to define an environment in which many such
 techniques may be integrated and made available to a broad community.
 
-VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
+{\bf VisMashup}~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
 are tools built on top of the VisTrails workflow management
@@
 in the process of creating experiments, we may be able to eliminate the need
-for programming teams to recreate their work in production which has a high
-penalty in time, expense and sometimes correctness.
+for programming teams to recreate their work in production, which has a high
+penalty in time, expense and accuracy.

### 2015-03-27 discussion.tex
working

@@
 considered both within and across collections of scripts (the latter being
-similar to Vistrails' enhanced recommendations by clustering multiple
+similar to VisTrails' enhanced recommendations by clustering multiple
 workflows).  It seems valuable to know which packages are frequently used
@@
 Though caching is usually implicit, in some environments, such as
-Vistrails, programmers may also have some explicit control over cache
+VisTrails, programmers may also have some explicit control over cache
 management. This may be desirable to ensure the repeatability of

### 2015-03-27 relatedwork.tex
yes

@@
-\begin{figure}
-\includegraphics[width=\linewidth]{fig/system/system.pdf}
-\caption{\label{fig:system}A diagram of RCloud's architecture. Dashed
-  lines represent features which not yet implemented. \carlos{Must remove anything that isn't implemented}}
-\end{figure}
+\begin{figure*}
+  \centering
+\includegraphics[width=.75\linewidth]{fig/system/system.pdf}
+\caption{\label{fig:system}A diagram of RCloud's architecture. }
+\end{figure*}
 
@@
 
-\stephen{In addition to the immense foundation of work on systems
-for statistical computing,}
 There is much previous work that inspired or influenced ours.
@@
 
+\stephen{We have a problem we already went over some of this.}
 Our work has been inspired by and benefited from other efforts
-to improve data analysis environments and processes,
-including RStudio \cite{RStudio:2013:SWA},
+to improve data analysis environments and processes. Building in
+the immense foundation of systems for statistical graphics,
+recent work includes RStudio \cite{RStudio:2013:SWA},
 the R packages Markdown \cite{Allaire:2014:MMR},
@@
 One overall goal is to reduce the gap between implementers and deployers
-of technology in visual analytics. The fusion of development with production
+in visual analytics. The fusion of development with production
 operations in software release management (``DevOps'' \cite{Httermann:2012:DD}
@@
 By making it convenient for data scientists to expend just a little more effort
-in the process of creating experiments, we may be able to eliminate the need
-for programming teams to recreate their work in production, which has a high
-penalty in time, expense and accuracy.
+when creating experiments, we may be able to eliminate the need for programming
+teams to recreate their work to deploy it in production,
+which has a high cost in time, expense and accuracy.

### 2015-03-27 introduction.tex

@@
 
-Over the past two years, we prototyped RCloud and deployed it to a
-community of data scientists, business analysts and other colleagues.
-The platform today has over 300 active accounts; about 20 people use
-it regularly, and about 50 people use it more than once a week.  Most of
-the active users are members of AT\&T Labs, but some are data scientists
-and other collaborators in other business units. \gordon{This doesn't measure how
-intensely people use it...}
+Over the past two \gordon{three?} years, we prototyped RCloud and deployed it to
+a community of data scientists, business analysts and other colleagues.  The
+platform today has over 300 active accounts; about 20 people use it regularly,
+another 30 people use it more than once a week, and another 50 use it more than
+once a month.  There are over 5000 notebooks, of which more than 500 have been
+starred and more than 350 have been forked. \gordon{Is this the right place for
+the notebook stats? We haven't defined forking or starring yet.}  Most of the
+active users are members of AT\&T Labs, but some are data scientists and other
+collaborators in other business units.
 

### 2015-03-28 interviews.tex
further narrate the interviews

@@
 analysts and business needs.
-
-\gordon{Add user statistics here: number of notebooks, forks, stars, users
-who have done one of these things. Number of users who use RCloud a lot,
-a moderate amount, or just tried it out.}
 
@@
 provide a starting point. Clint says, ``I think the best part about it is how
-easily you can share code... you can find a working example, rather than wearing
+easily you can share code [...] you can find a working example, rather than wearing
 out Google and finding questionable examples that may or may not work.'' Raif
@@
 much package development. I want people to see how the package works, so I
-clearly want them to see the code... I kind of write it just like I would write
+clearly want them to see the code [...] I kind of write it just like I would write
 GitHub Markdown, where you have little code snippets and text, but RCloud lets
@@
 best way for communication results - it's kind of like writing a blog post with
-very limited functionality... I have to save every
+very limited functionality [...] I have to save every
 picture and post it as an image'' and ``I can't share all of the code because it
@@
 \subsection{Forking}
-Forking is another popular feature. 
-Clint: ``I use [forking] on my own work quite a bit, and I use it on Research a
-ton. Honestly, it's one of the handiest functions, because instead of
-having to find it, copy, paste it, you just hit Fork, you rename it, and it's
-done. It's pretty amazing.''
+The ability to start work where someone else let off, by forking, has proved to
+be a popular feature. In fact, almost twice as many (131) users have forked
+someone else's notebook as have starred (75).  Clint: ``I use [forking] on my
+own work quite a bit, and I use it on Research a ton. Honestly, it's one of the
+handiest functions, because instead of having to find it, copy, paste it, you
+just hit Fork, you rename it, and it's done. It's pretty amazing.''
 
-Although we always intended forking to be used to improve others' code, we
-we initially didn't support forking one's own notebooks, and this has proved very useful. Rick says ``I fork my own notebooks because I'm going off and doing some other
-analogous project, so I've sort of got interesting content that I've already
-done in a previous analysis that I want to start from and then tweak to match a
-new set of data.''
+Although we always intended forking to be used to improve others' code, we we
+initially didn't support forking one's own notebooks, and this has proved very
+useful. Rick says ``I fork my own notebooks because I'm going off and doing some
+other analogous project, so I've sort of got interesting content that I've
+already done in a previous analysis that I want to start from and then tweak to
+match a new set of data.''
 
-A somewhat more problematic and extremely common use of forking is just to change parameters. Raif: ``I've been forking other people's notebooks myself, because essentially
-they have done some work, but then I want to run it on a different part of data,
-or I want to change some parts, I don't want to see this column, I want a
-different column, things like that.'' Horace: on working with other group using his notebooks, ``I basically point them to a few different parameters that they need to fix.''
-
-But forking also provides a way for others to troubleshoot when something goes wrong.
-Horace continues, ``I'll teach people
-to intercept the result in the middle and see what you should see, and what's
-not being returned, that sort of thing... If the computation went wrong in the
-middle, and something blew up, or you're just returning NULL or empty data,
-where should we check those things?''
+Forking also provides a way for others to troubleshoot when something goes
+wrong.  When Horace works with the users of his notebooks, ``I'll teach people to intercept the result in the
+middle,'' to ``insert print statements here and there and check values''
 
@@
 
+An extremely common but more problematic use of forking is to
+change parameters. Raif: ``I've been forking other people's notebooks [because]
+I want to run it on a different part of data, or I want to change some parts, I
+don't want to see this column, I want a different column, things like that.''
+
+Our users, especially those who build general purpose tools, see a need for better ways
+to parameterize notebooks.  (Currently parameters can be added to the URL, but adding
+user interface elements to do the same thing requires expertise.) Taylor complains that
+a notebook might say `` `This is a report of the volume of all of
+our feeds for this month', and someone would want to look at it for the next
+month or the previous month, so they'd fork it to change the month.''
+
+Jyothi calls it web-enabling the notebook: users ``can always fork the
+notebooks and make changes, but I feel that if the owner of the notebook
+web-enables it, then whoever is running it, it's easier to run. Instead of
+forking it, if they can set options. It's probably more efficient, and also they
+can [still] fork it if they want to.''
+
+%% Clint: ``For the things that I'll do multiple times, I'll try to parameterize
+%% it, and I'm working now on putting some packages together that will operate in
+%% some fashion outside of RCloud, straight from the command line''
+
 \subsection{Automatic source control}
-Rick: ``when I've finished something, there's a nice clean record of it.''
+%% Rick: ``when I've finished something, there's a nice clean record of it.''
+%% (I think this may be more about the publish/markdown than about source control.)
 
@@
 this subdirectory and I hope I've got them backed up, if they're on RCloud I
-know they are'' ``I don't have to worry about it if I drop my computer in the
+know they are [...] I don't have to worry about it if I drop my computer in the
 bathtub, everythings gone, because I'm not great about uploading my own for
@@
 
-Bart: ``I like the fact that it has a built-in editor, so if you need to
-fix a typo in a link, or an extra line break or any of this other nonsense,
-you can just switch to the edit view, pull up your asset, type something, it's
-automatically saved, committed, everything. You don't have to go back to your
-source code, change it, commit it to the repo, pull the repo to your
-distribution version, none of that's necessary. A lot of web development is
-fixing stupid little bugs, and you can do that instantly without a lot
-of overhead, which is nice.''
+Bart points out that the automatic versioning works well for dealing with the
+minutiae of web development. ``I like the fact that it has a built-in editor, so
+if you need to fix a typo in a link, or an extra line break or any of this other
+nonsense, you can just switch to the edit view, pull up your asset, type
+something, it's automatically saved, committed, everything. You don't have to go
+back to your source code, change it, commit it to the repo, pull the repo to
+your distribution version. None of that's necessary. A lot of web development is
+fixing stupid little bugs, and you can do that instantly without a lot of
+overhead, which is nice.''
 
-\subsection{Search versus Browsing}
-Raif: ``the fact that we have all the notebooks there, searchable, it saves
-me from replicating what other people have done'' ``One way of thinking of it is
-like Google Code, you know you just search for something and you find something
-useful and you just use it. It’s a repository of code''
+On the other hand, saving every change leads to many fine-grained
+versions. Rick says that for this reason, the history feature is not all that
+helpful: ``I don't need something that keeps track of every mistake I've made or
+every direction I've tried.'' Tagging these versions with names wouldn't
+necessarily help because ``namespaces are already crowded, so trying to remember
+the names of notebooks is hard, much less the names of states within each
+notebook.''
 
-But global search doesn't work for everyone, and may need to be refined. Other users
-prefer to just browse the notebooks of the experts they know. Rick says, ``usually I
-know somebody’s notebooks that I want to search through, like I’ll search
-through Simon Urbanek’s first, then I’ll go off to somebody else’s, because I
-know that the kind of thing I’m looking for something more obscure than I’m
-likely to find in some random person’s thing''
+\subsection{Discovering others' work}
+Many users find the search function useful. Raif says ``the fact that we have
+all the notebooks there, searchable, saves me from replicating what other people
+have done. [...] One way of thinking of it is like Google Code: you just search
+for something and you find something useful and you just use it.''
+
+But other users prefer to browse just the notebooks of the experts they
+know. Rick says, ``usually I know somebody's notebooks that I want to search
+through, like I'll search through Simon Urbanek's first, then I'll go off to
+somebody else's, because I know that the kind of thing I'm looking for something
+more obscure than I'm likely to find in some random person's.''
+
+More selective ways to search will be needed as the corpus of notebooks
+continues to grow. We are considering tagging notebooks filtering the notebook
+tree by tag, measuring the quality of notebooks by their forks as well as stars,
+and searching within a single user's notebooks.
 
 \subsection{Web integration}
-Horace ``that also means the other guys who want to do analytics on the data,
-they can first pull the data, do the analytics on it, and then feed the viewer
-the data. So if anyone from the stats group wants to insert some code in between
-the data pulling procedure, [they] can always insert something in between. We
-were trying to make such an example so that they don't have to rewrite the GUI
-part and they know how to get the data source, once you get the data into
-RCloud, then you have a dataframe to work with, and then they know how to
-produce another dataframe. We give them the specification of the dataframes, we
-give them the data input, they do whatever they want, they can reuse the data
-connection part and the GUI part. We can keep changing these components and
-that's a very nice thing.''
+Integrating an analysis language into a web development environment is something
+the tool developers really appreciate.  Horace: ``that also means the other guys
+who want to do analytics on the data, they can first pull the data, do the
+analytics on it, and then feed the viewer the data. So if anyone from the stats
+group wants to insert some code in between the data pulling procedure, [they]
+can always insert something in between. We were trying to make such an example
+so that they don't have to rewrite the GUI part and they know how to get the
+data source, once you get the data into RCloud, then you have a dataframe to
+work with, and then they know how to produce another dataframe. We give them the
+specification of the dataframes, we give them the data input, they do whatever
+they want, they can reuse the data connection part and the GUI part. We can keep
+changing these components and that's a very nice thing.''
 
@@
 applications need to be doing something with the data, so having an open session
-where I can run R command or R functions without having to invoke an API or send
-out a request and then wait for the request to come back to do the results.  So
+where I can run R commands or R functions without having to invoke an API or send
+out a request and then wait for the request to come back to do the results,
 that kind of seamless integration is extremely helpful in writing the
@@
 
-Jyothi: ``another thing I really like about RCloud is your publish feature, so
-that non-RCloud users, the anonymous login, non-RCloud users can access the
-notebooks without having to login to RCloud''
+%% Jyothi: ``another thing I really like about RCloud is your publish feature, so
+%% that non-RCloud users, the anonymous login, non-RCloud users can access the
+%% notebooks without having to login to RCloud''
 
-accidental feature: ``some of my jobs are huge and I like to just call them and
-the fact that the user can just shut down his browser or shut down his computer
-and the job still runs in the background is actually a very neat thing. So what,
-um, the way I have coded is, whenever I have a huge job to run, I just display
-to the user, you go do whatever you want, we'll send you an email when the job
-is done, um that's another feature of RCloud that's very helpful, that the job
-will continue to run in the background''
+%% accidental feature: ``some of my jobs are huge and I like to just call them and
+%% the fact that the user can just shut down his browser or shut down his computer
+%% and the job still runs in the background is actually a very neat thing. So what,
+%% um, the way I have coded is, whenever I have a huge job to run, I just display
+%% to the user, you go do whatever you want, we'll send you an email when the job
+%% is done. That's another feature of RCloud that's very helpful, that the job
+%% will continue to run in the background''
+
+\gordon{Not sure this belongs here}
+Running anything on the Web can also be a source of frustration.  Poor network
+links can cause web sockets to die, and Corey complains ``What frustrates me is
+that if I do leave this window for what feels like 5 or 7 minutes, it
+disconnects and I have to reconnect and run the whole notebook again.'' Raif
+whines that ``being web-based, it's obviously slower than something you would be
+running on your machine.'' For Byers: even the little lag of sending a command
+to a server and receiving the result is intolerable
 
 \subsection{Other workflows}
-Rick: ``I find that the style of editing is clunky. I would never edit like that
+Although most of the analysts we spoke to use and appreciates the sharing
+features, RCloud is less popular as a tool for exploratory data analysis and
+development.
+
+Much of this is because of familiarity with other tools. Horace: ``for other
+people who use RCloud vs RStudio, they will always think that they want to
+prototype on RStudio. Maybe because it's more familiar for them. So, they work
+on RStudio, on a small data set, and they fix up everything, make sure
+everything is right, and then they place it on RCloud for sharing.''
+
+Rick says, ``I find that the style of editing is clunky. I would never edit like that
 if I'm doing ordinary work. I usually use vi.'' ``I mean it's nice to have it
-saved, but you know, there's sort of this trade-off between it making it easier
-for me to present something or to save something, and my easy of typing and
-correcting and things in a plain old editor window.'' ``I just have the
-discipline of having a file that describes what I'm doing with the commands that
-I'm using so that I can go back and recreate it or pass it on to someone else,
-so it's a little bit like having an RCloud notebook, it's not necessarily
-executable but most of the commands that I've typed are in there.''
+saved, but there's this trade-off between it making it easier for me to present
+something or to save something, and my ease of typing and correcting and things
+in a plain old editor window.''
 
-Kenny: the UI doesn't work for his workflow, which is pasting snippets from a
-file into the console in the R GUI.  he tends to type type type into the file,
-paste paste paste to the console, rearrange so the good code is at the top of
-the file, then save the file and call it quits once he's got the right code at
-the top of the file.
+Many of the analysts we spoke to preferred using a text editor to refine and
+store their work, pasting it to an R command line to try it out. Rick describes,
+``I have the discipline of having a file that describes what I'm doing with the
+commands that I'm using so that I can go back and recreate it or pass it on to
+someone else, so it's a little bit like having an RCloud notebook, it's not
+necessarily executable but most of the commands that I've typed are in there.''
 
-Taylor: ``The big one is I don't know why would I want to use RCloud over my
-current setup. If it's just me, I like my text editor and terminal. There's
-nothing that I want that those two don't give me'' ``In R, when I'm doing
-development work, I copy and paste in like 5-10 lines of code, so when something
-breaks, I get an error message on that one line, and I can up-arrow and change
-it and fix it, whereas in RCloud I have to run a whole cell, so the only way to
-get that same functionality is if every line's in one cell. Because if it runs
-like lines 1,2,3, then 4 fails, I have to either run the whole cell again, or
-have line 4 in its own cell. I can't rerun just one line.'' ``There's nothing
-that I want that's not in the terminal. I don't think you should try to support
-all the things I want to do in the terminal.''
+Kenny works much the same way, rearranging the file so the good code is at the top of
+the file. Then he saves the file once he's got the right code at
+the top of the file (and often shares the work by pasting it into an RCloud notebook).
 
-Horace: ``It's not for myself, but for other people who use RCloud vs RStudio,
-they will always think that they want to prototype on RStudio. Maybe because
-it's more familiar for them. So, they work on RStudio, on a small data set, and
-they fix up everything, make sure everything is right, and then they place it on
-RCloud for sharing.''
+Taylor asks, ``why would I want to use RCloud over my
+current setup? If it's just me, I like my text editor and terminal. There's
+nothing that I want that those two don't give me.''
 
-Byers: installing random packages is part of exploratory data analysis
+Working in a shared environment also entails compromises about what you can
+install. Byers says that installing random packages is part of exploratory data
+analysis.
 
 \subsection{Proliferation of notebook cells}
-Rick: ``Making new cells and taking some stuff in it, then having it do
-something, then having to make a new cell, and I don't know, it just gets in my
-way, so if I'm doing really straightforward simple stuff, I'll just get into R
-and do it myself.'' ``I'm always scrolling up and down the screen'' ``every time
-I type more stuff, the notebook gets longer and longer and it's harder to deal
-with'' ``I've got to do a lot of scrolling around to get to where I'm going,
-whereas if I'm doing it in vi, I can search for things in my typing, and get to
-places I want to be without having to scroll through it and try to read it''
-``times when I just want to type in a couple of quick and get some results that
-are going to tell me what to do next, and they're not necessarily archival in
-any sense'' ``being literate about anything typically takes a lot of rewriting
-and going back over things... I just want to explore a few things and then I'll
-know what I want to write.'' ``I'll often type in expressions that allow me to
-check that I'm the right track, so I know that if the computation worked out,
-then the sum of the residuals is gonna be zero. So I'll type in sum(resid) and
-if it prints out zero, then I'm all happy, but I don't want that in my
-notebook.'' ``it tempts me to type a big long thing and then run the whole
-thing, as opposed to, type a few little pieces that I'm thinking of on the inner
-parts and then put them together to make the big thing''
+RCloud's notebook interface combines editable Markdown with a command line
+interface.
 
-Ivan has the complaint that too many cells are created to use the notebook as a
-console.  This has a lot to do with the extra real estate that gets taken up by
-the cells controls, blank space, etc of cells versus a straight command line.
+The interface takes some learning. Clint says ``it took me a couple of weeks of
+looking at it to become comfortable because I saw the cells and it just threw me
+for a total loop. I mean, it's a good idea because [...] I can run it all at
+once if I want or I can break these into sections for either debugging or
+staging purposes. I really like it now, but when I first saw it, [it was] very
+confusing [...] I think I prefer it now, because you can cut your code up.''
 
-KC and Ivan both complained that when there are long results or plots, it causes
-the code to scroll off the screen.
+Just the difference between a cell and the command prompt can confuse and slow
+down users. Rick: ``if I was typing into one of the cells near that top, I had
+to think `I'm editing this cell and then I have to execute it', and if was doing
+the one at the bottom I could type it but then it would automatically execute
+[by pressing enter], but then it became a standard cell and I have to edit
+it. It was just another set of modes that I didn't like''
 
-Kenny: ``It saves everything I do like everything is gold, but most of it is
-junk not meant to be saved'' ``the console is purely disposable''
+For some users, the cells getting saved feels incompatible with exploratory data
+analysis. Rick notes that Markdown is a form of literate programming, and
+``being literate about anything typically takes a lot of rewriting and going
+back over things [...] I just want to explore a few things and then I'll know
+what I want to write.''
 
-Raif: ``whatever you type on the command line, it becomes a part of your
-notebook, and I find it a little bit annoying, because I really type a lot of
-things to check my data, and I don't want that to become a part of the
-notebook'' ``Like you're kind of checking out what your data is, or you make a
-plot of the data. Things that should not really become part of a notebook, but
-things that help you understand your data better.''
+Much of the time, commands typed at the R command line just serve a transitory
+purpose, so having RCloud save commands into cells bothers experienced
+users. Kenny notes, ``It saves everything I do like everything is gold, but most
+of it is junk not meant to be saved.'' Rick says ``I just want to type in a
+couple of quick and get some results that are going to tell me what to do next,
+and they're not necessarily archival in any sense''
 
-Horace: ``Well, the nice thing is if you have the whole notebook, you can run it
-step by step, and try to mess around in between, interactively.''
+The material that is not appropriate to save includes ``expressions that allow
+me to check that I'm the right track'' (Rick), ``checking out what your data is,
+or you make a plot of the data. Things that should not really become part of a
+notebook, but things that help you understand your data better.'' (Raif)
+
+Users also felt that cells do not capture the right level of granularity: they
+either hold too much code or too little. Rick says that RCloud's cells structure
+``tempts me to type a big long thing and then run the whole thing, as opposed to
+typing a few little pieces that I'm thinking of on the inner parts and then put
+them together to make the big thing'' as he would do on the command line. When
+Taylor uses the R command line, he copies and pastes ``like 5-10 lines of code,
+so when something breaks, I get an error message on that one line, and I can
+up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell,
+so the only way to get that same functionality is if every line's in one
+cell. Because if it runs lines 1, 2, 3, then 4 fails, I have to either run the
+whole cell again, or have line 4 in its own cell. I can't rerun just one line.''
+
+Cells can also take up substantial vertical space, requiring scrolling. Rick
+complains that ``every time I type more stuff, the notebook gets longer and
+longer and it's harder to deal with.'' Ivan says that cells' controls and blank
+space take extra real estate compared to a straight command line, and he and
+KCKC both complained that when there are long results or plots, it causes the
+code to scroll off the screen. Raif: ``for example imagine I run a cell whose
+output is huge, and now that output would block the whole window, and in order
+to find the next cell, I have to scroll down, and find where that cell starts,
+right? So, I'm losing the continuity of my code to the output''
+
+Some users think users should have the option to keep results separate from the
+code.  Ivan thinks that RStudio's layout is more helpful because the charts
+are kept in another pane and stay in one place while doing analysis. Raif says
+``cells are really useful'' but ``you want to see your output in a different
+window [or] on a separate part of the window''.
+
+The cell structure also can be problematic if some cells take a long time to
+run. In Kenny's work, there is often a ``long tail''.  The first cell may take
+seconds to run, the next cell minutes, and the last cell 5 hours.  In this
+situation, the ``Run whole notebook'' button is ``dangerous''. When converting
+his work to notebooks, Kenny ends up with a lot of comments that say ``this cell
+takes a long time to run.'' This is also a huge
+issue for demoing code.
+
+The solution is probably something like knitr's caching of results, which Kenny
+is doing in an ad-hoc way, ``littering the notebook with little switches that
+comment out'' the slow parts and ``load the object instead'', ``save the object
+to disk''. One of the authors of this paper manually writes cells that check if
+a result file exists, and perform the calculation only if it's not there.
 
@@
 
-\subsection{Misc. notebook interface annoyance}
-Rick: ``if I was typing into one of the cells near that top, I had to think of
-`I'm editing this cell and then I have to execute it', and if was doing the one
-at the bottom I could type it but then it would automatically execute, but then
-it became a standard cell and I have to edit it. It was just another set of
-modes that I didn't like''
+RCloud combines the functions of scratchpad with notebook assets. KC thinks
+that having a scratch pad that doesn't persist between different notebooks is
+``useless''.
 
-Clint: ``it took me a couple of weeks of looking at it to become comfortable
-because I saw the cells and it just threw me for a total loop. I mean, it's a
-good idea because it's essentially like I can have this part and this part and
-this part. I can run it all at once if I want or I can break these into sections
-for either debugging or staging purposes which, I really like it now, but when I
-first saw it, [it was] very confusing... I think I prefer it now, because you
-can cut your code up.''
+%% The one-window design also makes some users feel that the side panes are taking
+%% up too much space. Ivan: having ``long term'' code (which would be in assets)
+%% above the command prompt, like in RStudio, means that both can be sized very
+%% wide to accomodate long lines.
 
-Corey: In the previous version, it switched between code and results. Now you
-see both - ``and there are times when that's good, but there's certainly times
-when it'd be nice to - you know, you have 10 or 15 cells and you just want to
-get back and there's all this output and you're scrolling all over the place so,
-if there were a way to control both, that'd be great. There are times when I do
-want to see the output, but there are times when I've seen enough of the same
-thing.''
+%% Kenny: Problem with screen real estate: side panes take up too much
+%% space. Doesn't need the notebook tree pane.
 
-Jyothi: ``It's good that I can see the results but if I'm working and I need
-more room and I want to hide the results, I can't.'' Could we have ``an option
-to close the results window?''
+%% Corey: In the previous version, it switched between code and results. Now you
+%% see both - ``and there are times when that's good, but there's certainly times
+%% when it'd be nice to - you know, you have 10 or 15 cells and you just want to
+%% get back and there's all this output and you're scrolling all over the place so,
+%% if there were a way to control both, that'd be great. There are times when I do
+%% want to see the output, but there are times when I've seen enough of the same
+%% thing.''
 
-Ivan: RStudio's layout is more helpful because - the charts are kept in another
-pane and stay in one place while doing analysis - having ``long term'' code
-(which would be in assets) above the command prompt means that both can be sized
-very wide to accomodate long lines
-
-Raif: ``cells are really useful'' but ``you want to see your output in a
-different window, on a separate part of the window'' ``for example imagine I run
-a cell whose output is huge, right, and now that output would block the whole
-window, and in order to find the next cell, I have to scroll down, and find
-where that cell starts, right? So, I'm losing the continuity of my code to the
-output''
-
-Kenny: Problem with screen real estate: side panes take up too much
-space. Doesn't need the notebook tree pane.
-
-KC thought that having a scratch pad that doesn't persist between different
-notebooks is ``useless''.
-
-Kenny: Problem of long-running cells.  ``The long tail.''  First cell takes
-seconds to run, next cell minutes, next cell 5 hours.  In this scenario, which
-is common for Kenny, the ``Run whole notebook'' button is dangerous. This is
-also a huge issue for demoing code.  Suggests something like knitr's caching of
-results. When converting his work to notebooks, Kenny ends up with a lot of
-comments that say ``this cell takes a long time to run.'' He ends up ``littering
-the notebook with little switches that comment out'' the slow parts and ``load
-the object instead'', ``save the object to disk''.
+%% Jyothi: ``It's good that I can see the results but if I'm working and I need
+%% more room and I want to hide the results, I can't.'' Could we have ``an option
+%% to close the results window?''
 
 \subsection{A sea of notebooks}
+RCloud is starting to be a victim of its own success, as there are now too many
+notebooks.  Rick doesn't find the notebook tree very useful for this reason: ``I
+don't necessarily need to see everybody's notebook that uses RCloud'' ``every
+time I do something new, I get a new notebook and so now my notebooks are maybe
+50 or 60. That's enough to think about just on my own, but if I've got
+everybody's 50 or 60 sitting on my display, I find that it's more than I want to
+know about.''
 
-Byers: Problem with forking a notebook and correcting it, is that the original
-notebook still exists with the error.  We need ways to see that there is a more
-recent fork of notebook and to fetch in the changes from it.
+Although RCloud promises an environment where notebooks stay working, our users
+have not all learned the habits that make this a reality. As Byers puts it,
+there is still a big problem with ``bitrot'' - notebooks often stop working
+because the user changed the structure of their data, changed a filename or a
+database.  He says we need ``organizational protocols'' to catch up with the
+technology.
 
-Taylor: ``I'll make a notebook of a little tutorial of, here's the basic
-functions and what they do, and then I'll send them out, and then people will
-give me feedback on them, or have some new feature, or I'll realize something's
-wrong, and then I'll go back and change it and update that page, but then what
-happens is people have in the meantime, forked it, and then they have a version
-that... and usually what it'll be is actually that I'll have to change a package
-as well, so their old fork stops working, and then they complain, and then, oh
-no, you have to go get the original [notebook]. Yeah, that's the sort of... This
-probably bites me more than other people, because I'm doing development so
-things are changing as they're looking at the code. Like, iotools, a lot of the
-functions change names or change syntax.''
+There are technical problems to be solved here as well. Byers continues that the
+problem with forking a notebook and correcting it, is that the original notebook
+still exists with the error.  We need a way to see that there is a more recent
+fork of notebook and to fetch in the changes from it.
 
-Taylor: ``That's just a problem because people will have old notebooks that
+Taylor, who writes tools and example notebooks for those tools, seems to be
+haunted the most by dead notebooks.  At first, he had to learn how to write safe
+notebooks: ``It would be a file inside a directory that only I had [...] access
+to and so someone else couldn't run it, [...] a temporary file on something that
+was read-only to everyone else. But I've gotten careful about that.''
+
+But even correcting notebooks can come too late: ``I'll make a notebook of a
+little tutorial of basic functions and what they do, and then I'll send [it]
+out, and then people will give me feedback on [it], or have some new feature, or
+I'll realize something's wrong, and then I'll go back and change it and update
+that page, but then what happens is people have in the meantime forked it. [...]
+Usually what it'll be is actually that I'll have to change a package as well, so
+their old fork stops working, and then they complain. [...] This probably bites
+me more than other people, because I'm doing development so things are changing
+as they're looking at the code. Like, iotools, a lot of the functions change
+names or change syntax.''
+
+Taylor tried to keep his notebook tree tidy, but this didn't help either, because
+notebooks still exist when they have left the tree. ``That's just a problem because people will have old notebooks that
 don't work anymore. It's annoying when it's a notebook that I still have and the
@@
 doesn't exist anymore on my thing.'' ``Now I'm really gun-shy about sharing
-notebooks because, I don't want to like, it's like, do I want to support this
-forever?''
+notebooks because it's like, do I want to support this forever?''
 
@@
 know why they were running it, they said `oh it doesn't work'. I think that's
-the biggest issue I'm having right now, all these forks''
+the biggest issue I'm having right now, all these forks.''
 
-Taylor: ``there's no way for both of us to have ownership of a notebook, so the
-only way is to fork it back and forth, and so we have dozens of old copies. [We
-  end up] deleting all the old ones, but people still have links to them,
-because they don't actually disappear.'' ``The UI makes it easy to go back to
-the old stale versions, because you can say `forked from' `forked from' `forked
-from', but there's actually no way to go the way that we want people to go,
-which is, like, up the tree.''
+Taylor and Horace work together on a notebook. ``There's no way for both of us
+to have ownership of a notebook, so the only way is to fork it back and forth,
+and so we have dozens of old copies. [We end up] deleting all the old ones, but
+people still have links to them, because they don't actually disappear. [...]
+The UI makes it easy to go back to the old stale versions, because you can say
+`forked from' `forked from' `forked from', but there's actually no way to go the
+way that we want people to go, which is, like, up the tree.''
 
-Rick: ``I don't necessarily need to see everybody's notebook that uses RCloud''
-``every time I do something new, I get a new notebook and so now my notebooks
-are maybe 50 or 60. That's enough to think about just on my own, but if I've got
-everybody's 50 or 60 sitting on my display, I find that it's more than I want to
-know about.''
-
-\subsection{Versioning not useful}
-Rick: ``I don't need something that keeps track of every mistake I've made or
-every direction I've tried.'' ``namespaces are already crowded, so trying to
-remember the names of notebooks is hard, much less the names of states within
-each notebook.''
-
-\subsection{Web snafus}
-Corey: ``What frustrates me is that if I do leave this window for what feels
-like 5 or 7 minutes, it disconnects and I have to reconnect and run the whole
-notebook again.''
-
-Kenny: it really sucks when you get disconnected
-
-Raif: ``being web-based, it's obviously slower than something you would be
-running on your machine''
-
-Byers: even the little lag of sending a command to a server and receiving the
-result is intolerable
-
-\subsection{Parameterization}
-Clint: ``For the things that I'll do multiple times, I'll try to parameterize
-it, and I'm working now on putting some packages together that will operate in
-some fashion outside of RCloud, straight from the command line''
-
-Taylor: ``like the top of it would be, `This is a report of the volume of all of
-our feeds for this month', and someone would want to look at it for the next
-month or the previous month, so they'd fork it to change the month''
-
-Jyothi: ``I think that if a notebook is web-enabled [parameterized] then it
-gives the person who is running the notebook, I mean they can always fork the
-notebooks and make changes, but I feel that if the owner of the notebook
-web-enables it, then whoever is running it, it's easier to run, instead of
-forking it, if they can set options, it's probably more efficient, and also they
-can fork it if they want to.''
-
-\subsection{Other observations}
-Byers: There is also a big problem with ``bitrot'' - notebooks often stop
-working because the user changed the structure of their data, changed a filename
-or a database.  We need ``organizational protocols'' to catch up with the
-technology.  There is a ``hysteresis between the technical and the social'' and
-we have not yet adapted.
-
-Taylor: `` It would be a file inside a directory that only I had read access to
-and so someone else couldn't run it... That was usually what it was, it was a
-temporary file on something that was read-only to everyone else. But I've gotten
-careful about that.''
-
-Bart: ``One UI notebook and a bunch of other notebooks that do various tasks and
-are called with API calls. The problem is, if you clone all the notebooks, then
-you have to go update every single notebook, because all the IDs are different
-now. There's no sort of relative path type stuff that you can do. You can do it
-with static assets, but you can't do it with Ajax calls''
+%% Bart: ``One UI notebook and a bunch of other notebooks that do various tasks and
+%% are called with API calls. The problem is, if you clone all the notebooks, then
+%% you have to go update every single notebook, because all the IDs are different
+%% now. There's no sort of relative path type stuff that you can do. You can do it
+%% with static assets, but you can't do it with Ajax calls''

### 2015-03-28 interviews.tex
use block quotes some places

@@
 
-But even correcting notebooks can come too late: ``I'll make a notebook of a
+But even correcting notebooks can come too late:
+\begin{quote}
+I'll make a notebook of a
 little tutorial of basic functions and what they do, and then I'll send [it]
@@
 as they're looking at the code. Like, iotools, a lot of the functions change
-names or change syntax.''
+names or change syntax.
+\end{quote}
 
 Taylor tried to keep his notebook tree tidy, but this didn't help either, because
-notebooks still exist when they have left the tree. ``That's just a problem because people will have old notebooks that
+notebooks still exist when they have left the tree.
+\begin{quote}
+That's just a problem because people will have old notebooks that
 don't work anymore. It's annoying when it's a notebook that I still have and the
@@
 know that you can't even delete notebooks, they'll have a fork of something that
-doesn't exist anymore on my thing.'' ``Now I'm really gun-shy about sharing
-notebooks because it's like, do I want to support this forever?''
+doesn't exist anymore on my thing [...] Now I'm really gun-shy about sharing
+notebooks because it's like, do I want to support this forever?
 
-Taylor: ``A good example is, someone was complaining about doing forking in an
+[...] A good example is, someone was complaining about doing forking in an
 RCloud notebook. They were trying to use a parallel package. And I said, no no I
@@
 know why they were running it, they said `oh it doesn't work'. I think that's
-the biggest issue I'm having right now, all these forks.''
+the biggest issue I'm having right now, all these forks.
+\end{quote}
 
-Taylor and Horace work together on a notebook. ``There's no way for both of us
+Taylor and Horace work together on a notebook.
+
+\begin{quote}
+There's no way for both of us
 to have ownership of a notebook, so the only way is to fork it back and forth,
@@
 `forked from' `forked from' `forked from', but there's actually no way to go the
-way that we want people to go, which is, like, up the tree.''
+way that we want people to go, which is, like, up the tree.
+\end{quote}
 

### 2015-03-28 casestudies.tex
case study update

@@
-\section{Case Study\label{sec:casestudy}: topic visualization in RCloud}
+\section{Case Study\label{sec:casestudy}: visual exploration of topic modeling in RCloud}
 
@@
 certain amount of manual intervention and
-interpretation~\cite{Siever:2014:LAM}.
+interpretation~\cite{Sievert:2014:LAM}.
 
@@
 LDAVis itself was developed by two members of the technical staff at
-AT\&T Labs, and originally was developed entirely in RStudio's
+AT\&T Labs, and originally targeted RStudio's
 Shiny~\cite{RStudio:2013:SWA}, a framework for developing R
-applications for the web. The main reason for moving the application
-to an RCloud development was precisely ease of deployment. Shiny
-provides ease of development, but when the application is finished,
-one needs to find a publicly available web server that can host the
-Shiny application; user management needs to be handled, and so
-on. Using RCloud, deployment happens transparently. This is a big win
-in terms of efficiency.
+applications for the web.  While Shiny provides unparalleled ease of
+development, discoverability and deployment turns out to be equally
+important aspects in the lifecycle of an internal application (see
+Section~\ref{sec:interviews}).  On these aspects, RCloud provides a
+very simple model: \emph{all developed notebooks are automatically
+  deployed}.
 
-Describe the application and the technologies. Runs code from an R
-package and custom analysis code; files are uploaded from the web
-client, results are provided in JSON format and visualized in the web
-client using d3~\cite{Bostock:2011:DDD}.
+LDAVis combines textual analysis, dimensionality reduction, and
+interactive visualization. The text analysis is performed via a
+combination of an existing R library that fits an LDA model using
+Gibbs sampling~\cite{} and some R code written specifically by the
+developers. The analysis module is, ultimately, a single R function
+that is exposed to the web application via the Javascript-R RPC
+mechanism described in Section~\ref{sec:system}; this way, the
+analysis is executed remotely on the RCloud servers.
+
+Each textual topic is a probability distribution over the all the
+words of the document. In order to expose patterns in the
+relationships between the topics, LDAVis employs a combination of
+interactive visualization and dimensionality reduction (letting users
+choose different measures for the topic distances and different
+dimensionality reduction techniques). The dimensionality reduction
+algorithms and distance measures are again implemented in R, which
+means they are executed on the RCloud servers as well.
+
+The result of the dimensionality reduction is a two-dimensional plot
+of the topic space, which can be seen in
+Figure~\ref{fig:textvis}. Describe d3 view.
+
+This application is simple, but highlights some of the unique features
+of RCloud. While RCloud notebooks allow deployment of R analyses over
+the web with no additional effor, RCloud \emph{applications} are more
+powerful, and are developed in a combination of Javascript and HTML
+for the front end. This requires additional knowledge than Shiny, but
+we argue that the RCloud model makes the analysis side simpler for the
+analysts (since they simply write the R code in the style they are
+used to), and the front-end visualization side simpler for the
+front-end developers (since they simply write the Javascript code in
+the style \emph{they} are used to). In addition, RCloud applications
+inherit the automatic deployment and discoverability features of
+regular RCloud notebooks.
 

### 2015-03-29 interviews.tex
trimming quotes a bit

@@
 To evaluate the effectiveness of RCloud, we interviewed 13 current and past
-users of RCloud. Of these, 9 are data analysts, and 4 build tools for data
+users of RCloud. Of these, 9 are data analysts and 4 build tools for data
 analysts and business needs.
@@
 
-Raif says ``I've been sharing my notebooks in order for other people to see what
-I've done. It's very convenient for that purpose, I would say.'' ``we haven't
+Raif's usage is typical:
+\begin{quote}
+I've been sharing my notebooks in order for other people to see what
+I've done. It's very convenient for that purpose. [...] We haven't
 been editing them together, but we've been just sharing, and I've been looking
@@
 developing the code, but there's sharing aspects for other people to see and for
-me to see other people's notebooks''
+me to see other people's notebooks.
+\end{quote}
 
@@
 found through the notebook tree or search. However, the method for sharing that
-users mentioned most is by sending links through email. Clint says, ``I can
-share [my notebook] with people and copy it pretty readily. So if my supervisor
+users mentioned most is by sending links through email. Clint says, ``If my supervisor
 wants to see what I've done or QA it, I can just send her a link.''
@@
 provide a starting point. Clint says, ``I think the best part about it is how
-easily you can share code [...] you can find a working example, rather than wearing
+easily you can share code. [...] You can find a working example, rather than wearing
 out Google and finding questionable examples that may or may not work.'' Raif
@@
 Taylor develops packages for analysts: ``I use it more for sharing code with
-other people, and for doing tutorials for iotools or hmr'' or explaining issues
-with data sources. ``A lot of times I'll share the code because I'm doing so
-much package development. I want people to see how the package works, so I
-clearly want them to see the code [...] I kind of write it just like I would write
-GitHub Markdown, where you have little code snippets and text, but RCloud lets
-me actually run the snippets'' and display the results.
+other people, and for doing tutorials for iotools or hmr'', the packages he
+maintains. ``I want people to see how the package works, so I clearly want them
+to see the code [...] I kind of write it just like I would write GitHub
+Markdown, where you have little code snippets and text, but RCloud lets me
+actually run the snippets'' and display the results. He also uses RCloud for
+explaining issues with data sources.
 
 Some users who tried out RCloud but are not able to use it for organizational
-reasons, already miss the functionality. Shelly likes
-``the concept of being able to create notebooks and share them.'' A ``wiki is not the
-best way for communication results - it's kind of like writing a blog post with
-very limited functionality [...] I have to save every
-picture and post it as an image'' and ``I can't share all of the code because it
-would just get crowded and wouldn't look right on a wiki.'' Bart says, ``if I could
-make a folder on
-RCloud and have Python notebooks and also Pig notebooks there, and execute them
-from RCloud, that would be much better than my current thing, because that would
-free me from manual documentation and version control and also telling people
-where my code was. It would be just, hey, go look on RCloud, here's the stuff.''
+reasons, already miss the functionality. Shelly likes ``the concept of being
+able to create notebooks and share them.'' A ``wiki is not the best way for
+communicating results - it's kind of like writing a blog post with very limited
+functionality [...] I have to save every picture and post it as an image'' and
+``I can't share all of the code because it would just get crowded and wouldn't
+look right on a wiki.'' Bart says, ``if I could make a folder on RCloud and have
+Python notebooks and also Pig notebooks there, and execute them from RCloud,
+that would be much better than my current thing, because that would free me from
+manual documentation and version control and also telling people where my code
+was. It would be just, hey, go look on RCloud, here's the stuff.''
 
@@
 be a popular feature. In fact, almost twice as many (131) users have forked
-someone else's notebook as have starred (75).  Clint: ``I use [forking] on my
-own work quite a bit, and I use it on Research a ton. Honestly, it's one of the
+someone else's notebook as have starred one (75).  Clint says, ``it's one of the
 handiest functions, because instead of having to find it, copy, paste it, you
@@
 useful. Rick says ``I fork my own notebooks because I'm going off and doing some
-other analogous project, so I've sort of got interesting content that I've
-already done in a previous analysis that I want to start from and then tweak to
-match a new set of data.''
+other analogous project, so I've got interesting content that I've already done
+in a previous analysis that I want to start from and then tweak to match a new
+set of data.''
 
 Forking also provides a way for others to troubleshoot when something goes
-wrong.  When Horace works with the users of his notebooks, ``I'll teach people to intercept the result in the
-middle,'' to ``insert print statements here and there and check values''
+wrong.  When Horace works with the users of his notebooks, ``I'll teach people
+to intercept the result in the middle,'' to ``insert print statements here and
+there and check values''
 
@@
 
-Our users, especially those who build general purpose tools, see a need for better ways
-to parameterize notebooks.  (Currently parameters can be added to the URL, but adding
-user interface elements to do the same thing requires expertise.) Taylor complains that
-a notebook might say `` `This is a report of the volume of all of
-our feeds for this month', and someone would want to look at it for the next
-month or the previous month, so they'd fork it to change the month.''
+Currently parameters can be added to the URL, but adding user interface elements
+to do the same thing requires expertise. The tool-builders see a need for better
+ways to parameterize notebooks. Taylor complains that a notebook might say ``
+`This is a report of the volume of all of our feeds for this month', and someone
+would want to look at it for the next month or the previous month, so they'd
+fork it to change the month.''
 
-Jyothi calls it web-enabling the notebook: users ``can always fork the
+Jyothi calls it ``web-enabling'' the notebook: users ``can always fork the
 notebooks and make changes, but I feel that if the owner of the notebook
-web-enables it, then whoever is running it, it's easier to run. Instead of
-forking it, if they can set options. It's probably more efficient, and also they
+web-enables it [...] it's easier to run. Instead of
+forking it, if they can set options, it's probably more efficient, and also they
 can [still] fork it if they want to.''
@@
 
-Clint: ``instead of looking back and saying I've got a billion files here in
+Automatic source control is also a popular feature. Clint says, ``instead of looking back and saying I've got a billion files here in
 this subdirectory and I hope I've got them backed up, if they're on RCloud I
-know they are [...] I don't have to worry about it if I drop my computer in the
+know they are. [...] I don't have to worry about it, if I drop my computer in the
 bathtub, everythings gone, because I'm not great about uploading my own for
-versioning''
+versioning.''
 
 Bart points out that the automatic versioning works well for dealing with the
-minutiae of web development. ``I like the fact that it has a built-in editor, so
+minutiae of web development:
+\begin{quote}
+I like the fact that it has a built-in editor, so
 if you need to fix a typo in a link, or an extra line break or any of this other
@@
 fixing stupid little bugs, and you can do that instantly without a lot of
-overhead, which is nice.''
+overhead, which is nice.
+\end{quote}
 
@@
 
-More selective ways to search will be needed as the corpus of notebooks
+More selective ways to search will be needed as the volume of notebooks
 continues to grow. We are considering tagging notebooks filtering the notebook
@@
 Integrating an analysis language into a web development environment is something
-the tool developers really appreciate.  Horace: ``that also means the other guys
-who want to do analytics on the data, they can first pull the data, do the
-analytics on it, and then feed the viewer the data. So if anyone from the stats
-group wants to insert some code in between the data pulling procedure, [they]
-can always insert something in between. We were trying to make such an example
-so that they don't have to rewrite the GUI part and they know how to get the
-data source, once you get the data into RCloud, then you have a dataframe to
-work with, and then they know how to produce another dataframe. We give them the
-specification of the dataframes, we give them the data input, they do whatever
-they want, they can reuse the data connection part and the GUI part. We can keep
-changing these components and that's a very nice thing.''
+the tool developers really appreciate.  The structure of Horace's visualization
+notebook means
+\begin{quote}
+the other guys who want to do analytics on the data, they can first pull the
+data, do the analytics on it, and then feed the viewer the data. So [...]
+anyone from the stats group [...] can always insert something in between. We
+were trying to make such an example so that they don't have to rewrite the GUI
+part and they know how to get the data source. Once you get the data into
+RCloud, then you have a dataframe to work with, and then they know how to
+produce another dataframe.
+\end{quote}
 
-Jyothi: ``Having an open session with R certainly helps too, because both of my
+Integrating R also helps Jyothi's topic modeling tool, described in the last section.
+\begin{quote}
+Having an open session with R certainly helps too, because both of my
 applications need to be doing something with the data, so having an open session
-where I can run R commands or R functions without having to invoke an API or send
-out a request and then wait for the request to come back to do the results,
+where I can run R commands or R functions without having to invoke an API or
+send out a request and then wait for the request to come back to do the results,
 that kind of seamless integration is extremely helpful in writing the
-application.''
+application.
+\end{quote}
 
@@
 
-\gordon{Not sure this belongs here}
-Running anything on the Web can also be a source of frustration.  Poor network
-links can cause web sockets to die, and Corey complains ``What frustrates me is
-that if I do leave this window for what feels like 5 or 7 minutes, it
-disconnects and I have to reconnect and run the whole notebook again.'' Raif
-whines that ``being web-based, it's obviously slower than something you would be
-running on your machine.'' For Byers: even the little lag of sending a command
-to a server and receiving the result is intolerable
+\subsection{Other workflows}
 
-\subsection{Other workflows}
 Although most of the analysts we spoke to use and appreciates the sharing
@@
 
-Rick says, ``I find that the style of editing is clunky. I would never edit like that
-if I'm doing ordinary work. I usually use vi.'' ``I mean it's nice to have it
-saved, but there's this trade-off between it making it easier for me to present
-something or to save something, and my ease of typing and correcting and things
-in a plain old editor window.''
-
 Many of the analysts we spoke to preferred using a text editor to refine and
@@
 
+But the web interface of RCloud doesn't cut it for him. ``I find that the style of editing is clunky. I would never edit like that
+if I'm doing ordinary work. I usually use vi.'' ``I mean it's nice to have it
+saved, but there's this trade-off between it making it easier for me to present
+something or to save something, and my ease of typing and correcting and things
+in a plain old editor window.''
+
 Kenny works much the same way, rearranging the file so the good code is at the top of
@@
 the top of the file (and often shares the work by pasting it into an RCloud notebook).
+
+Running anything on the Web can also be a source of frustration.  Poor network
+connections can cause web sockets to die, and Corey complains ``What frustrates me is
+that if I do leave this window for what feels like 5 or 7 minutes, it
+disconnects and I have to reconnect and run the whole notebook again.'' Raif
+whines that ``being web-based, it's obviously slower than something you would be
+running on your machine.'' For Byers: even the little lag of sending a command
+to a server and receiving the result is intolerable
 

### 2015-03-30 casestudies.tex
nits

@@
 applications for the web.  While Shiny provides unparalleled ease of
-development, discoverability and deployment turns out to be equally
+development, discoverability and deployment turn out to be equally
 important aspects in the lifecycle of an internal application (see
@@
 of RCloud. While RCloud notebooks allow deployment of R analyses over
-the web with no additional effor, RCloud \emph{applications} are more
+the web with no additional effort, RCloud \emph{applications} are more
 powerful, and are developed in a combination of Javascript and HTML
-for the front end. This requires additional knowledge than Shiny, but
+for the front end. This requires additional knowledge over Shiny, but
 we argue that the RCloud model makes the analysis side simpler for the

### 2015-03-30 system.tex
capitalize Gist in this context

@@
 In our case, even though this approach involved learning standards not directly
-related to visual analytics, our early decision to use the GitHub gist API
+related to visual analytics, our early decision to use the GitHub Gist API
 turned out to be successful. Future projects could gain many of the same benefits

### 2015-03-30 discussion.tex
pare down interviews more, discuss limitations raised

@@
 
-\gordon{This section now overlaps with the Interview Study quite a bit - we can
-  use that to motivate this, but that would mean not immediately responding to
-  the complaints raised there. On the other hand, trying to merge these two
-  sections would be a beast. The good news is, we anticipated most of the
-  problems that people are facing now. :-/}
-
 \subsection{Reflection on Design Considerations}
@@
 
-{\bf Shared artifacts and artifact histories} are a central feature
-of RCloud, through its shared workbooks of experiments and analyses.
-We observed that hackers readily shared work through RCloud. They
-used it to discuss ideas and results in data analysis, to ask for help
-with programming problems, and to publish results by sharing URLs.
-Artifact histories are provided through a version management system (git).
-\stephen{Do analysts type raw git commands?}
+{\bf Shared artifacts and artifact histories} are a central feature of
+RCloud, through its shared workbooks of experiments and analyses.  We
+observed that hackers readily shared work through RCloud. They used it to
+discuss ideas and results in data analysis, to ask for help with
+programming problems, and to publish results by sharing URLs.  Artifact
+histories can be accessed through the notebook tree or through GitHub's web
+interface.
 
@@
 
-{\bf Multiple language support.}
-We did not foresee the demand to relax the assumption that programming
-is done mainly in R. In retrospect, it is clear that many data science
-problems are solved with cross-language approaches (for example, scikit-learn
-in Python, weka in Java).\stephen{Isn't this fixed in some way now?}
-\gordon{Should we have a section on ways that we've responded to popular demand?}
+{\bf Exploration is hard.}
+R is a popular language for exploratory data analysis, and we built RCloud
+as a way for data scientists to seamlessly transition from EDA to sharing
+and presentation. But the interface we implemented for combining sharing and presentation
+into the R command line interferes with the exploration aspect.
+
+In existing R tools, the command prompt is used to try things out. It is very
+fast and there is no commitment: nothing gets saved and there is nothing to
+later clean up. In contrast, in RCloud's notebook interface, the extra cells need
+to get deleted later, and there seems to be a mental burden that ``bad stuff
+is getting saved.''
+
+One thing to try here is making deletion easier, for example with a
+shortcut key to remove the last cell, or a way to keep a few cells and cull
+the rest. A far more ambitious solution, suggested by one of our
+interviewees, is to implement auditing of data analyses as it existed in
+S \cite{Becker:1988:Auditing}, so that the code which did not lead to a
+result could be discarded automatically.
+
+The interface is also perceived to be slower, both in terms of reaction
+time---RCloud necessarily saves a cell to its repository before executing
+it---and because it requires mouse interaction.
+
+Other complaints mostly focus on layout and can probably be addressed with
+some iteration and customization features.
+
+{\bf Discovery is hard.}
+While we have provided tools for searching and curating notebooks, the
+search function does not help for finding methodology (and does not protect
+searchers from erroneous notebooks), and the notebook tree starts to get
+unwieldly with hundreds of users with hundreds of notebooks. In addition,
+our choice to retain notebooks even when they have been deleted has
+exacerbated the situation, since obsolete and broken notebooks stay around
+forever.
+
+One thing to try is tagging and filtering the tree on tags, so that users
+can mark their own notebooks or those of others to make them easier to
+find.
+
+However, this relies on users making a conscious effort, and it may be
+better to rely more on passive metrics such as how often a notebook has
+been run.
+
+The system also needs to help users curate their notebooks. If a notebook
+has been superceded by one of its forks, a user who opens it should be
+warned. If a notebook has been deleted, but someone still has a link to it,
+it should be clear that the notebook is dead. And operations on multiple
+notebooks and folders of notebooks should be supported to make curation
+easier.
+
+{\bf Collaboration is hard.}
+While forking provides a simple way to continue someone else's work, working
+in teams is a lot more complicated. The forking feature is popular but it
+poses a problem, too, as dozens of versions of a notebook start proliferating.
+
+One solution we are mulling over is to allow overwriting the current state
+of a notebook with the current state of another fork of the notebook. More
+general solutions would be serious engineering endeavors: even GitHub's web
+interface does not support merging where there are any conflicts. Actual
+sharing of notebooks would probably require moving toward git's notion of
+decentralized code repositories, because GitHub gists do not support shared
+ownership.
 
@@
 informal channels to coordinate work.
+
 Versioning needs to be managed in several places: in scripts;

### 2015-03-30 interviews.tex
pare down interviews more, discuss limitations raised

@@
 
-Taylor develops packages for analysts: ``I use it more for sharing code with
-other people, and for doing tutorials for iotools or hmr'', the packages he
-maintains. ``I want people to see how the package works, so I clearly want them
+Taylor develops packages for analysts, and uses RCloud ``more for sharing code with
+other people, and for doing tutorials for iotools or hmr'', his packages.
+``I want people to see how the package works, so I clearly want them
 to see the code [...] I kind of write it just like I would write GitHub
@@
 actually run the snippets'' and display the results. He also uses RCloud for
-explaining issues with data sources.
+explaining problems with data sources.
 
@@
 to intercept the result in the middle,'' to ``insert print statements here and
-there and check values''
+there and check values.''
 
@@
 Currently parameters can be added to the URL, but adding user interface elements
-to do the same thing requires expertise. The tool-builders see a need for better
-ways to parameterize notebooks. Taylor complains that a notebook might say ``
+to do the same thing requires expertise. The tool-builders see a need for
+facilities to make it easier. Taylor complains that a notebook might say ``
 `This is a report of the volume of all of our feeds for this month', and someone
@@
 notebooks and make changes, but I feel that if the owner of the notebook
-web-enables it [...] it's easier to run. Instead of
+web-enables it, [...] it's easier to run. Instead of
 forking it, if they can set options, it's probably more efficient, and also they
@@
 
-Automatic source control is also a popular feature. Clint says, ``instead of looking back and saying I've got a billion files here in
-this subdirectory and I hope I've got them backed up, if they're on RCloud I
-know they are. [...] I don't have to worry about it, if I drop my computer in the
-bathtub, everythings gone, because I'm not great about uploading my own for
-versioning.''
+Automatic source control is also a popular feature. Clint says, ``instead of
+looking back and saying I've got a billion files here in this subdirectory and I
+hope I've got them backed up, if they're on RCloud I know they are. [...] I
+don't have to worry about it, if I drop my computer in the bathtub, everythings
+gone, because I'm not great about uploading my own for versioning.''
 
@@
 all the notebooks there, searchable, saves me from replicating what other people
-have done. [...] One way of thinking of it is like Google Code: you just search
-for something and you find something useful and you just use it.''
+have done.''
 
@@
 
-\subsection{Web integration}
+\subsection{Integrated analysis}
 Integrating an analysis language into a web development environment is something
@@
 seconds to run, the next cell minutes, and the last cell 5 hours.  In this
-situation, the ``Run whole notebook'' button is ``dangerous''. When converting
+situation, the run whole notebook button is ``dangerous''. When converting
 his work to notebooks, Kenny ends up with a lot of comments that say ``this cell
-takes a long time to run.'' This is also a huge
-issue for demoing code.
+takes a long time to run.''
 
-The solution is probably something like knitr's caching of results, which Kenny
-is doing in an ad-hoc way, ``littering the notebook with little switches that
-comment out'' the slow parts and ``load the object instead'', ``save the object
-to disk''. One of the authors of this paper manually writes cells that check if
-a result file exists, and perform the calculation only if it's not there.
+To avert this danger, users are forced to write code within the notebook to
+cache results. Kenny ends up ``littering the notebook with little switches that
+comment out the slow parts'' and either ``load the object instead'' or ``save
+the object to disk''. One of the authors of this paper manually writes cells
+that check if a result file exists, and perform the calculation only if it's not
+there.
 
-Rick on auditing (possible solution): ``you didn't see your mistakes, they were
-in there but you never really saw them because a mistake never led to an answer
-at the end'' ``the real result of this analysis was these three plots, so go
-back and figure out everything that I did that was involved in creating those
-three plots. So that I could start from the raw data and create those three
-plots.''
+%% Rick on auditing (possible solution): ``you didn't see your mistakes, they were
+%% in there but you never really saw them because a mistake never led to an answer
+%% at the end'' ``the real result of this analysis was these three plots, so go
+%% back and figure out everything that I did that was involved in creating those
+%% three plots. So that I could start from the raw data and create those three
+%% plots.''
 
@@
 
-Taylor, who writes tools and example notebooks for those tools, seems to be
+Taylor, who writes packages and example notebooks for them, seems to be
 haunted the most by dead notebooks.  At first, he had to learn how to write safe
@@
 
-But even correcting notebooks can come too late:
-\begin{quote}
-I'll make a notebook of a
-little tutorial of basic functions and what they do, and then I'll send [it]
-out, and then people will give me feedback on [it], or have some new feature, or
-I'll realize something's wrong, and then I'll go back and change it and update
+But even correcting notebooks can come too late: ``I'll go back and change it and update
 that page, but then what happens is people have in the meantime forked it. [...]
-Usually what it'll be is actually that I'll have to change a package as well, so
-their old fork stops working, and then they complain. [...] This probably bites
-me more than other people, because I'm doing development so things are changing
-as they're looking at the code. Like, iotools, a lot of the functions change
-names or change syntax.
-\end{quote}
+Usually what it'll be is that I'll have to change a package as well, so
+their old fork stops working, and then they complain.''
 
-Taylor tried to keep his notebook tree tidy, but this didn't help either, because
-notebooks still exist when they have left the tree.
-\begin{quote}
-That's just a problem because people will have old notebooks that
-don't work anymore. It's annoying when it's a notebook that I still have and the
-format has changed just a bit. More than that is when people have, now that I
-know that you can't even delete notebooks, they'll have a fork of something that
-doesn't exist anymore on my thing [...] Now I'm really gun-shy about sharing
-notebooks because it's like, do I want to support this forever?
+Taylor tried to keep his notebook tree tidy, but this didn't help, because
+people either keep old links in their email, or forked a notebook which is now
+obsolete. By design, notebooks that are deleted in the user interface are not
+purged from the repository. ``Now I'm really gun-shy about sharing notebooks
+because it's like, do I want to support this forever?''
 
-[...] A good example is, someone was complaining about doing forking in an
-RCloud notebook. They were trying to use a parallel package. And I said, no no I
-think it works, and I did like a one-cell thing that just did some sweep kind of
-thing to show that it was actually working in parallel, and then I just deleted
-it, thinking it was gone. And then it saved something in a directory that I then
-deleted, because it was just a temporary directory. Three months later, I don't
-know why they were running it, they said `oh it doesn't work'. I think that's
-the biggest issue I'm having right now, all these forks.
-\end{quote}
-
-Taylor and Horace work together on a notebook.
-
+Taylor and Horace work together on a notebook:
 \begin{quote}
@@
 `forked from' `forked from' `forked from', but there's actually no way to go the
-way that we want people to go, which is, like, up the tree.
+way that we want people to go, which is up the tree.
 \end{quote}

### 2015-03-30 conclusion.tex
comment on what i think conclusion should be

@@
 not some little toy you provide out on the fringes.
-Infratsructure is hard.}
+Infrastructure is hard.}
+
+\gordon{I think that except for the DevOps slogan, this is redundant and
+too dense.  I would replace the first three paragraphs with something
+like this:} We designed a prototype environment for exploration, sharing,
+presentation, and publishing. The exploration part didn't work out so well,
+but people like sharing stuff and integrating web and analysis languages.
 
@@
 
-Possible next steps are to support cross-language development,
+Possible next steps are
 to incorporate richer recommendation techniques, to provide fine-grained

### 2015-03-30 introduction.tex
move notebook stats to interview section

@@
 another 30 people use it more than once a week, and another 50 use it more than
-once a month.  There are over 5000 notebooks, of which more than 500 have been
-starred and more than 350 have been forked. \gordon{Is this the right place for
-the notebook stats? We haven't defined forking or starring yet.}  Most of the
+once a month.  Most of the
 active users are members of AT\&T Labs, but some are data scientists and other

### 2015-03-30 interviews.tex
move notebook stats to interview section
minor edits

@@
 Working in a shared environment also entails compromises about what you can
-install. Byers says that installing random packages is part of exploratory data
-analysis.
+install. Byers says installing alternative or nonstandard packages is
+intrinsic to exploratory data analysis.
 
@@
 of it is junk not meant to be saved.'' Rick says ``I just want to type in a
-couple of quick and get some results that are going to tell me what to do next,
-and they're not necessarily archival in any sense''
+couple of quick commands and get some results that are going to tell
+me what to do next, and they're not necessarily archival in any sense.''
 
@@
 Users also felt that cells do not capture the right level of granularity: they
-either hold too much code or too little. Rick says that RCloud's cells structure
+either hold too much code or too little. Rick says that RCloud's cell structure
 ``tempts me to type a big long thing and then run the whole thing, as opposed to
@@
 code to scroll off the screen. Raif: ``for example imagine I run a cell whose
-output is huge, and now that output would block the whole window, and in order
-to find the next cell, I have to scroll down, and find where that cell starts,
-right? So, I'm losing the continuity of my code to the output''
+output is huge, and now that output would block the whole window, and to
+find the next cell, I have to scroll down, and find where that cell starts,
+right? So, I'm losing the continuity of my code.''
 
-Some users think users should have the option to keep results separate from the
-code.  Ivan thinks that RStudio's layout is more helpful because the charts
+Some users would like the option to keep results separate from code. 
+Ivan thinks that RStudio's layout is more helpful because charts
 are kept in another pane and stay in one place while doing analysis. Raif says
@@
 seconds to run, the next cell minutes, and the last cell 5 hours.  In this
-situation, the run whole notebook button is ``dangerous''. When converting
+situation, the Run Whole Notebook button is ``dangerous''. When converting
 his work to notebooks, Kenny ends up with a lot of comments that say ``this cell
@@
 RCloud is starting to be a victim of its own success, as there are now too many
-notebooks.  Rick doesn't find the notebook tree very useful for this reason: ``I
+notebooks.  There are over 5000 notebooks on the research instance, of which
+more than 500 have been starred and more than 350 have been forked.
+
+Rick doesn't find the notebook tree very useful for this reason: ``I
 don't necessarily need to see everybody's notebook that uses RCloud'' ``every

### 2015-03-30 casestudies.tex
update case study and bib

@@
 
-The result of the dimensionality reduction is a two-dimensional plot
-of the topic space, which can be seen in
-Figure~\ref{fig:textvis}. Describe d3 view.
+The result of the dimensionality reduction process is a
+two-dimensional plot of the topic space, which can be seen in
+Figure~\ref{fig:textvis}. The interactive view is implemented with SVG
+and Javascript through d3~\cite{Bostock:2011:DDD}. The most popular
+web-based visualization libraries for R is ggvis~\cite{ggvis}, and so
+a natural question becomes: could ggvis have been used instead of
+custom Javascript?  In this case, the interactive features of ggvis
+(and, by extension, Shiny) are a subset of those of Vega.  The custom
+interactions in LDAVis (hovering over a single topic, hovering over a
+topic cluster, hovering over a word in the distribution, etc) do not
+appear to be available in the current version of Vega~\cite{vega},
+although the required components for reactive interaction have been
+recently described by Satyanarayan et
+al.~\cite{Satyanarayan:2014:DID}. As a result, custom Javascript was
+required in this case, making RCloud uniquely positioned among systems
+available today.
 

### 2015-03-31 interviews.tex
further tweaking and pruning

@@
 found through the notebook tree or search. However, the method for sharing that
-users mentioned most is by sending links through email. Clint says, ``If my supervisor
+users mentioned most is sending links through email. Clint says, ``If my supervisor
 wants to see what I've done or QA it, I can just send her a link.''
@@
 More selective ways to search will be needed as the volume of notebooks
-continues to grow. We are considering tagging notebooks filtering the notebook
-tree by tag, measuring the quality of notebooks by their forks as well as stars,
-and searching within a single user's notebooks.
+continues to grow. We explore some ideas in the Discussion below.
 
@@
 
-\subsection{Other workflows}
+\subsection{Exploring elsewhere}
 
-Although most of the analysts we spoke to use and appreciates the sharing
-features, RCloud is less popular as a tool for exploratory data analysis and
-development.
+Although most of the analysts use and appreciate the sharing features,
+RCloud is less popular as a tool for exploratory data analysis. Every
+analyst with prior R experience still does analysis in another tool and
+pastes their code into RCloud for sharing.
 
-Much of this is because of familiarity with other tools. Horace: ``for other
-people who use RCloud vs RStudio, they will always think that they want to
-prototype on RStudio. Maybe because it's more familiar for them. So, they work
-on RStudio, on a small data set, and they fix up everything, make sure
-everything is right, and then they place it on RCloud for sharing.''
+This is partly out of familiarity. Taylor asks, ``why would I want to use
+RCloud over my current setup? If it's just me, I like my text editor and
+terminal. There's nothing that I want that those two don't give me.''
 
-Many of the analysts we spoke to preferred using a text editor to refine and
-store their work, pasting it to an R command line to try it out. Rick describes,
-``I have the discipline of having a file that describes what I'm doing with the
-commands that I'm using so that I can go back and recreate it or pass it on to
-someone else, so it's a little bit like having an RCloud notebook, it's not
-necessarily executable but most of the commands that I've typed are in there.''
+%% Much of this is because of familiarity with other tools. Horace: ``for other
+%% people who use RCloud vs RStudio, they will always think that they want to
+%% prototype on RStudio. Maybe because it's more familiar for them. So, they work
+%% on RStudio, on a small data set, and they fix up everything, make sure
+%% everything is right, and then they place it on RCloud for sharing.''
 
-But the web interface of RCloud doesn't cut it for him. ``I find that the style of editing is clunky. I would never edit like that
-if I'm doing ordinary work. I usually use vi.'' ``I mean it's nice to have it
-saved, but there's this trade-off between it making it easier for me to present
-something or to save something, and my ease of typing and correcting and things
-in a plain old editor window.''
+%% Many of the analysts we spoke to preferred using a text editor to refine and
+%% store their work, pasting it to an R command line to try it out.
 
-Kenny works much the same way, rearranging the file so the good code is at the top of
-the file. Then he saves the file once he's got the right code at
-the top of the file (and often shares the work by pasting it into an RCloud notebook).
+But there are interface reasons as well. Rick describes a common workflow:
+\begin{quote}
+I have the discipline of having a file that describes what I'm doing, with the
+commands that I'm using, so that I can go back and recreate it or pass it on to
+someone else. So it's a little bit like having an RCloud notebook. It's not
+necessarily executable but most of the commands that I've typed are in there.
+\end{quote}
 
-Running anything on the Web can also be a source of frustration.  Poor network
-connections can cause web sockets to die, and Corey complains ``What frustrates me is
-that if I do leave this window for what feels like 5 or 7 minutes, it
-disconnects and I have to reconnect and run the whole notebook again.'' Raif
-whines that ``being web-based, it's obviously slower than something you would be
-running on your machine.'' For Byers: even the little lag of sending a command
-to a server and receiving the result is intolerable
+However, the web interface of RCloud doesn't cut it for him:
+\begin{quote}
+I find that the style of editing is clunky. I would never edit like that if
+I'm doing ordinary work. I usually use vi. [...] I mean, it's nice to have
+it saved, but there's this trade-off between it making it easier for me to
+present something or to save something, and my ease of typing and
+correcting and things in a plain old editor window.
+\end{quote}
 
-Taylor asks, ``why would I want to use RCloud over my
-current setup? If it's just me, I like my text editor and terminal. There's
-nothing that I want that those two don't give me.''
+Kenny also works with text file and command line, rearranging a file so the
+good code is at the top of the file. Then he saves the file once he's got
+the right code at the top, and often shares the work by pasting it into an
+RCloud notebook.
+
+%% Running anything on the Web can also be a source of frustration.  Poor network
+%% connections can cause web sockets to die, and Corey complains ``What frustrates me is
+%% that if I do leave this window for what feels like 5 or 7 minutes, it
+%% disconnects and I have to reconnect and run the whole notebook again.'' Raif
+%% whines that ``being web-based, it's obviously slower than something you would be
+%% running on your machine.'' For Byers, even the little lag of sending a command
+%% to a server and receiving the result is intolerable
 
@@
 
-\subsection{Proliferation of notebook cells}
+\subsection{Cells versus the command line}
 RCloud's notebook interface combines editable Markdown with a command line
@@
 
-The interface takes some learning. Clint says ``it took me a couple of weeks of
+The interface takes some learning. Clint says
+\begin{quote}
+It took me a couple of weeks of
 looking at it to become comfortable because I saw the cells and it just threw me
@@
 staging purposes. I really like it now, but when I first saw it, [it was] very
-confusing [...] I think I prefer it now, because you can cut your code up.''
+confusing [...] I think I prefer it now, because you can cut your code up.
+\end{quote}
 
 Just the difference between a cell and the command prompt can confuse and slow
-down users. Rick: ``if I was typing into one of the cells near that top, I had
-to think `I'm editing this cell and then I have to execute it', and if was doing
+down users. Rick explains:
+\begin{quote}
+If I was typing into one of the cells near that top, I had
+to think ``I'm editing this cell and then I have to execute it,'' and if was doing
 the one at the bottom I could type it but then it would automatically execute
 [by pressing enter], but then it became a standard cell and I have to edit
-it. It was just another set of modes that I didn't like''
+it. It was just another set of modes that I didn't like
+\end{quote}
 
@@
 so the only way to get that same functionality is if every line's in one
-cell. Because if it runs lines 1, 2, 3, then 4 fails, I have to either run the
-whole cell again, or have line 4 in its own cell. I can't rerun just one line.''
+cell.''
 
@@
 RCloud is starting to be a victim of its own success, as there are now too many
-notebooks.  There are over 5000 notebooks on the research instance, of which
+notebooks.  There are over 5200 notebooks on the research instance, of which
 more than 500 have been starred and more than 350 have been forked.
@@
 Rick doesn't find the notebook tree very useful for this reason: ``I
-don't necessarily need to see everybody's notebook that uses RCloud'' ``every
+don't necessarily need to see everybody's notebook that uses RCloud [...] Every
 time I do something new, I get a new notebook and so now my notebooks are maybe

### 2015-03-31 discussion.tex
move hard parts to end of discussion
further tweaking and pruning

@@
 
-{\bf Exploration is hard.}
-R is a popular language for exploratory data analysis, and we built RCloud
-as a way for data scientists to seamlessly transition from EDA to sharing
-and presentation. But the interface we implemented for combining sharing and presentation
-into the R command line interferes with the exploration aspect.
-
-In existing R tools, the command prompt is used to try things out. It is very
-fast and there is no commitment: nothing gets saved and there is nothing to
-later clean up. In contrast, in RCloud's notebook interface, the extra cells need
-to get deleted later, and there seems to be a mental burden that ``bad stuff
-is getting saved.''
-
-One thing to try here is making deletion easier, for example with a
-shortcut key to remove the last cell, or a way to keep a few cells and cull
-the rest. A far more ambitious solution, suggested by one of our
-interviewees, is to implement auditing of data analyses as it existed in
-S \cite{Becker:1988:Auditing}, so that the code which did not lead to a
-result could be discarded automatically.
-
-The interface is also perceived to be slower, both in terms of reaction
-time---RCloud necessarily saves a cell to its repository before executing
-it---and because it requires mouse interaction.
-
-Other complaints mostly focus on layout and can probably be addressed with
-some iteration and customization features.
-
-{\bf Discovery is hard.}
-While we have provided tools for searching and curating notebooks, the
-search function does not help for finding methodology (and does not protect
-searchers from erroneous notebooks), and the notebook tree starts to get
-unwieldly with hundreds of users with hundreds of notebooks. In addition,
-our choice to retain notebooks even when they have been deleted has
-exacerbated the situation, since obsolete and broken notebooks stay around
-forever.
-
-One thing to try is tagging and filtering the tree on tags, so that users
-can mark their own notebooks or those of others to make them easier to
-find.
-
-However, this relies on users making a conscious effort, and it may be
-better to rely more on passive metrics such as how often a notebook has
-been run.
-
-The system also needs to help users curate their notebooks. If a notebook
-has been superceded by one of its forks, a user who opens it should be
-warned. If a notebook has been deleted, but someone still has a link to it,
-it should be clear that the notebook is dead. And operations on multiple
-notebooks and folders of notebooks should be supported to make curation
-easier.
-
-{\bf Collaboration is hard.}
-While forking provides a simple way to continue someone else's work, working
-in teams is a lot more complicated. The forking feature is popular but it
-poses a problem, too, as dozens of versions of a notebook start proliferating.
-
-One solution we are mulling over is to allow overwriting the current state
-of a notebook with the current state of another fork of the notebook. More
-general solutions would be serious engineering endeavors: even GitHub's web
-interface does not support merging where there are any conflicts. Actual
-sharing of notebooks would probably require moving toward git's notion of
-decentralized code repositories, because GitHub gists do not support shared
-ownership.
-
 {\bf Versioning.}
@@
 system to the breaking point, is that the idea?}
+
+{\bf Exploration is hard.}
+R is a popular language for exploratory data analysis, and we built RCloud
+as a way for data scientists to seamlessly transition from EDA to sharing
+and presentation. But the combined notebook and command line interface we designed interferes with the exploration aspect.
+
+In existing R tools, the command prompt is used to try things out. It is very
+fast and there is no commitment: nothing gets saved and there is nothing to
+later clean up. In contrast, in RCloud's notebook interface, the extra cells need
+to get deleted later, and there seems to be a mental burden that \textit{bad stuff
+is getting saved}.
+
+One thing to try here is making deletion easier, for example with a
+shortcut key to remove the last cell, or a way to keep a few cells and cull
+the rest. A far more ambitious solution, suggested by one of our
+interviewees, is to implement auditing of data analyses as it existed in S
+\cite{Becker:1988:Auditing}, so that a result could be selected and code
+which did not lead to it could be discarded automatically.
+
+We got here from an insistence on reproducibility: it is a core principle
+of RCloud that no command should be run without being saved, even if it
+gets deleted later. This is why we did not implement a simple command line
+as many users request. We could imagine making an exception for commands
+that have no side-effects, but in the general case this is no easier than
+auditing because R is not a purely functional language.
+
+The interface is also perceived to be slower, both for computer and for
+human. First, there is a slower reaction to commands being run, because
+RCloud necessarily saves a cell to its repository before executing it. It
+may be possible to safely write to the repository behind execution of the
+same code. Second, and perhaps more detrimental, the interface requires
+mouse interaction and switching between the keyboard and mouse, which are
+known to be slower. \gordon{we don't need to back that up, right? i hate
+  using passive voice.} Shortcut keys can be devised if people are willing
+to learn them, but it would be better to reduce the number of such
+operations needed.
+
+Other complaints mostly focus on layout and can probably be addressed with
+some iteration and customization features.
+
+{\bf Discovery is hard.}
+While we have provided tools for searching and curating notebooks, the
+search function does not help for finding methodology (and does not protect
+searchers from erroneous notebooks), and the notebook tree starts to get
+unwieldly with hundreds of users with hundreds of notebooks. In addition,
+our choice to retain notebooks even when they have been deleted has
+exacerbated the situation, since obsolete and broken notebooks stay around
+forever.
+
+One thing to try is tagging and filtering the tree on tags, so that users
+can mark their own notebooks or those of others to make them easier to
+find. However, this relies on users making a conscious effort, and it may be
+better to rely more on passive metrics such as how often a notebook has
+been run.
+
+The system also needs to help users curate their notebooks. If a notebook
+has been superceded by one of its forks, a user who opens it should be
+warned. If a notebook has been deleted, but someone still has a link to it,
+it should be clear that the notebook is dead. And operations on multiple
+notebooks and folders of notebooks should be supported to make curation
+easier.
+
+{\bf Collaboration is hard.}
+While forking provides a simple way to continue someone else's work, working
+in teams is a lot more complicated. The forking feature is popular but it
+poses a problem, too, as dozens of versions of a notebook start proliferating.
+
+One solution we are considering is to allow overwriting the current state
+of a notebook with the current state of another fork of the notebook.
+
+More general solutions would be serious engineering endeavors. Even
+GitHub's generous web interface does not support merging where there are
+conflicts. Actual sharing of notebooks would probably require moving toward
+git's notion of decentralized code repositories, because GitHub gists do
+not support shared ownership.

### 2015-03-31 interviews.tex
refine further

@@
 analysts and business needs.
+
+Generally, analysts were more annoyed by the cell interface and the
+perceived slowness of hte interface than web/tool developers. Analysts
+switching to R from another language were more likely to use RCloud for
+exploratory data analysis. Web/tool developers were more likely to see the
+need for parameterization.
 
@@
 know. Rick says, ``usually I know somebody's notebooks that I want to search
-through, like I'll search through Simon Urbanek's first, then I'll go off to
-somebody else's, because I know that the kind of thing I'm looking for something
+through [...] because I know that the kind of thing I'm looking for is something
 more obscure than I'm likely to find in some random person's.''
@@
 More selective ways to search will be needed as the volume of notebooks
-continues to grow. We explore some ideas in the Discussion below.
+continues to grow. We explore some ideas in Section~\ref{sec:discussion}.
 
@@
 \begin{quote}
-the other guys who want to do analytics on the data, they can first pull the
-data, do the analytics on it, and then feed the viewer the data. So [...]
-anyone from the stats group [...] can always insert something in between. We
-were trying to make such an example so that they don't have to rewrite the GUI
-part and they know how to get the data source. Once you get the data into
-RCloud, then you have a dataframe to work with, and then they know how to
-produce another dataframe.
+the other guys who want to do analytics on the data, they can first pull
+the data, do the analytics on it, and then feed the viewer the data. So
+[...] anyone from the stats group [...] can always insert something in
+between. [...] They don't have to rewrite the GUI part and they know how to
+get the data source. Once you get the data into RCloud, then you have a
+dataframe to work with, and then they know how to produce another
+dataframe.
 \end{quote}
@@
 
-However, the web interface of RCloud doesn't cut it for him:
-\begin{quote}
-I find that the style of editing is clunky. I would never edit like that if
-I'm doing ordinary work. I usually use vi. [...] I mean, it's nice to have
+However, the web interface of RCloud doesn't work for him. ``It's nice to have
 it saved, but there's this trade-off between it making it easier for me to
 present something or to save something, and my ease of typing and
-correcting and things in a plain old editor window.
-\end{quote}
+correcting and things in a plain old editor window.''
 
 Kenny also works with text file and command line, rearranging a file so the
-good code is at the top of the file. Then he saves the file once he's got
-the right code at the top, and often shares the work by pasting it into an
-RCloud notebook.
+good code is at the top of the file. When he's done, all the the right is
+code at the top, and often shares the work by pasting it into an RCloud
+notebook.
 
@@
 
-The interface takes some learning. Clint says
+The interface takes some learning. Clint says:
 \begin{quote}
@@
 staging purposes. I really like it now, but when I first saw it, [it was] very
-confusing [...] I think I prefer it now, because you can cut your code up.
+confusing.
 \end{quote}
@@
 [by pressing enter], but then it became a standard cell and I have to edit
-it. It was just another set of modes that I didn't like
+it.
 \end{quote}
@@
 ``being literate about anything typically takes a lot of rewriting and going
-back over things [...] I just want to explore a few things and then I'll know
+back over things. [...] I just want to explore a few things and then I'll know
 what I want to write.''
@@
 Much of the time, commands typed at the R command line just serve a transitory
-purpose, so having RCloud save commands into cells bothers experienced
-users. Kenny notes, ``It saves everything I do like everything is gold, but most
+purpose, so having RCloud save commands into cells can be annoying. Kenny notes, ``It saves everything I do like everything is gold, but most
 of it is junk not meant to be saved.'' Rick says ``I just want to type in a
@@
 or you make a plot of the data. Things that should not really become part of a
-notebook, but things that help you understand your data better.'' (Raif)
+notebook, but things that help you understand your data better'' (Raif).
 
@@
 ``tempts me to type a big long thing and then run the whole thing, as opposed to
-typing a few little pieces that I'm thinking of on the inner parts and then put
+typing a few little pieces [...] on the inner parts and then put
 them together to make the big thing'' as he would do on the command line. When
@@
 longer and it's harder to deal with.'' Ivan says that cells' controls and blank
-space take extra real estate compared to a straight command line, and he and
-KCKC both complained that when there are long results or plots, it causes the
-code to scroll off the screen. Raif: ``for example imagine I run a cell whose
-output is huge, and now that output would block the whole window, and to
-find the next cell, I have to scroll down, and find where that cell starts,
-right? So, I'm losing the continuity of my code.''
+space take extra screen space compared to a straight command line, and he and
+KC both complained that when there are long results or plots, it causes the
+code to scroll off the screen. Raif explains, ``Imagine I run a cell whose
+output is huge. [...] To find the next cell, I have to scroll down and find
+where that cell starts. So I'm losing the continuity of my code.''
 
@@
 
-To avert this danger, users are forced to write code within the notebook to
+To avert this danger, some users write code within the cells to
 cache results. Kenny ends up ``littering the notebook with little switches that
@@
 
-RCloud combines the functions of scratchpad with notebook assets. KC thinks
-that having a scratch pad that doesn't persist between different notebooks is
-``useless''.
+%% RCloud combines the functions of scratchpad with notebook assets. KC thinks
+%% that having a scratch pad that doesn't persist between different notebooks is
+%% ``useless''.
 
@@
 \subsection{A sea of notebooks}
-RCloud is starting to be a victim of its own success, as there are now too many
-notebooks.  There are over 5200 notebooks on the research instance, of which
-more than 500 have been starred and more than 350 have been forked.
+RCloud is starting to be a victim of its own success, as it is becoming
+difficult to navigate all the notebooks.  There are over 5200 notebooks on
+the research instance, of which more than 500 have been starred and more
+than 350 have been forked.
 
-Rick doesn't find the notebook tree very useful for this reason: ``I
-don't necessarily need to see everybody's notebook that uses RCloud [...] Every
+For this reason, Rick doesn't find the notebook tree very useful:
+\begin{quote}
+I don't necessarily need to see everybody's notebook that uses RCloud [...] Every
 time I do something new, I get a new notebook and so now my notebooks are maybe
@@
 everybody's 50 or 60 sitting on my display, I find that it's more than I want to
-know about.''
+know about.
+\end{quote}
 
-Although RCloud promises an environment where notebooks stay working, our users
-have not all learned the habits that make this a reality. As Byers puts it,
-there is still a big problem with ``bitrot'' - notebooks often stop working
-because the user changed the structure of their data, changed a filename or a
-database.  He says we need ``organizational protocols'' to catch up with the
-technology.
+Although RCloud promises an environment where notebooks stay working, our
+users have not all learned the habits that make this a reality. As Byers
+puts it, there is still a big problem with ``bitrot'' -- notebooks often
+stop working because the user changed the structure of their data, or
+changed a filename or a database.  He says we need ``organizational
+protocols'' to catch up with the technology. Even if one forks someone
+else's notebook and corrects it, the original notebook still exists with
+the error.
 
-There are technical problems to be solved here as well. Byers continues that the
-problem with forking a notebook and correcting it, is that the original notebook
-still exists with the error.  We need a way to see that there is a more recent
-fork of notebook and to fetch in the changes from it.
+%% At first, he had to learn how to write safe
+%% notebooks: ``It would be a file inside a directory that only I had [...] access
+%% to and so someone else couldn't run it, [...] a temporary file on something that
+%% was read-only to everyone else. But I've gotten careful about that.''
 
 Taylor, who writes packages and example notebooks for them, seems to be
-haunted the most by dead notebooks.  At first, he had to learn how to write safe
-notebooks: ``It would be a file inside a directory that only I had [...] access
-to and so someone else couldn't run it, [...] a temporary file on something that
-was read-only to everyone else. But I've gotten careful about that.''
+haunted the most by dead notebooks. Taylor and Horace work together on a notebook:
+\begin{quote}
+There's no way for both of us to have ownership of a notebook, so the only
+way is to fork it back and forth, and so we have dozens of old copies. [We
+end up] deleting all the old ones, but people still have links to them,
+because they don't actually disappear.
+\end{quote}
 
-But even correcting notebooks can come too late: ``I'll go back and change it and update
-that page, but then what happens is people have in the meantime forked it. [...]
-Usually what it'll be is that I'll have to change a package as well, so
-their old fork stops working, and then they complain.''
+The problem of dead notebooks is compounded by changing
+packages and their example notebooks at the same time. Taylor will ``go back and
+change it and update the page, but then what happens is people have in the
+meantime forked it. [...]  I'll have to change a package as well, so their
+old fork stops working, and then they complain.''
 
 Taylor tried to keep his notebook tree tidy, but this didn't help, because
-people either keep old links in their email, or forked a notebook which is now
+people either kept old links in their email, or forked a notebook which is now
 obsolete. By design, notebooks that are deleted in the user interface are not
@@
 because it's like, do I want to support this forever?''
-
-Taylor and Horace work together on a notebook:
-\begin{quote}
-There's no way for both of us
-to have ownership of a notebook, so the only way is to fork it back and forth,
-and so we have dozens of old copies. [We end up] deleting all the old ones, but
-people still have links to them, because they don't actually disappear. [...]
-The UI makes it easy to go back to the old stale versions, because you can say
-`forked from' `forked from' `forked from', but there's actually no way to go the
-way that we want people to go, which is up the tree.
-\end{quote}
 

### 2015-03-31 relatedwork.tex

@@
 \centering
-\includegraphics[width=.8\linewidth]{fig/notebook/notebook.pdf}
+\includegraphics[width=.95\linewidth]{fig/notebook/notebook.pdf}
 \caption{\label{fig:notebook}An RCloud notebook is a sequence of
-  \emph{cells}, each a snippet of source in R or Markdown. Notebooks
-  are stored in GitHub \emph{gists}, which provide the version-control
-  primitives needed in RCloud during editing (left) or viewing and
-  executing (right).}
+  \emph{cells}, each a snippet of source in one of the supported languages (typically R, but Python and others are also supported) or Markdown.
+The main creation workflow involves editing notebooks, which are transparently stored as git repositories in GitHub, which provides us with easy access primitives for version tracking.
+Notebooks can be executed as they're edited (left), or in a standalone viewer (right), via a slightly different URL.
+This lightweight provides a very low-friction mechanism for sharing results which we discuss in full in Section~\ref{where}.
+}
 \end{figure*}

### 2015-03-31 relatedwork.tex
too many whichs

@@
   \emph{cells}, each a snippet of source in one of the supported languages (typically R, but Python and others are also supported) or Markdown.
-The main creation workflow involves editing notebooks, which are transparently stored as git repositories in GitHub, which provides us with easy access primitives for version tracking.
+The main creation workflow involves editing notebooks, which are transparently stored as git repositories in GitHub, providing us with easy access primitives for version tracking.
 Notebooks can be executed as they're edited (left), or in a standalone viewer (right), via a slightly different URL.

### 2015-03-31 discussion.tex

@@
 later clean up. In contrast, in RCloud's notebook interface, the extra cells need
-to get deleted later, and there seems to be a mental burden that \textit{bad stuff
+to get deleted later, and there seems to be a mental burden that \emph{bad stuff
 is getting saved}.
@@
 The system also needs to help users curate their notebooks. If a notebook
-has been superceded by one of its forks, a user who opens it should be
-warned. If a notebook has been deleted, but someone still has a link to it,
-it should be clear that the notebook is dead. And operations on multiple
-notebooks and folders of notebooks should be supported to make curation
-easier.
+has been deleted or superceded by one of its forks, a user who opens it
+should be warned. And operations on multiple notebooks and folders of
+notebooks should be supported to make curation easier.
 

### 2015-03-31 discussion.tex
more tweaking

@@
 RCloud, through its shared workbooks of experiments and analyses.  We
-observed that hackers readily shared work through RCloud. They used it to
-discuss ideas and results in data analysis, to ask for help with
-programming problems, and to publish results by sharing URLs.  Artifact
-histories can be accessed through the notebook tree or through GitHub's web
-interface.
+observed that hackers readily share work through RCloud. They use it to
+demonstrate techniques, share results with peers and managers, and transfer algorithms
+to other groups.  Artifact histories can be accessed through the notebook
+tree or through GitHub's web interface.
 
@@
 encourages making incremental changes at low cost and without disrupting
-the work of others. \gordon{But this design choice means that developing
-larger, multi-page apps is less well supported.}
+the work of others.
 
@@
 {\bf Discussion.} Annotation and commenting was another central goal.
-Commenting is supported through github, but our hackers found it
+Commenting is supported through GitHub, but our hackers found it
 awkward and did not exercise it as much as we expected.
@@
 {\bf Content export} is not a capability we aimed at supporting,
-and R already has many packages for this. Instead, our goal was to
-encourage ways of working without needing as much explicit export.
-We did assume that RCloud notebooks should play well in the ecosystem
-of other tools, and this worked as intended, though we learned it also
-creates a temptation to absorb features almost without limit. (Should
-RCloud provide templates for personal web pages and graphical
-editing? Should it incorporate an email client and emacs?
-Should it push results to Facebook?) Experience shows that interfaces
-to external systems and other features almost always win over architectural
-purity in the real world. \gordon{If you don't export, doesn't that mean
-you have to support every possible operation downstream? Due to popular demand,
-we did add plot export.}
+and R already has many packages for this. Recently, due to popular
+demand, we added user interface support for exporting plot images.
+We assume that RCloud notebooks should play well in the ecosystem
+of other tools, but sometimes it is difficult to know whether adding
+a compatibility feature will offload complexity or bring more in.
 
-{\bf Social-psychological incentives} and {\bf voting and ranking} are supported
-through starring. This mechanism was employed often by our test
-audience. \gordon{actually, the notebook tree quickly became so unwieldy that
-no one uses this.} An obvious next step would be to enhance recommendations using
-relationships discovered by static and dynamic code analysis. This may be
-considered both within and across collections of scripts (the latter being
-similar to VisTrails' enhanced recommendations by clustering multiple
+{\bf Social-psychological incentives} and {\bf voting and ranking} are
+supported through starring and forking. These mechanisms are employed often
+on the platform. An obvious next step would be to enhance recommendations
+using relationships discovered by static and dynamic code analysis. This
+may be considered both within and across collections of scripts (the latter
+being similar to VisTrails' enhanced recommendations by clustering multiple
 workflows).  It seems valuable to know which packages are frequently used
-together, or appear in proximity to a certain record types or data feeds.  In
-general, trivial or passive mechanisms to collect data for recommendations are
-essential.
+together, or appear in proximity to a certain record types or data feeds.
+In general, trivial or passive mechanisms to collect data for
+recommendations are essential.
 
@@
 separate the development and deployment environments. More than that,
-ever version of every prototype is shared by default. Although this encourages
+every version of every prototype is shared by default. Although this encourages
 collaboration on work in progress it also quickly exposes errors such as
@@
 as a way for data scientists to seamlessly transition from EDA to sharing
-and presentation. But the combined notebook and command line interface we designed interferes with the exploration aspect.
+and presentation. But the combined notebook and command line interface we
+designed interferes with the use of R for exploration.
 

### 2015-03-31 casestudies.tex
work on conclusions

@@
 
-In this section, we present a simple, but real-world example
-application currently deployed using RCloud.
+In this section, as an example we present a simple real-world
+application that was deployed under RCloud.
 
-This example application is an implementation of LDAVis, a
-recently-published method for visualizing large collections of
-text. It was specifically designed to enable non-expert users to
-explore collections of short text documents using topic modeling and
+The application is an implementation of LDAVis, a
+recently-published method for visualizing large text corpora.
+It was specifically designed to assist non-experts to explore
+collections of short text documents using topic modeling and
 visualization. Topic modeling~\cite{Blei:2003:LDA} is a standard
@@
 
-LDAVis itself was developed by two members of the technical staff at
+LDAVis was developed by two technical staff members at
 AT\&T Labs, and originally targeted RStudio's
 Shiny~\cite{RStudio:2013:SWA}, a framework for developing R
-applications for the web.  While Shiny provides unparalleled ease of
+applications for the web.  While Shiny provides outstanding ease of
 development, discoverability and deployment turn out to be equally
@@
 
-LDAVis combines textual analysis, dimensionality reduction, and
-interactive visualization. The text analysis is performed via a
+LDAVis combines text analysis, dimensionality reduction and
+interactive visualization. Text analysis is performed via a
 combination of an existing R library that fits an LDA model using

### 2015-03-31 conclusion.tex
work on conclusions

@@
 \section{Conclusions and Future Work}
-
-\stephen{Is this the kind of bad conclusion that contains
-a lot of material that should have been explained at the
-beginning of the paper to move it along faster and tell
-people why they should be interested in the work? What
-would be our conclusion?  These capabilties are important,
-but they are really hard to program, and adoption is slow
-because you're meddling with people's work process and
-toolset, which is the heart of their environment and
-not some little toy you provide out on the fringes.
-Infrastructure is hard.}
 
@@
 Building on previous work to define requirements for
-visual analytics, we designed a prototype environment, RCloud.
-RCloud supports collaboration and communication
-through shared experimental workbooks with version
-control, searching, recommending, annotating, and publishing
-experiments as web sites and as reusable services.
+visual analytics, we designed an environment
+for exploration, sharing, presentation and publishing.
+We implemented it in the RCloud prototype.
 
-We also deployed RCloud in a community of working data scientists.
+We deployed RCloud in a community of working data scientists.
 Experience with the RCloud prototype provides evidence that data
@@
 having capabilities that support collaboration and integrate
-the entire visual analytics process.
+the entire visual analytics process. Specifically, features for
+sharing and publishing were eagerly adopted. Features for data
+single-user data exploration, that compete with existing mature
+tools, were not accepted as readily. Some experienced users
+fashioned their own workflow so they could keep using familiar EDA tools.
 
-RCloud is a step toward practical ``DevOps for data science'' and
-reproducible, publishable experiments.
-
-Possible next steps are
+The study has provided a step toward practical ``DevOps for data science''
+and reproducible, publishable experiments.  Possible next steps are
 to incorporate richer recommendation techniques, to provide fine-grained

### 2015-03-31 conclusion.tex
yes

@@
 
-\gordon{I think that except for the DevOps slogan, this is redundant and
-too dense.  I would replace the first three paragraphs with something
-like this:} We designed a prototype environment for exploration, sharing,
-presentation, and publishing. The exploration part didn't work out so well,
-but people like sharing stuff and integrating web and analysis languages.
+%\gordon{I think that except for the DevOps slogan, this is redundant and too dense.
+I would replace the first three paragraphs with something like this:} We designed a prototype environment for exploration, sharing, presentation, and publishing.
+The exploration part didn't work out so well, but people like sharing stuff and integrating web and analysis languages.
 
@@
 We deployed RCloud in a community of working data scientists.
-Experience with the RCloud prototype provides evidence that data
+Experience with the prototype provides evidence that data
 science teams and the organizations in which they work benefit from
-having capabilities that support collaboration and integrate
-the entire visual analytics process. Specifically, features for
-sharing and publishing were eagerly adopted. Features for data
+having capabilities to support collaboration and to integrate
+the entire visual analytics process. We found that features for
+sharing and publishing were eagerly adopted. Features for 
 single-user data exploration, that compete with existing mature
@@
 
-The study has provided a step toward practical ``DevOps for data science''
+This study has provided a step toward practical ``DevOps for data science''
 and reproducible, publishable experiments.  Possible next steps are

### 2015-03-31 introduction.tex
more tweaking
intro and related work cleanup, no more missing references

@@
 what was barely imaginable then. 
-We argue, on the other hand, that we \emph{develop} our data analysis
-solutions, and context in which they become part of the broader issues
-being solved in organizations, has not changed as much: the S language
-was developed essentially alongside Unix~\cite{TheSSystem}, after all.
+We argue, on the other hand, that how we \emph{develop} our data analysis
+solutions has not changed as much: the S language
+was developed essentially alongside Unix~\cite{TheSSystem} and,
+although environments such as RStudio include a variety of modern
+features, they still follow the basic metaphor of terminal, text
+editor, and source files stored in file systems.
 We turn our attention to this opportunity to use computation
@@
 The high level architecture of the prototype system
-is shown in figure~\ref{fig:high-level-architecture}.
+is shown in figure~\ref{fig:system}.
 We chose R as the foundation for our prototype because it is already
@@
 
-Over the past two \gordon{three?} years, we prototyped RCloud and deployed it to
+Over the past three years, we prototyped RCloud and deployed it to
 a community of data scientists, business analysts and other colleagues.  The

### 2015-03-31 relatedwork.tex
intro and related work cleanup, no more missing references

@@
 
-There is much previous work that inspired or influenced ours.
-Indeed, many studies have addressed requirements for future visual
-analytics environments.
-Previous work that we learned from includes:
+In some sense or another, the entire field of information
+visualization and visual analytics revolve around improving how users
+solve problems using data. In this section, we focus specifically on
+system work relating to problem-solving enviroments and multiple users.
 
-{\bf VisTrails} as a landmark prototype of a scientific workflow
-and provenance management system. This prototype demonstrated
-the value of capturing aspects of the processes that
-surround data analysis experiments and tools, including detailed
-history, collaboration, and deployment.
+\paragraph*{Social Data exploration and Analysis}
+{\bf ManyEyes}~\cite{Viegas:2007:MAS} was a landmark system designed
+for the crowdsourced creation and publication of data
+visualizations. Although ManyEyes only supported a limited number of
+visual encodings, the system's success was both a precursor to more
+sophisticated solutions and an early indication that the world-wide
+web was a suitable platform for data visualization. One of the main
+challenges we faced in designing RCloud was providing an experience
+for \emph{consuming} visualizations as seamless as ManyEyes's, while
+sacrificing as little as possible on the generality of the analyses
+themselves.
 
-{\bf iPython notebooks (Jupyter)} are the closest work we are aware of.
-The goal of this project is to provide shared notebooks and a remote
-execution environment for interpreted scripts. 
-\stephen{Are there significant differences involving versioning,
-provenance, annotation and other broad process support?}
-
-{\bf RStudio, RShiny} and other packages based on R for creating and
-publishing web content such as knitR, Rpubs, rCharts, and gg2v.
-These tools are not collaborative, but they are aimed at better
-graphics and web applications.
-
-{\bf bl.ocks, jsfiddle, plotly} and other web services for sharing code
-and demos.
-
+\paragraph*{Notebooks as a medium for data analysis dissemination}
+The concept of a ``notebook'' as we use it here can be traced all the
+way back to Knuth's literate programming~\cite{Knuth:1984:LP}. In
+literate programming, a comprehensive description in prose of the
+behavior of the program is ``weaved'' together with the source code,
+yielding both an executable program and a human-readable document.  A
+notebook represented as a collection of short, executable cells,
+originated with Mathematica, and in R, literate programming is
+supported by packages such as knitr and RMarkdown~\cite{Xie:2013:DDW}.
+Project Jupyter~\cite{jupyter} (originally implemented as part of
+IPython) offer some notebook features, but lack a transparent
+mechanism for sharing and deployment in multiple-user settings.
+%
 Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
@@
 visual analytics teams.
+%
+Although RStudio offers publication of literate R programs as a free
+service on their website, the workflow is somewhat disconnected from
+the development of those programs. Once they're uploaded, it's hard
+for other users to build off of the work published (or even for the
+original author to update new versions). In other words, RStudio
+handles \emph{publication}, but not \emph{deployment} and sharing.
 
+\paragraph*{Provenance and versioning} As
+we mentioned, one central problem in exploratory analysis is that
+problems change quickly over time, often in the course of developing a
+solution. As a result, systems should provide adequate support for
+tracking \emph{changes} of the data analyses scripts.  VisTrails was
+one of the original systems for managing \emph{process provenance} ,
+and demonstrated the value of capturing aspects of the processes that
+surround data analysis experiments and tools, including detailed
+history, collaboration, and deployment~\cite{Callahan:2006:VVM}.
+
+{\bf VisMashup}~\cite{Santos:2009:VST} VisMashup defines a schema and
+semantics for automatically deriving user interfaces from workflows,
+while Crowdlabs exposes these capabilities on a website
+feature~\cite{Mates:2011:CSA} workflow upload and remote execution. In
+our view, the impedance mismatch between a dataflow pipeline
+specification and the power of a general-purpose language is too great
+for the type of general exploratory work in data science teams. At the
+expense of ease of use for non-programmers, RCloud tries to provide a
+closer match for analysts accustomed to creating and executing R and
+Python code, while retaining attractive properties like transparent
+provenance tracking and interactive data visualization on the web.
+
+\paragraph*{Web-based tools for sharing code snippets}
+There have been a variety of tools recently developed for quickly
+sharing small programs on the Web, including but not limited to:
+bl.ocks~\cite{blocks}, jsfiddle~\cite{jsfiddle}, and
+plot.ly~\cite{plotly}.  bl.ocks and jsfiddle are both designed to
+share Javascript programs, which means that their deployment happens
+automatically through the web browser. If and when Javascript becomes
+the lingua franca of exploratory data analysis, then we can foresee
+building a much simpler version of RCloud where all execution happens
+either on the client side or via web services. Unfortunately that is
+not a realistic assumption for the present, making these solutions by
+themselves unsuitable to our use case. Plot.ly is notable in that it
+provides API support for publishing \emph{from} scripts: in other
+words, it is possible to generate a plot.ly visualization from inside
+a separate program. Although this is an intriguing idea, it
+nevertheless creates a disconnect between the analysis and the
+resulting visualization. In RCloud, we wanted to make sure that every
+visualization was transparently linked to the source code that
+generated it.
+
+%% publishing web content such as knitR, Rpubs,
+%% rCharts, and gg2v.  These tools are not collaborative, but they are
+%% aimed at better graphics and web applications.
+%% .  {\bf iPython
+%%   notebooks (Jupyter)} are the closest work we are aware of.  The goal
+%% of this project is to provide shared notebooks and a remote execution
+%% environment for interpreted scripts.  \stephen{Are there significant
+%%   differences involving versioning, provenance, annotation and other
+%%   broad process support?}
+
+%% {\bf bl.ocks, jsfiddle, plotly} and other web services for sharing code
+%% and demos.
+
+\paragraph*{Needs of data analysts}
 Kandel et al.'s interview study points out the typical ``explore'',
@@
 
-{\bf Manyeyes} \cite{Viegas:2007:MAS} was a landmark system for the integration
-of social media with visualization and data publishing. We build on this
-work by defining a rich interface for collaboration about code, and for
-operating on metadata.
-
 The need for integrating statistics and visualization has been
@@
 techniques may be integrated and made available to a broad community.
-
-{\bf VisMashup}~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
-are tools built on top of the VisTrails workflow management
-system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
-semantics for automatically deriving user interfaces from workflows,
-while Crowdlabs exposes these capabilities on a website feature
-workflow upload and remote execution. In our view, the impedance
-mismatch between a dataflow pipeline specification and the power of a
-general-purpose language is too great for the type of general
-exploratory work in data science teams. As a result, RCloud tries to
-provide a closer match for analysts accustomed to creating and
-executing R and Python code, while retaining attractive
-properties like automatic versioning and management.
-
 \begin{figure*}
@@
 Notebooks can be executed as they're edited (left), or in a standalone viewer (right), via a slightly different URL.
-This lightweight provides a very low-friction mechanism for sharing results which we discuss in full in Section~\ref{where}.
+This lightweight provides a very low-friction mechanism for sharing results which we discuss in full in Section~\ref{sec:interviews}.
 }
@@
 
-\stephen{We have a problem we already went over some of this.}
-Our work has been inspired by and benefited from other efforts
-to improve data analysis environments and processes. Building in
-the immense foundation of systems for statistical graphics,
-recent work includes RStudio \cite{RStudio:2013:SWA},
-the R packages Markdown \cite{Allaire:2014:MMR},
-knitr \cite{Xie:2013:DDW}
-and Shiny \cite{RStudio:2013:SWA},
-and iPython notebooks \cite{Perez:2007:IAS}
-to name a few. RStudio aims at providing an integrated development environment
-for R, with support for publishing code in packages. Markdown,
-knitR and Shiny augment R with sophisticated reporting capabilities, including
-interactive web interfaces. IPython \cite{Perez:2007:IAS}
-shares many of our goals, such as providing a comprehensive environment
-for analysis and programming, and share-able documents on the web.
-\stephen{Do we now need to introduce one of those remarks where we
-explain that we are not the same as IPython.}
+%% \stephen{We have a problem we already went over some of this.}
+%% Our work has been inspired by and benefited from other efforts
+%% to improve data analysis environments and processes. Building in
+%% the immense foundation of systems for statistical graphics,
+%% recent work includes RStudio \cite{RStudio:2013:SWA},
+%% the R packages Markdown \cite{Allaire:2014:MMR},
+%% knitr \cite{Xie:2013:DDW}
+%% and Shiny \cite{RStudio:2013:SWA},
+%% and iPython notebooks \cite{Perez:2007:IAS}
+%% to name a few. RStudio aims at providing an integrated development environment
+%% for R, with support for publishing code in packages. Markdown,
+%% knitR and Shiny augment R with sophisticated reporting capabilities, including
+%% interactive web interfaces. IPython \cite{Perez:2007:IAS}
+%% shares many of our goals, such as providing a comprehensive environment
+%% for analysis and programming, and share-able documents on the web.
+%% \stephen{Do we now need to introduce one of those remarks where we
+%% explain that we are not the same as IPython.}
 
-One overall goal is to reduce the gap between implementers and deployers
-in visual analytics. The fusion of development with production
-operations in software release management (``DevOps'' \cite{Httermann:2012:DD}
-or ``continuous integration'' \cite{Fowler:2006:Continuous}) is a trend in web
-services and related fields.
-By making it convenient for data scientists to expend just a little more effort
-when creating experiments, we may be able to eliminate the need for programming
-teams to recreate their work to deploy it in production,
-which has a high cost in time, expense and accuracy.
+One overall goal is to reduce the gap between implementers and
+deployers in visual analytics. The fusion of development with
+production operations in software release management (``DevOps''
+\cite{Httermann:2012:DD} or ``continuous integration''
+\cite{Fowler:2006:Continuous}) is a trend in web services and related
+fields.  By making it convenient for data scientists to expend just a
+little more effort when creating experiments, we may be able to
+eliminate the need for programming teams to recreate their work to
+deploy it in production, which has a high cost in time, expense and
+accuracy. We now turn to a description of the system architecture
+itself, and how it enables the features we have been highlighting.

### 2015-03-31 interviews.tex

@@
 
-Raif's usage is typical:
-\begin{quote}
-I've been sharing my notebooks in order for other people to see what
-I've done. It's very convenient for that purpose. [...] We haven't
-been editing them together, but we've been just sharing, and I've been looking
-at things that other people do. So there's no collaborative aspect of it, like
-developing the code, but there's sharing aspects for other people to see and for
-me to see other people's notebooks.
-\end{quote}
+%% Raif's usage is typical:
+%% \begin{quote}
+%% I've been sharing my notebooks in order for other people to see what
+%% I've done. It's very convenient for that purpose. [...] We haven't
+%% been editing them together, but we've been just sharing, and I've been looking
+%% at things that other people do. So there's no collaborative aspect of it, like
+%% developing the code, but there's sharing aspects for other people to see and for
+%% me to see other people's notebooks.
+%% \end{quote}
 
@@
 
-An extremely common but more problematic use of forking is to
-change parameters. Raif: ``I've been forking other people's notebooks [because]
-I want to run it on a different part of data, or I want to change some parts, I
-don't want to see this column, I want a different column, things like that.''
+An extremely common but more problematic use of forking is to change
+parameters. Raif: ``I've been forking other people's notebooks [because] I
+want to run it on a different part of data, or I want to change some parts,
+I don't want to see this column, I want a different column, things like
+that.''  Taylor complains that a notebook might say ```This is a report of
+the volume of all of our feeds for this month', and someone would want to
+look at it for the next month or the previous month, so they'd fork it to
+change the month.''
 
-Currently parameters can be added to the URL, but adding user interface elements
-to do the same thing requires expertise. The tool-builders see a need for
-facilities to make it easier. Taylor complains that a notebook might say ``
-`This is a report of the volume of all of our feeds for this month', and someone
-would want to look at it for the next month or the previous month, so they'd
-fork it to change the month.''
-
-Jyothi calls it ``web-enabling'' the notebook: users ``can always fork the
-notebooks and make changes, but I feel that if the owner of the notebook
-web-enables it, [...] it's easier to run. Instead of
-forking it, if they can set options, it's probably more efficient, and also they
-can [still] fork it if they want to.''
+Currently parameters can be added to the URL, but adding user interface
+elements to do the same thing requires expertise. The tool-builders see a
+need for facilities to make it easier. Jyothi calls it ``web-enabling'' the
+notebook: users ``can always fork the notebooks and make changes, but I
+feel that if the owner of the notebook web-enables it, [...] it's easier to
+run. Instead of forking it, if they can set options, it's probably more
+efficient, and also they can [still] fork it if they want to.''
 
@@
 looking back and saying I've got a billion files here in this subdirectory and I
-hope I've got them backed up, if they're on RCloud I know they are. [...] I
-don't have to worry about it, if I drop my computer in the bathtub, everythings
-gone, because I'm not great about uploading my own for versioning.''
+hope I've got them backed up, if they're on RCloud I know they are.''
+
+%%  [...] I
+%% don't have to worry about it, if I drop my computer in the bathtub, everythings
+%% gone, because I'm not great about uploading my own for versioning.''
 
@@
 back to your source code, change it, commit it to the repo, pull the repo to
-your distribution version. None of that's necessary. A lot of web development is
-fixing stupid little bugs, and you can do that instantly without a lot of
-overhead, which is nice.
+your distribution version. 
+%% None of that's necessary. A lot of web development is
+%% fixing stupid little bugs, and you can do that instantly without a lot of
+%% overhead, which is nice.
 \end{quote}
@@
 helpful: ``I don't need something that keeps track of every mistake I've made or
-every direction I've tried.'' Tagging these versions with names wouldn't
-necessarily help because ``namespaces are already crowded, so trying to remember
-the names of notebooks is hard, much less the names of states within each
-notebook.''
+every direction I've tried.''
+%% Tagging these versions with names wouldn't
+%% necessarily help because ``namespaces are already crowded, so trying to remember
+%% the names of notebooks is hard, much less the names of states within each
+%% notebook.''
 
@@
 [...] anyone from the stats group [...] can always insert something in
-between. [...] They don't have to rewrite the GUI part and they know how to
-get the data source. Once you get the data into RCloud, then you have a
+between. [...] Once you get the data into RCloud, then you have a
 dataframe to work with, and then they know how to produce another
@@
 
-Integrating R also helps Jyothi's topic modeling tool, described in the last section.
-\begin{quote}
-Having an open session with R certainly helps too, because both of my
-applications need to be doing something with the data, so having an open session
-where I can run R commands or R functions without having to invoke an API or
-send out a request and then wait for the request to come back to do the results,
-that kind of seamless integration is extremely helpful in writing the
-application.
-\end{quote}
+Integrating R also helps Jyothi's topic modeling tool, described in the
+last section: ``Having an open session where I can run R commands or R
+functions without having to invoke an API or send out a request and then
+wait for the request to come back [...] is extremely helpful in writing the
+application.''
 
@@
 
-But there are interface reasons as well. Rick describes a common workflow:
-\begin{quote}
-I have the discipline of having a file that describes what I'm doing, with the
-commands that I'm using, so that I can go back and recreate it or pass it on to
-someone else. So it's a little bit like having an RCloud notebook. It's not
-necessarily executable but most of the commands that I've typed are in there.
-\end{quote}
+%% But there are interface reasons as well. Rick describes a common workflow:
+%% \begin{quote}
+%% I have the discipline of having a file that describes what I'm doing, with the
+%% commands that I'm using, so that I can go back and recreate it or pass it on to
+%% someone else. So it's a little bit like having an RCloud notebook. It's not
+%% necessarily executable but most of the commands that I've typed are in there.
+%% \end{quote}
 
-However, the web interface of RCloud doesn't work for him. ``It's nice to have
+The web interface of RCloud doesn't work for Rick: ``It's nice to have
 it saved, but there's this trade-off between it making it easier for me to
@@
 good code is at the top of the file. When he's done, all the the right is
-code at the top, and often shares the work by pasting it into an RCloud
-notebook.
+code at the top. RCloud does not readily support this workflow, but Kenny
+often shares the work by pasting it into an RCloud notebook when he's done.
 
@@
 
-The interface takes some learning. Clint says:
+The interface takes some learning. Clint says at first,
 \begin{quote}
-It took me a couple of weeks of
-looking at it to become comfortable because I saw the cells and it just threw me
+%% It took me a couple of weeks of
+%% looking at it to become comfortable because 
+I saw the cells and it just threw me
 for a total loop. I mean, it's a good idea because [...] I can run it all at
@@
 
-For some users, the cells getting saved feels incompatible with exploratory data
-analysis. Rick notes that Markdown is a form of literate programming, and
-``being literate about anything typically takes a lot of rewriting and going
-back over things. [...] I just want to explore a few things and then I'll know
-what I want to write.''
+%% For some users, the cells getting saved feels incompatible with exploratory data
+%% analysis. Rick notes that Markdown is a form of literate programming, and
+%% ``being literate about anything typically takes a lot of rewriting and going
+%% back over things. [...] I just want to explore a few things and then I'll know
+%% what I want to write.''
 

### 2015-03-31 relatedwork.tex
oops

@@
 
-{\bf VisMashup}~\cite{Santos:2009:VST} VisMashup defines a schema and
+VisMashup~\cite{Santos:2009:VST}  defines a schema and
 semantics for automatically deriving user interfaces from workflows,

### 2015-03-31 introduction.tex
fixes

@@
 \item It is hard to find data, metadata, and knowledge about them.
-\item Most coordination is done through meetings, whose content is not related to other artifacts and may not even be stored.
+\item Most coordination is done through meetings, whose content is not linked to other artifacts and may not even be stored.
 \item Production deployment requires porting code to a different environment or even completely rewriting it, thwarting continuous release.
@@
 In view of the opportunity to improve this situation,
-we created a software environment that supports the end to end
+we created a software environment that supports the end-to-end
 visual analytics process for individuals and teams.

### 2015-03-31 interviews.tex
remove reference to case study

@@
 
-Integrating R also helps Jyothi's topic modeling tool, described in the
-last section: ``Having an open session where I can run R commands or R
-functions without having to invoke an API or send out a request and then
-wait for the request to come back [...] is extremely helpful in writing the
-application.''
+Integrating R also helps Jyothi's application: ``Having an open session
+where I can run R commands or R functions without having to invoke an API
+or send out a request and then wait for the request to come back [...] is
+extremely helpful.''
 

### 2015-03-31 interviews.tex
names

@@
 
-%% Raif's usage is typical:
+%% Wendy's usage is typical:
 %% \begin{quote}
@@
 found through the notebook tree or search. However, the method for sharing that
-users mentioned most is sending links through email. Clint says, ``If my supervisor
+users mentioned most is sending links through email. Lilo says, ``If my supervisor
 wants to see what I've done or QA it, I can just send her a link.''
@@
 Besides providing a way to present work to others, notebook sharing can also
-provide a starting point. Clint says, ``I think the best part about it is how
+provide a starting point. Lilo says, ``I think the best part about it is how
 easily you can share code. [...] You can find a working example, rather than wearing
-out Google and finding questionable examples that may or may not work.'' Raif
+out Google and finding questionable examples that may or may not work.'' Wendy
 notes, if ``some person has done something similar, then you're able to just
@@
 
-Taylor develops packages for analysts, and uses RCloud ``more for sharing code with
+Evelyn develops packages for analysts, and uses RCloud ``more for sharing code with
 other people, and for doing tutorials for iotools or hmr'', his packages.
@@
 Some users who tried out RCloud but are not able to use it for organizational
-reasons, already miss the functionality. Shelly likes ``the concept of being
+reasons, already miss the functionality. Leith likes ``the concept of being
 able to create notebooks and share them.'' A ``wiki is not the best way for
@@
 ``I can't share all of the code because it would just get crowded and wouldn't
-look right on a wiki.'' Bart says, ``if I could make a folder on RCloud and have
+look right on a wiki.'' Iris says, ``if I could make a folder on RCloud and have
 Python notebooks and also Pig notebooks there, and execute them from RCloud,
@@
 
-%% Corey: ``being able to share notebooks or codes, or collaborate on something,
+%% Kaylee: ``being able to share notebooks or codes, or collaborate on something,
 %% easily being able to share that information. That's an obvious advantage and
@@
 
-%% Rick: ``I like being able to look at other people's notebooks.''
+%% Kenyon: ``I like being able to look at other people's notebooks.''
 
@@
 be a popular feature. In fact, almost twice as many (131) users have forked
-someone else's notebook as have starred one (75).  Clint says, ``it's one of the
+someone else's notebook as have starred one (75).  Lilo says, ``it's one of the
 handiest functions, because instead of having to find it, copy, paste it, you
@@
 
-Although we always intended forking to be used to improve others' code, we we
+Although we always intended forking to be used to improve others' code, we
 initially didn't support forking one's own notebooks, and this has proved very
-useful. Rick says ``I fork my own notebooks because I'm going off and doing some
+useful. Kenyon says ``I fork my own notebooks because I'm going off and doing some
 other analogous project, so I've got interesting content that I've already done
@@
 Forking also provides a way for others to troubleshoot when something goes
-wrong.  When Horace works with the users of his notebooks, ``I'll teach people
+wrong.  When Hugh works with the users of his notebooks, ``I'll teach people
 to intercept the result in the middle,'' to ``insert print statements here and
@@
 
-%% Shelly: ``gives the audience a way to try their own analysis, like maybe Shelly
+%% Leith: ``gives the audience a way to try their own analysis, like maybe Leith
 %% should have used this parameter and they can just change the parameters''
@@
 An extremely common but more problematic use of forking is to change
-parameters. Raif: ``I've been forking other people's notebooks [because] I
+parameters. Wendy says, ``I've been forking other people's notebooks [because] I
 want to run it on a different part of data, or I want to change some parts,
 I don't want to see this column, I want a different column, things like
-that.''  Taylor complains that a notebook might say ```This is a report of
+that.''  Evelyn complains that a notebook might say ```This is a report of
 the volume of all of our feeds for this month', and someone would want to
@@
 elements to do the same thing requires expertise. The tool-builders see a
-need for facilities to make it easier. Jyothi calls it ``web-enabling'' the
+need for facilities to make it easier. Allison calls it ``web-enabling'' the
 notebook: users ``can always fork the notebooks and make changes, but I
@@
 
-%% Clint: ``For the things that I'll do multiple times, I'll try to parameterize
+%% Lilo: ``For the things that I'll do multiple times, I'll try to parameterize
 %% it, and I'm working now on putting some packages together that will operate in
@@
 \subsection{Automatic source control}
-%% Rick: ``when I've finished something, there's a nice clean record of it.''
+%% Kenyon: ``when I've finished something, there's a nice clean record of it.''
 %% (I think this may be more about the publish/markdown than about source control.)
 
-Automatic source control is also a popular feature. Clint says, ``instead of
+Automatic source control is also a popular feature. Lilo says, ``instead of
 looking back and saying I've got a billion files here in this subdirectory and I
@@
 
-Bart points out that the automatic versioning works well for dealing with the
+Iris points out that the automatic versioning works well for dealing with the
 minutiae of web development:
@@
 On the other hand, saving every change leads to many fine-grained
-versions. Rick says that for this reason, the history feature is not all that
+versions. Kenyon says that for this reason, the history feature is not all that
 helpful: ``I don't need something that keeps track of every mistake I've made or
@@
 \subsection{Discovering others' work}
-Many users find the search function useful. Raif says ``the fact that we have
+Many users find the search function useful. Wendy says ``the fact that we have
 all the notebooks there, searchable, saves me from replicating what other people
@@
 But other users prefer to browse just the notebooks of the experts they
-know. Rick says, ``usually I know somebody's notebooks that I want to search
+know. Kenyon says, ``usually I know somebody's notebooks that I want to search
 through [...] because I know that the kind of thing I'm looking for is something
@@
 Integrating an analysis language into a web development environment is something
-the tool developers really appreciate.  The structure of Horace's visualization
+the tool developers really appreciate.  The structure of Hugh's visualization
 notebook means
@@
 
-Integrating R also helps Jyothi's application: ``Having an open session
+Integrating R also helps Allison's application: ``Having an open session
 where I can run R commands or R functions without having to invoke an API
@@
 
-%% Jyothi: ``another thing I really like about RCloud is your publish feature, so
+%% Allison: ``another thing I really like about RCloud is your publish feature, so
 %% that non-RCloud users, the anonymous login, non-RCloud users can access the
@@
 
-This is partly out of familiarity. Taylor asks, ``why would I want to use
+This is partly out of familiarity. Evelyn asks, ``why would I want to use
 RCloud over my current setup? If it's just me, I like my text editor and
@@
 
-%% Much of this is because of familiarity with other tools. Horace: ``for other
+%% Much of this is because of familiarity with other tools. Hugh: ``for other
 %% people who use RCloud vs RStudio, they will always think that they want to
@@
 
-%% But there are interface reasons as well. Rick describes a common workflow:
+%% But there are interface reasons as well. Kenyon describes a common workflow:
 %% \begin{quote}
@@
 
-The web interface of RCloud doesn't work for Rick: ``It's nice to have
+The web interface of RCloud doesn't work for Kenyon: ``It's nice to have
 it saved, but there's this trade-off between it making it easier for me to
@@
 
-Kenny also works with text file and command line, rearranging a file so the
+Coby also works with text file and command line, rearranging a file so the
 good code is at the top of the file. When he's done, all the the right is
-code at the top. RCloud does not readily support this workflow, but Kenny
+code at the top. RCloud does not readily support this workflow, but Coby
 often shares the work by pasting it into an RCloud notebook when he's done.
@@
 %% Running anything on the Web can also be a source of frustration.  Poor network
-%% connections can cause web sockets to die, and Corey complains ``What frustrates me is
+%% connections can cause web sockets to die, and Kaylee complains ``What frustrates me is
 %% that if I do leave this window for what feels like 5 or 7 minutes, it
-%% disconnects and I have to reconnect and run the whole notebook again.'' Raif
+%% disconnects and I have to reconnect and run the whole notebook again.'' Wendy
 %% whines that ``being web-based, it's obviously slower than something you would be
-%% running on your machine.'' For Byers, even the little lag of sending a command
+%% running on your machine.'' For Joy, even the little lag of sending a command
 %% to a server and receiving the result is intolerable
@@
 Working in a shared environment also entails compromises about what you can
-install. Byers says installing alternative or nonstandard packages is
+install. Joy says installing alternative or nonstandard packages is
 intrinsic to exploratory data analysis.
@@
 
-The interface takes some learning. Clint says at first,
+The interface takes some learning. Lilo says, at first,
 \begin{quote}
@@
 Just the difference between a cell and the command prompt can confuse and slow
-down users. Rick explains:
+down users. Kenyon explains:
 \begin{quote}
@@
 %% For some users, the cells getting saved feels incompatible with exploratory data
-%% analysis. Rick notes that Markdown is a form of literate programming, and
+%% analysis. Kenyon notes that Markdown is a form of literate programming, and
 %% ``being literate about anything typically takes a lot of rewriting and going
@@
 Much of the time, commands typed at the R command line just serve a transitory
-purpose, so having RCloud save commands into cells can be annoying. Kenny notes, ``It saves everything I do like everything is gold, but most
-of it is junk not meant to be saved.'' Rick says ``I just want to type in a
+purpose, so having RCloud save commands into cells can be annoying. Coby notes, ``It saves everything I do like everything is gold, but most
+of it is junk not meant to be saved.'' Kenyon says ``I just want to type in a
 couple of quick commands and get some results that are going to tell
@@
 The material that is not appropriate to save includes ``expressions that allow
-me to check that I'm the right track'' (Rick), ``checking out what your data is,
+me to check that I'm the right track'' (Kenyon), ``checking out what your data is,
 or you make a plot of the data. Things that should not really become part of a
-notebook, but things that help you understand your data better'' (Raif).
+notebook, but things that help you understand your data better'' (Wendy).
 
 Users also felt that cells do not capture the right level of granularity: they
-either hold too much code or too little. Rick says that RCloud's cell structure
+either hold too much code or too little. Kenyon says that RCloud's cell structure
 ``tempts me to type a big long thing and then run the whole thing, as opposed to
@@
 them together to make the big thing'' as he would do on the command line. When
-Taylor uses the R command line, he copies and pastes ``like 5-10 lines of code,
+Evelyn uses the R command line, he copies and pastes ``like 5-10 lines of code,
 so when something breaks, I get an error message on that one line, and I can
@@
 
-Cells can also take up substantial vertical space, requiring scrolling. Rick
+Cells can also take up substantial vertical space, requiring scrolling. Kenyon
 complains that ``every time I type more stuff, the notebook gets longer and
-longer and it's harder to deal with.'' Ivan says that cells' controls and blank
+longer and it's harder to deal with.'' Gerrard says that cells' controls and blank
 space take extra screen space compared to a straight command line, and he and
-KC both complained that when there are long results or plots, it causes the
-code to scroll off the screen. Raif explains, ``Imagine I run a cell whose
+Tina both complained that when there are long results or plots, it causes the
+code to scroll off the screen. Wendy explains, ``Imagine I run a cell whose
 output is huge. [...] To find the next cell, I have to scroll down and find
@@
 Some users would like the option to keep results separate from code.
-Ivan thinks that RStudio's layout is more helpful because charts
-are kept in another pane and stay in one place while doing analysis. Raif says
+Gerrard thinks that RStudio's layout is more helpful because charts
+are kept in another pane and stay in one place while doing analysis. Wendy says
 ``cells are really useful'' but ``you want to see your output in a different
@@
 The cell structure also can be problematic if some cells take a long time to
-run. In Kenny's work, there is often a ``long tail''.  The first cell may take
+run. In Coby's work, there is often a ``long tail''.  The first cell may take
 seconds to run, the next cell minutes, and the last cell 5 hours.  In this
 situation, the Run Whole Notebook button is ``dangerous''. When converting
-his work to notebooks, Kenny ends up with a lot of comments that say ``this cell
+his work to notebooks, Coby ends up with a lot of comments that say ``this cell
 takes a long time to run.''
@@
 To avert this danger, some users write code within the cells to
-cache results. Kenny ends up ``littering the notebook with little switches that
+cache results. Coby ends up ``littering the notebook with little switches that
 comment out the slow parts'' and either ``load the object instead'' or ``save
@@
 
-%% Rick on auditing (possible solution): ``you didn't see your mistakes, they were
+%% Kenyon on auditing (possible solution): ``you didn't see your mistakes, they were
 %% in there but you never really saw them because a mistake never led to an answer
@@
 
-%% RCloud combines the functions of scratchpad with notebook assets. KC thinks
+%% RCloud combines the functions of scratchpad with notebook assets. Tina thinks
 %% that having a scratch pad that doesn't persist between different notebooks is
@@
 %% The one-window design also makes some users feel that the side panes are taking
-%% up too much space. Ivan: having ``long term'' code (which would be in assets)
+%% up too much space. Gerrard: having ``long term'' code (which would be in assets)
 %% above the command prompt, like in RStudio, means that both can be sized very
@@
 
-%% Kenny: Problem with screen real estate: side panes take up too much
+%% Coby: Problem with screen real estate: side panes take up too much
 %% space. Doesn't need the notebook tree pane.
 
-%% Corey: In the previous version, it switched between code and results. Now you
+%% Kaylee: In the previous version, it switched between code and results. Now you
 %% see both - ``and there are times when that's good, but there's certainly times
@@
 
-%% Jyothi: ``It's good that I can see the results but if I'm working and I need
+%% Allison: ``It's good that I can see the results but if I'm working and I need
 %% more room and I want to hide the results, I can't.'' Could we have ``an option
@@
 
-For this reason, Rick doesn't find the notebook tree very useful:
+For this reason, Kenyon doesn't find the notebook tree very useful:
 \begin{quote}
@@
 Although RCloud promises an environment where notebooks stay working, our
-users have not all learned the habits that make this a reality. As Byers
+users have not all learned the habits that make this a reality. As Joy
 puts it, there is still a big problem with ``bitrot'' -- notebooks often
 stop working because the user changed the structure of their data, or
-changed a filename or a database.  He says we need ``organizational
+changed a filename or a database.  She says we need ``organizational
 protocols'' to catch up with the technology. Even if one forks someone
@@
 
-Taylor, who writes packages and example notebooks for them, seems to be
-haunted the most by dead notebooks. Taylor and Horace work together on a notebook:
+Evelyn, who writes packages and example notebooks for them, seems to be
+haunted the most by dead notebooks. Evelyn and Hugh work together on a notebook:
 \begin{quote}
@@
 The problem of dead notebooks is compounded by changing
-packages and their example notebooks at the same time. Taylor will ``go back and
+packages and their example notebooks at the same time. Evelyn will ``go back and
 change it and update the page, but then what happens is people have in the
@@
 
-Taylor tried to keep his notebook tree tidy, but this didn't help, because
+Evelyn tried to keep his notebook tree tidy, but this didn't help, because
 people either kept old links in their email, or forked a notebook which is now
@@
 
-%% Bart: ``One UI notebook and a bunch of other notebooks that do various tasks and
+%% Iris: ``One UI notebook and a bunch of other notebooks that do various tasks and
 %% are called with API calls. The problem is, if you clone all the notebooks, then

### 2015-03-31 introduction.tex

@@
 
-From a more principled or theoretical perspective, visual analytics goals
-or objectives for emerging systems, described in previous work, closely match
-the goals of our prototype. Adopting the terminology of X Y and Z, we noted
-the following.
+From a more principled or theoretical perspective, certain visual analytics
+goals or objectives for emerging systems, described in previous work,
+closely match ours. We noted the following central themes.
 

### 2015-03-31 interviews.tex
working

--- 2015-03-31 interviews.tex
@@
 
-To evaluate the effectiveness of RCloud, we interviewed 13 current and past
-users of RCloud. Of these, 9 are data analysts and 4 build tools for data
+To evaluate the effectiveness of RCloud, we interviewed 13 current or
+recent users of RCloud. 9 are data analysts, and 4 build tools for data
 analysts and business needs.
 
-Generally, analysts were more annoyed by the cell interface and the
-perceived slowness of hte interface than web/tool developers. Analysts
-switching to R from another language were more likely to use RCloud for
-exploratory data analysis. Web/tool developers were more likely to see the
-need for parameterization.
+The subjects differed in their evaluation of its benefits and disadvantages,
+as described below.
+
+%Analysts were more annoyed by the cell interface and its perceived
+%slowness than were web/tool developers. Analysts
+%switching to R from another language were more likely to use RCloud for
+%exploratory data analysis. Web/tool developers were more likely to see the
+%need for parameterization.
 
@@
 Sharing of notebooks is the core feature of RCloud, and a popular one. All
-active users we spoke to praised this feature.
+the subjects praised this feature.
 
@@
 
-By default, all notebooks are publicly visible in RCloud, and notebooks can be
-found through the notebook tree or search. However, the method for sharing that
-users mentioned most is sending links through email. Clint says, ``If my supervisor
-wants to see what I've done or QA it, I can just send her a link.''
+By default, all RCloud notebooks are publicly visible, and notebooks can be
+found by navigating the notebook tree, or by searching. However, users most
+often mentioned sharing by sending links through email. Clint says,
+``If my supervisor wants to see what I've done or QA it, I can just send her a link.''
 
-Besides providing a way to present work to others, notebook sharing can also
-provide a starting point. Clint says, ``I think the best part about it is how
-easily you can share code. [...] You can find a working example, rather than wearing
+Besides providing a way to present work, notebook sharing can provide a
+starting point for coding. Clint says, ``The best part is how easily you can
+share code. You can find a working example, rather than wearing
 out Google and finding questionable examples that may or may not work.'' Raif
-notes, if ``some person has done something similar, then you're able to just
-edit that, and that's saved a lot of work time for me''
+notes, ``[If] some person has done something similar, then you're able to just
+edit that, and that's saved a lot of work time for me.''
 
 Taylor develops packages for analysts, and uses RCloud ``more for sharing code with
-other people, and for doing tutorials for iotools or hmr'', his packages.
+other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
 ``I want people to see how the package works, so I clearly want them
-to see the code [...] I kind of write it just like I would write GitHub
+to see the code... I write it just like I would write GitHub
 Markdown, where you have little code snippets and text, but RCloud lets me
-actually run the snippets'' and display the results. He also uses RCloud for
-explaining problems with data sources.
+actually run the snippets [and display the results].'' He also uses RCloud for
+describing and debugging  problems with data sources.
 
-Some users who tried out RCloud but are not able to use it for organizational
-reasons, already miss the functionality. Shelly likes ``the concept of being
-able to create notebooks and share them.'' A ``wiki is not the best way for
+Some users, who tried RCloud and were not able to continue for organizational
+reasons, miss certain capabilities. Shelly ``[likes] the concept of being
+able to create notebooks and share them... A wiki is not the best way for
 communicating results - it's kind of like writing a blog post with very limited
-functionality [...] I have to save every picture and post it as an image'' and
+functionality. I have to save every picture and post it as an image.'' She said,
 ``I can't share all of the code because it would just get crowded and wouldn't
-look right on a wiki.'' Bart says, ``if I could make a folder on RCloud and have
-Python notebooks and also Pig notebooks there, and execute them from RCloud,
-that would be much better than my current thing, because that would free me from
-manual documentation and version control and also telling people where my code
-was. It would be just, hey, go look on RCloud, here's the stuff.''
+look right on a wiki.'' Bart explained, ``If I could make a folder on RCloud and
+have Python notebooks and also Pig notebooks there, and execute them from RCloud,
+that would be much better than my current [environment], because that would free
+me from manual documentation and version control and also telling people where
+my code was. It would be just, hey, go look on RCloud, here's the stuff.''
 
@@
 \subsection{Forking}
-The ability to start work where someone else let off, by forking, has proved to
+The ability to start work where someone else let off, by forking, proved to
 be a popular feature. In fact, almost twice as many (131) users have forked
-someone else's notebook as have starred one (75).  Clint says, ``it's one of the
+someone else's notebook as have starred one (75).  Clint says, ``It's one of the
 handiest functions, because instead of having to find it, copy, paste it, you
-just hit Fork, you rename it, and it's done. It's pretty amazing.''
+just hit Fork, rename it, and it's done. It's pretty amazing.''
 
-Although we always intended forking to be used to improve others' code, we we
-initially didn't support forking one's own notebooks, and this has proved very
-useful. Rick says ``I fork my own notebooks because I'm going off and doing some
-other analogous project, so I've got interesting content that I've already done
-in a previous analysis that I want to start from and then tweak to match a new
-set of data.''
+Although we intended forking to be available to improve others' code, we 
+initially didn't anticipate support forking one's own notebooks, which
+proved very useful. Rick says, ``I fork my own notebooks because I'm going off
+and doing some other analogous project, so I've got interesting content that I've
+already done in a previous analysis, that I want to start from and then tweak
+to match a new set of data.''
 
@@
 wrong.  When Horace works with the users of his notebooks, ``I'll teach people
-to intercept the result in the middle,'' to ``insert print statements here and
+to intercept the result in the middle, to insert print statements here and
 there and check values.''
@@
 
-An extremely common but more problematic use of forking is to change
+A common but somewhat problematic use of forking is to change
 parameters. Raif: ``I've been forking other people's notebooks [because] I
-want to run it on a different part of data, or I want to change some parts,
+want to run them on a different part of data, or I want to change some parts,
 I don't want to see this column, I want a different column, things like
@@
 
-Currently parameters can be added to the URL, but adding user interface
-elements to do the same thing requires expertise. The tool-builders see a
+Parameters can be added to a notebook URL, but adding user interface
+elements to do the same thing requires expertise. Tool builders see a
 need for facilities to make it easier. Jyothi calls it ``web-enabling'' the
-notebook: users ``can always fork the notebooks and make changes, but I
-feel that if the owner of the notebook web-enables it, [...] it's easier to
+notebook: ``[users] can always fork the notebooks and make changes, but I
+feel that if the owner of the notebook web-enables it, it's easier to
 run. Instead of forking it, if they can set options, it's probably more
-efficient, and also they can [still] fork it if they want to.''
+efficient, and they can [still] fork it if they want to.''
 
@@
 
-Automatic source control is also a popular feature. Clint says, ``instead of
+Automatic source control is also a popular feature. Clint says, ``Instead of
 looking back and saying I've got a billion files here in this subdirectory and I
@@
 I like the fact that it has a built-in editor, so
-if you need to fix a typo in a link, or an extra line break or any of this other
+if you need to fix a typo in a link, or an extra line break or other
 nonsense, you can just switch to the edit view, pull up your asset, type
@@
 On the other hand, saving every change leads to many fine-grained
-versions. Rick says that for this reason, the history feature is not all that
+versions. Rick says that for this reason, the history feature is not that
 helpful: ``I don't need something that keeps track of every mistake I've made or
@@
 \subsection{Discovering others' work}
-Many users find the search function useful. Raif says ``the fact that we have
+Many users find the search function valuable. Raif says, ``The fact that we have
 all the notebooks there, searchable, saves me from replicating what other people
@@
 
-But other users prefer to browse just the notebooks of the experts they
-know. Rick says, ``usually I know somebody's notebooks that I want to search
-through [...] because I know that the kind of thing I'm looking for is something
+But other users prefer to browse just the notebooks of experts they
+know. Rick says, ``Usually I know somebody's notebooks that I want to search
+through, because I know that the kind of thing I'm looking for is something
 more obscure than I'm likely to find in some random person's.''
 
-More selective ways to search will be needed as the volume of notebooks
-continues to grow. We explore some ideas in Section~\ref{sec:discussion}.
+More selective ways to search will be needed as the number of notebooks
+grows further. We explore some ideas in Section~\ref{sec:discussion}.
 
@@
 Integrating an analysis language into a web development environment is something
-the tool developers really appreciate.  The structure of Horace's visualization
+tool developers really appreciate.  The structure of Horace's visualization
 notebook means
 \begin{quote}
-the other guys who want to do analytics on the data, they can first pull
-the data, do the analytics on it, and then feed the viewer the data. So
-[...] anyone from the stats group [...] can always insert something in
-between. [...] Once you get the data into RCloud, then you have a
-dataframe to work with, and then they know how to produce another
-dataframe.
+the other guys who want to do analytics on the data can first pull
+the data, do the analytics on it, and then feed the viewer the data.
+Anyone from the stats group can insert something in between.
+Once you get the data into RCloud, then you have a dataframe
+to work with, and then they can produce another dataframe.
 \end{quote}
 
-Integrating R also helps Jyothi's application: ``Having an open session
-where I can run R commands or R functions without having to invoke an API
-or send out a request and then wait for the request to come back [...] is
-extremely helpful.''
+Integrating R also helps Jyothi's topic modeling tool, described in the
+last section: ``Having an open session where I can run R commands or
+functions without having to invoke an API or send a request and then
+wait for the response is extremely helpful in writing the application.''
 
@@
 RCloud is less popular as a tool for exploratory data analysis. Every
-analyst with prior R experience still does analysis in another tool and
-pastes their code into RCloud for sharing.
+analyst with prior R experience still prefers to do analysis in another
+tool, and paste code into RCloud for sharing.
 
-This is partly out of familiarity. Taylor asks, ``why would I want to use
+This is mainly out of familiarity. Taylor asks, ``Why would I want to use
 RCloud over my current setup? If it's just me, I like my text editor and
@@
 
-The web interface of RCloud doesn't work for Rick: ``It's nice to have
+The web interface of RCloud isn't satisfactory to Rick: ``It's nice to have
 it saved, but there's this trade-off between it making it easier for me to
 present something or to save something, and my ease of typing and
-correcting and things in a plain old editor window.''
+correcting and things in a plain editor window.''
 
-Kenny also works with text file and command line, rearranging a file so the
-good code is at the top of the file. When he's done, all the the right is
-code at the top. RCloud does not readily support this workflow, but Kenny
-often shares the work by pasting it into an RCloud notebook when he's done.
+Kenny also works with text files and command line, rearranging a file so the
+good code is at the top of the file. When he's done, all the right code is
+at the top. RCloud does not readily support this workflow, but Kenny
+often shares work by pasting it into an RCloud notebook when he's done.
 
@@
 
-The interface takes some learning. Clint says at first,
+The interface takes some learning. Clint says,
 \begin{quote}
@@
 
-Just the difference between a cell and the command prompt can confuse and slow
-down users. Rick explains:
+Simply the difference between a cell and the command prompt can confuse some people.
+Rick explains:
 \begin{quote}
-If I was typing into one of the cells near that top, I had
-to think ``I'm editing this cell and then I have to execute it,'' and if was doing
-the one at the bottom I could type it but then it would automatically execute
-[by pressing enter], but then it became a standard cell and I have to edit
-it.
+If I was typing into one of the cells near that top, I had to think ``I'm editing
+this cell and then I have to execute it,'' and if I was editing one at the bottom,
+I could type it but then it would automatically execute [by pressing enter].
+%, then it became a standard cell and I have to edit it.
 \end{quote}
@@
 
-Much of the time, commands typed at the R command line just serve a transitory
-purpose, so having RCloud save commands into cells can be annoying. Kenny notes, ``It saves everything I do like everything is gold, but most
-of it is junk not meant to be saved.'' Rick says ``I just want to type in a
+Much of the time, commands typed at the R command line serve only a transitory
+purpose, so having RCloud persist commands in cells can be annoying.
+Kenny notes, ``It saves everything I do like everything is gold, but most
+of it is junk not meant to be saved.'' Rick says, ``I just want to type in a
 couple of quick commands and get some results that are going to tell
@@
 
-The material that is not appropriate to save includes ``expressions that allow
+Material that is not appropriate to save includes ``expressions that allow
 me to check that I'm the right track'' (Rick), ``checking out what your data is,
@@
 
-Users also felt that cells do not capture the right level of granularity: they
+Users felt that cells do not capture the right level of granularity: they
 either hold too much code or too little. Rick says that RCloud's cell structure
 ``tempts me to type a big long thing and then run the whole thing, as opposed to
-typing a few little pieces [...] on the inner parts and then put
-them together to make the big thing'' as he would do on the command line. When
-Taylor uses the R command line, he copies and pastes ``like 5-10 lines of code,
+typing a few little pieces and then put them together'' as he would on the command line.
+When Taylor uses the R command line, he ``copies and pastes 5-10 lines of code,
 so when something breaks, I get an error message on that one line, and I can
 up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell,
-so the only way to get that same functionality is if every line's in one
-cell.''
+so the only way to get that functionality is if every line's in one cell.''
+Another complaint related to the human interface was that cells can take up
+substantial vertical space, requiring scrolling.
+% Rick complains, ``Every time I type more stuff, the notebook gets longer and
+% longer and it's harder to deal with.'' Ivan says that cells' controls and blank
+% space take extra screen space compared to a straight command line, and he and
+% KC both complained that when there are long results or plots, it causes the
+% code to scroll off the screen. Raif explains, ``Imagine I run a cell whose
+% output is huge. [...] To find the next cell, I have to scroll down and find
+% where that cell starts. So I'm losing the continuity of my code.''
 
-Cells can also take up substantial vertical space, requiring scrolling. Rick
-complains that ``every time I type more stuff, the notebook gets longer and
-longer and it's harder to deal with.'' Ivan says that cells' controls and blank
-space take extra screen space compared to a straight command line, and he and
-KC both complained that when there are long results or plots, it causes the
-code to scroll off the screen. Raif explains, ``Imagine I run a cell whose
-output is huge. [...] To find the next cell, I have to scroll down and find
-where that cell starts. So I'm losing the continuity of my code.''
-
-Some users would like the option to keep results separate from code.
-Ivan thinks that RStudio's layout is more helpful because charts
-are kept in another pane and stay in one place while doing analysis. Raif says
-``cells are really useful'' but ``you want to see your output in a different
-window [or] on a separate part of the window''.
+Some users would like a way to keep results separate from code.
+Ivan thinks RStudio's layout is more helpful because charts
+are shown in another pane that stays in place while doing analysis. Raif says
+``Cells are really useful, [but] you want to see your output in a different
+window or on a separate part of the window''.
 
 The cell structure also can be problematic if some cells take a long time to
-run. In Kenny's work, there is often a ``long tail''.  The first cell may take
-seconds to run, the next cell minutes, and the last cell 5 hours.  In this
-situation, the Run Whole Notebook button is ``dangerous''. When converting
-his work to notebooks, Kenny ends up with a lot of comments that say ``this cell
-takes a long time to run.''
+execute. In Kenny's work, there is often a ``long tail''.  The first cell may
+take just seconds to run, the next cell minutes, and the last cell 5 hours.
+In this situation, the Run Whole Notebook button is dangerous.
+Kenny reports that when converting his work to notebooks, he
+ends up with a lot of comments saying ``This cell takes a long time to run.''
 
-To avert this danger, some users write code within the cells to
-cache results. Kenny ends up ``littering the notebook with little switches that
-comment out the slow parts'' and either ``load the object instead'' or ``save
-the object to disk''. One of the authors of this paper manually writes cells
-that check if a result file exists, and perform the calculation only if it's not
-there.
+To avert this pitfall, some users write code within cells to explicitly cache results.
+Kenny ends up ``littering the notebook with little switches that comment out the
+slow parts'' and either load or save the object to disk. One of the authors of this paper
+manually writes cells that check if a result file exists, and perform a calculation only
+if it's needed.
 
@@
 \subsection{A sea of notebooks}
-RCloud is starting to be a victim of its own success, as it is becoming
-difficult to navigate all the notebooks.  There are over 5200 notebooks on
-the research instance, of which more than 500 have been starred and more
+RCloud is becoming a victim of its own success, as it is becoming
+difficult to navigate all the notebooks.  There are over 5200 notebooks in
+the research instance, of which more than 500 have been starred, and more
 than 350 have been forked.
 
-For this reason, Rick doesn't find the notebook tree very useful:
+For this reason, Rick doesn't find the notebook tree satisfactory:
 \begin{quote}
-I don't necessarily need to see everybody's notebook that uses RCloud [...] Every
-time I do something new, I get a new notebook and so now my notebooks are maybe
-50 or 60. That's enough to think about just on my own, but if I've got
-everybody's 50 or 60 sitting on my display, I find that it's more than I want to
-know about.
+I don't necessarily need to see everybody's notebook that uses RCloud. Every
+time I do something new, I get a new notebook, so now I have 50 or 60 notebooks.
+That's enough to think about just on my own, but if everybody has
+50 or 60 sitting on my display, it's more than I want to know about.
 \end{quote}
 
-Although RCloud promises an environment where notebooks stay working, our
-users have not all learned the habits that make this a reality. As Byers
+Although RCloud promises an environment where notebooks should keep working,
+not all our users have learned the habits that make this a reality. As Byers
 puts it, there is still a big problem with ``bitrot'' -- notebooks often
@@
 
-Taylor, who writes packages and example notebooks for them, seems to be
-haunted the most by dead notebooks. Taylor and Horace work together on a notebook:
+Taylor, who writes packages and example notebooks for them, mentioned the
+accumulation of dead notebooks. When Taylor and Horace work together on a notebook:
 \begin{quote}
 There's no way for both of us to have ownership of a notebook, so the only
-way is to fork it back and forth, and so we have dozens of old copies. [We
-end up] deleting all the old ones, but people still have links to them,
+way is to fork it back and forth, and so we have dozens of old copies. We
+end up deleting all the old ones, but people still have links to them,
 because they don't actually disappear.
@@
 The problem of dead notebooks is compounded by changing
-packages and their example notebooks at the same time. Taylor will ``go back and
+packages and example notebooks at the same time. Taylor will ``go back and
 change it and update the page, but then what happens is people have in the
-meantime forked it. [...]  I'll have to change a package as well, so their
-old fork stops working, and then they complain.''
+meantime forked it. I'll have to change a package as well, so their
+old fork stops working, and they complain.''
 
-Taylor tried to keep his notebook tree tidy, but this didn't help, because
-people either kept old links in their email, or forked a notebook which is now
-obsolete. By design, notebooks that are deleted in the user interface are not
-purged from the repository. ``Now I'm really gun-shy about sharing notebooks
-because it's like, do I want to support this forever?''
+Taylor tried to keep his notebook tree well organized, but this didn't help,
+because some people kept old links in their email, or forked notebooks which 
+had become obsolete. By design, notebooks that are deleted in the user interface
+are not purged from the repository. This caused some subjects to express
+caution about sharing notebooks too freely.
+%``Now I'm really gun-shy about sharing notebooks
+%because it's like, do I want to support this forever?''
 

### 2015-03-31 introduction.tex
working, also spell checked

@@
 This situation can affect how data science teams; as analysts and
-toolbuilders, we have experienced this first-hand in our environment, and
+tool builders, we have experienced this first-hand in our environment, and
 others report similar findings. Gutierrez collected interviews with data

### 2015-03-31 conclusion.tex
yes

@@
 visual analytics, we designed an environment
-for exploration, sharing, presentation and publishing.
+for exploration, sharing, presenting and publishing.
 We implemented it in the RCloud prototype.
@@
 science teams and the organizations in which they work benefit from
-having capabilities to support collaboration and to integrate
+capabilities to support collaboration and to integrate
 the entire visual analytics process. We found that features for

### 2015-03-31 system.tex
intro and related work cleanup, no more missing references
system

--- 2015-03-21 system.tex
@@
 
-\subsection{High-level Architecture\label{sec:highlevelarchitecture}}
+One of the starting points in our design goals was to leverage, as
+much as possible, the current infrastructure of the world-wide
+web. HTTP, with all its warts, has become the lingua franca of
+distributed interprocess communication. This meant that if our system
+spoke HTTP natively, then it was possible to provide a low-friction
+path from development of an interactive data analysis script to an
+automated \emph{web service}, essentially a function call over the
+web, which can then be used in higher-level tools. An example of this
+kind of transition is described in Section~\ref{sec:casestudy}.
 
-% Stephen: Let's work from front to back, from human interface to the
-% services and frameworks that support it.
-
-The internal computing infrastructure of organizations has changed
-radically in the last fifteen years. The shift from large
-servers toward scalable, lower-cost, distributed systems (``the cloud'')
-led to a software ecosystem of processes distributed over a
-network, usually communicating via HTTP. HTTP is dominant because
-web browsers and servers are ubiquitous and available in almost
-all hardware devices, from tiny sensors, to handheld devices and
-laptops, to rack-mounted servers.
-
-As a result, HTTP is the lingua franca of interprocess communication
-(IPC). One of the design goals for RCloud was to provide a productive
-environment for creators of data-analysis scripts, that also behaves
-as a first-class citizen in the pre-existing ecosystem of computer
-services and networks in an organization.
-
-%% Design for cloud-friendly? This goes back to the point in the
-%% introduction about playing nice with the rest of the ecosystem.
-
-\subsection{Human Interface\label{sec:humaninterface}}
-
-%\stephen{What??}
-%We designed RCloud around the front-facing API, which provides roughly
-%one entry point to correspond with each requirement.
-
-Figure~\ref{fig:something} shows the RCloud developer interface.
-The center panel is a notebook for code and visualizations.
-Code is edited in this panel, and visualizations are rendered.
-An inventory of supplementary ``asset'' files, that are part of
-the notebook, though not in its executable flow, are edited on the right.
-Controls at the top of the screen allow the user to run, view, and
-share the notebook. The result of code execution is the final cell of
-the notebook.  On the left is a browsing area for searching and viewing
-other users' notebooks, and a help system.
-
-The call mechanism is the URL itself; arguments are read from the URL
-and the final cell is rendered as the result.
+As a result, the entire high-level infrastructure of RCloud, shown in
+Figure~\ref{sec:system} uses web standards. The communication between
+the web browser and the active R session as a user edits a notebook is
+a combination of HTTP and Websockets, while all other IPC happens
+through HTTP, from notebook versioning in GitHub to building and
+maintaining full-text indices through SOLR.
+The most peculiar feature of RCloud's runtime system is its tight integration between the web browser and the backend R process, which we describe in Section~\ref{sec:Rinbrowser}.
 
@@
 
-The main unit of computation in RCloud is a \emph{notebook}.
-A notebook holds a sequence of \emph{cells}, each of which contains a
-snippet of code or hypertext in Markdown. This is not a novel idea;
-executable documents structured this way are a feature of many
-other systems, including Mathematica, IPython and Sage.
-\stephen{Code is executed when...?}
+The main unit of computation in RCloud is a \emph{notebook}.  A
+notebook holds a sequence of \emph{cells}, each of which is a snippet
+of code or hypertext in Markdown. As mentioned in
+Section~\ref{sec:related}, this is not a novel idea: executable
+documents structured this way are a feature of many other systems,
+including Mathematica, IPython and Sage.
 
-One of the main contributions of RCloud is the idea that notebooks
-are ``always deployed''. In other words, the most recent version of
-a notebook is immediately available.\stephen{What if it is broken?}
-This makes it convenient to share and modify experiments and compose
-results without binding to a specific version of a notebook.
-On the other hand, there are situations where it might be important
-to name specific versions. We do not expect designers to decide which
-versions need to be preserved, but we embrace \emph{transparent} versioning.
-This is similar to models like Jankun-Kelly et al.'s p-set calculus \cite{Jankun-Kelly:2007:MFV}
-and VisTrails's version tree \cite{Callahan:2006:VVM}, where every change in the state of the system is tracked.
+Notebooks can be executed one cell at a time during an interactive
+editing session, providing a similar experience to traditional
+read-eval-print loops, or can also be executed all at once, providing
+a similar experience to running a shell script.
 
-To implement this, we built RCloud on top of Github's \emph{gists}~\cite{Github:2014:GG}.
-Github offers a HTTP interface for creating simplified git repositories, the main limitation
-being a restriction to text-only files in a single directory. The GitHub web-service
-API provides most of the semantics we need for the versioning portion of the storage back end:
-access to previous versions, comments, starring, and forking.
+One of the main contributions of RCloud is the idea that notebooks are
+``always deployed''. In other words, the most recent version of a
+notebook is immediately available to all other users of the system as
+the notebook is being developed. One way to describe this is that
+RCloud lacks a ``save'' button: any notebook cell that runs is always
+associated to a notebook version serialized to disk. Although one of
+the interviewed users reports that this sometimes leads to an
+excessively fine-grained sequence of versions (see
+Section~\ref{sec:interviews}), we believe that the alternative of
+losing automatic sharing is worse (and see
+Section~\ref{sec:interviews} again for users reporting their
+satisfaction over the low-friction shareability).
 
-Using GitHub for storage and versioning also exposes other capabilities
-that can be invoked with JSON and HTTP.
-Particularly, we can provide full text search using Apache SOLR.
-SOLR is scalable, can index in near-realtime, supports multiple character sets,
-indexes several common types of documents, and has schemas and faceted search.
-%
-Integrating existing services and software components, especially open source,
-instead of implementing custom code is a trend in the software industry, and
-is very positive for systems research and prototyping in visual analytics.
-By adopting existing technology, small teams with limited resoures can explore
-novel ideas toward the improvement of large, complex software ecosystems.
-In our case, even though this approach involved learning standards not directly
-related to visual analytics, our early decision to use the GitHub gist API
-turned out to be successful. Future projects could gain many of the same benefits
-by adopting this strategy. In fact, it will be essential for technology adoption
-and transfer, because it is almost impossible for any one tool or platform to
-``own'' or support the entire visual analytics process.
-\stephen{Are we preaching too much?}
+When we combine this with the ability to refer to the latest version
+of a notebook by name, we get notebooks that are always live, but
+sometimes broken. When stability is important, we also
+allow any previous version of a notebook to be tagged and referred
+to. These are similar to models like Jankun-Kelly et al.'s p-set calculus
+\cite{Jankun-Kelly:2007:MFV} and VisTrails's version tree
+\cite{Callahan:2006:VVM}, where every change in the state of the
+system is tracked.
+
+Our implementation of the versioning mechanism is built on
+top of GitHub's \emph{gists}~\cite{GitHub:2014:GG}, which are an HTTP
+interface for simplified repositories. The GitHub web-service API
+provides most of the semantics we need for the versioning portion of
+the storage back end: access to previous versions, comments, starring,
+and forking. This provides the additional benefeit that every RCloud
+notebook can also be manipulated like a git repository. Advanced users
+then have access to features like command-line checkout, history
+editing, and so on.
 
@@
 
-Information retrieval based on collecting usage and recommendations
-is a cornerstone of modern web services. We would like to help data
-scientists to find workbooks (and therefore items in their contents
-such as code, data, and colleagues with specific expertise) with
-the benefit of such information.
+One side advantage of centralizing the execution and storage of
+notebooks is that it becomes feasible to collect usage information
+that would otherwise be lost. In the case of social data visualization
+platforms, we would like to leverage this usage data to help analysts
+find content of interest, whether this content is actual source code
+or accompanying data. The standard way to achieve this is through
+\emph{user-generated curation} and \emph{automatic recommendations}.
 
-In RCloud, reputation and interest are a relationship between
-\emph{notebooks} and \emph{users}, rather than a relationship between
-user pairs. We chose this approach because we expect initial 
-RCloud deployments to have relatively few users, but some users to
-create many notebooks. Under that assumption, assigning interest
-to users would not provide sufficiently ``high-resolution'' data.
-
+Automatic recommendations have become famous in the user experience
+provided by companies such as Amazon and Netflix (``if you've liked
+this film, then you'll like this other film''). In order to build such
+recommendations, we need users to \emph{curate} the collection, by
+providing explicit reviews or some other method of indicating interest.
+%%
+%% In RCloud, reputation and interest are a relationship between
+%% \emph{notebooks} and \emph{users}, rather than a relationship between
+%% user pairs. We chose this approach because we expect initial 
+%% RCloud deployments to have relatively few users, but some users to
+%% create many notebooks. Under that assumption, assigning interest
+%% to users would not provide sufficiently ``high-resolution'' data.
+%%
 We incorporate both explicit and implicit indications of interest
@@
 always readily available, to encourage its use.
+%
+The current version of RCloud uses only simple aggregate counts of
+stars to indicate overall notebook interest, mostly because the
+recommendation system approaches popularized by the Netflix prize
+require large recommendation sets to paying off. Nevertheless, the
+starring mechanism is sufficient to create personalized
+recommendations~\cite{Hu:2008:CFF}.
 
-Implicit signaling of interest is supported by keeping click-through counts
-\cite{Joachims:2005:AIC} and execution counts. (In addition to these
-standard techniques of collecting feedback from web search, we anticipate
-applying static and dynamic code analysis to infer fine-grained
-information about relationships, for example, which packages and data
-sets often appear together.)
-
-\subsection{Executing R in a web browser\label{sec:Rinbrowser}}
-
-One our main goals is to provide ubiquitous access to the statistical
-tools of the R programming language in the wider environment in which data
-science teams participate.
-%
-Given the recent developments of interactive visualization and visual
-analytics in HTML5 and the web, we decided to create an
-\emph{interconnect} between the two environments.
-%
-Every RCloud session spawns a new R execution process in a remote
-server.
-
-%% \subsubsection{Object capabilities: R to Javascript RPC and access
-%%   control}
-
-HTTP and web services are convenient but much of the protocol is
-stateless. (For example, GET requests are required to not change the
-remote state, and results can be cached in transparent proxies along
-the network). In the case of close communication between a running
-R process and a web browser, we need something that is more flexible
-than distinct URLs for every R process, and which embraces
-statefulness of both calculations and visualization parameters.
-
-\carlos{In-depth explanation of R<->web interconnect, and why it's
-  necessary or important. Main point: make programming in the R side
-  as close as possible to what R programming feels like, and make
-  programming in the Javascript side as close as possible to what
-  Javascript programming feels like. We do this because this part of
-  the system is targeted at programmers; Shiny targets web programming
-  at non-web-programmers. We don't attempt such a thing and land at a
-  different point in the spectrum.}
-
-\carlos{compare against https://www.opencpu.org, compare against
-  IPython notebooks. Two-way communication? Actually describe what it
-  is. IPython manipulate notebooks? Understand and contrast}
-
-\subsection{Interactive notebooks\label{sec:interactivenotebooks}}
-
-%% How do we do things that are not trivial to do with IPython (for
-%% example)
-
-%% What is relationship to a standard CMS with R/Python
-
-%% dcplot. two-way communication between between backend session and
-%% frontend session.
-
-A key engineering decision in RCloud was to rely on full two-way
-communication between the web client human interface, and data and
-computing resources in the cloud. It is easier to engineer a design
-in which the human interface is pushed one-way from the server
-and thereafter all interaction takes place within the client.
-Two-way communication is necessary, though, where the size of
-the data or computational performance makes it impractical to
-move all computation to the client.
-
-In addition, R is well suited as a language for analysis, and
-JavaScript for interactive visualization. To draw on
-the strengths of both languages and both environments, the connection
-between the languages is not just procedural: it makes \emph{closures}
-and \emph{first-class functions} available across the network.
-This provides considerable flexibility, so that for example, a chart built
-with dc.js or leaflet.js can call back to analysis functions in R
-without having to formalize the protocol between the processes.
+Implicit signaling of interest is supported by keeping click-through
+counts~\cite{Joachims:2005:AIC} and execution counts. In addition to
+these standard techniques of collecting feedback from web search, we
+anticipate applying static and dynamic code analysis to infer
+fine-grained information about relationships, for example, which
+packages and data sets often appear together, in the style of Fast et
+al's Codex system.~\cite{Fast:2014:ECS}.
 
@@
 
-Every notebook in RCloud is named by a URL, and notebooks by
-default are visible by the entire organization. This is deliberate.
-As pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS},
-broad access to analysis outputs (in their case, for NameVoyager)
-increases long-term engagement in part through cross-references on
-the web. Although our prototype RCloud deployment is only visible
-inside a corporate intranet, we nevertheless found anecdotal support
-for this notion by discovering links to RCloud notebooks in internal
-discussion fora and mailing lists.
+Every notebook in RCloud is named by a URL, and notebooks by default
+are visible by the entire organization. This is deliberate.  As
+pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS}, broad
+access to analysis outputs (in their case, for NameVoyager) increases
+long-term engagement in part through cross-references on the
+web. Although our prototype RCloud deployment is only visible inside a
+corporate intranet, we nevertheless found support for this notion by
+discovering links to RCloud notebooks in internal discussion fora and
+mailing lists. In addition, as we describe in
+Section~\ref{sec:interviews}, users have almost unanimously adopted
+``share-by-URL'' as their default communication mechanism, as opposed
+to ``share-by-screenshot'', which we consider to be an encouraging
+validation of the system.
 
-Because notebooks are continually published, any work is immediately
-available both as a subroutine and a visual component. Close colleagues
-can start on the next stage of analysis, or delve into the data,
-even while the original author is polishing an algorithm or its
-presentation. \stephen{We glossed over the things that can go wrong.
-We should at least discuss that briefly.}
-The code is the page, and can either be shaped into
-a function of inputs and outputs, or the linear cells of the notebook
-can be reworked into a full-fledged HTML layout.
+\subsection{Executing R through a web browser, and Javascript through an R process\label{sec:Rinbrowser}}
+
+As mentioned before, the other main goal in RCloud was to provide, as
+much as possible, full access to the R programming language during the
+\emph{development} of a data analysis notebook.
+At the same time, when notebooks are \emph{deployed} (and potentially
+accessed by anyone with a web browser), we'd like to allow a web
+browser to invoke only a very limited subset of R, namely those
+notebooks that have been published.
+
+The solution we have developed is simple and general, and is directly
+inspired by Miller's \emph{object
+  capabilities}~\cite{Miller:2006:RCT}. Specifically, the R layer that
+communicates with javascript does not expose naked evaluation of
+arbitrary functions. Instead, every function that the R layer intends
+to expose is associated with a large, cryptographically-safe random
+number (informally, a ``hash''). This random number is then sent
+across the wire. This number is interpreted as an opaque identifier to
+a function. Because these identifiers are cryptographically safe, all
+that the Javascript layer can do is return them to the R side, in a
+message requesting that this function be called on its behalf. In that
+case, the results of this function call might include \emph{new}
+opaque identifiers, exposing new ``capabilities'' to the web browser.
+
+The same idea of exposing functionality via hashes can be used to give
+the server-side of RCloud access to Javascript functions. This allows
+R libraries to respond to user input in the browser, giving them
+access to features ranging from password prompts to the currently
+selected set of points in an interactive linked brushing
+visualization.
+
+As a result, the features required to provide safe access to R from
+the client side, and Javascript access from the R side, also enable
+full two-way communication between the languages.  This provides
+considerable flexibility so that for example, a chart built with dc.js
+or leaflet.js can call back to analysis functions in R without having
+to define an additional protocol between the processes. From the
+Javascript side, an RPC call into the R process is just another HTTP
+request. From the R process, a call into Javascript looks is just
+another function call.

### 2015-03-31 casestudies.tex
some work

--- 2015-03-30 casestudies.tex
@@
 
-In this section, we present a simple, but real-world example
-application currently deployed using RCloud.
+As an example, we present a simple real-world application
+deployed under RCloud.
 
-This example application is an implementation of LDAVis, a
-recently-published method for visualizing large collections of
-text. It was specifically designed to enable non-expert users to
-explore collections of short text documents using topic modeling and
+The application is an implementation of LDAVis,
+a recent method for visualizing large text corpora.
+It was designed to help non-experts to explore
+collections of short text documents using topic modeling and
 visualization. Topic modeling~\cite{Blei:2003:LDA} is a standard
-technique for text summarization which, although powerful, requires a
-certain amount of manual intervention and
-interpretation~\cite{Sievert:2014:LAM}.
+technique for text summarization which, although powerful, requires
+some manual intervention and interpretation~\cite{Sievert:2014:LAM}.
 
@@
 
-LDAVis itself was developed by two members of the technical staff at
-AT\&T Labs, and originally targeted RStudio's
+LDAVis was developed by two technical staff members at
+AT\&T Labs, and originally was written in RStudio's
 Shiny~\cite{RStudio:2013:SWA}, a framework for developing R
-applications for the web.  While Shiny provides unparalleled ease of
-development, discoverability and deployment turn out to be equally
+applications for the web.  While Shiny provides outstanding ease of
+development, it turns out that discoverability and deployment are equally
 important aspects in the lifecycle of an internal application (see
-Section~\ref{sec:interviews}).  On these aspects, RCloud provides a
-very simple model: \emph{all developed notebooks are automatically
-  deployed}.
+Section~\ref{sec:interviews}). In these aspects, RCloud provides a
+simple solution: \emph{all developed notebooks are automatically deployed}.
 
-LDAVis combines textual analysis, dimensionality reduction, and
-interactive visualization. The text analysis is performed via a
-combination of an existing R library that fits an LDA model using
-Gibbs sampling~\cite{} and some R code written specifically by the
+LDAVis combines text analysis, dimensionality reduction and
+interactive visualization. Text analysis is performed by
+combining a standard R library to fit LDA models using
+Gibbs sampling~\cite{} with custom R code written by the
 developers. The analysis module is, ultimately, a single R function
-that is exposed to the web application via the Javascript-R RPC
-mechanism described in Section~\ref{sec:system}; this way, the
-analysis is executed remotely on the RCloud servers.
+exposed to the web application via the Javascript-R RPC
+mechanism described in Section~\ref{sec:system};
+thus analysis is performed remotely on an RCloud server.
 
-Each textual topic is a probability distribution over the all the
-words of the document. In order to expose patterns in the
-relationships between the topics, LDAVis employs a combination of
-interactive visualization and dimensionality reduction (letting users
-choose different measures for the topic distances and different
-dimensionality reduction techniques). The dimensionality reduction
-algorithms and distance measures are again implemented in R, which
-means they are executed on the RCloud servers as well.
+Each text topic is a probability distribution over the all the
+words of the document. To expose patterns in the relationships
+between topics, LDAVis employs a combination of
+interactive visualization and dimensionality reduction,
+allowing users to adjust measures for topic distances and the choice
+of dimensionality reduction technique. The dimensionality reduction
+algorithms and distance measures are implemented in R, which
+means they can be executed on the RCloud servers as well.
 
-The result of the dimensionality reduction is a two-dimensional plot
-of the topic space, which can be seen in
-Figure~\ref{fig:textvis}. Describe d3 view.
+The result of the dimensionality reduction process is a
+two-dimensional plot of the topic space, an example of which is shown in
+Figure~\ref{fig:textvis}. The interactive view is implemented in SVG
+and Javascript through D3~\cite{Bostock:2011:DDD}. One of the most
+popular web-based visualization libraries for R is ggvis~\cite{ggvis},
+so it is natural to ask if ggvis could have been used instead of
+custom Javascript. In this case, the interactive features of ggvis
+(and, by extension, Shiny) are a subset of those in Vega~\cite{vega}.
+Custom interactions in LDAVis (such as hovering over a topic,
+topic cluster, or word) do not appear to be available yet in Vega~\cite{vega},
+although the required components for reactive interaction were
+recently described by Satyanarayan et
+al.~\cite{Satyanarayan:2014:DID}.
+Although custom Javascript was required, the flexibility is welcomed
+by many web developers.
 
-This application is simple, but highlights some of the unique features
+The LDAVis application is simple, but highlights some unique features
 of RCloud. While RCloud notebooks allow deployment of R analyses over
 the web with no additional effort, RCloud \emph{applications} are more
-powerful, and are developed in a combination of Javascript and HTML
+powerful, and are developed with a combination of Javascript and HTML
 for the front end. This requires additional knowledge over Shiny, but
-we argue that the RCloud model makes the analysis side simpler for the
-analysts (since they simply write the R code in the style they are
-used to), and the front-end visualization side simpler for the
-front-end developers (since they simply write the Javascript code in
+we argue that the RCloud model makes the analysis side simpler 
+(since analysts simply write R in the style they are used to),
+and the front end visualization side is simpler for the web
+developers (since they simply write the Javascript code in
 the style \emph{they} are used to). In addition, RCloud applications
 inherit the automatic deployment and discoverability features of
-regular RCloud notebooks.
+all RCloud notebooks.
 

### 2015-03-31 system_old.tex

\section{System\label{sec:system}}

\subsection{High-level Architecture\label{sec:highlevelarchitecture}}

% Stephen: Let's work from front to back, from human interface to the
% services and frameworks that support it.

The internal computing infrastructure of organizations has changed
radically in the last fifteen years. The shift from large
servers toward scalable, lower-cost, distributed systems (``the cloud'')
led to a software ecosystem of processes distributed over a
network, usually communicating via HTTP. HTTP is dominant because
web browsers and servers are ubiquitous and available in almost
all hardware devices, from tiny sensors, to handheld devices and
laptops, to rack-mounted servers.

As a result, HTTP is the lingua franca of interprocess communication
(IPC). One of the design goals for RCloud was to provide a productive
environment for creators of data-analysis scripts, that also behaves
as a first-class citizen in the pre-existing ecosystem of computer
services and networks in an organization.

%% Design for cloud-friendly? This goes back to the point in the
%% introduction about playing nice with the rest of the ecosystem.

\subsection{Human Interface\label{sec:humaninterface}}

%\stephen{What??}
%We designed RCloud around the front-facing API, which provides roughly
%one entry point to correspond with each requirement.

Figure~\ref{fig:notebook} shows the RCloud developer interface.  The center
panel is a notebook for code and visualizations.  Code is edited in this panel,
and visualizations are rendered.  An inventory of supplementary ``asset'' files,
that are part of the notebook, though not in its executable flow, are edited on
the right.  Controls at the top of the screen allow the user to run, view, and
share the notebook. \gordon{Why are we talking about the callable interface
here?} The result of code execution is the final cell of the notebook.  On the
left is a browsing area for searching and viewing other users' notebooks, and a
help system.

The call mechanism is the URL itself; arguments are read from the URL
and the final cell is rendered as the result.

\subsection{Notebooks\label{sec:notebooks}}

The main unit of computation in RCloud is a \emph{notebook}.
A notebook holds a sequence of \emph{cells}, each of which contains a
snippet of code or hypertext in Markdown. This is not a novel idea;
executable documents structured this way are a feature of many
other systems, including Mathematica, IPython and Sage.
\stephen{Code is executed when...?}

One of the main contributions of RCloud is the idea that notebooks
are ``always deployed''. In other words, the most recent version of
a notebook is immediately available.\stephen{What if it is broken?}
This makes it convenient to share and modify experiments and compose
results without binding to a specific version of a notebook.
On the other hand, there are situations where it might be important
to name specific versions. We do not expect designers to decide which
versions need to be preserved, but we embrace \emph{transparent} versioning.
This is similar to models like Jankun-Kelly et al.'s p-set calculus \cite{Jankun-Kelly:2007:MFV}
and VisTrails's version tree \cite{Callahan:2006:VVM}, where every change in the state of the system is tracked.

To implement this, we built RCloud on top of GitHub's \emph{gists}~\cite{GitHub:2014:GG}.
GitHub offers a HTTP interface for creating simplified git repositories, the main limitation
being a restriction to text-only files in a single directory. The GitHub web-service
API provides most of the semantics we need for the versioning portion of the storage back end:
access to previous versions, comments, starring, and forking.

Using GitHub for storage and versioning also exposes other capabilities
that can be invoked with JSON and HTTP.
Particularly, we can provide full text search using Apache SOLR.
SOLR is scalable, can index in near-realtime, supports multiple character sets,
indexes several common types of documents, and has schemas and faceted search.
%
Integrating existing services and software components, especially open source,
instead of implementing custom code is a trend in the software industry, and
is very positive for systems research and prototyping in visual analytics.
By adopting existing technology, small teams with limited resoures can explore
novel ideas toward the improvement of large, complex software ecosystems.
In our case, even though this approach involved learning standards not directly
related to visual analytics, our early decision to use the GitHub Gist API
turned out to be successful. Future projects could gain many of the same benefits
by adopting this strategy. In fact, it will be essential for technology adoption
and transfer, because it is almost impossible for any one tool or platform to
``own'' or support the entire visual analytics process.
\stephen{Are we preaching too much?}

\subsection{Reputation and Interest: starring\label{sec:starring}}

Information retrieval based on collecting usage and recommendations
is a cornerstone of modern web services. We would like to help data
scientists to find workbooks (and therefore items in their contents
such as code, data, and colleagues with specific expertise) with
the benefit of such information.

In RCloud, reputation and interest are a relationship between
\emph{notebooks} and \emph{users}, rather than a relationship between
user pairs. We chose this approach because we expect initial 
RCloud deployments to have relatively few users, but some users to
create many notebooks. Under that assumption, assigning interest
to users would not provide sufficiently ``high-resolution'' data.

We incorporate both explicit and implicit indications of interest
in notebooks. Explicit interest is indicated by ``starring,'' or
clicking on a button that marks a notebook as interesting.
This makes explicit indication of interest a nearly trivial operation,
always readily available, to encourage its use.

Implicit signaling of interest is supported by keeping click-through counts
\cite{Joachims:2005:AIC} and execution counts. (In addition to these
standard techniques of collecting feedback from web search, we anticipate
applying static and dynamic code analysis to infer fine-grained
information about relationships, for example, which packages and data
sets often appear together.)

\subsection{Executing R in a web browser\label{sec:Rinbrowser}}

One our main goals is to provide ubiquitous access to the statistical
tools of the R programming language in the wider environment in which data
science teams participate.
%
Given the recent developments of interactive visualization and visual
analytics in HTML5 and the web, we decided to create an
\emph{interconnect} between the two environments.
%
Every RCloud session spawns a new R execution process in a remote
server.

%% \subsubsection{Object capabilities: R to Javascript RPC and access
%%   control}

HTTP and web services are convenient but much of the protocol is
stateless. (For example, GET requests are required to not change the
remote state, and results can be cached in transparent proxies along
the network). In the case of close communication between a running
R process and a web browser, we need something that is more flexible
than distinct URLs for every R process, and which embraces
statefulness of both calculations and visualization parameters.

\carlos{In-depth explanation of R<->web interconnect, and why it's
  necessary or important. Main point: make programming in the R side
  as close as possible to what R programming feels like, and make
  programming in the Javascript side as close as possible to what
  Javascript programming feels like. We do this because this part of
  the system is targeted at programmers; Shiny targets web programming
  at non-web-programmers. We don't attempt such a thing and land at a
  different point in the spectrum.}

\carlos{compare against https://www.opencpu.org, compare against
  IPython notebooks. Two-way communication? Actually describe what it
  is. IPython manipulate notebooks? Understand and contrast}

\subsection{Interactive notebooks\label{sec:interactivenotebooks}}

%% How do we do things that are not trivial to do with IPython (for
%% example)

%% What is relationship to a standard CMS with R/Python

%% dcplot. two-way communication between between backend session and
%% frontend session.

A key engineering decision in RCloud was to rely on full two-way
communication between the web client human interface, and data and
computing resources in the cloud. It is easier to engineer a design
in which the human interface is pushed one-way from the server
and thereafter all interaction takes place within the client.
Two-way communication is necessary, though, where the size of
the data or computational performance makes it impractical to
move all computation to the client.

In addition, R is well suited as a language for analysis, and
JavaScript for interactive visualization. To draw on
the strengths of both languages and both environments, the connection
between the languages is not just procedural: it makes \emph{closures}
and \emph{first-class functions} available across the network.
This provides considerable flexibility, so that for example, a chart built
with dc.js or leaflet.js can call back to analysis functions in R
without having to formalize the protocol between the processes.

\subsection{Deployment of notebooks\label{sec:deployment}}

Every notebook in RCloud is named by a URL, and notebooks by
default are visible by the entire organization. This is deliberate.
As pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS},
broad access to analysis outputs (in their case, for NameVoyager)
increases long-term engagement in part through cross-references on
the web. Although our prototype RCloud deployment is only visible
inside a corporate intranet, we nevertheless found anecdotal support
for this notion by discovering links to RCloud notebooks in internal
discussion fora and mailing lists.

Because notebooks are continually published, any work is immediately
available both as a subroutine and a visual component. Close colleagues
can start on the next stage of analysis, or delve into the data,
even while the original author is polishing an algorithm or its
presentation. \stephen{We glossed over the things that can go wrong.
We should at least discuss that briefly.}
The code is the page, and can either be shaped into
a function of inputs and outputs, or the linear cells of the notebook
can be reworked into a full-fledged HTML layout.

### 2015-03-31 discussion.tex

@@
 considerations, and indicated areas for further exploration. We adopt
-their taxonomy for the following discussion.
+their taxonomy in the following discussion.
 
@@
 observed that hackers readily share work through RCloud. They use it to
-demonstrate techniques, share results with peers and managers, and transfer algorithms
-to other groups.  Artifact histories can be accessed through the notebook
+demonstrate techniques, share results with peers and managers, and to
+transfer algorithms to other groups. Artifact histories can be accessed through the notebook
 tree or through GitHub's web interface.
@@
 searchers from erroneous notebooks), and the notebook tree starts to get
-unwieldly with hundreds of users with hundreds of notebooks. In addition,
+unwieldy with hundreds of users with hundreds of notebooks. In addition,
 our choice to retain notebooks even when they have been deleted has
@@
 The system also needs to help users curate their notebooks. If a notebook
-has been deleted or superceded by one of its forks, a user who opens it
+has been deleted or superseded by one of its forks, a user who opens it
 should be warned. And operations on multiple notebooks and folders of

### 2015-03-31 relatedwork.tex

@@
 solve problems using data. In this section, we focus specifically on
-system work relating to problem-solving enviroments and multiple users.
+system work relating to problem-solving environments and multiple users.
 
@@
 we find much frustration among practitioners that this knowledge
-is difficult to find and often isnot recorded or available in a
+is difficult to find and often is not recorded or available in a
 reusable form even within the same organization.

### 2015-03-31 discussion.tex
working

@@
 and to apply algorithms and metadata to organize them.
-\stephen{I think we need to defend that a bit more.}
 
@@
 lists, which tend to be cumbersome.
-Information security in RCloud should be improved, but more
-sophisticated approaches (such as information flow analysis,
-and query languages that ensure differential privacy) are
-active research topics, and many difficult problems remain
+Information security in RCloud should be improved. More
+sophisticated approaches, such as information flow analysis,
+and query languages that ensure differential privacy, are
+active research topics, and many problems remain
 \cite{Moore:2011:SAF}. We are hopeful that having a richer
@@
 new approaches to information security in visual analytics.
-\gordon{This sounds like we think we are helping by pushing the
-system to the breaking point, is that the idea?}
+%\gordon{This sounds like we think we are helping by pushing the
+%system to the breaking point, is that the idea?}
 
@@
 The interface is also perceived to be slower, both for computer and for
-human. First, there is a slower reaction to commands being run, because
-RCloud necessarily saves a cell to its repository before executing it. It
+human. First, there is higher latency when commands run, because
+RCloud saves a cell to a repository before executing it. It
 may be possible to safely write to the repository behind execution of the
 same code. Second, and perhaps more detrimental, the interface requires
-mouse interaction and switching between the keyboard and mouse, which are
-known to be slower. \gordon{we don't need to back that up, right? i hate
-  using passive voice.} Shortcut keys can be devised if people are willing
-to learn them, but it would be better to reduce the number of such
-operations needed.
+switching between the keyboard and mouse, which can be cumbersome.
+%\gordon{we don't need to back that up, right? i hate using passive voice.}
+Shortcut keys can defined as a workaround, but it would be better
+to reduce the need for them.
 
-Other complaints mostly focus on layout and can probably be addressed with
-some iteration and customization features.
+Other complaints mostly focus on layout, and can probably be addressed by
+offering better customization, and designs such as tabbed views.
 
 {\bf Discovery is hard.}
-While we have provided tools for searching and curating notebooks, the
+While we provided tools for searching and curating notebooks, the
 search function does not help for finding methodology (and does not protect
 searchers from erroneous notebooks), and the notebook tree starts to get
-unwieldy with hundreds of users with hundreds of notebooks. In addition,
-our choice to retain notebooks even when they have been deleted has
-exacerbated the situation, since obsolete and broken notebooks stay around
+unwieldy with hundreds of users and many hundreds of notebooks. In addition,
+our choice to retain notebooks even when they have been deleted 
+exacerbates the situation, since obsolete and broken notebooks stay around
 forever.
 
-One thing to try is tagging and filtering the tree on tags, so that users
+One thing to try is tagging and filtering on tags, so that users
 can mark their own notebooks or those of others to make them easier to
-find. However, this relies on users making a conscious effort, and it may be
-better to rely more on passive metrics such as how often a notebook has
-been run.
+find. However, this relies on users making a conscious effort, which
+is not always reliable. It seems better if possible to lean more on
+passive metrics, such as how often a notebook has been run.
 
-The system also needs to help users curate their notebooks. If a notebook
+The system also needs to help users to curate notebooks. If a notebook
 has been deleted or superseded by one of its forks, a user who opens it
@@
 in teams is a lot more complicated. The forking feature is popular but it
-poses a problem, too, as dozens of versions of a notebook start proliferating.
+poses a problem, too, as dozens of versions of a notebook proliferate.
 
@@
 
-More general solutions would be serious engineering endeavors. Even
-GitHub's generous web interface does not support merging where there are
-conflicts. Actual sharing of notebooks would probably require moving toward
-git's notion of decentralized code repositories, because GitHub gists do
-not support shared ownership.
+A more general solutions would be a serious engineering endeavor. Even
+GitHub's admirable web interface does not support merging where there are
+conflicts. Unrestricted sharing of notebooks would probably require moving
+toward git's notion of decentralized code repositories, because GitHub gists
+do not support shared ownership.

### 2015-03-31 interviews.tex

@@
 found by navigating the notebook tree, or by searching. However, users most
-often mentioned sharing by sending links through email. Clint says,
+often mentioned sharing by sending links through email. Lilo says,
 ``If my supervisor wants to see what I've done or QA it, I can just send her a link.''
@@
 Besides providing a way to present work, notebook sharing can provide a
-starting point for coding. Clint says, ``The best part is how easily you can
+starting point for coding. Lilo says, ``The best part is how easily you can
 share code. You can find a working example, rather than wearing
-out Google and finding questionable examples that may or may not work.'' Raif
+out Google and finding questionable examples that may or may not work.'' Wendy
 notes, ``[If] some person has done something similar, then you're able to just
@@
 
-Taylor develops packages for analysts, and uses RCloud ``more for sharing code with
+Evelyn develops packages for analysts, and uses RCloud ``more for sharing code with
 other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
@@
 Some users, who tried RCloud and were not able to continue for organizational
-reasons, miss certain capabilities. Shelly ``[likes] the concept of being
+reasons, miss certain capabilities. Leith ``[likes] the concept of being
 able to create notebooks and share them... A wiki is not the best way for
@@
 ``I can't share all of the code because it would just get crowded and wouldn't
-look right on a wiki.'' Bart explained, ``If I could make a folder on RCloud and
+look right on a wiki.'' Iris explains, ``If I could make a folder on RCloud and
 have Python notebooks and also Pig notebooks there, and execute them from RCloud,
@@
 be a popular feature. In fact, almost twice as many (131) users have forked
-someone else's notebook as have starred one (75).  Clint says, ``It's one of the
+someone else's notebook as have starred one (75).  Lilo says, ``It's one of the
 handiest functions, because instead of having to find it, copy, paste it, you
@@
 initially didn't anticipate support forking one's own notebooks, which
-proved very useful. Rick says, ``I fork my own notebooks because I'm going off
+proved very useful. Kenyon says, ``I fork my own notebooks because I'm going off
 and doing some other analogous project, so I've got interesting content that I've
@@
 Forking also provides a way for others to troubleshoot when something goes
-wrong.  When Horace works with the users of his notebooks, ``I'll teach people
+wrong.  When Hugh works with the users of his notebooks, ``I'll teach people
 to intercept the result in the middle, to insert print statements here and
@@
 A common but somewhat problematic use of forking is to change
-parameters. Raif: ``I've been forking other people's notebooks [because] I
+parameters. Wendy says, ``I've been forking other people's notebooks [because] I
 want to run them on a different part of data, or I want to change some parts,
 I don't want to see this column, I want a different column, things like
-that.''  Taylor complains that a notebook might say ```This is a report of
+that.''  Evelyn complains that a notebook might say ```This is a report of
 the volume of all of our feeds for this month', and someone would want to
@@
 elements to do the same thing requires expertise. Tool builders see a
-need for facilities to make it easier. Jyothi calls it ``web-enabling'' the
+need for facilities to make it easier. Allison calls it ``web-enabling'' the
 notebook: ``[users] can always fork the notebooks and make changes, but I
@@
 \subsection{Automatic source control}
-%% Rick: ``when I've finished something, there's a nice clean record of it.''
+%% Kenyon: ``when I've finished something, there's a nice clean record of it.''
 %% (I think this may be more about the publish/markdown than about source control.)
 
-Automatic source control is also a popular feature. Clint says, ``Instead of
+Automatic source control is also a popular feature. Lilo says, ``Instead of
 looking back and saying I've got a billion files here in this subdirectory and I
@@
 On the other hand, saving every change leads to many fine-grained
-versions. Rick says that for this reason, the history feature is not that
+versions. Kenyon says that for this reason, the history feature is not that
 helpful: ``I don't need something that keeps track of every mistake I've made or
@@
 \subsection{Discovering others' work}
-Many users find the search function valuable. Raif says, ``The fact that we have
+Many users find the search function valuable. Wendy says, ``The fact that we have
 all the notebooks there, searchable, saves me from replicating what other people
@@
 But other users prefer to browse just the notebooks of experts they
-know. Rick says, ``Usually I know somebody's notebooks that I want to search
+know. Kenyon says, ``Usually I know somebody's notebooks that I want to search
 through, because I know that the kind of thing I'm looking for is something
@@
 Integrating an analysis language into a web development environment is something
-tool developers really appreciate.  The structure of Horace's visualization
+tool developers really appreciate.  The structure of Hugh's visualization
 notebook means
@@
 
-Integrating R also helps Jyothi's topic modeling tool, described in the
-last section: ``Having an open session where I can run R commands or
+Integrating R also helps Allison's application: ``Having an open session where I can run R commands or
 functions without having to invoke an API or send a request and then
@@
 
-This is mainly out of familiarity. Taylor asks, ``Why would I want to use
+This is mainly out of familiarity. Evelyn asks, ``Why would I want to use
 RCloud over my current setup? If it's just me, I like my text editor and
@@
 
-The web interface of RCloud isn't satisfactory to Rick: ``It's nice to have
+The web interface of RCloud isn't satisfactory to Kenyon: ``It's nice to have
 it saved, but there's this trade-off between it making it easier for me to
@@
 
-Kenny also works with text files and command line, rearranging a file so the
+Coby also works with text files and command line, rearranging a file so the
 good code is at the top of the file. When he's done, all the right code is
-at the top. RCloud does not readily support this workflow, but Kenny
+at the top. RCloud does not readily support this workflow, but Coby
 often shares work by pasting it into an RCloud notebook when he's done.
@@
 
-The interface takes some learning. Clint says,
+The interface takes some learning. Lilo says,
 \begin{quote}
@@
 Simply the difference between a cell and the command prompt can confuse some people.
-Rick explains:
+Kenyon explains:
 \begin{quote}
@@
 purpose, so having RCloud persist commands in cells can be annoying.
-Kenny notes, ``It saves everything I do like everything is gold, but most
-of it is junk not meant to be saved.'' Rick says, ``I just want to type in a
+Coby notes, ``It saves everything I do like everything is gold, but most
+of it is junk not meant to be saved.'' Kenyon says, ``I just want to type in a
 couple of quick commands and get some results that are going to tell
@@
 Material that is not appropriate to save includes ``expressions that allow
-me to check that I'm the right track'' (Rick), ``checking out what your data is,
+me to check that I'm the right track'' (Kenyon), ``checking out what your data is,
 or you make a plot of the data. Things that should not really become part of a
-notebook, but things that help you understand your data better'' (Raif).
+notebook, but things that help you understand your data better'' (Wendy).
 
 Users felt that cells do not capture the right level of granularity: they
-either hold too much code or too little. Rick says that RCloud's cell structure
+either hold too much code or too little. Kenyon says that RCloud's cell structure
 ``tempts me to type a big long thing and then run the whole thing, as opposed to
 typing a few little pieces and then put them together'' as he would on the command line.
-When Taylor uses the R command line, he ``copies and pastes 5-10 lines of code,
+When Evelyn uses the R command line, he ``copies and pastes 5-10 lines of code,
 so when something breaks, I get an error message on that one line, and I can
@@
 Some users would like a way to keep results separate from code.
-Ivan thinks RStudio's layout is more helpful because charts
-are shown in another pane that stays in place while doing analysis. Raif says
+Gerrard thinks RStudio's layout is more helpful because charts
+are shown in another pane that stays in place while doing analysis. Wendy says
 ``Cells are really useful, [but] you want to see your output in a different
@@
 The cell structure also can be problematic if some cells take a long time to
-execute. In Kenny's work, there is often a ``long tail''.  The first cell may
+execute. In Coby's work, there is often a ``long tail''.  The first cell may
 take just seconds to run, the next cell minutes, and the last cell 5 hours.
 In this situation, the Run Whole Notebook button is dangerous.
-Kenny reports that when converting his work to notebooks, he
+Coby reports that when converting his work to notebooks, he
 ends up with a lot of comments saying ``This cell takes a long time to run.''
@@
 To avert this pitfall, some users write code within cells to explicitly cache results.
-Kenny ends up ``littering the notebook with little switches that comment out the
+Coby ends up ``littering the notebook with little switches that comment out the
 slow parts'' and either load or save the object to disk. One of the authors of this paper
@@
 
-For this reason, Rick doesn't find the notebook tree satisfactory:
+For this reason, Kenyon doesn't find the notebook tree satisfactory:
 \begin{quote}
@@
 Although RCloud promises an environment where notebooks should keep working,
-not all our users have learned the habits that make this a reality. As Byers
+not all our users have learned the habits that make this a reality. As Joy
 puts it, there is still a big problem with ``bitrot'' -- notebooks often
 stop working because the user changed the structure of their data, or
-changed a filename or a database.  He says we need ``organizational
+changed a filename or a database.  She says we need ``organizational
 protocols'' to catch up with the technology. Even if one forks someone
@@
 
-Taylor, who writes packages and example notebooks for them, mentioned the
-accumulation of dead notebooks. When Taylor and Horace work together on a notebook:
+Evelyn, who writes packages and example notebooks for them, mentioned the
+accumulation of dead notebooks. When Evelyn and Hugh work together on a notebook:
 \begin{quote}
@@
 The problem of dead notebooks is compounded by changing
-packages and example notebooks at the same time. Taylor will ``go back and
+packages and example notebooks at the same time. Evelyn will ``go back and
 change it and update the page, but then what happens is people have in the
@@
 
-Taylor tried to keep his notebook tree well organized, but this didn't help,
+Evelyn tried to keep his notebook tree well organized, but this didn't help,
 because some people kept old links in their email, or forked notebooks which 
 had become obsolete. By design, notebooks that are deleted in the user interface
-are not purged from the repository. This caused some subjects to express
-caution about sharing notebooks too freely.
-%``Now I'm really gun-shy about sharing notebooks
-%because it's like, do I want to support this forever?''
+are not purged from the repository. Now he says he is scared to share notebooks:
+``Do I want to support this forever?''
 
-%% Bart: ``One UI notebook and a bunch of other notebooks that do various tasks and
+%% Iris: ``One UI notebook and a bunch of other notebooks that do various tasks and
 %% are called with API calls. The problem is, if you clone all the notebooks, then

### 2015-03-31 paper.tex
intro and related work
update case study and bib
fixed fig 1 caption

@@
 
-\title{Social Visual Analytics: DevOps for Data Scientists}
+\title{Collaborative Visual Analysis on the Web with RCloud}
+% \title{Integrating Social Visual Analysis and Deployment with RCloud}
+% \title{Social Visual Analytics: DevOps for Data Scientists}
 
@@
     An overview of features in RCloud.
-RCloud was built to support an environment in which a large number of loosely-related problems in data and visual analytics are solved by a small number of \emph{hackers} and \emph{scripters}.
-In such a shared environment, \emph{discoverability} is also a problem.
-RCloud supports search, annotation, recommendation, and commenting for all notebooks, enabling an overview mode where users can \emph{browse} popular and recent analyses.
-When problems and data sources change constantly, deployment becomes a time sink; RCloud supports \emph{transparent and automatic} deployment of analyses as web pages and web services, allowing a seamless transition from exploratory work to production code.
+RCloud supports an environment in which a large number of loosely-related problems in data and visual analytics are solved by a small number of \emph{hackers} and \emph{scripters}.
+In such a shared environment, \emph{discoverability} is a concern.
+RCloud supports search, annotation, recommendation, and commenting for all notebooks, and provides an overview where users can \emph{browse} popular and recent analyses.
+When problems and data sources change frequently, deployment can become very costly; RCloud supports \emph{transparent and automatic} deployment of analyses as web pages and web services, allowing a seamless transition from data exploration work to production web services.
   }

### 2015-03-31 paper.tex
fix abstract a little

@@
   Second, these environments often assume a single-developer perspective,
-  while data scientist teams would get much benefit from easier sharing
+  while data scientist teams could get much benefit from easier sharing
   of scripts and data feeds, experiments, annotations, and automated
@@
   collaborative data analysis, visualization and web deployment.
-  The biggest deployment of RCloud has been in active use for more than a year,
+  The biggest deployment of RCloud has been in active use for more than two years,
   and has about fifty active users. We report on interviews with some of these users, and

### 2015-03-31 introduction.tex
working
copy editing

--- 2015-03-31 introduction.tex
@@
 Knowledge is shared through informal conversations, emails,
-meetings, phone calls, source code itself, in wikis, in project documents and
+meetings, phone calls, source code, in wikis, project documents and
 by other means. Results of experiments are often shared first by copying static
-output (usually a simple screen shot or image output). By the time
-decisions are made based on the results of that screenshot, the data
-and the processes might have changed enough to bring to question the
-original decision.
+output, usually a screen shot or image output. By the time
+decisions are made based on the results of that screenshot, data
+and processes may have changed so much that the decision can be wrong.
 
@@
 
-This situation can affect how data science teams; as analysts and
-toolbuilders, we have experienced this first-hand in our environment, and
+This situation can affect how data science teams work; as analysts and
+tool builders, we have experienced this first-hand in our environment, and
 others report similar findings. Gutierrez collected interviews with data
@@
 \item (upon being given access to other code and data analysis, in
-order to learn about Hadoop) ``[...] Not only that, I could also look at other people’s code and
-play with their code and data sets as well''
+order to learn about Hadoop) ``I could look at other people’s code
+and play with their code and data sets as well''
 \item (discussing the value of sharing prototypes rather than static
-data) ``[...] Finally, prototyping our products so that internal
+data) ``Prototyping our products so that internal
 customers can use them early on has been crucial for
-our success. [...] Now we can shoot off a URL to internal
-customers and it allows them to [...] provide feedback way before
+our success. Now we can shoot off a URL to internal
+customers and it allows them to provide feedback way before
 we're talking about getting it into production.''
 \item (on sharing more than just a finished product) ``We also share
-exploratory analyses and reports [...] so that we can still exchange
+exploratory analyses and reports so that we can still exchange
 knowledge even if the work didn't make it into a larger project.''
 \item (on changing analyses and processes) ``It is important to
-have testing frameworks for all of your data, so [...] you can go back and test all
-of your data. 
+have testing frameworks so you can go back and test all
+of your data.''
 \end{itemize}
 
-Currently, these concerns need to be addressed \emph{alongside} the
-visual analytics environment. Some symptoms of the current situation are:
+Currently, these concerns are being addressed \emph{alongside} the
+visual analytics environment. Some symptoms of the situation are:
 \begin{itemize}
@@
 A key point is that, for the most part, the improved capabilities are
-available by default, without much explicit intervention on the part of
+available by default, without much explicit involvement on the part of
 data scientists and other customers. Visual analytics experiments are
@@
 goals or objectives for emerging systems, described in previous work,
-closely match ours. We noted the following central themes.
+closely match ours. We identified the following central themes.
 
@@
 
-\carlos{we need a segue into related work}.
+%\carlos{we need a segue into related work}.
+%\stephen{maybe not?}

### 2015-03-31 relatedwork.tex
copy editing

@@
 
-\section{Related Work}
+\section{Related Work\label{sec:related}}
 
-In some sense or another, the entire field of information
-visualization and visual analytics revolve around improving how users
-solve problems using data. In this section, we focus specifically on
+Broadly, the entire field of information visualization and visual analytics
+revolves around improving how users solve problems involving data.
+In this section, we focus specifically on
 system work relating to problem-solving environments and multiple users.
@@
 behavior of the program is ``weaved'' together with the source code,
-yielding both an executable program and a human-readable document.  A
-notebook represented as a collection of short, executable cells,
+yielding both an executable program and a human-readable document.
+A notebook represented as a collection of short, executable cells,
 originated with Mathematica, and in R, literate programming is
@@
 %
-Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
+Further afield, {Electronic Lab Notebooks} are applications for organizing
 and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
@@
 \paragraph*{Provenance and versioning} As
-we mentioned, one central problem in exploratory analysis is that
+mentioned, one central issue in exploratory analysis is that
 problems change quickly over time, often in the course of developing a
 solution. As a result, systems should provide adequate support for
-tracking \emph{changes} of the data analyses scripts.  VisTrails was
+tracking \emph{changes} of the data analysis scripts. VisTrails was
 one of the original systems for managing \emph{process provenance} ,
@@
 \paragraph*{Web-based tools for sharing code snippets}
-There have been a variety of tools recently developed for quickly
-sharing small programs on the Web, including but not limited to:
+There have been quite a few tools recently developed for quickly
+sharing small programs on the web, including
 bl.ocks~\cite{blocks}, jsfiddle~\cite{jsfiddle}, and
-plot.ly~\cite{plotly}.  bl.ocks and jsfiddle are both designed to
-share Javascript programs, which means that their deployment happens
-automatically through the web browser. If and when Javascript becomes
-the lingua franca of exploratory data analysis, then we can foresee
-building a much simpler version of RCloud where all execution happens
+plot.ly~\cite{plotly}. bl.ocks and jsfiddle are designed to
+share Javascript programs, which means that deployment happens
+automatically through the web browser. If Javascript eventually becomes
+the lingua franca of exploratory data analysis, we can foresee
+building a simpler version of RCloud where all execution happens
 either on the client side or via web services. Unfortunately that is
-not a realistic assumption for the present, making these solutions by
-themselves unsuitable to our use case. Plot.ly is notable in that it
+not a realistic assumption at present, making these stand-alone
+solutions unsuitable for our purpose. Plot.ly is notable in that it
 provides API support for publishing \emph{from} scripts: in other
 words, it is possible to generate a plot.ly visualization from inside
-a separate program. Although this is an intriguing idea, it
-nevertheless creates a disconnect between the analysis and the
-resulting visualization. In RCloud, we wanted to make sure that every
-visualization was transparently linked to the source code that
-generated it.
+another program. Although this is an intriguing idea, it
+nevertheless creates a disconnection between the analysis and the
+resulting visualization. In RCloud, we wanted to ensure that every
+visualization is transparently linked to the source code that
+generates it.
 
@@
 Heer and Agarwala identify many design considerations for
-collaborative visual analytics~\cite{Heer:2008:DCF}.
+collaborative visual analytics~\cite{Heer:2008:DCF} that
+influenced our work.
 RCloud notebooks, and the integrated version control system for them,
@@
 
-There has been noteworthy work on specific techniques such as
-social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
-and crowdsourcing \cite{Fast:2014:ECS} to support collaborative
-or social development or analysis processes.
+There has been noteworthy work on specific techniques
+to support collaborative or social code development and data analysis,
+such as social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
+and crowdsourcing \cite{Fast:2014:ECS}.
 Similarly, there are computational methods to support high
@@
 \cite{Httermann:2012:DD} or ``continuous integration''
-\cite{Fowler:2006:Continuous}) is a trend in web services and related
+\cite{Fowler:2006:Continuous}) is a trend in web services and similar
 fields. By making it convenient for data scientists to expend just a
-little more effort when creating experiments, we may be able to
+little additional effort when creating experiments, we may be able to
 eliminate the need for programming teams to recreate their work to
 deploy it in production, which has a high cost in time, expense and
-accuracy. We now turn to a description of the system architecture
-itself, and how it enables the features we have been highlighting.
+accuracy.
+
+We next give a description of the system architecture,
+and how it enables capabilities that satisfy the requirements as described.

### 2015-03-31 relatedwork.tex
fix stupid inconsistency

@@
 \paragraph*{Social Data exploration and Analysis}
-{\bf ManyEyes}~\cite{Viegas:2007:MAS} was a landmark system designed
+ManyEyes~\cite{Viegas:2007:MAS} was a landmark system designed
 for the crowdsourced creation and publication of data

### 2015-03-31 system.tex
fix stupid inconsistency

@@
 
-One of the starting points in our design goals was to leverage, as
-much as possible, the current infrastructure of the world-wide
-web. HTTP, with all its warts, has become the lingua franca of
+The starting point in our design was to take advantage of the current
+infrastructure of web software as much as possible. HTTP, despite its
+deficiencies, has become the lingua franca of
 distributed interprocess communication. This meant that if our system
-spoke HTTP natively, then it was possible to provide a low-friction
+could speak HTTP natively, then it would be possible to provide a low-friction
 path from development of an interactive data analysis script to an
-automated \emph{web service}, essentially a function call over the
-web, which can then be used in higher-level tools. An example of this
-kind of transition is described in Section~\ref{sec:casestudy}.
+automated \emph{web service}, essentially a remote function call over
+the web, which can be invoked by higher-level tools. An example of
+this sort of progression is described in Section~\ref{sec:casestudy}.
 
-As a result, the entire high-level infrastructure of RCloud, shown in
-Figure~\ref{sec:system} uses web standards. The communication between
-the web browser and the active R session as a user edits a notebook is
-a combination of HTTP and Websockets, while all other IPC happens
-through HTTP, from notebook versioning in GitHub to building and
-maintaining full-text indices through SOLR.
-The most peculiar feature of RCloud's runtime system is its tight integration between the web browser and the backend R process, which we describe in Section~\ref{sec:Rinbrowser}.
+In consequence, the entire high-level infrastructure of RCloud, shown
+in Figure~\ref{sec:system}, uses web standards. The communication between
+a web browser and an active R session as a user edits a notebook is
+performed by a combination of HTTP and Websockets, while all other IPC
+is done through HTTP, from notebook versioning in GitHub to building and
+maintaining full-text indices through SOLR. The most novel aspect
+of RCloud's runtime system is its tight integration between the web client
+and the backend R process, described in Section~\ref{sec:Rinbrowser}.
 
@@
 of code or hypertext in Markdown. As mentioned in
-Section~\ref{sec:related}, this is not a novel idea: executable
-documents structured this way are a feature of many other systems,
+Section~\ref{sec:related}, this is not novel; executable
+documents like this are a feature of many other systems,
 including Mathematica, IPython and Sage.
@@
 read-eval-print loops, or can also be executed all at once, providing
-a similar experience to running a shell script.
+an experience similar to running a shell script.
 
-One of the main contributions of RCloud is the idea that notebooks are
+One of the main contributions of RCloud is the notion that notebooks are
 ``always deployed''. In other words, the most recent version of a
 notebook is immediately available to all other users of the system as
-the notebook is being developed. One way to describe this is that
+the notebook is being developed. Another way to describe this is that
 RCloud lacks a ``save'' button: any notebook cell that runs is always
 associated to a notebook version serialized to disk. Although one of
-the interviewed users reports that this sometimes leads to an
+our interview subjects reported that this sometimes leads to an
 excessively fine-grained sequence of versions (see
@@
 
-When we combine this with the ability to refer to the latest version
-of a notebook by name, we get notebooks that are always live, but
-sometimes broken. When stability is important, we also
-allow any previous version of a notebook to be tagged and referred
-to. These are similar to models like Jankun-Kelly et al.'s p-set calculus
+When we combine immediate deployment with the ability to refer to the
+latest version of a notebook by name, we get notebooks that are always live,
+but sometimes broken. Because stability is important, we also
+allow any previous version of a notebook to be tagged and referenced.
+Our notebook scheme is similar in many ways to models like
+Jankun-Kelly et al.'s p-set calculus
 \cite{Jankun-Kelly:2007:MFV} and VisTrails's version tree
@@
 
-Our implementation of the versioning mechanism is built on
+RCloud's implementation of the versioning mechanism is built on
 top of GitHub's \emph{gists}~\cite{GitHub:2014:GG}, which are an HTTP
@@
 the storage back end: access to previous versions, comments, starring,
-and forking. This provides the additional benefeit that every RCloud
+and forking. This provides the additional benefit that every RCloud
 notebook can also be manipulated like a git repository. Advanced users
@@
 that would otherwise be lost. In the case of social data visualization
-platforms, we would like to leverage this usage data to help analysts
-find content of interest, whether this content is actual source code
+platforms, we would like to exploit usage data to help analysts
+find content of interest, whether the content is actual source code
 or accompanying data. The standard way to achieve this is through
@@
 Automatic recommendations have become famous in the user experience
-provided by companies such as Amazon and Netflix (``if you've liked
-this film, then you'll like this other film''). In order to build such
+provided by companies such as Amazon and Netflix (``If you liked
+that film, then you should like this one too''). To compute such
 recommendations, we need users to \emph{curate} the collection, by
@@
 This makes explicit indication of interest a nearly trivial operation,
-always readily available, to encourage its use.
+always available, to encourage its use.
 %
-The current version of RCloud uses only simple aggregate counts of
-stars to indicate overall notebook interest, mostly because the
-recommendation system approaches popularized by the Netflix prize
-require large recommendation sets to paying off. Nevertheless, the
-starring mechanism is sufficient to create personalized
-recommendations~\cite{Hu:2008:CFF}.
+RCloud today uses only simple counts of stars to measure overall
+notebook interest, mostly because standard
+recommender system algorithms 
+require reasonably large training sets to pay off.
+Nevertheless, the starring mechanism is sufficient
+to create personalized recommendations~\cite{Hu:2008:CFF}.
 
@@
 Every notebook in RCloud is named by a URL, and notebooks by default
-are visible by the entire organization. This is deliberate.  As
-pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS}, broad
-access to analysis outputs (in their case, for NameVoyager) increases
+are visible in the entire organization. This is deliberate.
+As pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS},
+broad access to analysis outputs (in their case, for NameVoyager) increases
 long-term engagement in part through cross-references on the
@@
 
-As mentioned before, the other main goal in RCloud was to provide, as
-much as possible, full access to the R programming language during the
+As mentioned, the other main goal in RCloud was to provide
+full access to the R statistical programming language during the
 \emph{development} of a data analysis notebook.
 At the same time, when notebooks are \emph{deployed} (and potentially
-accessed by anyone with a web browser), we'd like to allow a web
+accessed by anyone with a web browser), we'd like to allow the
 browser to invoke only a very limited subset of R, namely those
@@
 
-The solution we have developed is simple and general, and is directly
+The solution we developed is simple and general, and was directly
 inspired by Miller's \emph{object
   capabilities}~\cite{Miller:2006:RCT}. Specifically, the R layer that
-communicates with javascript does not expose naked evaluation of
+communicates with Javascript does not expose unprotected evaluation of
 arbitrary functions. Instead, every function that the R layer intends
 to expose is associated with a large, cryptographically-safe random
-number (informally, a ``hash''). This random number is then sent
+number (a ``hash''). This random number is then sent
 across the wire. This number is interpreted as an opaque identifier to
@@
 R libraries to respond to user input in the browser, giving them
-access to features ranging from password prompts to the currently
+access to features ranging from password prompts, to the currently
 selected set of points in an interactive linked brushing
@@
 As a result, the features required to provide safe access to R from
-the client side, and Javascript access from the R side, also enable
+the client, and Javascript access from the R side, also enable
 full two-way communication between the languages.  This provides
-considerable flexibility so that for example, a chart built with dc.js
+considerable flexibility, so that for example, a chart built with dc.js
 or leaflet.js can call back to analysis functions in R without having
@@
 Javascript side, an RPC call into the R process is just another HTTP
-request. From the R process, a call into Javascript looks is just
+request. From the R process, a call into Javascript looks like just
 another function call.

### 2015-03-31 discussion.tex
fix discussion headings

@@
 
-{\bf Shared artifacts and artifact histories} are a central feature of
+\paragraph*{Shared artifacts and artifact histories} are a central feature of
 RCloud, through its shared workbooks of experiments and analyses. We
@@
 
-{\bf Modularity and granularity.} The ability to partition work into
+\paragraph*{Modularity and granularity} The ability to partition work into
 independent units (modularity) is a key to working productively in teams.
@@
 
-{\bf View sharing, bookmarking.} Most resources in RCloud are named
+\paragraph*{View sharing, bookmarking} Most resources in RCloud are named
 and accessed as URLs. This proved to be an effective mechanism for
@@
 
-{\bf Discussion.} Annotation and commenting was another central goal.
+\paragraph*{Discussion} Annotation and commenting was another central goal.
 Commenting is supported through GitHub, but our hackers found it
@@
 
-{\bf Content export} is not a capability we aimed at supporting,
+\paragraph*{Content export} is not a capability we aimed at supporting,
 and R already has many packages for this. Recently, due to popular
@@
 
-{\bf Social-psychological incentives} and {\bf voting and ranking} are
+\paragraph*{Social-psychological incentives} and {\bf voting and ranking} are
 supported through starring and forking. These mechanisms are employed often
@@
 
-{\bf Group management, size and diversity.} This area needs better
+\paragraph*{Group management, size and diversity} This area needs better
 support in RCloud, but is clearly important to working in teams.
@@
 
-{\bf Curation.} Even without formal group management, users may
+\paragraph*{Curation} Even without formal group management, users may
 curate groups of related notebooks using notebook tree folders.
@@
 
-{\bf Versioning.}
+\paragraph*{Versioning}
 Some important aspects of the environment are more difficult to manage
@@
 
-{\bf Caching.}
+\paragraph*{Caching}
 An important consideration is how and where to introduce caching
@@
 
-{\bf Security.}
+\paragraph*{Security}
 It is essential to provide security in an environment for
@@
 
-{\bf Exploration is hard.}
+\paragraph*{Exploration is hard}
 R is a popular language for exploratory data analysis, and we built RCloud
@@
 
-{\bf Discovery is hard.}
+\paragraph*{Discovery is hard}
 While we provided tools for searching and curating notebooks, the
@@
 
-{\bf Collaboration is hard.}
+\paragraph*{Collaboration is hard}
 While forking provides a simple way to continue someone else's work, working

### 2015-06-14 interviews.tex
nix anonymized gender confusion

@@
 
-Evelyn develops packages for analysts, and uses RCloud ``more for sharing code with
+Eric develops packages for analysts, and uses RCloud ``more for sharing code with
 other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
@@
 I don't want to see this column, I want a different column, things like
-that.''  Evelyn complains that a notebook might say ```This is a report of
+that.''  Eric complains that a notebook might say ```This is a report of
 the volume of all of our feeds for this month', and someone would want to
@@
 
-This is mainly out of familiarity. Evelyn asks, ``Why would I want to use
+This is mainly out of familiarity. Eric asks, ``Why would I want to use
 RCloud over my current setup? If it's just me, I like my text editor and
@@
 typing a few little pieces and then put them together'' as he would on the command line.
-When Evelyn uses the R command line, he ``copies and pastes 5-10 lines of code,
+When Eric uses the R command line, he ``copies and pastes 5-10 lines of code,
 so when something breaks, I get an error message on that one line, and I can
@@
 
-Evelyn, who writes packages and example notebooks for them, mentioned the
-accumulation of dead notebooks. When Evelyn and Hugh work together on a notebook:
+Eric, who writes packages and example notebooks for them, mentioned the
+accumulation of dead notebooks. When Eric and Hugh work together on a notebook:
 \begin{quote}
@@
 The problem of dead notebooks is compounded by changing
-packages and example notebooks at the same time. Evelyn will ``go back and
+packages and example notebooks at the same time. Eric will ``go back and
 change it and update the page, but then what happens is people have in the
@@
 
-Evelyn tried to keep his notebook tree well organized, but this didn't help,
+Eric tried to keep his notebook tree well organized, but this didn't help,
 because some people kept old links in their email, or forked notebooks which 

### 2015-06-15 introduction.tex
working

@@
 Powerful, interactive environments for analysis and programming were
-the goal, together with an unflinching (and, at the time, somewhat
-heretical) insistence in keeping humans as a central part of the
-discovery process. His now-famous quip that ``the picture-examining eye is the
-best finder we have of the wholly unanticipated'' has come to define
-much of visual analytics and exploratory visualization~\cite{TukeyEDA}.
+the goal, with unswerving insistence on visualization as a
+central part of the discovery process. His now-famous quip that
+``the picture-examining eye is the best finder we have of the
+wholly unanticipated'' has come to define much of visual analytics
+and exploratory visualization~\cite{TukeyEDA}.
 
-In some way, we have moved far beyond what Tukey imagined:
-computing and networking capabilities today far exceed
-what was barely imaginable then. 
-We argue, on the other hand, that how we \emph{develop} our data analysis
-solutions has not changed as much: the S language
-was developed essentially alongside Unix~\cite{TheSSystem} and,
-although environments such as RStudio include a variety of modern
-features, they still follow the basic metaphor of terminal, text
-editor, and source files stored in file systems.
-We turn our attention to this opportunity to use computation
-to support, not only individuals, but entire teams and their
-work within larger organizations. In this paper, we contribute the
-design of RCloud, together with an interview study conducted over the
-course of its development and deployment at AT\&T Labs, where RCloud has
-been in use for about two years.
+Computing has moved far beyond what Tukey imagined.
+Processing and networking capabilities today far exceed
+what was barely imaginable then. Sophisticated technologies
+for processing, storing, publishing, searching and sharing
+information and analyses are commonplace.
+But how we \emph{develop} analytic solutions has not changed much. 
+Although environments such as RStudio include many features to
+graphically interact with a local workspace, still follow the metaphor
+of terminal, text editor, and source files stored in file systems.
 
-Consider the role of a data science team within
-a business or technical organization today.
+The work of data scientists in teams is also changing.
 Project teams vary in size, from just a few people to dozens or
-more, even within one project's duration. Assignments are often
-broad, and include tasks such as problem identification,
+more.
+Assignments are broad and include tasks such as problem identification,
 data wrangling, modeling, analysis, visualization, summarization,
@@
 actions to help clients to realize the benefits of the analysis.
-Eventually, knowledge or working prototypes created by these teams
-are transferred to other organizations to employ them in production.
-There are many details to these tasks. For example, data wrangling can involve
-finding data, understanding its syntax and semantics, assessing data quality,
-performing normalization and data quality remediation, and making the data
-available to other tasks that will follow.
+Knowledge or software prototypes created by these teams are transferred
+to other organizations to employ them in production.
+This work depends extensively on communication and collaboration.
+At almost every step, collaborators share detailed knowledge about tasks,
+data and code.
+Further, data scientists are increasingly asked to work closely
+with less technically-oriented business or domain specialists.
 
-Visual analytics depends greatly on communication and collaboration.
-At almost every step, detailed knowledge about data, code and tasks
-is shared by collaborators. 
-Further, data scientists are increasingly asked to work more closely
-with business or domain specialists who may be less technically oriented.
-Thus, data scientists and developers are being asked to become
-very broad, and integrate work across the spectrum.
+Previous work has identified important areas for improvement in
+visual analytics systems to address this situation.
 
-%% Interviews in {\it Data Scientists at Work} by Sebastian Gutierrez illuminate
-%% the teamwork and sharing in the current practice of visual analytics in industry.
-%% We particularly appreciated the remarks by Jonathan Lenaghan about the potential
-%% for continuous integration and publication.
+\begin{enumerate}
+\item Discoverability. A chronic complaint of data analysis teams is
+inefficiency due to a lack of transparency in accessing the work of
+other team memnbers.  The processes and technologies supporting
+data analysis and visual analytics today are fragmented.
+Knowledge is shared imperfectly through informal conversations, emails,
+meetings, phone calls, source code, wikis and project documents.
+This causes work to be repeated unnecessarily
+(``How can I connect to database X?'', ``How do you get clean data
+from column Y from feed Z?'') due to lack of communication about
+previously solved problems. 
+Our goal is support and
+promote transparency of work between team members.
 
-%% % \begin{addmargin}[1em]{1em}  % left and right margin change
-%% % part1.txt
-%% Erin Shellman [of Nordstrom] described the benefits of
-%% continuous deployment of experiments, and sharing knowledge
-%% through software and data analysis artifacts.
-%% ``Finally, prototyping our products so that internal
-%% customers can use them early on has been crucial for
-%% our success. It doesn't even have to be something
-%% fancy. For instance, our recommendations preview
-%% tool doesn't have a particularly interesting
-%% visualization, but its enough to show the result of our
-%% algorithms. Now we can shoot off a URL to internal
-%% customers and it allows them to sit at their desk and
-%% understand the behavior of our product and
-%% experiment with it, and provide feedback way before
-%% we're talking about getting it into production. This
-%% has been super helpful and has been a really great
-%% way to get people excited about what we're working
-%% on...
-%% %Building and maintaining those relationships are
-%% %just like anything else-- you always have to be
-%% %working on them. Nurturing those relationships are
-%% %part of the job, and if you leave them unattended,
-%% %they might not be there later.
-%% %
-%% % part2.txt
-%% %Gutierrez: How do you share the knowledge you are
-%% %building with others?
-%% We’re obsessed with Confluence from
-%% Atlassian, and the Data Lab has a very active
-%% Confluence space. Recommendo is fully documented
-%% on Confluence, so anyone in the company could learn
-%% how it’s built, where to find it, and how to use it.
-%% We also share exploratory analyses and reports on
-%% Confluence so that we can still exchange knowledge
-%% even if the work didn't make it into a larger project.''
+Results of experiments are often shared first by copying screen
+shots or static output. By the time decisions are made based on
+the results of that screenshot, data and processes may have changed
+so much that the decision can be wrong.
 
-%% % part3.txt
-%% John Foreman commented on the scope of the work in
-%% data science teams.
-%% ``I lead the data science team at
-%% MailChimp, and I like to get my hands dirty too.
-%% Some of the big pieces of my day involve working
-%% with my team to take stock of current projects and
-%% figure out where to go next, doing my own work and
-%% prototyping things-- I do projects just like my peers
-%% do and then also my talking with other teams,
-%% talking with management, and planning for the
-%% future.
-%% On our team, we've got different folks facing different
-%% % part4.txt
-%% kinds of projects. We've got one person who really
-%% owns compliance and looks at our compliance
-%% processes. We've got another data scientist who
-%% focuses on more of the user experience side of the
-%% house and understanding our customers. I help them
-%% and others as is needed while I do the three things
-%% mentioned earlier-- build data products, be a
-%% translator, and have conversations with the data
-%% science community.''
+\item Technology transfer.
+In most organizations, development of analyses and visualizations 
+s done by \emph{hackers} and \emph{scripters} (adopting Kandel et al.'s
+terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
+gain the benefits of the tools developed by hackers and scripters.
+Results of experiments are often shared first by copying screen
+shots or static output.  By the time decisions are made based on
+the results of that screenshot, data and processes may have changed
+so much that the decision can be wrong.
+\stephen{Transitional sentence here.}
+Generally, the connection between these two worlds is made by IT staff,
+who port or rewrite code so it can run as a stable production service.
+In an environment where business needs can change rapidly, this
+process does not scale. Our goal is to merge the worlds of EDA and
+production.
 
-%% % part5.txt
-%% Jonathan Lenaghan from PlaceIQ discussed to the potential
-%% for continuous deployment in data science.
-%% ``If you look at dev ops right now, they have things
-%% such as continuous integration, continuous build,
-%% automated testing, and test harnesses-- a1l of which
-%% map very well from the dev ops world to the data ops
-%% (a phrase I stole from Red Monk) world very easily. I
-%% think this is a very powerful notion. It is important to
-%% have testing frameworks for all of your data, so that if
-%% you make a code change, you can go back and test all
-%% of your data.''
+\item Coexistence. In current ecosystems, the isolation of exploratory
+visualization and data analysis environments hinders wider adoption of
+modern techniques from each.
+The richness of interactive visualization tools is still somewhat
+separated from the power of statistical programming environments.
+There is an opportunity to provide a framework so that developments
+on each side are more easily adopted by the other, and made available
+in production services.
 
-%% % part6.txt
-%% Anna Smith from Bitly spoke about the benefits of access to
-%% a rich environment of shared data and code.
-%% ``In addition to Bitly's extraordinary data sets,
-%% I was given all the resources that I needed. These were
-%% resources that I’d only heard about but had never
-%% seen. I went from asking what a Hadoop cluster was,
-%% as my school didn't have one, to being able to work
-%% with one. Now I could play with one to figure out how
-%% to make my work better and how to change my
-%% algorithms so they were cleaner and ran faster on
-%% Hadoop. The whole transition was eye opening. Not
-%% only that, I could also look at other people’s code and
-%% play with their code and data sets as well. It was
-%% fantastic and I was convinced that working with data was my thing.''
-%% % \end{addmargin}
+\end{enumerate}
 
-%% % (Hiring ``full stack developers'' to perform integration seems to be a trend, but does not achieve the potential benefits of specialization and division of labor in larger teams, and is not a practice to be encouraged.)
-
-Unfortunately, the processes and technologies supporting data analysis visual analytics are fragmented.
-Knowledge is shared through informal conversations, emails,
-meetings, phone calls, source code, in wikis, project documents and
-by other means. Results of experiments are often shared first by copying static
-output, usually a screen shot or image output. By the time
-decisions are made based on the results of that screenshot, data
-and processes may have changed so much that the decision can be wrong.
-
-%% This disjointed, inconsistent process inhibits the use of tools to bridge
-%% the many gaps and to help people find resources.
-%% For example, even when code is shared, we still not may know where to
-%% find examples that show how to use it.
-%% When we find a useful data source, we don't know what packages or
-%% other data sources usually go along with it, what problems with the data
-%% others have already discovered and corrected, or where to obtain online feeds
-%% to get updates.
-
-This situation can affect how data science teams work; as analysts and
-tool builders, we have experienced this first-hand in our environment, and
-others report similar findings. Gutierrez collected interviews with data
-scientists~\cite{Gutierrez:2014:DSA}, which include the following:
+Fortunately, there is abundant evidance that technology can
+solve many of these problems.
+Gutierrez collected interviews with data scientists~\cite{Gutierrez:2014:DSA},
+which include the following:
 
@@
 
-Currently, these concerns are being addressed \emph{alongside} the
-visual analytics environment. Some symptoms of the situation are:
-\begin{itemize}
-\item It is hard to find data, metadata, and knowledge about them.
-\item Most coordination is done through meetings, whose content is not linked to other artifacts and may not even be stored.
-\item Production deployment requires porting code to a different environment or even completely rewriting it, thwarting continuous release.
 \item Many tools adopt a standalone or single developer perspective,
@@
 we created a software environment that supports the end-to-end
-visual analytics process for individuals and teams.
+visual analytics process for individuals and teams and
+their work within larger organizations.
+
+In this paper, we contribute the
+design of RCloud, with an interview study based on the
+course of its development and deployment at AT\&T Labs for 
+about three years.
+
 This environment knits together some familiar tools, and 
@@
 
-From a more principled or theoretical perspective, certain visual analytics
-goals or objectives for emerging systems, described in previous work,
-closely match ours. We identified the following central themes.
-
-\begin{enumerate}
-
-\item Technology transfer.
-In most organizations, development of analyses and visualizations 
-s done by \emph{hackers} and \emph{scripters} (adopting Kandel et al.'s
-terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
-gain the benefits of the tools developed by hackers and scripters.
-Generally, the connection between these two worlds is made by IT staff,
-who port or even rewrite code so it can run as a stable production service.
-In an environment where business needs can change rapidly, this
-process does not scale. Our objective is to merge the worlds
-of EDA and production.
-
-\item Coexistence. In the current data science ecosystem, the
-isolation of exploratory visualization and data analysis
-environments hinders wider adoption of modern techniques from each.
-The richness of interactive visualization tools is still somewhat
-separated from the power of statistical programming environments.
-There is an opportunity to provide a framework so that developments
-on each side are more easily adopted by the other, and made available
-in production services.
-
-\item Discoverability. A chronic complaint of data analysis teams is
-repeated work (``How can I connect to database X?'', ``How do you get
-clean data from column Y from feed Z?'') This work arises from lack of
-communication about previously solved subproblems. Our goal is encourage
-and support transparency of work between team members.
-
-\end{enumerate}
-
 Over the past three years, we prototyped RCloud and deployed it to
@@
 collaborators in other business units.
-
-%\carlos{we need a segue into related work}.
-%\stephen{maybe not?}

### 2015-06-15 discussion.tex
working

@@
 (``hackers'') gave us some insight into whether the proposed
-requirements are appropriate, and how well they were met.
+requirements were appropriate, and were met.
 
@@
 
-\paragraph*{Shared artifacts and artifact histories} are a central feature of
-RCloud, through its shared workbooks of experiments and analyses. We
-observed that hackers readily share work through RCloud. They use it to
-demonstrate techniques, share results with peers and managers, and to
-transfer algorithms to other groups. Artifact histories can be accessed through the notebook
-tree or through GitHub's web interface.
+\paragraph*{Shared artifacts and artifact histories} are a central feature
+of RCloud, through shared workbooks. We observed that hackers readily
+share experiments, demonstrate techniques, publish results to peers and
+managers, and transfer algorithms to other groups. Artifact histories
+can be accessed through the notebook tree or through GitHub's web interface.
+But a probem is that it is easy to create a large number of artifacts
+and the system does not help enough in their organization and mangement.
 
@@
 independent units (modularity) is a key to working productively in teams.
-It is good if the units can be kept small, so team members can realize
-benefits at least proportional to the work on the units (granularity).
+It is good if the units can be kept small or granular, so team members can
+realize benefits at least proportional to the work on the units.
 RCloud's notebook and versioning capabilities allow work to be divided
@@
 \paragraph*{View sharing, bookmarking} Most resources in RCloud are named
-and accessed as URLs. This proved to be an effective mechanism for
-sharing and for integrating analyses with external processes and systems.
-It is particularly advantageous for work to be shared as URLs that
-provide access to code and experiments, instead of by pasting static
-screenshots into documents and presentations.
+and accessed as URLs. This proved to be effective for sharing and integrating
+analyses externally. It is particularly advantageous for work to be shared
+as URLs that provide access to code and experiments, instead of by pasting
+screenshots into reports.
 

### 2015-06-16 system.tex
yes

@@
 
-The starting point in our design was to take advantage of the current
-infrastructure of web software as much as possible. HTTP, despite its
-deficiencies, has become the lingua franca of
-distributed interprocess communication. This meant that if our system
-could speak HTTP natively, then it would be possible to provide a low-friction
-path from development of an interactive data analysis script to an
-automated \emph{web service}, essentially a remote function call over
+Our design takes advantage of existing web software as much as possible.
+HTTP, despite its deficiencies, is the lingua franca of distributed
+interprocess communication. By speaking HTTP natively, our system
+provides a low-friction path from experimental data analysis scripts
+to automated \emph{web services}. 
+essentially a remote function call over
 the web, which can be invoked by higher-level tools. An example of
@@
 of code or hypertext in Markdown. As mentioned in
-Section~\ref{sec:related}, this is not novel; executable
-documents like this are a feature of many other systems,
-including Mathematica, IPython and Sage.
+Section~\ref{sec:related}, executable documents like
+this are a feature of previous systems, including Mathematica,
+IPython and Sage.
 
-Notebooks can be executed one cell at a time during an interactive
-editing session, providing a similar experience to traditional
-read-eval-print loops, or can also be executed all at once, providing
-an experience similar to running a shell script.
+Notebooks can be executed one cell at a time in an interactive
+session, similar to traditional read-eval-print
+loops, or can all be executed concurrently,
+similar to running a shell script.
 
@@
 ``always deployed''. In other words, the most recent version of a
-notebook is immediately available to all other users of the system as
-the notebook is being developed. Another way to describe this is that
+notebook is immediately available to all other users of the system.
+Another way to describe this is that
 RCloud lacks a ``save'' button: any notebook cell that runs is always
@@
 
-When we combine immediate deployment with the ability to refer to the
-latest version of a notebook by name, we get notebooks that are always live,
+In this design, we get notebooks that are always live,
 but sometimes broken. Because stability is important, we also
@@
 
-One side advantage of centralizing the execution and storage of
+One side benefit of centralizing the execution and storage of
 notebooks is that it becomes feasible to collect usage information
-that would otherwise be lost. In the case of social data visualization
+that might be lost in a conventional environment.
+In the case of social data visualization
 platforms, we would like to exploit usage data to help analysts
@@
 provided by companies such as Amazon and Netflix (``If you liked
-that film, then you should like this one too''). To compute such
+that film, then you should like this one too''). To make such
 recommendations, we need users to \emph{curate} the collection, by
@@
 This makes explicit indication of interest a nearly trivial operation,
-always available, to encourage its use.
+always available and easy to use.
 %
@@
 counts~\cite{Joachims:2005:AIC} and execution counts. In addition to
-these standard techniques of collecting feedback from web search, we
-anticipate applying static and dynamic code analysis to infer
-fine-grained information about relationships, for example, which
-packages and data sets often appear together, in the style of Fast et
-al's Codex system.~\cite{Fast:2014:ECS}.
+these standard techniques, we hope to apply static and dynamic code
+analysis to infer fine-grained information about relationships,
+for example, which packages and data sets often appear together,
+in the style of Fast et al's Codex system.~\cite{Fast:2014:ECS}.
 

### 2015-06-16 relatedwork.tex
fix typo in figure caption
yes

@@
 
-Broadly, the entire field of information visualization and visual analytics
-revolves around improving how users solve problems involving data.
-In this section, we focus specifically on
-system work relating to problem-solving environments and multiple users.
+Previous work has identified some requirements for supporting the
+visual analytics process, including the work of multiple users.
 
 \paragraph*{Social Data exploration and Analysis}
-ManyEyes~\cite{Viegas:2007:MAS} was a landmark system designed
-for the crowdsourced creation and publication of data
-visualizations. Although ManyEyes only supported a limited number of
-visual encodings, the system's success was both a precursor to more
+ManyEyes~\cite{Viegas:2007:MAS} was a landmark system for the
+crowdsourced creation and publication of data
+visualizations. Although ManyEyes supported a fixed set of
+visual encodings, the system's success was a precursor to more
 sophisticated solutions and an early indication that the world-wide
@@
 mentioned, one central issue in exploratory analysis is that
-problems change quickly over time, often in the course of developing a
+problems can change quickly, often in the course of developing a
 solution. As a result, systems should provide adequate support for
@@
 
-%% publishing web content such as knitR, Rpubs,
-%% rCharts, and gg2v. These tools are not collaborative, but they are
-%% aimed at better graphics and web applications.
-%% . {\bf iPython
-%%   notebooks (Jupyter)} are the closest work we are aware of.  The goal
-%% of this project is to provide shared notebooks and a remote execution
-%% environment for interpreted scripts.  \stephen{Are there significant
-%%   differences involving versioning, provenance, annotation and other
-%%   broad process support?}
-
-%% {\bf bl.ocks, jsfiddle, plotly} and other web services for sharing code
-%% and demos.
-
 \paragraph*{Needs of data analysts}
@@
 RCloud notebooks, and the integrated version control system for them,
-described in Section~\ref{sec:notebooks}, address modularity and granularity,
-and artifact histories.
+described in Section~\ref{sec:notebooks}, address {\it modularity and
+granularity}, and {\it artifact histories}.
 \emph{Starring}, the means for signaling interest in notebooks, described in
@@
 Notebooks can be executed as they're edited (left), or in a standalone viewer (right), via a slightly different URL.
-This lightweight provides a very low-friction mechanism for sharing results which we discuss in full in Section~\ref{sec:interviews}.
+This provides a lightweight, low-friction mechanism for sharing results which we discuss in detail in Section~\ref{sec:interviews}.
 }
 \end{figure*}
-
-%% \stephen{We have a problem we already went over some of this.}
-%% Our work has been inspired by and benefited from other efforts
-%% to improve data analysis environments and processes. Building in
-%% the immense foundation of systems for statistical graphics,
-%% recent work includes RStudio \cite{RStudio:2013:SWA},
-%% the R packages Markdown \cite{Allaire:2014:MMR},
-%% knitr \cite{Xie:2013:DDW}
-%% and Shiny \cite{RStudio:2013:SWA},
-%% and iPython notebooks \cite{Perez:2007:IAS}
-%% to name a few. RStudio aims at providing an integrated development environment
-%% for R, with support for publishing code in packages. Markdown,
-%% knitR and Shiny augment R with sophisticated reporting capabilities, including
-%% interactive web interfaces. IPython \cite{Perez:2007:IAS}
-%% shares many of our goals, such as providing a comprehensive environment
-%% for analysis and programming, and share-able documents on the web.
-%% \stephen{Do we now need to introduce one of those remarks where we
-%% explain that we are not the same as IPython.}
 

### 2015-06-16 discussion.tex
working

@@
 
-\paragraph*{Modularity and granularity} The ability to partition work into
-independent units (modularity) is a key to working productively in teams.
+\paragraph*{Modularity and granularity} Modularity, or the ability
+to partition work into independent units is a key to working
+productively in teams.
 It is good if the units can be kept small or granular, so team members can
@@
 of other tools, but sometimes it is difficult to know whether adding
-a compatibility feature will offload complexity or bring more in.
+a feature for compatibility will offload complexity, or increase it.
 
@@
 fast and there is no commitment: nothing gets saved and there is nothing to
-later clean up. In contrast, in RCloud's notebook interface, the extra cells need
-to get deleted later, and there seems to be a mental burden that \emph{bad stuff
+clean up. In contrast, in RCloud's notebook interface, the extra cells need
+to be deleted later, and there seems to be a mental burden that \emph{bad stuff
 is getting saved}.

### 2015-06-16 introduction.tex
working

@@
 In an environment where business needs can change rapidly, this
-process does not scale. Our goal is to merge the worlds of EDA and
-production.
+process does not scale. It is also difficult to trace problems
+from production back to EDA.
+Our goal is to merge the worlds of EDA and production to close this gap.
 
@@
 
-\item Many tools adopt a standalone or single developer perspective,
-not suited to collaboration and web deployment. 
-\item Analysts find insufficient support for tracing events or issues from production back to the EDA process.
-\end{itemize}
-
-In view of the opportunity to improve this situation,
+In consideration of the opportunity to improve this situation,
 we created a software environment that supports the end-to-end

### 2015-06-16 conclusion.tex
yes

@@
 \section{Conclusions and Future Work}
+We designed a prototype environment for exploration, sharing,
+presentation, and publishing.
+This study has provided a step toward practical ``DevOps for data science''
+and reproducible, publishable experiments. Our users found that the
+benefits of sharing
 
-%\gordon{I think that except for the DevOps slogan, this is redundant and too dense.
-I would replace the first three paragraphs with something like this:} We designed a prototype environment for exploration, sharing, presentation, and publishing.
-The exploration part didn't work out so well, but people like sharing stuff and integrating web and analysis languages.
-
-We presented a case for an environment that supports the
-visual analytics process for work by teams. The process
-includes data acquisition, exploratory data analysis,
-code development and deployment.
-
-Building on previous work to define requirements for
-visual analytics, we designed an environment
-for exploration, sharing, presenting and publishing.
-We implemented it in the RCloud prototype.
-
-We deployed RCloud in a community of working data scientists.
-Experience with the prototype provides evidence that data
-science teams and the organizations in which they work benefit from
-capabilities to support collaboration and to integrate
-the entire visual analytics process. We found that features for
-sharing and publishing were eagerly adopted. Features for 
-single-user data exploration, that compete with existing mature
+We found that features for sharing and publishing were eagerly adopted.
+Features for single-user data exploration, that compete with existing mature
 tools, were not accepted as readily. Some experienced users
@@
 
-This study has provided a step toward practical ``DevOps for data science''
-and reproducible, publishable experiments.  Possible next steps are
-to incorporate richer recommendation techniques, to provide fine-grained
-information security, and to improve the usability of the human interface.
+Possible next steps to improve the existing prototype are to incorporate
+richer recommendation techniques, to provide fine-grained information
+security, and to improve the usability of the human interface.
+
+There is a clear need for better software infrastructure to support
+experiments in visual analytics systems. We did benefit greatly from
+building RCloud on exisitng software and services, but the implementation
+was still difficult with much trial and error. Possibly some analogue
+of grammar-based visualization methods would help the visual analytics process.
 

### 2015-06-16 discussion.tex
yes

@@
 But a probem is that it is easy to create a large number of artifacts
-and the system does not help enough in their organization and mangement.
+and the system does not help enough to organize and navigate them.
 
 \paragraph*{Modularity and granularity} Modularity, or the ability
-to partition work into independent units is a key to working
+to partition work into independent units, is a key to working
 productively in teams.
-It is good if the units can be kept small or granular, so team members can
-realize benefits at least proportional to the work on the units.
+It is a benefit if the units can be kept small or granular, so team members
+can realize benefits at least proportional to the work on the units.
 RCloud's notebook and versioning capabilities allow work to be divided
@@
 \paragraph*{View sharing, bookmarking} Most resources in RCloud are named
-and accessed as URLs. This proved to be effective for sharing and integrating
-analyses externally. It is particularly advantageous for work to be shared
-as URLs that provide access to code and experiments, instead of by pasting
-screenshots into reports.
+and accessed as URLs. This proved to be effective for sharing analyses
+and integrating them with external processes.  It is particularly
+advantageous for work to be shared as URLs that provide access to code
+and experiments, instead of by pasting screenshots into reports.
 

### 2015-06-16 introduction.tex
yes

@@
 Processing and networking capabilities today far exceed
-what was barely imaginable then. Sophisticated technologies
-for processing, storing, publishing, searching and sharing
-information and analyses are commonplace.
+what was barely imaginable then. There has been a lot of
+work on distributed computing frameworks, and social computing
+technologies for publishing, searching and sharing information
+are common on the internet.
 But how we \emph{develop} analytic solutions has not changed much. 
 Although environments such as RStudio include many features to
-graphically interact with a local workspace, still follow the metaphor
-of terminal, text editor, and source files stored in file systems.
+graphically interact with a local workspace, they still follow the
+metaphor of terminal, text editor, and source files stored in file systems.
 

### 2015-06-16 relatedwork.tex
yes

@@
 
-Previous work has identified some requirements for supporting the
-visual analytics process, including the work of multiple users.
+Previous work has identified key requirements for supporting the
+visual analytics process, including by multiple users.
 
@@
 ManyEyes~\cite{Viegas:2007:MAS} was a landmark system for the
-crowdsourced creation and publication of data
-visualizations. Although ManyEyes supported a fixed set of
-visual encodings, the system's success was a precursor to more
-sophisticated solutions and an early indication that the world-wide
-web was a suitable platform for data visualization. One of the main
-challenges we faced in designing RCloud was providing an experience
-for \emph{consuming} visualizations as seamless as ManyEyes's, while
-sacrificing as little as possible on the generality of the analyses
-themselves.
+crowdsourced creation and publication of data visualizations.
+Although ManyEyes supported only a fixed set of visual encodings,
+the system's success was an early indication of the potential to
+combine social media on the web with visual analytics.
 
 \paragraph*{Notebooks as a medium for data analysis dissemination}
-The concept of a ``notebook'' as we use it here can be traced all the
-way back to Knuth's literate programming~\cite{Knuth:1984:LP}. In
-literate programming, a comprehensive description in prose of the
-behavior of the program is ``weaved'' together with the source code,
-yielding both an executable program and a human-readable document.
-A notebook represented as a collection of short, executable cells,
+The concept of a ``notebook'' as we use it can be traced
+back to Knuth's literate programming~\cite{Knuth:1984:LP}. In
+literate programming, a prose description of the behavior of
+a program is ``weaved'' with its source code, yielding both an
+executable program and a human-readable document.
+A notebook represented as a collection of small executable cells,
 originated with Mathematica, and in R, literate programming is
@@
 Project Jupyter~\cite{jupyter} (originally implemented as part of
-IPython) offer some notebook features, but lack a transparent
-mechanism for sharing and deployment in multiple-user settings.
+IPython) offers some notebook features, though it lacks a transparent
+mechanism for multi-user sharing and deployment.
 %
@@
 %
+\stephen{RStudio is highly acclaimed for its...}
 Although RStudio offers publication of literate R programs as a free

### 2015-06-18 casestudies.tex
try to clean up text

@@
 
+To illustrate ...
 As an example, we present a simple real-world application
 deployed under RCloud.
-
 The application is an implementation of LDAVis,
-a recent method for visualizing large text corpora.
-It was designed to help non-experts to explore
-collections of short text documents using topic modeling and
-visualization. Topic modeling~\cite{Blei:2003:LDA} is a standard
-technique for text summarization which, although powerful, requires
-some manual intervention and interpretation~\cite{Sievert:2014:LAM}.
+a method for visualizing large text corpora.
+LDAVis helps non-experts to explore collections of
+short text documents using topic modeling and
+visualization. Topic modeling~\cite{Blei:2003:LDA}, although
+powerful, often requires human interaction
+and interpretation~\cite{Sievert:2014:LAM}.
 
@@
 Gibbs sampling~\cite{} with custom R code written by the
-developers. The analysis module is, ultimately, a single R function
+developers. The analysis module is a single R function
 exposed to the web application via the Javascript-R RPC
@@
 
-The LDAVis application is simple, but highlights some unique features
-of RCloud. While RCloud notebooks allow deployment of R analyses over
+LDAVis highlights some unique features of RCloud.
+While RCloud notebooks allow deployment of R analyses over
 the web with no additional effort, RCloud \emph{applications} are more
@@
 for the front end. This requires additional knowledge over Shiny, but
-we argue that the RCloud model makes the analysis side simpler 
-(since analysts simply write R in the style they are used to),
-and the front end visualization side is simpler for the web
+the RCloud model makes the analysis side simpler (since analysts
+simply write R in the style they already know)
+and front end visualization side is simpler for the web
 developers (since they simply write the Javascript code in
@@
 
-%% Text analysis in
-
-%% To illustrate some aspects of RCloud, we present an example notebook
-%% for stock price analysis.\footnote{Unfortunately this report could not
-%% include production notebooks containing proprietary information.}
-%% The example, although simplified, shows key steps in the development
-%% and deployment of an application.
-
-%% \begin{figure}
-%% \includegraphics[width=\linewidth]{fig/casestudy1/casestudy1.pdf}
-%% \caption{\label{fig:stockvis}Iterations of a stock ticker
-%%   visualization, based on an example by Hadley Wickham. A
-%%   simple, static visualization of the closing price of a single stock
-%%   is progressively developed into a configurable display suitable
-%%   for a dashboard, then into an interactive visualization that
-%%   compares the volatility and volume of two stocks, and finally
-%%   into an API for data access by other RCloud notebooks. The notebooks
-%%   in this example can all be loaded as web pages. When a notebook
-%%   corresponding to a function call is displayed a web page,
-%%   its associated documentation is displayed.}
-%% \end{figure}
-
-%% It shows a sequence of visualizations of the performance
-%% of financial stocks over several years. The first visualization
-%% in this example uses ggplot2 and was written by Hadley Wickham.
-%% It reads data provided by Yahoo! Finance as a web service, and shows
-%% the price for a single trading symbol.
-
-%% The webpage that produces that visualization includes a link to
-%% the underlying source code. From this link, a user can fork the notebook.
-%% In the example,a configurable ticker is added, based on the URL of
-%% the notebook. The change to the notebook is minor.
-
-%% Next, the notebook author or a collaborator may decide to extend
-%% the application to also provide convenient access to the data,
-%% apart from its visualization.
-%% This is done by creating a notebook that defines a function.
-%% That notebook then becomes a \emph{subroutine} for other notebooks.
-%% It can be invoked from interactive notebooks, such as dashboards.
-%% The data access notebook is version-controlled like other notebooks.
-
-%% Finally, consider the situation where an analyst wants to understand
-%% the price dynamics of stocks with respects to other attributes and
-%% time ranges. For this, interactive visualization may be very helpful.
-%% In our example, the analyst creates an interactive tool with multiple
-%% linked views.
-
-%% Demonstrating the integration of R and JavaScript, here the analyst
-%% writes an R function that passes the dataframe and description of
-%% the charts to the dcplot library.  Simple R expressions
-%% are captured as trees to generate JavaScript expressions.
-%% The terse chart description language, with sensible defaults inspired by
-%% ggplot2, provides a simple yet powerful interface to the grouping
-%% and reduction functionality of the well-accepted charting libraries
-%% crossfilter, dc.js and d3.js.
-
-% \subsection{Text analysis\label{sec:textvis}}
-% Same.
-
 % IMPORTANT: what is unique about RCloud here

### 2015-06-18 discussion.tex
try to clean up text

@@
 \paragraph*{Security}
-It is essential to provide security in an environment for
+Security is an important concern in an environment for
 collaboration and data publishing. To provide some
@@
 active research topics, and many problems remain
-\cite{Moore:2011:SAF}. We are hopeful that having a richer
-environment for collaboration and information sharing will encourage
-new approaches to information security in visual analytics.
+\cite{Moore:2011:SAF}.
+% We are hopeful that having a richer
+% environment for collaboration and information sharing will encourage
+% new approaches to information security in visual analytics.
 %\gordon{This sounds like we think we are helping by pushing the
@@
 \paragraph*{Exploration is hard}
-R is a popular language for exploratory data analysis, and we built RCloud
-as a way for data scientists to seamlessly transition from EDA to sharing
-and presentation. But the combined notebook and command line interface we
-designed interferes with the use of R for exploration.
+% R is a popular language for exploratory data analysis, and we built RCloud
+% as a way for data scientists to seamlessly transition from EDA to sharing
+% and presentation.
+Users reported that RCloud's combined notebook and command line interface
+interferes with the familiar use of R for exploration.
+In current R practice, the command prompt is used to try things out.
+It is very fast and there is often no commitment: nothing gets saved
+and there is nothing to clean up.
+In contrast, in RCloud's notebook interface, the extra cells need
+to be deleted later, and there seems to be a mental burden
+that \emph{scratch work is getting saved}.
 
-In existing R tools, the command prompt is used to try things out. It is very
-fast and there is no commitment: nothing gets saved and there is nothing to
-clean up. In contrast, in RCloud's notebook interface, the extra cells need
-to be deleted later, and there seems to be a mental burden that \emph{bad stuff
-is getting saved}.
-
-One thing to try here is making deletion easier, for example with a
-shortcut key to remove the last cell, or a way to keep a few cells and cull
-the rest. A far more ambitious solution, suggested by one of our
+One thing to try is to make deletion easier, for example with a
+shortcut key that removes the last cell, or a way to keep a few cells
+and cull the rest. A far more ambitious solution, suggested by one of our
 interviewees, is to implement auditing of data analyses as it existed in S
@@
 
-\paragraph*{Discovery is hard}
-While we provided tools for searching and curating notebooks, the
-search function does not help for finding methodology (and does not protect
-searchers from erroneous notebooks), and the notebook tree starts to get
-unwieldy with hundreds of users and many hundreds of notebooks. In addition,
-our choice to retain notebooks even when they have been deleted 
-exacerbates the situation, since obsolete and broken notebooks stay around
-forever.
+\paragraph*{Discovery is hard} There are several problems.
+With hundreds of users and thousands of notebooks, the notebook
+tree becomes unwieldy.
+The search function does not protect searchers from obsolete and
+erroneous notebooks, which by default are retained forever.
+It helps with finding notebooks, but not with acquiring
+other types of information, such as methodology.
 
-One thing to try is tagging and filtering on tags, so that users
-can mark their own notebooks or those of others to make them easier to
-find. However, this relies on users making a conscious effort, which
-is not always reliable. It seems better if possible to lean more on
-passive metrics, such as how often a notebook has been run.
+One thing to try is tagging, so that users can mark notebooks
+to make them easier to find. However, this relies on users making
+a conscious effort, which is not always reliable. It seems better
+to rely more on passive metrics, such as how often a notebook has been run.
 
@@
 has been deleted or superseded by one of its forks, a user who opens it
-should be warned. And operations on multiple notebooks and folders of
+should be informed. Operations on multiple notebooks and folders of
 notebooks should be supported to make curation easier.
@@
 in teams is a lot more complicated. The forking feature is popular but it
-poses a problem, too, as dozens of versions of a notebook proliferate.
-
-One solution we are considering is to allow overwriting the current state
+poses a problem, too, when dozens of versions of a notebook proliferate.
+A possible solution is to allow overwriting the current state
 of a notebook with the current state of another fork of the notebook.
 
-A more general solutions would be a serious engineering endeavor. Even
-GitHub's admirable web interface does not support merging where there are
-conflicts. Unrestricted sharing of notebooks would probably require moving
-toward git's notion of decentralized code repositories, because GitHub gists
-do not support shared ownership.
+Implementing a more general solution seems like a big job.
+Even GitHub's web interface does not support merging where there are
+conflicts. Unrestricted sharing of notebooks may involve moving
+toward git's notion of decentralized code repositories, because GitHub
+gists do not support shared ownership.

### 2015-06-18 interviews.tex
try to clean up text

@@
 notes, ``[If] some person has done something similar, then you're able to just
-edit that, and that's saved a lot of work time for me.''
+edit that, and that's saved a lot of work for me.''
 
@@
 \subsection{Forking}
-The ability to start work where someone else let off, by forking, proved to
-be a popular feature. In fact, almost twice as many (131) users have forked
+The ability to start work where someone else let off, by forking, was
+a popular feature. Almost twice as many (131) users have forked
 someone else's notebook as have starred one (75).  Lilo says, ``It's one of the
@@
 
-A common but somewhat problematic use of forking is to change
+A common but ill-advised use of forking is to change
 parameters. Wendy says, ``I've been forking other people's notebooks [because] I
@@
 Parameters can be added to a notebook URL, but adding user interface
-elements to do the same thing requires expertise. Tool builders see a
+elements to do the same thing takes more expertise. Tool builders see a
 need for facilities to make it easier. Allison calls it ``web-enabling'' the
-notebook: ``[users] can always fork the notebooks and make changes, but I
+notebook: ``[Users] can always fork the notebooks and make changes, but I
 feel that if the owner of the notebook web-enables it, it's easier to
@@
 
-Iris points out that the automatic versioning works well for dealing with the
-minutiae of web development:
+Iris points out that automatic versioning works well for coping with
+the details of web development:
 \begin{quote}
@@
 More selective ways to search will be needed as the number of notebooks
-grows further. We explore some ideas in Section~\ref{sec:discussion}.
+grows further. We mention possible approaches in Section~\ref{sec:discussion}.
 
@@
 
-Although most of the analysts use and appreciate the sharing features,
-RCloud is less popular as a tool for exploratory data analysis. Every
-analyst with prior R experience still prefers to do analysis in another
-tool, and paste code into RCloud for sharing.
-
-This is mainly out of familiarity. Eric asks, ``Why would I want to use
+Although most of the analysts appreciate RCloud's sharing features,
+it is not popular as a tool for exploratory data analysis.
+All analysts with R experience prefer to do analysis in another
+tool, then paste code into RCloud for sharing.
+It appears this is mainly due to familiarity.
+Eric asks, ``Why would I want to use
 RCloud over my current setup? If it's just me, I like my text editor and
@@
 
-Coby also works with text files and command line, rearranging a file so the
-good code is at the top of the file. When he's done, all the right code is
-at the top. RCloud does not readily support this workflow, but Coby
-often shares work by pasting it into an RCloud notebook when he's done.
+Coby also works with text files and commands, rearranging source files
+so the good code is at the top. RCloud does not readily support this
+workflow, so Coby often shares work by pasting it into an RCloud
+notebook when he's done.
 
@@
 install. Joy says installing alternative or nonstandard packages is
-intrinsic to exploratory data analysis.
+intrinsic to exploratory data analysis. This is a weakness in systems
+like RCloud.
 
@@
 
-Simply the difference between a cell and the command prompt can confuse some people.
+The difference between a cell and a command prompt confuses some people.
 Kenyon explains:
@@
 The cell structure also can be problematic if some cells take a long time to
-execute. In Coby's work, there is often a ``long tail''.  The first cell may
-take just seconds to run, the next cell minutes, and the last cell 5 hours.
-In this situation, the Run Whole Notebook button is dangerous.
+execute. High variance is common: cells may take anywhere from seconds to
+hours to run.  In this situation, the Run Whole Notebook button is dangerous.
 Coby reports that when converting his work to notebooks, he
@@
 
-To avert this pitfall, some users write code within cells to explicitly cache results.
+To avoid this pitfall, some users write code within cells to explicitly cache results.
 Coby ends up ``littering the notebook with little switches that comment out the
-slow parts'' and either load or save the object to disk. One of the authors of this paper
+slow parts'' and either load or save the object to disk. One of the authors of this study
 manually writes cells that check if a result file exists, and perform a calculation only

### 2015-06-23 discussion.tex
table 1 rework
update

@@
 
-\begin{table*}
+\begin{table}
   \centering
   \begin{tabular}{l|cccccc}
-           & Versioning/Forking & Collaboration & Deployment & Multilanguage & Integrated Reports & Integrated Analysis \\
+    & V./F. & C. & D. & M.L. & R. & A. \\
     \hline
@@
 Jupyter    &                    &               &            &      x        &           x        &           x         \\
-Tableau    &                     &      x        &    x       &               &          x         &                     
+Tableau    &                    &      x        &     x      &               &           x        &                     \\
+ManyEyes   &                    &      x        &     x      &               &                    &                     
   \end{tabular}
-  \caption{Comparison of system features.}
-\end{table*}
+  \caption{Comparison of system features. Columns: Versioning/Forking, Collaboration, Deployment, Multi-language support, Integrated Reports, and Integrated Analysis.}
+\end{table}
 % SN says Tableau: versioning=no  collaboration=yes  deployment=yes multilanguage=no  reports=yes  analysis=no

### 2015-06-23 conclusion.tex
reduced text in conclusion.

@@
 We designed a prototype environment for exploration, sharing,
-presentation, and publishing.
-This study has provided a step toward practical ``DevOps for data science''
-and reproducible, publishable experiments. Our users found that the
-benefits of sharing
+presentation, and publishing.  This study has provided a step toward
+practical ``DevOps for data science'' and reproducible, publishable
+experiments.
 
-We found that features for sharing and publishing were eagerly adopted.
-Features for single-user data exploration, that compete with existing mature
-tools, were not accepted as readily. Some experienced users
-fashioned their own workflow so they could keep using familiar EDA tools.
+RCloud has become a valid avenue for easy deployment of visualizations
+and interactive exploration tools inside our organization.  We found
+that features for sharing and publishing were eagerly adopted.
+Features for single-user data exploration, on the other hand, competed
+with existing mature tools, and were not accepted as readily.  In
+future work, we hope to study and to understand the degree to which
+this reluctance is caused by bad design decisions on our part, and how
+much comes from ``mere'' change aversion.
 
-Possible next steps to improve the existing prototype are to incorporate
-richer recommendation techniques, to provide fine-grained information
-security, and to improve the usability of the human interface.
+We still firmly believe that the current biggest barrier to the
+adoption of visual analytics is in software infrastructure.
+Just like structuring visualization software around
+simple but composable parts is a successful approach (like Vega, ggplot and the grammar of
+graphics), it is intriguing to consider what systems are
+possible if this philosophy were carried over to the 
+broader requirements of interactive analysis.
 
-There is a clear need for better software infrastructure to support
-experiments in visual analytics systems. We did benefit greatly from
-building RCloud on exisitng software and services, but the implementation
-was still difficult with much trial and error. Possibly some analogue
-of grammar-based visualization methods would help the visual analytics process.
-
-RCloud code is available at \url{github.com/att/rcloud/}
-under an MIT open source license.
+RCloud is open-source, and available at \url{github.com/att/rcloud/}.

### 2015-06-23 discussion.tex
reduced text in paragraphs of discussion (collaboration and discovery)

@@
 
-\paragraph*{Discovery is hard} There are several problems.
-With hundreds of users and thousands of notebooks, the notebook
-tree becomes unwieldy.
-The search function does not protect searchers from obsolete and
-erroneous notebooks, which by default are retained forever.
-It helps with finding notebooks, but not with acquiring
-other types of information, such as methodology.
-
-One thing to try is tagging, so that users can mark notebooks
-to make them easier to find. However, this relies on users making
-a conscious effort, which is not always reliable. It seems better
-to rely more on passive metrics, such as how often a notebook has been run.
-
-The system also needs to help users to curate notebooks. If a notebook
-has been deleted or superseded by one of its forks, a user who opens it
-should be informed. Operations on multiple notebooks and folders of
-notebooks should be supported to make curation easier.
+\paragraph*{Discovery is hard} 
+With hundreds of users and thousands of notebooks, the notebook tree
+becomes unwieldy. The search function does not protect searchers from
+obsolete and erroneous notebooks, which by default are retained
+forever.  It helps with finding notebooks, but not with finding other
+types of metadata. In other words, \emph{textual search} is easy, but
+\emph{relevance search} is what we need. Allowing users to \emph{tag}
+notebooks is a natural idea, but it relies on users making a conscious
+effort, This goes somewhat against our ideal of not asking the user to
+do work for the system. Rather, we believe a better alternative is to
+have the system work for the user, relying on passive metrics, such as
+how often a notebook has been run.  More generally, infrastructure for
+automatic \emph{curation} of the software artifacts generated by teams
+seems to be an important avenue for future work.
 
 \paragraph*{Collaboration is hard}
-While forking provides a simple way to continue someone else's work, working
-in teams is a lot more complicated. The forking feature is popular but it
-poses a problem, too, when dozens of versions of a notebook proliferate.
-A possible solution is to allow overwriting the current state
-of a notebook with the current state of another fork of the notebook.
-
-Implementing a more general solution seems like a big job.
-Even GitHub's web interface does not support merging where there are
-conflicts. Unrestricted sharing of notebooks may involve moving
-toward git's notion of decentralized code repositories, because GitHub
-gists do not support shared ownership.
+While forking provides a simple way to continue someone else's work,
+teams working together generate workflows for which forking is not
+ideal.  Although popular, forking poses a problem when dozens of
+versions of a notebook proliferate. A fully general solution seems
+intractable. Nevertheless, we expect future versions of the
+notebook infrastructure to move toward git's ideal of decentralized
+code repositories, instead of the exclusive ownership mode we
+currently employ.

### 2015-06-23 discussion.tex
reduced text on caching and security

@@
 \paragraph*{Caching}
-An important consideration is how and where to introduce caching
-in RCloud's distributed computation model. Caching can dramatically
+An important consideration is how and where to introduce caching in
+RCloud's distributed computation model. Caching can dramatically
 improve performance in a way that is otherwise transparent to
 application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
-Though caching is usually implicit, in some environments, such as
-VisTrails, programmers may also have some explicit control over cache
-management. This may be desirable to ensure the repeatability of
-computations that rely on volatile or unreliable data sources.
-For example, the stock ticker use case is one for which the data
-might temporarily become unavailable, so caching could improve
-reliability and consistency of results. Alternatively, in situations
-involving relatively expensive computation (for example, analysis of large
-text corpora, clustering multivariate time series or deep learning algorithms)
-caching may be essential to adequate performance. Currently we would
-program this in RCloud by explicitly saving an analysis in a persistent database,
-but it seems better to implement this capability in a general purpose associative
-cache, instead of application-specific libraries. We envision cache management
-being implemented in a new middle layer to be added in the future.
+Currently, RCloud offers no support for implicit caching, but since it
+supports the entirety of the R language, it's possible to incorporate
+caching in R libraries.  Still, since the reuse of intermediate
+results is a recurring problem in data exploration environments, it
+seems better to implement this capability in a general purpose
+associative cache. We envision cache management being implemented in a
+new middle layer to be added in the future.
 
 \paragraph*{Security}
-Security is an important concern in an environment for
-collaboration and data publishing. To provide some
-protection for RCloud workbooks in an organization's intranet,
-RCloud uses an object capability model
-\cite{Miller:2006:RCT}
-recently added to the Rserve protocol \cite{Urbanek:2003:AFW}
-that prevents unauthenticated clients from making
-unauthorized calls to the RCloud runtime environment.
+Because RCloud allows arbitrary R code to be run
+\emph{by design}, it's not straightforward to protect deployments
+from running unauthorized arbitrary code. RCloud uses an object capability model
+\cite{Miller:2006:RCT} recently added to the Rserve
+protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
+to remote procedure calls; this way, web browsers never directly instruct the
+RCloud backend to execute arbitrary code, preventing unauthenticated clients
+from making unauthorized calls to the RCloud runtime environment.
 
-Our back-end environment assumes a high degree of trust between users.
-Access control for information security is delegated to the host
-operating system and web server. In practice, most operating
-systems and web servers rely either on coarse permission models,
+Our back-end environment, on the other hand, assumes a high degree of trust between users:
+access control for information security is delegated to the host
+operating system and web server. In practice, operating
+systems rely either on coarse permission models,
 which tend to be ineffective, or on detailed access control
-lists, which tend to be cumbersome.
-Information security in RCloud should be improved. More
+lists, which tend to be cumbersome. This means that information security
+in RCloud is a delicate, unsolved matter. More
 sophisticated approaches, such as information flow analysis,
 and query languages that ensure differential privacy, are
-active research topics, and many problems remain
-\cite{Moore:2011:SAF}.
+still research topics~\cite{Moore:2011:SAF}.
 % We are hopeful that having a richer
@@
 
-\paragraph*{Exploration is hard}
-% R is a popular language for exploratory data analysis, and we built RCloud
-% as a way for data scientists to seamlessly transition from EDA to sharing
-% and presentation.
-Users reported that RCloud's combined notebook and command line interface
-interferes with the familiar use of R for exploration.
-In current R practice, the command prompt is used to try things out.
-It is very fast and there is often no commitment: nothing gets saved
-and there is nothing to clean up.
-In contrast, in RCloud's notebook interface, the extra cells need
-to be deleted later, and there seems to be a mental burden
-that \emph{scratch work is getting saved}.
+\paragraph*{Nonforgetful, effortless exploration is hard}
+Users reported that RCloud's combined notebook and command line
+interface interferes with their familiar use of R for exploration.  In
+current R practice, the command prompt is used to try things out.  It
+is very fast and there is often no commitment: nothing gets saved by
+default, but clean up is automatic.  In contrast, in RCloud's
+notebook interface, cells generated doing exploration are saved in the
+notebook.  We arrived here from an insistence on reproducibility: it is a
+core principle of RCloud that no command should be run without being
+saved. In our experience some history steps are only found to be
+important in hindsight, and deleting history by default makes it too
+easy to lose those steps by accident.
 
-One thing to try is to make deletion easier, for example with a
-shortcut key that removes the last cell, or a way to keep a few cells
-and cull the rest. A far more ambitious solution, suggested by one of our
-interviewees, is to implement auditing of data analyses as it existed in S
-\cite{Becker:1988:Auditing}, so that a result could be selected and code
-which did not lead to it could be discarded automatically.
+Surprisingly, we found that when users realize that \emph{scratch work
+  is getting saved}, the extra versions become a mental burden for
+them. Clearly, RCloud has now moved too far to the other side, and the
+right solution is somewhere in the middle. An intriguing possibility,
+suggested by one of our interviewees, is to implement auditing of data
+analyses as it existed in S~\cite{Becker:1988:Auditing}, tying the
+data artifacts to the processes that generated them. As a result, the
+bookkeeping is reduced to selecting the \emph{data} to be kept, and
+the code which generates it can be derived and saved transparently.
 
-We got here from an insistence on reproducibility: it is a core principle
-of RCloud that no command should be run without being saved, even if it
-gets deleted later. This is why we did not implement a simple command line
-as many users request. We could imagine making an exception for commands
-that have no side-effects, but in the general case this is no easier than
-auditing because R is not a purely functional language.
+%% . This is why we did not implement a simple command line
+%% as many users request. We could imagine making an exception for commands
+%% that have no side-effects, but in the general case this is no easier than
+%% auditing because R is not a purely functional language.
 
-The interface is also perceived to be slower, both for computer and for
-human. First, there is higher latency when commands run, because
-RCloud saves a cell to a repository before executing it. It
-may be possible to safely write to the repository behind execution of the
-same code. Second, and perhaps more detrimental, the interface requires
-switching between the keyboard and mouse, which can be cumbersome.
-%\gordon{we don't need to back that up, right? i hate using passive voice.}
-Shortcut keys can defined as a workaround, but it would be better
-to reduce the need for them.
+%% The interface is also perceived to be slower, both for computer and for
+%% human. First, there is higher latency when commands run, because
+%% RCloud saves a cell to a repository before executing it. It
+%% may be possible to safely write to the repository behind execution of the
+%% same code. Second, and perhaps more detrimental, the interface requires
+%% switching between the keyboard and mouse, which can be cumbersome.
+%% %\gordon{we don't need to back that up, right? i hate using passive voice.}
+%% Shortcut keys can defined as a workaround, but it would be better
+%% to reduce the need for them.
 
-Other complaints mostly focus on layout, and can probably be addressed by
-offering better customization, and designs such as tabbed views.
+%% Other complaints mostly focus on layout, and can probably be addressed by
+%% offering better customization, and designs such as tabbed views.
 
@@
 obsolete and erroneous notebooks, which by default are retained
-forever.  It helps with finding notebooks, but not with finding other
-types of metadata. In other words, \emph{textual search} is easy, but
-\emph{relevance search} is what we need. Allowing users to \emph{tag}
-notebooks is a natural idea, but it relies on users making a conscious
-effort, This goes somewhat against our ideal of not asking the user to
-do work for the system. Rather, we believe a better alternative is to
-have the system work for the user, relying on passive metrics, such as
-how often a notebook has been run.  More generally, infrastructure for
-automatic \emph{curation} of the software artifacts generated by teams
-seems to be an important avenue for future work.
+forever.  In addition, some important metadata (has this notebook been
+recently run, or edited?) is also not easily searchable. In other words,
+\emph{textual search} is easy, but \emph{relevance search} is what we
+need. Allowing users to \emph{tag} notebooks is a natural idea, but it
+relies on users making a conscious effort, which goes somewhat against
+our ideal of not asking the user to do work for the system. Rather, we
+believe a better alternative is to have the system work for the user,
+relying on passive metrics, such as how often a notebook has been run.
+More generally, infrastructure for automatic \emph{curation} of the
+software artifacts generated by teams seems to be an important avenue
+for future work.
 

### 2015-06-23 discussion.tex
reduced text in Versioning ("and Upgrading", as it's now called). setion 6.2 pass is done.

@@
 
-\paragraph*{Versioning}
+\paragraph*{Versioning and Upgrading}
 Some important aspects of the environment are more difficult to manage
-in RCloud than in conventional systems. RCloud does not explicitly
-separate the development and deployment environments. More than that,
-every version of every prototype is shared by default. Although this encourages
-collaboration on work in progress it also quickly exposes errors such as
-misconfiguration, mismatches between versions of packages, and programming
-errors that can unintentionally affect production websites. Such errors
-are difficult to completely avoid, but the power of convenient sharing is often
-worth it (as proven true with distributed systems in general).
-Still, effective control over versioning when sharing problem needs more
-attention if RCloud would be scaled up to a wider number of users lacking
-informal channels to coordinate work.
+in RCloud than in conventional systems. RCloud does not separate the
+development and deployment environments, and every version of every
+prototype is shared by default. Although this encourages
+collaboration, it also quickly exposes errors such as
+misconfiguration, mismatches between versions of packages, and
+programming errors that can unintentionally affect production
+websites. For example, if the RCloud website is distributed across
+servers in different machines, different versions of system software
+can cause errors that are hard to find and fix.
 
-Versioning needs to be managed in several places: in scripts;
-in the R environment such as the installed libraries and packages;
-and in the external environment such as the operating system and
-protocols spoken by remote services. At one end of the scale, we have
-full control over the versioning of scripts via git, along with conventions
-to name stable or working versions of scripts, and versions known to work together.
-The R environment itself is under the control of its package manager
-that has rules to ensure reasonable consistency.
-It is at least possible for RCloud to access information about
-package versions and configurations, but support for checking compatibility
-and maintaining multiple versions in the same environment is not strong.
-At the other end of the scale, there is not much reason to expect version
-control for the external environment. Most programmers rely on clean living
-and a careful approach to system upgrades.
-
-On top of this, RCloud allows or even encourages hackers to fork experiments
-to try new ideas quickly. So far we have not done much more to address the
-problem of having a lot of bits and pieces of code and data lying around,
-though we have provided a framework in which it should be easier to find them
-and to apply algorithms and metadata to organize them.
+Versioning of different software pieces must be managed in several
+levels: in data analysis and visualization scripts (that is, the
+notebooks themselves); in the R environment such as the installed
+libraries and packages; and in the external environment such as the
+operating system, libraries, and protocols spoken by remote
+services. At one end of the scale, we have full control over the
+versions via git, along with conventions that fix stable notebook
+versions.  The R environment itself is under the control of its
+package manager that has rules to ensure reasonable consistency.  It
+is at least possible for RCloud to access information about package
+versions and configurations, but support for checking compatibility
+and maintaining multiple versions in the same environment is not
+strong.  At the other end of the scale, there is not much reason to
+expect version control for the external environment, and consider what
+would happend if a web service that is used by a notebook changes the
+format of its output: notebooks now stop working, but nothing in
+RCloud has changed.
 
@@
 from making unauthorized calls to the RCloud runtime environment.
-
+%
 Our back-end environment, on the other hand, assumes a high degree of trust between users:

### 2015-06-24 casestudies.tex
fix table columns

@@
 
-To illustrate ...
-As an example, we present a simple real-world application
-deployed under RCloud.
-The application is an implementation of LDAVis,
-a method for visualizing large text corpora.
 LDAVis helps non-experts to explore collections of
-short text documents using topic modeling and
-visualization. Topic modeling~\cite{Blei:2003:LDA}, although
-powerful, often requires human interaction
+short text documents, using topic modeling and
+visualization. Topic modeling~\cite{Blei:2003:LDA},
+although powerful, often requires human guidance
 and interpretation~\cite{Sievert:2014:LAM}.
@@
 LDAVis was developed by two technical staff members at
-AT\&T Labs, and originally was written in RStudio's
-Shiny~\cite{RStudio:2013:SWA}, a framework for developing R
-applications for the web.  While Shiny provides outstanding ease of
-development, it turns out that discoverability and deployment are equally
-important aspects in the lifecycle of an internal application (see
-Section~\ref{sec:interviews}). In these aspects, RCloud provides a
-simple solution: \emph{all developed notebooks are automatically deployed}.
+AT\&T Labs, originally in RStudio Shiny~\cite{RStudio:2013:SWA},
+a framework for developing web applications in R.
+While Shiny offers outstanding ease of development,
+discoverability and deployment are also very important aspects
+in an application's lifecycle (see Section~\ref{sec:interviews}).
+RCloud's assumption that \emph{all developed notebooks are automatically
+deployed} simplifies this for LDAVis.
 
@@
 for the front end. This requires additional knowledge over Shiny, but
-the RCloud model makes the analysis side simpler (since analysts
+the RCloud model makes analysis simpler (since analysts
 simply write R in the style they already know)
-and front end visualization side is simpler for the web
+and front end visualization side is simpler for web
 developers (since they simply write the Javascript code in
-the style \emph{they} are used to). In addition, RCloud applications
+the style \emph{they} know). In addition, RCloud applications
 inherit the automatic deployment and discoverability features of

### 2015-06-25 conclusion.tex
fixed spelling errors

@@
 \section{Conclusions and Future Work}
-We designed a prototype environment for exploration, sharing,
-presentation, and publishing.  This study has provided a step toward
-practical ``DevOps for data science'' and reproducible, publishable
-experiments.
+We designed and deployed a prototype environment for visual analytics
+exploration, sharing, presentation, and publishing. It is a step toward
+practical ``DevOps for data science'' and reproducible, publishable data
+science experiments.
 
-RCloud has become a valid avenue for easy deployment of visualizations
-and interactive exploration tools inside our organization.  We found
-that features for sharing and publishing were eagerly adopted.
-Features for single-user data exploration, on the other hand, competed
-with existing mature tools, and were not accepted as readily.  In
-future work, we hope to study and to understand the degree to which
+RCloud was accepted as a platform for easy deployment of visualizations
+and interactive exploration tools inside our organization.  
+Sharing and publishing were eagerly adopted.
+Features for single-user data exploration, on the other hand, that
+compete with existing mature tools, were not readily accepted.
+In future work, we hope to study and to understand the degree to which
 this reluctance is caused by bad design decisions on our part, and how
-much comes from ``mere'' change aversion.
+much comes from ``mere'' change aversion. It is also clear we need
+need better ways to manage the persistence of workbooks and to cope
+with potential information overload.
 
-We still firmly believe that the current biggest barrier to the
-adoption of visual analytics is in software infrastructure.
-Just like structuring visualization software around
-simple but composable parts is a successful approach (like Vega, ggplot and the grammar of
-graphics), it is intriguing to consider what systems are
-possible if this philosophy were carried over to the 
+We find much evidence that the biggest barrier to the adoption of
+visual analytics is inadequate software infrastructure.
+Just as structuring visualization software around simple,
+composable parts is highly successful (exemplified by
+Grammar of Graphics approaches like Vega and ggplot)
+it is intriguing to consider what systems will be
+possible when this philosophy is extended to the
 broader requirements of interactive analysis.

### 2015-06-25 discussion.tex
fix table columns
fixed spelling errors

@@
   \begin{tabular}{l|cccccc}
-    & V./F. & C. & D. & M.L. & R. & A. \\
+    & V/F & C & D & ML & R & A \\
     \hline
@@
 can be accessed through the notebook tree or through GitHub's web interface.
-But a probem is that it is easy to create a large number of artifacts
+But a problem is that it is easy to create a large number of artifacts
 and the system does not help enough to organize and navigate them.
@@
 expect version control for the external environment, and consider what
-would happend if a web service that is used by a notebook changes the
+would happened if a web service that is used by a notebook changes the
 format of its output: notebooks now stop working, but nothing in

### 2015-06-25 introduction.tex
fixed spelling errors

@@
 inefficiency due to a lack of transparency in accessing the work of
-other team memnbers.  The processes and technologies supporting
+other team members. The processes and technologies supporting
 data analysis and visual analytics today are fragmented.
@@
 terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
-gain the benefits of the tools developed by hackers and scripters.
+benefit from tools developed by hackers and scripters.
 Results of experiments are often shared first by copying screen
 shots or static output. By the time decisions are made based on
-the results of that screenshot, data and processes may have changed
+the results of those screenshots, data and processes may have changed
 so much that the decision can be wrong.
@@
 
-Fortunately, there is abundant evidance that technology can
+Fortunately, there is abundant evidence that technology can
 solve many of these problems.
@@
 
-In consideration of the opportunity to improve this situation,
+In view of the opportunity to improve this situation,
 we created a software environment that supports the end-to-end
@@
 
-In this paper, we contribute the
-design of RCloud, with an interview study based on the
-course of its development and deployment at AT\&T Labs for 
-about three years.
+In this paper, we contribute the design of RCloud, and an
+interview study based on its development and deployment
+at AT\&T Labs for about three years.
 
-This environment knits together some familiar tools, and 
+RCloud knits together some familiar tools, and 
 provides new features to find data and code; create experimental workbooks;

### 2015-06-25 discussion.tex
some copy editing

@@
 is getting saved}, the extra versions become a mental burden for
-them. Clearly, RCloud has now moved too far to the other side, and the
-right solution is somewhere in the middle. An intriguing possibility,
+them. Clearly, RCloud is an extreme, and the right solution is
+in the middle. An intriguing possibility,
 suggested by one of our interviewees, is to implement auditing of data
-analyses as it existed in S~\cite{Becker:1988:Auditing}, tying the
-data artifacts to the processes that generated them. As a result, the
-bookkeeping is reduced to selecting the \emph{data} to be kept, and
-the code which generates it can be derived and saved transparently.
+analyses as it existed in S~\cite{Becker:1988:Auditing}, tying
+data artifacts to the processes that generated them. As a result,
+bookkeeping can be reduced to selecting the \emph{data} to be kept,
+and the code which generates it can be derived and saved transparently.
 
@@
 \paragraph*{Discovery is hard} 
-With hundreds of users and thousands of notebooks, the notebook tree
+With hundreds of users and thousands of notebooks, a global notebook tree
 becomes unwieldy. The search function does not protect searchers from
 obsolete and erroneous notebooks, which by default are retained
-forever. In addition, some important metadata (has this notebook been
+forever. In addition, important metadata (has this notebook been
 recently run, or edited?) is also not easily searchable. In other words,
 \emph{textual search} is easy, but \emph{relevance search} is what we
-need. Allowing users to \emph{tag} notebooks is a natural idea, but it
-relies on users making a conscious effort, which goes somewhat against
-our ideal of not asking the user to do work for the system. Rather, we
-believe a better alternative is to have the system work for the user,
+need. Allowing users to \emph{tag} notebooks is natural, but it
+relies on users making a conscious effort, which goes against
+our ideal of not asking the user to do work for the system.
+It is better to have the system work for the user,
 relying on passive metrics, such as how often a notebook has been run.
-More generally, infrastructure for automatic \emph{curation} of the
+More generally, infrastructure for automatic \emph{curation} of
 software artifacts generated by teams seems to be an important avenue
@@
 notebook infrastructure to move toward git's ideal of decentralized
-code repositories, instead of the exclusive ownership mode we
-currently employ.
+code repositories, instead of the exclusive ownership mode RCloud has now.

### 2015-06-25 introduction.tex
some copy editing

@@
 RCloud knits together some familiar tools, and 
-provides new features to find data and code; create experimental workbooks;
+provides new features to find data and code; create experimental notebooks;
 run experiments; annotate and deploy experiments as end-user websites or
@@
 searched immediately. Recommendation is as easy as clicking a star
-on a useful workbook.
+on a useful notebook.
 
@@
 is shown in figure~\ref{fig:system}.
-We chose R as the foundation for our prototype because it is already
+We chose R as the foundation because it is already
 the dominant statistical computing language in our lab.

### 2015-06-25 relatedwork.tex
figure squeezing
some copy editing

@@
 \centering
-\includegraphics[width=.75\linewidth]{fig/system/system.pdf}
+\includegraphics[width=.9\linewidth]{fig/system/system.pdf}
 \caption{\label{fig:system}A diagram of RCloud's architecture. }
@@
 the system's success was an early indication of the potential to
-combine social media on the web with visual analytics.
+combine social media with visual analytics.
 
@@
 The concept of a ``notebook'' as we use it can be traced
-back to Knuth's literate programming~\cite{Knuth:1984:LP}. In
-literate programming, a prose description of the behavior of
+back to Knuth's literate programming~\cite{Knuth:1984:LP}.
+In literate programming, a prose description of the behavior of
 a program is ``weaved'' with its source code, yielding both an
@@
 %
-\stephen{RStudio is highly acclaimed for its...}
+RStudio is highly acclaimed for its polished and powerful human interface.
 Although RStudio offers publication of literate R programs as a free
-service on their website, the workflow is somewhat disconnected from
+service on its website, the workflow is somewhat disconnected from
 the development of those programs. Once they're uploaded, it's hard
-for other users to build off of the work published (or even for the
-original author to update new versions). In other words, RStudio
+for other users to build off the work published, or even for the
+original author to update new versions. In other words, RStudio
 handles \emph{publication}, but not \emph{deployment} and sharing.
@@
 mentioned, one central issue in exploratory analysis is that
-problems can change quickly, often in the course of developing a
-solution. As a result, systems should provide adequate support for
-tracking \emph{changes} of the data analysis scripts. VisTrails was
-one of the original systems for managing \emph{process provenance} ,
-and demonstrated the value of capturing aspects of the processes that
+problems may change quickly, often while in the course of
+developing a solution. As a result, systems need to provide adequate
+support for tracking \emph{changes} of data analysis scripts. VisTrails
+was one of the original systems for managing \emph{process provenance},
+and it demonstrated the value of capturing aspects of the processes that
 surround data analysis experiments and tools, including detailed
@@
 \paragraph*{Web-based tools for sharing code snippets}
-There have been quite a few tools recently developed for quickly
+Quite a few tools were recently developed for quickly
 sharing small programs on the web, including
@@
 automatically through the web browser. If Javascript eventually becomes
-the lingua franca of exploratory data analysis, we can foresee
+the lingua franca of exploratory data analysis, we foresee
 building a simpler version of RCloud where all execution happens
@@
 drives or email. Analysts rarely stored their analytic code in source
-control.'' Their study highlights the opportunity for better ways
-of supporting collaboration and sharing in data analysis teams.
+control.'' Their study highlights the opportunity to contribute better
+ways of supporting collaboration and sharing by data analysis teams.
 
@@
 
-One overall goal is to reduce the gap between implementers and
+One high-level goal is to reduce the gap between implementers and
 deployers in visual analytics. The fusion of development with
@@
 
-We next give a description of the system architecture,
-and how it enables capabilities that satisfy the requirements as described.
+% We next give a description of the system architecture,
+% and how it enables capabilities that satisfy the requirements as described.
+% \stephen{This should flow naturally, or have a more natural transition.}

### 2015-06-25 casestudies.tex
some copy editing

@@
 Each text topic is a probability distribution over the all the
-words of the document. To expose patterns in the relationships
-between topics, LDAVis employs a combination of
+words of a document. To expose patterns in the relationships
+between topics, LDAVis combines
 interactive visualization and dimensionality reduction,
@@
 algorithms and distance measures are implemented in R, which
-means they can be executed on the RCloud servers as well.
+means they can be executed on RCloud servers as well.
 
@@
 al.~\cite{Satyanarayan:2014:DID}.
-Although custom Javascript was required, the flexibility is welcomed
+Although custom Javascript was required, its flexibility is welcomed
 by many web developers.
@@
 and front end visualization side is simpler for web
-developers (since they simply write the Javascript code in
+developers (since they simply write Javascript in
 the style \emph{they} know). In addition, RCloud applications

### 2015-06-25 conclusion.tex
some copy editing

@@
 \section{Conclusions and Future Work}
-We designed and deployed a prototype environment for visual analytics
-exploration, sharing, presentation, and publishing. It is a step toward
+We designed. implemented and deployed a prototype visual analytics environment
+for data exploration, sharing, presentation, and publishing. It is a step toward
 practical ``DevOps for data science'' and reproducible, publishable data
@@
 
-RCloud was accepted as a platform for easy deployment of visualizations
-and interactive exploration tools inside our organization.  
-Sharing and publishing were eagerly adopted.
+RCloud was readily accepted as a platform for deployment of
+visualizations and interactive exploration tools inside our organization.  
+Notebook sharing and publishing were eagerly adopted.
 Features for single-user data exploration, on the other hand, that
 compete with existing mature tools, were not readily accepted.
-In future work, we hope to study and to understand the degree to which
+In the future we hope to study and understand the degree to which
 this reluctance is caused by bad design decisions on our part, and how
 much comes from ``mere'' change aversion. It is also clear we need
-need better ways to manage the persistence of workbooks and to cope
+need better ways to manage the persistence of notebooks and to cope
 with potential information overload.

### 2015-06-25 interviews.tex
fix table columns
some copy editing

@@
 \subsection{Forking}
-The ability to start work where someone else let off, by forking, was
-a popular feature. Almost twice as many (131) users have forked
+The ability to start work where someone else let off, by forking, was popular.
+Almost twice as many (131) users have forked
 someone else's notebook as have starred one (75). Lilo says, ``It's one of the
@@
 Although we intended forking to be available to improve others' code, we 
-initially didn't anticipate support forking one's own notebooks, which
+didn't anticipate forking one's own notebooks, which
 proved very useful. Kenyon says, ``I fork my own notebooks because I'm going off
@@
 Forking also provides a way for others to troubleshoot when something goes
-wrong.  When Hugh works with the users of his notebooks, ``I'll teach people
+wrong. When Hugh collaborates with users of his notebooks, ``I'll teach people
 to intercept the result in the middle, to insert print statements here and
@@
 elements to do the same thing takes more expertise. Tool builders see a
-need for facilities to make it easier. Allison calls it ``web-enabling'' the
-notebook: ``[Users] can always fork the notebooks and make changes, but I
+need for facilities to make that easier. Allison calls it ``web-enabling''
+notebooks: ``[Users] can always fork the notebooks and make changes, but I
 feel that if the owner of the notebook web-enables it, it's easier to
@@
 Coby also works with text files and commands, rearranging source files
-so the good code is at the top. RCloud does not readily support this
-workflow, so Coby often shares work by pasting it into an RCloud
-notebook when he's done.
+so active code is at the top. RCloud does not readily support this
+workflow, so Coby often shares work by pasting code into an RCloud
+notebook after he's done.
 
@@
 install. Joy says installing alternative or nonstandard packages is
-intrinsic to exploratory data analysis. This is a weakness in systems
-like RCloud.
+intrinsic to exploratory data analysis. This is a weakness in shared
+environments like RCloud.
 
@@
 Users felt that cells do not capture the right level of granularity: they
-either hold too much code or too little. Kenyon says that RCloud's cell structure
+hold either too much code or too little. Kenyon says that RCloud's cell structure
 ``tempts me to type a big long thing and then run the whole thing, as opposed to
@@
 so the only way to get that functionality is if every line's in one cell.''
-Another complaint related to the human interface was that cells can take up
-substantial vertical space, requiring scrolling.
+Another complaint was that cells can take up too much vertical space, requiring scrolling.
 % Rick complains, ``Every time I type more stuff, the notebook gets longer and
@@
 ends up with a lot of comments saying ``This cell takes a long time to run.''
-
 To avoid this pitfall, some users write code within cells to explicitly cache results.
 Coby ends up ``littering the notebook with little switches that comment out the
-slow parts'' and either load or save the object to disk. One of the authors of this study
-manually writes cells that check if a result file exists, and perform a calculation only
-if it's needed.
+slow parts'' and either load or save the object to disk. One of the authors of this study manually writes cells that check if a result file exists, and perform a calculation only when needed.
 
@@
 packages and example notebooks at the same time. Eric will ``go back and
-change it and update the page, but then what happens is people have in the
-meantime forked it. I'll have to change a package as well, so their
+change [a notebook] and update the page, but in the meantime people
+have forked it. I'll have to change a package as well, so their
 old fork stops working, and they complain.''
@@
 had become obsolete. By design, notebooks that are deleted in the user interface
-are not purged from the repository. Now he says he is scared to share notebooks:
+are not purged from the repository. He says he became reluctant to share notebooks:
 ``Do I want to support this forever?''

### 2015-06-25 interviews.tex
fixups

@@
 analysts and business needs.
-
+% no paragraphbreak needed! scn
 The subjects differed in their evaluation of its benefits and disadvantages,
@@
 
-Besides providing a way to present work, notebook sharing can provide a
-starting point for coding. Lilo says, ``The best part is how easily you can
+Besides providing a way to present work, notebook sharing become a starting point
+in coding. Lilo says, ``The best part is how easily you can
 share code. You can find a working example, rather than wearing
@@
 actually run the snippets [and display the results].'' He also uses RCloud for
-describing and debugging problems with data sources.
+describing and debugging data sources.
 
@@
 
-Automatic source control is also a popular feature. Lilo says, ``Instead of
+Automatic source control proved helpful. Lilo says, ``Instead of
 looking back and saying I've got a billion files here in this subdirectory and I
@@
 have done.''
-
 But other users prefer to browse just the notebooks of experts they
@@
 more obscure than I'm likely to find in some random person's.''
-
 More selective ways to search will be needed as the number of notebooks
@@
 
-Integrating R also helps Allison's application: ``Having an open session where I can run R commands or
+Integrating R helps Allison's application: ``Having an open session where I can run R commands or
 functions without having to invoke an API or send a request and then
@@
 
-Although most of the analysts appreciate RCloud's sharing features,
-it is not popular as a tool for exploratory data analysis.
-All analysts with R experience prefer to do analysis in another
+Although most analysts appreciated RCloud's sharing features,
+it was not a popular tool for exploratory data analysis.
+All analysts with R experience prefered to do analysis in another
 tool, then paste code into RCloud for sharing.
@@
 interface.
-
 The interface takes some learning. Lilo says,
@@
 ends up with a lot of comments saying ``This cell takes a long time to run.''
-To avoid this pitfall, some users write code within cells to explicitly cache results.
-Coby ends up ``littering the notebook with little switches that comment out the
-slow parts'' and either load or save the object to disk. One of the authors of this study manually writes cells that check if a result file exists, and perform a calculation only when needed.
+To avoid this pitfall, some users write code to explicitly cache results.
+% Coby ends up ``littering the notebook with little switches that comment out the
+% slow parts'' and either load or save the object to disk. One of the authors of this study manually writes cells that check if a result file exists, and perform a calculation only when needed.
 

### 2015-06-25 relatedwork.tex
fixups

@@
 The concept of a ``notebook'' as we use it can be traced
-back to Knuth's literate programming~\cite{Knuth:1984:LP}.
+to Knuth's literate programming~\cite{Knuth:1984:LP}.
 In literate programming, a prose description of the behavior of
@@
 executable program and a human-readable document.
-A notebook represented as a collection of small executable cells,
+A notebook represented as a collection of small executable cells
 originated with Mathematica, and in R, literate programming is
 supported by packages such as knitr and RMarkdown~\cite{Xie:2013:DDW}.
-Project Jupyter~\cite{jupyter} (originally implemented as part of
-IPython) offers some notebook features, though it lacks a transparent
-mechanism for multi-user sharing and deployment.
-%
-Further afield, {Electronic Lab Notebooks} are applications for organizing
-and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
-In a sense, we hope to adapt and extend this concept to the work of
-visual analytics teams.
-%
+Project Jupyter~\cite{jupyter} (originally part of
+IPython) offers some notebook features, though it lacks transparent
+multi-user sharing and deployment.
 RStudio is highly acclaimed for its polished and powerful human interface.
@@
 the development of those programs. Once they're uploaded, it's hard
-for other users to build off the work published, or even for the
+for others to build off the work published, or even for the
 original author to update new versions. In other words, RStudio
 handles \emph{publication}, but not \emph{deployment} and sharing.
+%
+% This may be something we can delete -- scn
+Further afield, {Electronic Lab Notebooks} organize
+and share data from scientific lab experiments\cite{Rubacha:2011:ELN}.
+% In a sense, we hope to adapt and extend this concept to the work of visual analytics teams.
+%
 
@@
 support for tracking \emph{changes} of data analysis scripts. VisTrails
-was one of the original systems for managing \emph{process provenance},
-and it demonstrated the value of capturing aspects of the processes that
-surround data analysis experiments and tools, including detailed
+was an early system for managing \emph{process provenance},
+that demonstrated the value of capturing aspects of the processes
+that surround data analysis experiments and tools, including detailed
 history, collaboration, and deployment~\cite{Callahan:2006:VVM}.
@@
 while Crowdlabs exposes these capabilities on a website
-feature~\cite{Mates:2011:CSA} workflow upload and remote execution. In
-our view, the impedance mismatch between a dataflow pipeline
-specification and the power of a general-purpose language is too great
-for the type of general exploratory work in data science teams. At the
-expense of ease of use for non-programmers, RCloud tries to provide a
-closer match for analysts accustomed to creating and executing R and
-Python code, while retaining attractive properties like transparent
+featuring workflow upload and remote execution~\cite{Mates:2011:CSA}.
+In our view, the impedance mismatch between dataflow pipeline
+specifications and powerful general-purpose languages is too great
+for exploratory work by data science teams. At the
+expense of ease of use for non-programmers, RCloud provides
+a closer match for analysts familiar with R and Python,
+while retaining attractive properties like transparent
 provenance tracking and interactive data visualization on the web.
@@
 share Javascript programs, which means that deployment happens
-automatically through the web browser. If Javascript eventually becomes
+automatically through the web browser. If Javascript becomes
 the lingua franca of exploratory data analysis, we foresee
@@
 either on the client side or via web services. Unfortunately that is
-not a realistic assumption at present, making these stand-alone
+not realistic today, making these stand-alone
 solutions unsuitable for our purpose. Plot.ly is notable in that it
-provides API support for publishing \emph{from} scripts: in other
-words, it is possible to generate a plot.ly visualization from inside
+provides API support for publishing \emph{from} scripts: 
+it is possible to generate a plot.ly visualization within
 another program. Although this is an intriguing idea, it
@@
 analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
-this cycle that cost time and effort to overcome. RCloud seeks to
-reduce this mismatch. Kandel et al also point out that larger teams
+this cycle that cost time and effort to overcome.
+% RCloud seeks to reduce this mismatch.
+Kandel et al also point out that larger teams
 are becoming more common in data analysis, that supporting
@@
 drives or email. Analysts rarely stored their analytic code in source
-control.'' Their study highlights the opportunity to contribute better
-ways of supporting collaboration and sharing by data analysis teams.
+control.''
+% Their study highlights the opportunity to contribute better
+% ways of supporting collaboration and sharing by data analysis teams.
 
@@
 (cleaning, parsing and transformation)is a major part of exploratory
-analysis and visualization~\cite{Kandel:2011:RDI}. We view this
-as attacking a different semantic level than ours, but also
-showing the need for an environment that enables better sharing
+analysis and visualization~\cite{Kandel:2011:RDI}.
+% We view this as attacking a different semantic level than ours, but also showing
+Although attacking a different semantic level from ours, it shows
+the need for an environment that enables better sharing
 of the knowledge, tools and processes to do this. Anecdotally
@@
 
-The need for integrating statistics and visualization has been
-highlighted in previous studies and is widely understood by
-various technical communities \cite{Perer:2008:ISA}
-Lucas and Roth were early advocates of combining
-data exploration with presentation and publication \cite{Lucas:1996:EIV}.
+% The need for integrating statistics and visualization has been
+% highlighted in previous studies and is widely understood by
+% various technical communities \cite{Perer:2008:ISA}
+% Lucas and Roth were early advocates of combining
+% data exploration with presentation and publication \cite{Lucas:1996:EIV}.
 

### 2015-06-25 conclusion.tex
fixups

@@
 \section{Conclusions and Future Work}
-We designed. implemented and deployed a prototype visual analytics environment
+We designed, implemented and deployed a prototype visual analytics environment
 for data exploration, sharing, presentation, and publishing. It is a step toward

### 2015-06-25 discussion.tex
fixups

@@
 \paragraph*{Curation} Even without formal group management, users may
-curate groups of related notebooks using notebook tree folders.
+organize groups of related notebooks using tree folders.
 We found this particularly effective in collecting and distributing
@@
 \paragraph*{Versioning and Upgrading}
-Some important aspects of the environment are more difficult to manage
-in RCloud than in conventional systems. RCloud does not separate the
-development and deployment environments, and every version of every
-prototype is shared by default. Although this encourages
+Some aspects of the programming environment are more difficult to
+manage in RCloud than in conventional systems. RCloud does not separate
+the development and deployment environments, and every version of every
+prototype is shared by default. While this encourages
 collaboration, it also quickly exposes errors such as
-misconfiguration, mismatches between versions of packages, and
-programming errors that can unintentionally affect production
-websites. For example, if the RCloud website is distributed across
-servers in different machines, different versions of system software
-can cause errors that are hard to find and fix.
+misconfiguration, mismatches between package versions, and
+programming errors, that can unintentionally affect production
+websites. For example, if an RCloud website is distributed across
+multiple servers, conflicting versions of system software
+can cause errors that are hard to debug.
 
-Versioning of different software pieces must be managed in several
-levels: in data analysis and visualization scripts (that is, the
+Versioning of software components must be managed in several
+levels: in data analysis and visualization scripts (the
 notebooks themselves); in the R environment such as the installed
@@
 operating system, libraries, and protocols spoken by remote
-services. At one end of the scale, we have full control over the
-versions via git, along with conventions that fix stable notebook
-versions.  The R environment itself is under the control of its
-package manager that has rules to ensure reasonable consistency.  It
-is at least possible for RCloud to access information about package
+services. At one end of the scale, we have full control over notebooks
+via git, along with conventions to fix stable notebook versions.
+The R environment is under the control of its package manager that
+has rules to ensure reasonable consistency.
+(It is possible for RCloud to access information about package
 versions and configurations, but support for checking compatibility
 and maintaining multiple versions in the same environment is not
-strong. At the other end of the scale, there is not much reason to
-expect version control for the external environment, and consider what
-would happened if a web service that is used by a notebook changes the
-format of its output: notebooks now stop working, but nothing in
-RCloud has changed.
+strong.) At the other end of the scale, there is hardly any version
+control for the external environment, and if an external service 
+used by a notebook changes its API, the notebook can stop working.
 
@@
 \paragraph*{Caching}
-An important consideration is how and where to introduce caching in
+An important issue is how and where to introduce caching in
 RCloud's distributed computation model. Caching can dramatically
-improve performance in a way that is otherwise transparent to
+improve performance in a way that is transparent to
 application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
-Currently, RCloud offers no support for implicit caching, but since it
-supports the entirety of the R language, it's possible to incorporate
-caching in R libraries. Still, since the reuse of intermediate
-results is a recurring problem in data exploration environments, it
-seems better to implement this capability in a general purpose
-associative cache. We envision cache management being implemented in a
-new middle layer to be added in the future.
+Currently, RCloud offers no support for implicit caching, but 
+it's possible to incorporate caching in individual R libraries.
+The reuse of intermediate results is a recurring problem
+in many data exploration environments.
+We envision general cache management being implemented in a
+new middle layer in the future.
 
@@
 protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
-to remote procedure calls; this way, web browsers never directly instruct the
+to remote procedure calls; web browsers never directly instruct the
 RCloud backend to execute arbitrary code, preventing unauthenticated clients
@@
 access control for information security is delegated to the host
-operating system and web server. In practice, operating
-systems rely either on coarse permission models,
+operating system and web server.
+Most operating systems rely either on coarse permission models,
 which tend to be ineffective, or on detailed access control
@@
 and query languages that ensure differential privacy, are
-still research topics~\cite{Moore:2011:SAF}.
+open research topics~\cite{Moore:2011:SAF}.
 % We are hopeful that having a richer
@@
 is very fast and there is often no commitment: nothing gets saved by
-default, but clean up is automatic. In contrast, in RCloud's
+default, and clean up is automatic. In contrast, in RCloud's
 notebook interface, cells generated doing exploration are saved in the
 notebook. We arrived here from an insistence on reproducibility: it is a
-core principle of RCloud that no command should be run without being
+core principle that no command should be run without being
 saved. In our experience some history steps are only found to be
@@
 Surprisingly, we found that when users realize that \emph{scratch work
-is getting saved}, the extra versions become a mental burden for
-them. Clearly, RCloud is an extreme, and the right solution is
+is getting saved}, the extra versions become burdensome to them.
+Clearly, RCloud is an extreme, and the right solution is
 in the middle. An intriguing possibility,
@@
 analyses as it existed in S~\cite{Becker:1988:Auditing}, tying
-data artifacts to the processes that generated them. As a result,
-bookkeeping can be reduced to selecting the \emph{data} to be kept,
-and the code which generates it can be derived and saved transparently.
+data artifacts to the processes that generated them. 
+Bookkeeping would be reduced to selecting the \emph{data} to be kept,
+and the code that generated it would be derived and saved transparently.
 
@@
 forever. In addition, important metadata (has this notebook been
-recently run, or edited?) is also not easily searchable. In other words,
+recently run, or edited?) is not easily searchable. In other words,
 \emph{textual search} is easy, but \emph{relevance search} is what we
-need. Allowing users to \emph{tag} notebooks is natural, but it
+need. Allowing users to \emph{tag} notebooks is natural, but
 relies on users making a conscious effort, which goes against
 our ideal of not asking the user to do work for the system.
-It is better to have the system work for the user,
-relying on passive metrics, such as how often a notebook has been run.
-More generally, infrastructure for automatic \emph{curation} of
+% It is better to have the system work for the user,
+% relying on passive metrics, such as how often a notebook has been run.
+% More generally, infrastructure for automatic
+Automatic
+\emph{curation} of
 software artifacts generated by teams seems to be an important avenue

### 2015-06-25 paper.tex
submitted
figure squeezing
fixups

@@
-%\documentclass[journal]{vgtc}                % final (journal style)
-\documentclass[review,journal]{vgtc}         % review (journal style)
+\documentclass[journal]{vgtc}                % final (journal style)
+%\documentclass[review,journal]{vgtc}         % review (journal style)
 %\documentclass[widereview]{vgtc}             % wide-spaced review
@@
 %% OnlineID. Otherwise, you may safely leave it at ``0''.
-\onlineid{}
+\onlineid{341}
 
@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{Stephen North and Carlos Scheidegger and Simon Urbanek and
+\author{Stephen North, Carlos Scheidegger, Simon Urbanek and
   Gordon Woodhull}
@@
 %% insert punctuation at end of each item
+\item 
+Stephen North is with Infovisible LLC, Oldwick, USA, and graphviz.org.
+\\Email: s.c.n@ieee.org
+\item
+Carlos Scheidegger is with the University of Arizona, USA.
+\\Email: cscheid@email.arizona.edu
+\item
+Simon Urbanek is with AT\&T Labs in Bedminster, USA.
+\\Email: urbanek@research.att.com
+\item
+Gordon Woodhull is with AT\&T Labs in Bedminster, USA.
+\\Email: gordon@research.att.com
 }
@@
   collaborative data analysis, visualization and web deployment.
-  The biggest deployment of RCloud has been in active use for more than two years,
-  and has about fifty active users. We report on interviews with some of these users, and
-  discuss the design decisions, tradeoffs and limitations, comparing RCloud
-  to other current proposals.
+  RCloud has been in active use for over two years by about 100 active users.
+  We report on interviews with some of these users, and
+  discuss the design decisions, tradeoffs and limitations.
+  % comparing RCloud to other current proposals.
 } % end of abstract
@@
   \centering
-  \includegraphics[width=.95\linewidth]{fig/teaser/teaser.pdf}
+  \includegraphics[width=.80\linewidth]{fig/teaser/teaser.pdf}
   % \vspace{-.5em}

### 2015-06-25 system.tex
fixups

@@
 for example, which packages and data sets often appear together,
-in the style of Fast et al's Codex system.~\cite{Fast:2014:ECS}.
+% in the style of Fast et al's Codex system.~\cite{Fast:2014:ECS}.
+in the style of Codex.~\cite{Fast:2014:ECS}.
 

### 2015-06-25 discussion.tex
working

@@
 can be accessed through the notebook tree or through GitHub's web interface.
-But a problem is that it is easy to create a large number of artifacts
+But it is easy to create a large number of artifacts
 and the system does not help enough to organize and navigate them.
@@
 can realize benefits at least proportional to the work on the units.
-RCloud's notebook and versioning capabilities allow work to be divided
-into units as fine as the underlying language allows, and versioning
-encourages making incremental changes at low cost and without disrupting
+RCloud's notebook and versioning allow work to be divided
+into units as fine as the underlying language allows, and
+encourage making incremental changes at low cost and without disrupting
 the work of others.
@@
 and integrating them with external processes.  It is particularly
-advantageous for work to be shared as URLs that provide access to code
-and experiments, instead of by pasting screenshots into reports.
+advantageous for work to be shared as URLs that provide access to live
+notebooks, instead of by pasting screenshots into reports.
 
@@
 demand, we added user interface support for exporting plot images.
-We assume that RCloud notebooks should play well in the ecosystem
-of other tools, but sometimes it is difficult to know whether adding
+RCloud notebooks should play well in the R ecosystem,
+but sometimes it is difficult to know whether adding
 a feature for compatibility will offload complexity, or increase it.
@@
 being similar to VisTrails' enhanced recommendations by clustering multiple
-workflows).  It seems valuable to know which packages are frequently used
-together, or appear in proximity to a certain record types or data feeds.
-In general, trivial or passive mechanisms to collect data for
-recommendations are essential.
+workflows). It seems helpful to help users find which packages are frequently
+used together, or when using record types or data feeds.
+Overall, simple, passive mechanisms to collect data for recommendations are preferred.
 
 \paragraph*{Group management, size and diversity} This area needs better
-support in RCloud, but is clearly important to working in teams.
+support, and is clearly important to working in teams.
 We rely on external administrative processes and social conventions
@@
 
-\paragraph*{Curation} Even without formal group management, users may
+\paragraph*{Curation} Even without formal group management, users often
 organize groups of related notebooks using tree folders.
@@
 \paragraph*{Versioning and Upgrading}
-Some aspects of the programming environment are more difficult to
+Some aspects of the environment are more difficult to
 manage in RCloud than in conventional systems. RCloud does not separate
-the development and deployment environments, and every version of every
-prototype is shared by default. While this encourages
-collaboration, it also quickly exposes errors such as
-misconfiguration, mismatches between package versions, and
-programming errors, that can unintentionally affect production
-websites. For example, if an RCloud website is distributed across
-multiple servers, conflicting versions of system software
-can cause errors that are hard to debug.
+its development and deployment environments, and every version of every
+prototype is shared by default. Although this encourages
+collaboration, it also quickly exposes misconfigurations,
+mismatches between package versions, and programming errors
+that can unintentionally affect production websites.
+If an RCloud website is distributed across multiple servers,
+conflicting versions of system software
+can also cause errors that are hard to debug.
 
@@
 has rules to ensure reasonable consistency.
-(It is possible for RCloud to access information about package
-versions and configurations, but support for checking compatibility
-and maintaining multiple versions in the same environment is not
-strong.) At the other end of the scale, there is hardly any version
-control for the external environment, and if an external service 
+% (It is possible for RCloud to access information about package
+% versions and configurations, but support for checking compatibility
+% and maintaining multiple versions in the same environment is not
+% strong.)
+At the other end of the scale, there is basically no
+control over the external environment, and if an external service 
 used by a notebook changes its API, the notebook can stop working.
@@
 Our back-end environment, on the other hand, assumes a high degree of trust between users:
-access control for information security is delegated to the host
-operating system and web server.
+access control is delegated to the host operating system and web server.
 Most operating systems rely either on coarse permission models,
 which tend to be ineffective, or on detailed access control
-lists, which tend to be cumbersome. This means that information security
-in RCloud is a delicate, unsolved matter. More
-sophisticated approaches, such as information flow analysis,
-and query languages that ensure differential privacy, are
-open research topics~\cite{Moore:2011:SAF}.
+lists, which tend to be cumbersome.
+Information security in RCloud today is an unsolved matter.
+More sophisticated approaches are open research topics~\cite{Moore:2011:SAF}.
 % We are hopeful that having a richer
@@
 
-Surprisingly, we found that when users realize that \emph{scratch work
-is getting saved}, the extra versions become burdensome to them.
+% Surprisingly, we found that when users realize that \emph{scratch work
+We found that when users realize that \emph{scratch work
+is getting saved}, it sheer volume becomes burdensome.
 Clearly, RCloud is an extreme, and the right solution is
-in the middle. An intriguing possibility,
+probably in the middle. An intriguing possibility,
 suggested by one of our interviewees, is to implement auditing of data

### 2015-06-25 interviews.tex
working

@@
 Some users, who tried RCloud and were not able to continue for organizational
-reasons, miss certain capabilities. Leith ``[likes] the concept of being
+reasons, miss certain capabilities. Leith ``[likes] \comment{the concept of} being
 able to create notebooks and share them... A wiki is not the best way for
-communicating results - it's kind of like writing a blog post with very limited
+communicating results - it's \comment{kind of} like writing a blog post with very limited
 functionality. I have to save every picture and post it as an image.'' She said,
@@
 old fork stops working, and they complain.''
-
+% we don't need a new paragraph here - -scn
 Eric tried to keep his notebook tree well organized, but this didn't help,
-because some people kept old links in their email, or forked notebooks which 
-had become obsolete. By design, notebooks that are deleted in the user interface
-are not purged from the repository. He says he became reluctant to share notebooks:
-``Do I want to support this forever?''
+because some people kept old links in email, or forked notebooks which 
+had become obsolete.
+% By design, notebooks that are deleted in the user interface
+% are not purged from the repository.
+He says he became reluctant to share notebooks: ``Do I want to support this forever?''
 

### 2015-06-25 introduction.tex
working

@@
 In most organizations, development of analyses and visualizations 
-s done by \emph{hackers} and \emph{scripters} (adopting Kandel et al.'s
+is done by \emph{hackers} and \emph{scripters} (adopting Kandel et al.'s
 terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
 benefit from tools developed by hackers and scripters.
-Results of experiments are often shared first by copying screen
+Results of experiments are often first shared by copying screen
 shots or static output. By the time decisions are made based on
-the results of those screenshots, data and processes may have changed
-so much that the decision can be wrong.
-\stephen{Transitional sentence here.}
-Generally, the connection between these two worlds is made by IT staff,
-who port or rewrite code so it can run as a stable production service.
+those screenshots, the data and processes may have
+changed so much that the decision can be wrong.
+Moreover, the connection between EDA and deployment
+are made by porting or rewriting code so it runs as a stable service.
 In an environment where business needs can change rapidly, this
@@
 from production back to EDA.
-Our goal is to merge the worlds of EDA and production to close this gap.
+Our goal is to merge the worlds of EDA and deployment, to close the gap.
 
@@
 run experiments; annotate and deploy experiments as end-user websites or
-as reusable, callable services; and to share, search and recommend these artifacts. The artifacts are stored in a version
+as reusable, callable services; and to share, search and recommend these artifacts.
+The artifacts are stored in a version
 control system that provides a common workspace, as well as needed control
@@
 
-A key point is that, for the most part, the improved capabilities are
-available by default, without much explicit involvement on the part of
+For the most part, the improved capabilities are available by default,
+without much explicit involvement on the part of
 data scientists and other customers. Visual analytics experiments are

### 2015-06-25 system.tex
working

@@
 excessively fine-grained sequence of versions (see
-Section~\ref{sec:interviews}), we believe that the alternative of
-losing automatic sharing is worse (and see
-Section~\ref{sec:interviews} again for users reporting their
-satisfaction over the low-friction shareability).
+Section~\ref{sec:interviews}), 
+losing information that could be shared is worse (see
+Section~\ref{sec:interviews} user comments on
+satisfaction with low-friction shareability).
 
@@
 
-One side benefit of centralizing the execution and storage of
+A side benefit of centralizing the execution and storage of
 notebooks is that it becomes feasible to collect usage information
-that might be lost in a conventional environment.
+that is lost in conventional environments.
 In the case of social data visualization
 platforms, we would like to exploit usage data to help analysts
-find content of interest, whether the content is actual source code
-or accompanying data. The standard way to achieve this is through
+find content of interest, whether that content is source code
+or data. Standard ways to achieve this are through
 \emph{user-generated curation} and \emph{automatic recommendations}.
@@
 provided by companies such as Amazon and Netflix (``If you liked
-that film, then you should like this one too''). To make such
+that film, you'll like this one too''). To make such
 recommendations, we need users to \emph{curate} the collection, by
@@
 % in the style of Fast et al's Codex system.~\cite{Fast:2014:ECS}.
-in the style of Codex.~\cite{Fast:2014:ECS}.
+in the style of Codex~\cite{Fast:2014:ECS}.
 

### 2015-06-25 casestudies.tex
working

@@
 custom Javascript. In this case, the interactive features of ggvis
-(and, by extension, Shiny) are a subset of those in Vega~\cite{vega}.
-Custom interactions in LDAVis (such as hovering over a topic,
-topic cluster, or word) do not appear to be available yet in Vega~\cite{vega},
-although the required components for reactive interaction were
-recently described by Satyanarayan et
+and Shiny are a subset of Vega's~\cite{vega}.
+Custom interactions in LDAVis (like hovering over a topic,
+topic cluster, or word) are not available yet in Vega~\cite{vega},
+although the required components were recently described by Satyanarayan et
 al.~\cite{Satyanarayan:2014:DID}.
-Although custom Javascript was required, its flexibility is welcomed
+Although LDAVis required custom Javascript, that flexibility is welcomed
 by many web developers.
 
-LDAVis highlights some unique features of RCloud.
-While RCloud notebooks allow deployment of R analyses over
-the web with no additional effort, RCloud \emph{applications} are more
-powerful, and are developed with a combination of Javascript and HTML
-for the front end. This requires additional knowledge over Shiny, but
+LDAVis demonstrates some unique features of RCloud.
+While RCloud notebooks allow deployment of analyses over
+the web with no extra effort, RCloud \emph{applications} are more
+powerful, and are developed by combining Javascript and HTML
+for the front end. This requires more expertise than Shiny, but
 the RCloud model makes analysis simpler (since analysts
 simply write R in the style they already know)
-and front end visualization side is simpler for web
+and the visualization front end is simpler for web
 developers (since they simply write Javascript in
-the style \emph{they} know). In addition, RCloud applications
-inherit the automatic deployment and discoverability features of
-all RCloud notebooks.
+the style \emph{they} know).
+LDAVis inherits the automatic deployment and discoverability
+of all RCloud applications.
+% In addition, RCloud applications
+% inherit the automatic deployment and discoverability features of
+% all RCloud notebooks.
 

### 2015-06-25 introduction.tex
merge

@@
 solve many of these problems.
-Gutierrez collected interviews with data scientists~\cite{Gutierrez:2014:DSA},
-which include the following:
+Gutierrez's interviews with data scientists~\cite{Gutierrez:2014:DSA}, 
+include the following:
 
@@
 
-In view of the opportunity to improve this situation,
-we created a software environment that supports the end-to-end
-visual analytics process for individuals and teams and
-their work within larger organizations.
-
-In this paper, we contribute the design of RCloud, and an
-interview study based on its development and deployment
-at AT\&T Labs for about three years.
+In order to tackle these issues, we created RCloud, a software environment
+supporting the end-to-end visual analytics process for individuals and
+teams and their work within larger organizations. In this paper, we
+contribute the design of RCloud, and an interview study based on the
+course of its development and deployment at AT\&T Labs for about three
+years.
 
@@
 
-For the most part, the improved capabilities are available by default,
-without much explicit involvement on the part of
-data scientists and other customers. Visual analytics experiments are
-conveniently shared and turned into production websites, without moving
-or porting code. All the published artifacts in the system can be
-searched immediately. Recommendation is as easy as clicking a star
-on a useful notebook.
+One of the key design criteria for RCloud was \emph{transparency}: for
+the most part, the novel features are
+available and usable by default, without explicit involvement on the part of
+data scientists and other customers. Visualizations are
+shared and turned into production websites without moving
+or porting code; all artifacts present in the system can be
+searched at all times; recommendation is as easy as clicking a star
+on a useful workbook.
 
-The high level architecture of the prototype system
+The high level architecture of RCloud
 is shown in figure~\ref{fig:system}.
@@
 once a month. Most of the
-active users are members of AT\&T Labs, but some are data scientists and other
+active users are members of AT\&T Labs, but some are data scientists and
 collaborators in other business units.

### 2015-06-25 casestudies.tex
yes

@@
-\section{Case Study\label{sec:casestudy}: visual exploration of topic modeling in RCloud}
+\section{Case Study\label{sec:casestudy}: visual exploration of topic modeling}
 
@@
 combining a standard R library to fit LDA models using
-Gibbs sampling~\cite{} with custom R code written by the
-developers. The analysis module is a single R function
+Gibbs sampling with custom developed R code.
+The analysis module is a single R function
 exposed to the web application via the Javascript-R RPC

### 2015-06-25 system.tex

@@
 number (a ``hash''). This random number is then sent
-across the wire. This number is interpreted as an opaque identifier to
-a function. Because these identifiers are cryptographically safe, all
-that the Javascript layer can do is return them to the R side, in a
-message requesting that this function be called on its behalf. In that
-case, the results of this function call might include \emph{new}
-opaque identifiers, exposing new ``capabilities'' to the web browser.
+across the wire and interpreted as an opaque function identifier.
+Because these identifiers are cryptographically safe, all
+the Javascript layer can do is send them to the R side, in a
+message requesting a function call. 
+The result of this function call might include \emph{new}
+opaque identifiers, exposing new ``capabilities'' to the client.
 

### 2015-06-26 interviews.tex
topic sentences; trim boring stuff
in "evaluation" section
also remove all comments to make this easier to edit

@@
-\section{Interview Study\label{sec:interviews}}
+\section{Evaluation\label{sec:interviews}}
 
@@
 recent users of RCloud. 9 are data analysts, and 4 build tools for data
-analysts and business needs.
-% no paragraphbreak needed! scn
-The subjects differed in their evaluation of its benefits and disadvantages,
-as described below.
-
-%Analysts were more annoyed by the cell interface and its perceived
-%slowness than were web/tool developers. Analysts
-%switching to R from another language were more likely to use RCloud for
-%exploratory data analysis. Web/tool developers were more likely to see the
-%need for parameterization.
+analysts and business needs. Users felt that sharing works well, but the notebook
+interface slows down exploration, and the browsing interface is not scaling well.
+%gw (trying to be specific)
 
@@
 the subjects praised this feature.
-
-%% Wendy's usage is typical:
-%% \begin{quote}
-%% I've been sharing my notebooks in order for other people to see what
-%% I've done. It's very convenient for that purpose. [...] We haven't
-%% been editing them together, but we've been just sharing, and I've been looking
-%% at things that other people do. So there's no collaborative aspect of it, like
-%% developing the code, but there's sharing aspects for other people to see and for
-%% me to see other people's notebooks.
-%% \end{quote}
 
@@
 
-Eric develops packages for analysts, and uses RCloud ``more for sharing code with
+Eric develops packages for analysts, and uses RCloud ``for sharing code with
 other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
@@
 communicating results - it's \comment{kind of} like writing a blog post with very limited
-functionality. I have to save every picture and post it as an image.'' She said,
+functionality. I have to save every picture and post it as an image.'' He adds,
 ``I can't share all of the code because it would just get crowded and wouldn't
@@
 
-%% Kaylee: ``being able to share notebooks or codes, or collaborate on something,
-%% easily being able to share that information. That's an obvious advantage and
-%% nice to have.''
-
-%% Kenyon: ``I like being able to look at other people's notebooks.''
-
 \subsection{Forking}
-The ability to start work where someone else let off, by forking, was popular.
+The ability to fork someone's notebook and continue their work was popular, but it
+should probably not be the only way to change notebooks.
 Almost twice as many (131) users have forked
@@
 
-%% ``a lot of the time we just
-%% insert print statements here and there and check values. Having debugging code
-%% inside a functioning program is helpful for others to understand what
-%% you're doing.''
-
-%% Leith: ``gives the audience a way to try their own analysis, like maybe Leith
-%% should have used this parameter and they can just change the parameters''
-
 A common but ill-advised use of forking is to change
-parameters. Wendy says, ``I've been forking other people's notebooks [because] I
-want to run them on a different part of data, or I want to change some parts,
-I don't want to see this column, I want a different column, things like
-that.'' Eric complains that a notebook might say ```This is a report of
+parameters. Eric complains that a notebook might say ```This is a report of
 the volume of all of our feeds for this month', and someone would want to
@@
 Parameters can be added to a notebook URL, but adding user interface
-elements to do the same thing takes more expertise. Tool builders see a
-need for facilities to make that easier. Allison calls it ``web-enabling''
-notebooks: ``[Users] can always fork the notebooks and make changes, but I
-feel that if the owner of the notebook web-enables it, it's easier to
+elements to do the same thing takes more expertise. Tool builders told us it
+should be easier. Allison explains: ``[Users] can always fork the notebooks
+and make changes, but I
+feel that if the owner of the notebook [adds a user interface], it's easier to
 run. Instead of forking it, if they can set options, it's probably more
@@
 
-%% Lilo: ``For the things that I'll do multiple times, I'll try to parameterize
-%% it, and I'm working now on putting some packages together that will operate in
-%% some fashion outside of RCloud, straight from the command line''
-
 \subsection{Automatic source control}
-%% Kenyon: ``when I've finished something, there's a nice clean record of it.''
-%% (I think this may be more about the publish/markdown than about source control.)
-
-Automatic source control proved helpful. Lilo says, ``Instead of
+Automatic source control is helpful, but it also leads to many commits. Lilo says, ``Instead of
 looking back and saying I've got a billion files here in this subdirectory and I
 hope I've got them backed up, if they're on RCloud I know they are.''
-
-%%  [...] I
-%% don't have to worry about it, if I drop my computer in the bathtub, everythings
-%% gone, because I'm not great about uploading my own for versioning.''
 
@@
 your distribution version.
-%% None of that's necessary. A lot of web development is
-%% fixing stupid little bugs, and you can do that instantly without a lot of
-%% overhead, which is nice.
 \end{quote}
@@
 every direction I've tried.''
-%% Tagging these versions with names wouldn't
-%% necessarily help because ``namespaces are already crowded, so trying to remember
-%% the names of notebooks is hard, much less the names of states within each
-%% notebook.''
 
 \subsection{Discovering others' work}
-Many users find the search function valuable. Wendy says, ``The fact that we have
+Users reported that the search function is more helpful for learning about
+particular functions than techniques. Wendy says, ``The fact that we have
 all the notebooks there, searchable, saves me from replicating what other people
@@
 know. Kenyon says, ``Usually I know somebody's notebooks that I want to search
-through, because I know that the kind of thing I'm looking for is something
+through, because \comment{I know that} the kind of thing I'm looking for is something
 more obscure than I'm likely to find in some random person's.''
@@
 
-%% Allison: ``another thing I really like about RCloud is your publish feature, so
-%% that non-RCloud users, the anonymous login, non-RCloud users can access the
-%% notebooks without having to login to RCloud''
-
-%% accidental feature: ``some of my jobs are huge and I like to just call them and
-%% the fact that the user can just shut down his browser or shut down his computer
-%% and the job still runs in the background is actually a very neat thing. So what,
-%% um, the way I have coded is, whenever I have a huge job to run, I just display
-%% to the user, you go do whatever you want, we'll send you an email when the job
-%% is done. That's another feature of RCloud that's very helpful, that the job
-%% will continue to run in the background''
-
-\subsection{Exploring elsewhere}
+\subsection{Blocked by the web}
 
@@
 All analysts with R experience prefered to do analysis in another
-tool, then paste code into RCloud for sharing.
-It appears this is mainly due to familiarity.
-Eric asks, ``Why would I want to use
-RCloud over my current setup? If it's just me, I like my text editor and
-terminal. There's nothing that I want that those two don't give me.''
-
-%% Much of this is because of familiarity with other tools. Hugh: ``for other
-%% people who use RCloud vs RStudio, they will always think that they want to
-%% prototype on RStudio. Maybe because it's more familiar for them. So, they work
-%% on RStudio, on a small data set, and they fix up everything, make sure
-%% everything is right, and then they place it on RCloud for sharing.''
-
-%% Many of the analysts we spoke to preferred using a text editor to refine and
-%% store their work, pasting it to an R command line to try it out.
-
-%% But there are interface reasons as well. Kenyon describes a common workflow:
-%% \begin{quote}
-%% I have the discipline of having a file that describes what I'm doing, with the
-%% commands that I'm using, so that I can go back and recreate it or pass it on to
-%% someone else. So it's a little bit like having an RCloud notebook. It's not
-%% necessarily executable but most of the commands that I've typed are in there.
-%% \end{quote}
+tool, then paste code into RCloud for sharing, because the web interface got in the
+way of their work.
 
@@
 
-%% Running anything on the Web can also be a source of frustration. Poor network
-%% connections can cause web sockets to die, and Kaylee complains ``What frustrates me is
-%% that if I do leave this window for what feels like 5 or 7 minutes, it
-%% disconnects and I have to reconnect and run the whole notebook again.'' Wendy
-%% whines that ``being web-based, it's obviously slower than something you would be
-%% running on your machine.'' For Joy, even the little lag of sending a command
-%% to a server and receiving the result is intolerable
-
 Working in a shared environment also entails compromises about what you can
@@
 RCloud's notebook interface combines editable Markdown with a command line
-interface.
-The interface takes some learning. Lilo says,
-\begin{quote}
-%% It took me a couple of weeks of
-%% looking at it to become comfortable because 
-I saw the cells and it just threw me
-for a total loop. I mean, it's a good idea because [...] I can run it all at
-once if I want or I can break these into sections for either debugging or
-staging purposes. I really like it now, but when I first saw it, [it was] very
-confusing.
-\end{quote}
-
-The difference between a cell and a command prompt confuses some people.
-Kenyon explains:
-\begin{quote}
-If I was typing into one of the cells near that top, I had to think ``I'm editing
-this cell and then I have to execute it,'' and if I was editing one at the bottom,
-I could type it but then it would automatically execute [by pressing enter].
-%, then it became a standard cell and I have to edit it.
-\end{quote}
-
-%% For some users, the cells getting saved feels incompatible with exploratory data
-%% analysis. Kenyon notes that Markdown is a form of literate programming, and
-%% ``being literate about anything typically takes a lot of rewriting and going
-%% back over things. [...] I just want to explore a few things and then I'll know
-%% what I want to write.''
+interface. Many users felt that this format, so effective for presentation,
+is incompatible with exploratory data analysis.
 
@@
 Coby notes, ``It saves everything I do like everything is gold, but most
-of it is junk not meant to be saved.'' Kenyon says, ``I just want to type in a
-couple of quick commands and get some results that are going to tell
-me what to do next, and they're not necessarily archival in any sense.''
+of it is junk not meant to be saved.''
 
@@
 
-Users felt that cells do not capture the right level of granularity: they
-hold either too much code or too little. Kenyon says that RCloud's cell structure
+Users felt that cells do not capture the right level of granularity.
+Kenyon says that RCloud's cell structure
 ``tempts me to type a big long thing and then run the whole thing, as opposed to
@@
 so the only way to get that functionality is if every line's in one cell.''
-Another complaint was that cells can take up too much vertical space, requiring scrolling.
-% Rick complains, ``Every time I type more stuff, the notebook gets longer and
-% longer and it's harder to deal with.'' Ivan says that cells' controls and blank
-% space take extra screen space compared to a straight command line, and he and
-% KC both complained that when there are long results or plots, it causes the
-% code to scroll off the screen. Raif explains, ``Imagine I run a cell whose
-% output is huge. [...] To find the next cell, I have to scroll down and find
-% where that cell starts. So I'm losing the continuity of my code.''
 
-Some users would like a way to keep results separate from code.
+\subsection{Too vertical}
+The strict vertical linearity of a notebook also poses space and time problems
+not encountered on the command line, where typically only the last few lines are looked.
+
+Scrolling between code and results is time consuming, and some users would
+prefer to keep results separate from code.
 Gerrard thinks RStudio's layout is more helpful because charts
@@
 To avoid this pitfall, some users write code to explicitly cache results.
-% Coby ends up ``littering the notebook with little switches that comment out the
-% slow parts'' and either load or save the object to disk. One of the authors of this study manually writes cells that check if a result file exists, and perform a calculation only when needed.
-
-%% Kenyon on auditing (possible solution): ``you didn't see your mistakes, they were
-%% in there but you never really saw them because a mistake never led to an answer
-%% at the end'' ``the real result of this analysis was these three plots, so go
-%% back and figure out everything that I did that was involved in creating those
-%% three plots. So that I could start from the raw data and create those three
-%% plots.''
-
-%% RCloud combines the functions of scratchpad with notebook assets. Tina thinks
-%% that having a scratch pad that doesn't persist between different notebooks is
-%% ``useless''.
-
-%% The one-window design also makes some users feel that the side panes are taking
-%% up too much space. Gerrard: having ``long term'' code (which would be in assets)
-%% above the command prompt, like in RStudio, means that both can be sized very
-%% wide to accomodate long lines.
-
-%% Coby: Problem with screen real estate: side panes take up too much
-%% space. Doesn't need the notebook tree pane.
-
-%% Kaylee: In the previous version, it switched between code and results. Now you
-%% see both - ``and there are times when that's good, but there's certainly times
-%% when it'd be nice to - you know, you have 10 or 15 cells and you just want to
-%% get back and there's all this output and you're scrolling all over the place so,
-%% if there were a way to control both, that'd be great. There are times when I do
-%% want to see the output, but there are times when I've seen enough of the same
-%% thing.''
-
-%% Allison: ``It's good that I can see the results but if I'm working and I need
-%% more room and I want to hide the results, I can't.'' Could we have ``an option
-%% to close the results window?''
 
@@
 not all our users have learned the habits that make this a reality. As Joy
-puts it, there is still a big problem with ``bitrot'' -- notebooks often
+puts it, ``bitrot'' is still a big problem -- notebooks often
 stop working because the user changed the structure of their data, or
-changed a filename or a database. She says we need ``organizational
-protocols'' to catch up with the technology. Even if one forks someone
+changed a filename or a database. Even if one forks someone
 else's notebook and corrects it, the original notebook still exists with
-the error.
-
-%% At first, he had to learn how to write safe
-%% notebooks: ``It would be a file inside a directory that only I had [...] access
-%% to and so someone else couldn't run it, [...] a temporary file on something that
-%% was read-only to everyone else. But I've gotten careful about that.''
+the error. She says we need ``organizational protocols'' to catch up with the technology.
 
@@
 
-The problem of dead notebooks is compounded by changing
-packages and example notebooks at the same time. Eric will ``go back and
-change [a notebook] and update the page, but in the meantime people
-have forked it. I'll have to change a package as well, so their
-old fork stops working, and they complain.''
-% we don't need a new paragraph here - -scn
-Eric tried to keep his notebook tree well organized, but this didn't help,
-because some people kept old links in email, or forked notebooks which 
-had become obsolete.
-% By design, notebooks that are deleted in the user interface
-% are not purged from the repository.
-He says he became reluctant to share notebooks: ``Do I want to support this forever?''
+Eric tried to keep his notebook
+tree well organized, but this didn't help, because people kept old links
+in email, or forked notebooks which had become obsolete.
+He says he became scared to share notebooks: ``Do I want to support this forever?''
 

### 2015-06-26 discussion.tex
bibliography fixes
fixes

@@
 productively in teams.
-It is a benefit if the units can be kept small or granular, so team members
-can realize benefits at least proportional to the work on the units.
+If the units can be kept small or granular, team members
+can realize benefits at least proportional to their work on the units.
 RCloud's notebook and versioning allow work to be divided
@@
 \paragraph*{Nonforgetful, effortless exploration is hard}
-Users reported that RCloud's combined notebook and command line
-interface interferes with their familiar use of R for exploration. In
-current R practice, the command prompt is used to try things out. It
-is very fast and there is often no commitment: nothing gets saved by
-default, and clean up is automatic. In contrast, in RCloud's
-notebook interface, cells generated doing exploration are saved in the
-notebook. We arrived here from an insistence on reproducibility: it is a
-core principle that no command should be run without being
-saved. In our experience some history steps are only found to be
-important in hindsight, and deleting history by default makes it too
-easy to lose those steps by accident.
+Users reported that RCloud's interface clashes with the
+typical use of R for exploration.
+In current R practice, the command prompt is used to try things out.
+It is very fast and there is often no commitment: nothing gets saved by
+default, and clean up is automatic. In contrast, in RCloud,
+cells generated during exploration are saved in notebooks.
+We arrived at this from an insistence on reproducibility: 
+no command should be run without being saved.
+In our experience, some history steps are only found to be
+important in hindsight, and deleting history by default makes
+it too easy to lose those steps by accident.
 
@@
 We found that when users realize that \emph{scratch work
-is getting saved}, it sheer volume becomes burdensome.
+is getting saved}, its sheer volume becomes burdensome.
 Clearly, RCloud is an extreme, and the right solution is

### 2015-06-26 relatedwork.tex
copy editing

@@
 the need for an environment that enables better sharing
-of the knowledge, tools and processes to do this. Anecdotally
+of wrangling knowledge, tools and processes. Anecdotally
 we find much frustration among practitioners that this knowledge
@@
 \emph{cells}, each a snippet of source in one of the supported languages (typically R, but Python and others are also supported) or Markdown.
-The main creation workflow involves editing notebooks, which are transparently stored as git repositories in GitHub, providing us with easy access primitives for version tracking.
+The main creation workflow involves editing notebooks, which are transparently stored as git repositories in GitHub, providing us with easy access to primitives for version tracking.
 Notebooks can be executed as they're edited (left), or in a standalone viewer (right), via a slightly different URL.
@@
 \cite{Fowler:2006:Continuous}) is a trend in web services and similar
-fields. By making it convenient for data scientists to expend just a
-little additional effort when creating experiments, we may be able to
-eliminate the need for programming teams to recreate their work to
-deploy it in production, which has a high cost in time, expense and
-accuracy.
+fields.
+%% By making it convenient for data scientists to expend just a
+%% little additional effort when creating experiments, we may be able to
+%% eliminate the need for programming teams to recreate their work to
+%% deploy it in production, which has a high cost in time, expense and
+%% accuracy.
+%gw{sentence was just too long}
+When programming teams recreate the work of data scientists to
+deploy it in production, it has a high cost in time, expense and
+accuracy. By making it convenient for data scientists to productionize
+their experiments, we may be able to eliminate the need for this costly step.
 

### 2015-06-26 system.tex
copy editing

@@
 provides a low-friction path from experimental data analysis scripts
-to automated \emph{web services}. 
-essentially a remote function call over
-the web, which can be invoked by higher-level tools. An example of
-this sort of progression is described in Section~\ref{sec:casestudy}.
+to automated \emph{web services}: essentially a remote function call over
+the web which can be invoked by higher-level tools.
+%An example of
+%this sort of progression is described in Section~\ref{sec:casestudy}.
 
@@
 and forking. This provides the additional benefit that every RCloud
-notebook can also be manipulated like a git repository. Advanced users
-then have access to features like command-line checkout, history
-editing, and so on.
+notebook can also be manipulated like a git repository, granting advanced users
+access to features like command-line checkout and history
+editing.
 
@@
 the server-side of RCloud access to Javascript functions. This allows
-R libraries to respond to user input in the browser, giving them
+R libraries to request user input in the browser, giving them
 access to features ranging from password prompts, to the currently
-selected set of points in an interactive linked brushing
-visualization.
+selected set of points in an interactive visualization.
 
@@
 considerable flexibility, so that for example, a chart built with dc.js
-or leaflet.js can call back to analysis functions in R without having
-to define an additional protocol between the processes. From the
-Javascript side, an RPC call into the R process is just another HTTP
-request. From the R process, a call into Javascript looks like just
-another function call.
+or leaflet.js can call analysis functions in R without having
+to define an additional protocol between the processes. Calls in either
+direction look like ordinary function calls.
+%% gw: inaccurate, it's websocket; we still don't support HTTP into a live process
+%% From the
+%% Javascript side, an RPC call into the R process is just another HTTP
+%% request. From the R process, a call into Javascript looks like just
+%% another function call.

### 2015-06-26 conclusion.tex
chaff

@@
 Notebook sharing and publishing were eagerly adopted.
-Features for single-user data exploration, on the other hand, that
+On the other hand, features for single-user data exploration that
 compete with existing mature tools, were not readily accepted.
@@
 
-RCloud is open-source, and available at \url{github.com/att/rcloud/}.
+RCloud is available at \url{github.com/att/rcloud/}.

### 2015-06-26 discussion.tex
remove comments
chaff

@@
 \end{table}
-% SN says Tableau: versioning=no  collaboration=yes  deployment=yes multilanguage=no  reports=yes  analysis=no
 
@@
 its development and deployment environments, and every version of every
-prototype is shared by default. Although this encourages
+notebook is shared by default. Although this encourages
 collaboration, it also quickly exposes misconfigurations,
@@
 that can unintentionally affect production websites.
-If an RCloud website is distributed across multiple servers,
-conflicting versions of system software
-can also cause errors that are hard to debug.
-
-Versioning of software components must be managed in several
-levels: in data analysis and visualization scripts (the
-notebooks themselves); in the R environment such as the installed
-libraries and packages; and in the external environment such as the
-operating system, libraries, and protocols spoken by remote
-services. At one end of the scale, we have full control over notebooks
-via git, along with conventions to fix stable notebook versions.
-The R environment is under the control of its package manager that
-has rules to ensure reasonable consistency.
-% (It is possible for RCloud to access information about package
-% versions and configurations, but support for checking compatibility
-% and maintaining multiple versions in the same environment is not
-% strong.)
-At the other end of the scale, there is basically no
-control over the external environment, and if an external service 
-used by a notebook changes its API, the notebook can stop working.
+%% If an RCloud website is distributed across multiple servers,
+%% conflicting versions of system software
+%% can also cause errors that are hard to debug.
+Versioning of software components must be managed in several levels: in
+the notebooks themselves, in the installed libraries and packages of the
+R environment, and in the external environment such as the operating
+system, libraries, and protocols spoken by remote services.
+%% At one end of the scale, we have full control over notebooks
+%% via git, along with conventions to fix stable notebook versions.
+%% The R environment is under the control of its package manager that
+%% has rules to ensure reasonable consistency.
+%% % (It is possible for RCloud to access information about package
+%% % versions and configurations, but support for checking compatibility
+%% % and maintaining multiple versions in the same environment is not
+%% % strong.)
+%% At the other end of the scale, there is basically no
+%% control over the external environment, and if an external service 
+%% used by a notebook changes its API, the notebook can stop working.
 
@@
 %recommendation systems, and navigation of the set of notebooks, etc.
-
-\paragraph*{Caching}
-An important issue is how and where to introduce caching in
-RCloud's distributed computation model. Caching can dramatically
-improve performance in a way that is transparent to
-application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
-Currently, RCloud offers no support for implicit caching, but 
-it's possible to incorporate caching in individual R libraries.
-The reuse of intermediate results is a recurring problem
-in many data exploration environments.
-We envision general cache management being implemented in a
-new middle layer in the future.
 
@@
 protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
-to remote procedure calls; web browsers never directly instruct the
+to remote procedure calls: web browsers never directly instruct the
 RCloud backend to execute arbitrary code, preventing unauthenticated clients
 from making unauthorized calls to the RCloud runtime environment.
-%
 Our back-end environment, on the other hand, assumes a high degree of trust between users:
@@
 More sophisticated approaches are open research topics~\cite{Moore:2011:SAF}.
-% We are hopeful that having a richer
-% environment for collaboration and information sharing will encourage
-% new approaches to information security in visual analytics.
-%\gordon{This sounds like we think we are helping by pushing the
-%system to the breaking point, is that the idea?}
 
@@
 
-% Surprisingly, we found that when users realize that \emph{scratch work
 We found that when users realize that \emph{scratch work
@@
 and the code that generated it would be derived and saved transparently.
-
-%% . This is why we did not implement a simple command line
-%% as many users request. We could imagine making an exception for commands
-%% that have no side-effects, but in the general case this is no easier than
-%% auditing because R is not a purely functional language.
-
-%% The interface is also perceived to be slower, both for computer and for
-%% human. First, there is higher latency when commands run, because
-%% RCloud saves a cell to a repository before executing it. It
-%% may be possible to safely write to the repository behind execution of the
-%% same code. Second, and perhaps more detrimental, the interface requires
-%% switching between the keyboard and mouse, which can be cumbersome.
-%% %\gordon{we don't need to back that up, right? i hate using passive voice.}
-%% Shortcut keys can defined as a workaround, but it would be better
-%% to reduce the need for them.
-
-%% Other complaints mostly focus on layout, and can probably be addressed by
-%% offering better customization, and designs such as tabbed views.
 
@@
 our ideal of not asking the user to do work for the system.
-% It is better to have the system work for the user,
-% relying on passive metrics, such as how often a notebook has been run.
-% More generally, infrastructure for automatic
 Automatic

### 2015-06-26 interviews.tex
remove comments
use same inline headlines
a few lines, hey
also improve some topic sentences

@@
 
-\subsection{Sharing of results}
-Sharing of notebooks is the core feature of RCloud, and a popular one. All
+\paragraph*{Sharing of results} is the core feature of RCloud, and a popular one. All
 the subjects praised this feature.
@@
 found by navigating the notebook tree, or by searching. However, users most
-often mentioned sharing by sending links through email. Lilo says,
-``If my supervisor wants to see what I've done or QA it, I can just send her a link.''
+often mentioned sharing by sending links through email.
 
@@
 
-\subsection{Forking}
-The ability to fork someone's notebook and continue their work was popular, but it
-should probably not be the only way to change notebooks.
+\paragraph*{Forking} The ability to fork someone's notebook and continue their
+work proves to be a useful feature, but it is a lousy way to change simple parameters.
 Almost twice as many (131) users have forked
@@
 Although we intended forking to be available to improve others' code, we
-didn't anticipate forking one's own notebooks, which
+didn't anticipate users forking their own notebooks, which
 proved very useful. Kenyon says, ``I fork my own notebooks because I'm going off
@@
 
-\subsection{Automatic source control}
-Automatic source control is helpful, but it also leads to many commits. Lilo says, ``Instead of
+\paragraph*{Automatic source control} is appreciated for its safety and ease,
+but the large number of commits can be hard to manage. Lilo says, ``Instead of
 looking back and saying I've got a billion files here in this subdirectory and I
@@
 
-\subsection{Discovering others' work}
-Users reported that the search function is more helpful for learning about
+\paragraph*{Discovering others' work} Users reported that the search function is
+more helpful for learning about
 particular functions than techniques. Wendy says, ``The fact that we have
@@
 have done.''
+
 But other users prefer to browse just the notebooks of experts they
@@
 
-\subsection{Integrated analysis}
-Integrating an analysis language into a web development environment is something
-tool developers really appreciate. The structure of Hugh's visualization
-notebook means
+\paragraph*{Integrated analysis} Integrating an analysis language into a web
+development environment is something tool developers really appreciate.
+The structure of Hugh's visualization notebook means
 \begin{quote}
@@
 
-Integrating R helps Allison's application: ``Having an open session where I can run R commands or
+Integrating R makes Allison's application development easier:
+``Having an open session where I can run R commands or
 functions without having to invoke an API or send a request and then
@@
 
-\subsection{Blocked by the web}
-
-Although most analysts appreciated RCloud's sharing features,
+\paragraph*{Web is limiting} Although most analysts appreciated RCloud's sharing features,
 it was not a popular tool for exploratory data analysis.
@@
 
-\subsection{Cells versus the command line}
-RCloud's notebook interface combines editable Markdown with a command line
+\paragraph*{Impermeable cells} RCloud's notebook interface combines editable
+Markdown with a command line
 interface. Many users felt that this format, so effective for presentation,
@@
 
-\subsection{Too vertical}
-The strict vertical linearity of a notebook also poses space and time problems
-not encountered on the command line, where typically only the last few lines are looked.
+\paragraph*{Too vertical} The vertical orientation of a notebook also poses problems
+not encountered on the command line, where typically only the last few page of output matters.
 
@@
 
-\subsection{A sea of notebooks}
-RCloud is becoming a victim of its own success, as it is becoming
+\paragraph*{A sea of notebooks} RCloud is becoming a victim of its own success, as it is becoming
 difficult to navigate all the notebooks. There are over 5200 notebooks in
@@
 He says he became scared to share notebooks: ``Do I want to support this forever?''
-
-%% Iris: ``One UI notebook and a bunch of other notebooks that do various tasks and
-%% are called with API calls. The problem is, if you clone all the notebooks, then
-%% you have to go update every single notebook, because all the IDs are different
-%% now. There's no sort of relative path type stuff that you can do. You can do it
-%% with static assets, but you can't do it with Ajax calls''

### 2015-06-26 relatedwork.tex
EDA in JavaScript is absurd

@@
 share Javascript programs, which means that deployment happens
-automatically through the web browser. If Javascript becomes
-the lingua franca of exploratory data analysis, we foresee
-building a simpler version of RCloud where all execution happens
-either on the client side or via web services. Unfortunately that is
-not realistic today, making these stand-alone
-solutions unsuitable for our purpose. Plot.ly is notable in that it
+automatically through the web browser. This provides a seamless
+way to share visualization techniques, but it does not help with
+analysis.
+%% If Javascript becomes
+%% the lingua franca of exploratory data analysis, we foresee
+%% building a simpler version of RCloud where all execution happens
+%% either on the client side or via web services. Unfortunately that is
+%% not realistic today, making these stand-alone
+%% solutions unsuitable for our purpose. 
+Plot.ly is notable in that it
 provides API support for publishing \emph{from} scripts: 

### 2015-06-26 interviews.tex
s/prefered/preferred/

@@
 it was not a popular tool for exploratory data analysis.
-All analysts with R experience prefered to do analysis in another
+All analysts with R experience prefeered to do analysis in another
 tool, then paste code into RCloud for sharing, because the web interface got in the

### 2015-06-26 paper.tex
working
fix abstract

@@
 \usepackage{amssymb}
+\newcommand{\comment}[1]{}
 
@@
   collaborative data analysis, visualization and web deployment.
-  RCloud has been in active use for over two years by about 100 active users.
+  About 100 people used RCloud for two years.
   We report on interviews with some of these users, and
-  discuss the design decisions, tradeoffs and limitations.
-  % comparing RCloud to other current proposals.
+  discuss design decisions, tradeoffs and limitations in
+  comparison to other approaches.
 } % end of abstract

### 2015-06-26 paper.tex
fix the title
work on abstract

@@
 
-\title{Collaborative Visual Analysis on the Web with RCloud}
+\title{Collaborative Visual Analysis with RCloud}
 % \title{Integrating Social Visual Analysis and Deployment with RCloud}
@@
 An overview of features in RCloud.
-RCloud supports an environment in which a large number of loosely-related problems in data and visual analytics are solved by a small number of \emph{hackers} and \emph{scripters}.
-In such a shared environment, \emph{discoverability} is a concern.
+RCloud is an environment in which \emph{hackers} and \emph{scripters}
+solve problems in data and visual analytics and create web applications
+and services.  It is a shared environment, in which \emph{discoverability} is a concern.
 RCloud supports search, annotation, recommendation, and commenting for all notebooks, and provides an overview where users can \emph{browse} popular and recent analyses.
-When problems and data sources change frequently, deployment can become very costly; RCloud supports \emph{transparent and automatic} deployment of analyses as web pages and web services, allowing a seamless transition from data exploration work to production web services.
+When problems and data sources change frequently, deployment can become very costly; RCloud supports \emph{transparent and automatic} deployment of analyses as web pages and web services, allowing a seamless transition from data exploration to production.
   }

### 2015-06-27 casestudies.tex
within page limits! still writing revision letter

@@
 \begin{figure*}
+  \vspace{-1.5em}
   \includegraphics[width=\linewidth]{fig/casestudytext/casestudytext.pdf}
-  \caption{\label{fig:textvis}An example application developed and deployed in RCloud.}
+  \vspace{-3.0em}
+  \caption{\label{fig:textvis}LDAVis, an example application developed
+    and deployed entirely in RCloud.  LDAVis allows analysts to
+    visually explore \emph{topic models}, statistical
+    representations of different topics in a document collection.
+  }\vspace{-1.5em}
 \end{figure*}

### 2015-06-27 discussion.tex
within page limits! still writing revision letter

@@
   \end{tabular}
-  \caption{Comparison of system features. Columns: Versioning/Forking, Collaboration, Deployment, Multi-language support, Integrated Reports, and Integrated Analysis.}
+  \caption{Summary of features for similar systems.
+The columns stand for, respectively, Versioning/Forking, Collaboration, Deployment, Multi-language support, Integrated Reports, and Integrated Analysis.}
 \end{table}
@@
 requirements were appropriate, and were met.
-
+%
 Our experience underscored the relevance of Heer and Agrawala's design
-considerations, and indicated areas for further exploration. We adopt
+considerations, and indicated areas for further exploration~\cite{Heer:2008:DCF}. We adopt
 their taxonomy in the following discussion.

### 2015-06-27 introduction.tex
accidentally duplicated text
within page limits! still writing revision letter

@@
 
-Results of experiments are often shared first by copying screen
-shots or static output. By the time decisions are made based on
-the results of that screenshot, data and processes may have changed
-so much that the decision can be wrong.
-
 \item Technology transfer.
@@
 
-Fortunately, there is abundant evidence that technology can
-solve many of these problems.
-Gutierrez's interviews with data scientists~\cite{Gutierrez:2014:DSA}, 
-include the following:
+%% Fortunately, there is abundant evidence that technology can
+%% solve many of these problems.
+%% Gutierrez's interviews with data scientists~\cite{Gutierrez:2014:DSA}, 
+%% include the following:
 
-\begin{itemize}
-\item (upon being given access to other code and data analysis, in
-order to learn about Hadoop) ``I could look at other people’s code
-and play with their code and data sets as well''
-\item (discussing the value of sharing prototypes rather than static
-data) ``Prototyping our products so that internal
-customers can use them early on has been crucial for
-our success. Now we can shoot off a URL to internal
-customers and it allows them to provide feedback way before
-we're talking about getting it into production.''
-\item (on sharing more than just a finished product) ``We also share
-exploratory analyses and reports so that we can still exchange
-knowledge even if the work didn't make it into a larger project.''
-\item (on changing analyses and processes) ``It is important to
-have testing frameworks so you can go back and test all
-of your data.''
-\end{itemize}
+%% \begin{itemize}
+%% \item (upon being given access to other code and data analysis, in
+%% order to learn about Hadoop) ``I could look at other people’s code
+%% and play with their code and data sets as well''
+%% \item (discussing the value of sharing prototypes rather than static
+%% data) ``Prototyping our products so that internal
+%% customers can use them early on has been crucial for
+%% our success. Now we can shoot off a URL to internal
+%% customers and it allows them to provide feedback way before
+%% we're talking about getting it into production.''
+%% \item (on sharing more than just a finished product) ``We also share
+%% exploratory analyses and reports so that we can still exchange
+%% knowledge even if the work didn't make it into a larger project.''
+%% \item (on changing analyses and processes) ``It is important to
+%% have testing frameworks so you can go back and test all
+%% of your data.''
+%% \end{itemize}
 

### 2015-06-27 revision_letter.tex
within page limits! still writing revision letter

\documentclass{article}
\usepackage{microtype}
\usepackage{color}
\title{Letter to Reviewers of Submission \#341}
\begin{document}
\maketitle

We thank the reviewers for their careful reading of the manuscript and
their suggestions. In this document, we summarize the changes we
performed to the original submission.

\paragraph*{Paper Length} We have reduced the length of the paper
to eight pages, removing only extended discussion in the limitation
and interviews sections. We achieved most of the space savings was
achieved by careful redesign of the figures and tables, and rewriting
of text.

\paragraph*{Section Lengths} The reviewers were concerned about the length
of the introduction. We have significantly reduced it.

\end{document}

### 2015-06-27 relatedwork.tex
fixed widow
within page limits! still writing revision letter

@@
 \includegraphics[width=.9\linewidth]{fig/system/system.pdf}
-\caption{\label{fig:system}A diagram of RCloud's architecture. }
+\caption{\label{fig:system}A diagram of RCloud's architecture.
+RCloud was specifically designed to leverage existing systems and standards, and so communication between most parts of the system happens through HTTP.
+As a result, some desirable features of RCloud become natural, such as native support for RCloud notebooks as web services.
+}
 \end{figure*}
@@
 \begin{figure*}
+  \vspace{-1.5em}
 \centering
 \includegraphics[width=.95\linewidth]{fig/notebook/notebook.pdf}
+  \vspace{-1.5em}
 \caption{\label{fig:notebook}An RCloud notebook is a sequence of
@@
 }
+  \vspace{-1.5em}
 \end{figure*}
@@
 accuracy. By making it convenient for data scientists to productionize
-their experiments, we may be able to eliminate the need for this costly step.
+experiments, we may be able to eliminate the need for this costly step.
 

### 2015-06-27 introduction.tex
it's only the right column of the 8th page.

--- 2015-06-25 introduction.tex
@@
 
-Fortunately, there is abundant evidence that technology can
-solve many of these problems.
-Gutierrez's interviews with data scientists~\cite{Gutierrez:2014:DSA}, 
-include the following:
+Fortunately, there is evidence that appropriate technology might
+solve some of these problems.  Gutierrez's interviews with data
+scientists~\cite{Gutierrez:2014:DSA} include the following quotes:
 

### 2015-06-27 revision_letter.tex
revision letter

@@
 
+Throughout the text, we have attempted to clarify that although RCloud
+itself does not present a novel visual encoding (and isn't a pure
+visual analysis system either), we still believe it presents an
+important contribution to VAST. That's because we have found that, in
+our experience in deploying visualization solutions in a larger
+organization, \textbf{the infrastructure for deploying a large number of
+  changing, exploratory visualizations is itself the bottleneck for
+  visualization adoption}. Once we accept that visualization is a part
+of a wider data analysis infrastructure in a larger organization,
+\textbf{failure to tackle the larger issues will mean that visualization
+solutions simply do not get adopted. We want to avoid this future}.
+RCloud is an attempt to fix the problems we have found to be the most
+pressing, namely of managing a large number of changing visual
+analysis solutions deeply connected to statistical models and data
+access layers, all while maintaining easy deployment and portability.
+
+\section*{Specific changes}
+
 \paragraph*{Paper Length} We have reduced the length of the paper
@@
 and interviews sections. We achieved most of the space savings was
-achieved by careful redesign of the figures and tables, and rewriting
-of text.
+achieved by a redesign of the figures and tables, and rewriting
+the text.
 
 \paragraph*{Section Lengths} The reviewers were concerned about the length
-of the introduction. We have significantly reduced it.
+of the introduction. We have significantly reduced it. Unfortunately,
+space constraints meant that we could not add more details to the
+system section.
+
+\paragraph*{Comparisons} We have added the comparison to ManyEyes to Table 1.
+
+\paragraph*{Captions} We have added appropriately-sized captions to Figures 2 and 4.
+
+\paragraph*{Typos, missing references} We have corrected all missing references and typos.
 

### 2015-06-27 relatedwork.tex
it's only the right column of the 8th page.
clarify what "wrangling" modifies

@@
 the need for an environment that enables better sharing
-of wrangling knowledge, tools and processes. Anecdotally
+of knowledge, tools and processes for wrangling. Anecdotally
 we find much frustration among practitioners that this knowledge
@@
 \begin{figure*}
-  \vspace{-1.5em}
 \centering
 \includegraphics[width=.95\linewidth]{fig/notebook/notebook.pdf}
-  \vspace{-1.5em}
 \caption{\label{fig:notebook}An RCloud notebook is a sequence of
@@
 }
-  \vspace{-1.5em}
 \end{figure*}

### 2015-06-27 revision_letter.tex
revision letter: "clarified evaluation"
worked on revision letter

@@
 We thank the reviewers for their careful reading of the manuscript and
-their suggestions. In this document, we summarize the changes we
-performed to the original submission.
+their suggestions. As requested, we here summarize the changes
+made to the previous draft that was reviewed.
 
-Throughout the text, we have attempted to clarify that although RCloud
+Throughout the text, we attempted to clarify that although RCloud
 itself does not present a novel visual encoding (and isn't a pure
-visual analysis system either), we still believe it presents an
-important contribution to VAST. That's because we have found that, in
+visual analysis system either), it presents a valuable contribution
+to VAST. That's because we have found that, in
 our experience in deploying visualization solutions in a larger
@@
 solutions simply do not get adopted. We want to avoid this future}.
-RCloud is an attempt to fix the problems we have found to be the most
+RCloud is an attempt to fix the problems we found to be the most
 pressing, namely of managing a large number of changing visual
@@
 
-\paragraph*{Paper Length} We have reduced the length of the paper
-to eight pages, removing only extended discussion in the limitation
-and interviews sections. We achieved most of the space savings was
-achieved by a redesign of the figures and tables, and rewriting
+\paragraph*{Paper Length} We reduced our submission to eight pages,
+mainly cutting the extended discussion in the limitation
+and interviews sections. We achieved most of the space savings
+by a redesign of the figures and tables, and tightening
 the text.
@@
 \paragraph*{Section Lengths} The reviewers were concerned about the length
-of the introduction. We have significantly reduced it. Unfortunately,
+of the introduction. We significantly reduced it. Unfortunately,
 space constraints meant that we could not add more details to the
@@
 
-\paragraph*{Comparisons} We have added the comparison to ManyEyes to Table 1.
+\paragraph*{Comparisons} We added a comparison to ManyEyes to Table 1.
 
-\paragraph*{Captions} We have added appropriately-sized captions to Figures 2 and 4.
+\paragraph*{Captions} We added appropriately-sized captions to Figures 2 and 4.
 

### 2015-06-28 interviews.tex
"informally" interviewed
spellchecker pass and full revision letter

--- 2015-06-26 interviews.tex
@@
 it was not a popular tool for exploratory data analysis.
-All analysts with R experience prefered to do analysis in another
+All analysts with R experience preferred to do analysis in another
 tool, then paste code into RCloud for sharing, because the web interface got in the

### 2015-06-28 relatedwork.tex
spellchecker pass and full revision letter

--- 2015-06-27 relatedwork.tex
@@
 
-Heer and Agarwala identify many design considerations for
+Heer and Agrawala identify many design considerations for
 collaborative visual analytics~\cite{Heer:2008:DCF} that
@@
 When programming teams recreate the work of data scientists to
-deploy it in production, it has a high cost in time, expense and
-accuracy. By making it convenient for data scientists to productionize
-experiments, we may be able to eliminate the need for this costly step.
+deploy it in production, there is a high cost in time, expense and
+accuracy. By making it convenient for data scientists to take
+experiments into production, we may be able to eliminate the need for this costly step.
 

### 2015-06-28 revision_letter.tex
spellchecker pass and full revision letter

--- 2015-06-27 revision_letter.tex
@@
 
-\section*{Specific changes}
+\section*{Specific changes, by review criteria}
+
+We will answer the primary reviewer, point by point.
+
+\paragraph*{1. Strengths and weaknesses}
+We have kept most of the content of the interviews, but structured it
+differently to provide more context, as suggested by a reviewer. On
+the other hand, we have chosen to keep the details about the version
+control and execution, because they are central to the system design
+of RCloud, and they enable the collaboration modes we wish the system
+support. As we've argued above, inadequate infrastructure for sharing
+and searching for different analyses and visualizations is becoming
+the bottleneck in large organizations, and we believe our
+infrastructure offers a novel attack on the problem.
+
+\paragraph*{2. Is the work novel, incremental, or previously published? 3. Is this work relevant for VAST?}
+We accept that there are no novel visual encodings in RCloud; our goal
+was to provide supporting infrastructure for visual and exploratory
+analysis as a central tool for discovery in a larger organization. We
+believe there are many previously published VAST papers falling into
+this broader category, and would like to argue for our submission to
+be considered similarly.
+
+\paragraph*{4. Is related work adequately referenced?}
+Even though the primary reviewer did not request any changes on this
+point, later in the review (specifically in point 7) they mentioned a
+missing comparison to ManyEyes in Table 1. We have added this
+comparison.
+
+\paragraph*{5. Are the technical results sound?}
+We apologize for the lack of other examples, but we chose LDAVis as
+one application to showcase RCloud because, to put it completely
+frankly, we are not allowed to disclose the other applications being
+developed at AT\&T Labs. LDAVis has been given publication clearance
+\emph{and} is developed in RCloud. We understand that this is not a
+completely satisfactory solution, but again, we believe the
+fundamental contribution of RCloud is to reduce the barriers for
+exploratory visual analysis in larger organizations, rather than any
+specific visual encodings. Since RCloud is built on top of R and
+HTML5, any technologies spanning this subspace are usable: LDAVis uses
+D3, but one can similarly develop visual encodings using other
+techniques such as WebGL.
+
+\paragraph*{6. Is the exposition clear? 7. Should anything be deleted or condensed?}
+We have reduced the length of the introduction, as requested, but kept
+the system description to most of its current length. We believe that
+exposition to be important for the completeness of the manuscript.
+
+\paragraph*{8. Are the figures informative?}
+We have extended the captions of the requested figures.
+
+\section*{Other issues}
 
@@
 and interviews sections. We achieved most of the space savings was
-achieved by a redesign of the figures and tables, and rewriting
-the text.
+achieved by a redesign of the figures and tables, and rewriting the
+text.
 
@@
 
-\paragraph*{Comparisons} We have added the comparison to ManyEyes to Table 1.
-
-\paragraph*{Captions} We have added appropriately-sized captions to Figures 2 and 4.
+\paragraph*{Should anything be deleted or condensed?}
+We have added the comparison to ManyEyes in Table 1, as suggested by the review.
 

### 2015-06-28 paper.tex

@@
   requirements for systems built to support current data science teams
-  and users: \emph{technology transfer}, \emph{coexistence}, and
-  \emph{discoverability}. In addition, we contribute the design and
+  and users: \emph{discoverability}, \emph{technology transfer},
+  and \emph{coexistence}. In addition, we contribute the design and
   implementation of RCloud, a system that supports the requirements of

### 2015-06-28 revision_letter.tex

@@
 
-\paragraph*{Paper Length} We have reduced the length of the paper
-to eight pages, removing only extended discussion in the limitation
-and interviews sections. We achieved most of the space savings was
-achieved by a redesign of the figures and tables, and rewriting the
-text.
+\paragraph*{Paper Length} We reduced our submission to eight pages,
+mainly cutting the extended discussion in the limitation
+and interviews sections. We achieved most of the space savings
+by a redesign of the figures and tables, and tightening
+the text.
 
 \paragraph*{Section Lengths} The reviewers were concerned about the length
-of the introduction. We have significantly reduced it. Unfortunately,
+of the introduction. We significantly reduced it. Unfortunately,
 space constraints meant that we could not add more details to the
@@
 
-\paragraph*{Should anything be deleted or condensed?}
-We have added the comparison to ManyEyes in Table 1, as suggested by the review.
+\paragraph*{Evaluation} We have clarified the evaluation section and added summaries of the results.
 

### 2015-07-13 interviews.tex
remove reference to iotools/hmr
not properly anonymized

@@
 
-To evaluate the effectiveness of RCloud, we interviewed 13 current or
+To evaluate the effectiveness of RCloud, we informally interviewed 13 current or
 recent users of RCloud. 9 are data analysts, and 4 build tools for data
@@
 Eric develops packages for analysts, and uses RCloud ``for sharing code with
-other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
+other people, and for doing tutorials for'' the packages he develops.
 ``I want people to see how the package works, so I clearly want them

### 2015-07-27 casestudies.tex
it's only the right column of the 8th page.
kenny's topic model web tool edits

--- 2015-06-25 casestudies.tex
@@
 
-LDAVis helps non-experts to explore collections of
-short text documents, using topic modeling and
+The Topic Model Web Tool (TMWT) helps non-experts to explore collections of
+text documents using topic modeling and
 visualization. Topic modeling~\cite{Blei:2003:LDA},
-although powerful, often requires human guidance
-and interpretation~\cite{Sievert:2014:LAM}.
+although powerful, often requires human guidance in both the 
+data preprocessing step as well as the
+interpretation of topics once a model is fit to the data.
+%~\cite{Sievert:2014:LAM}.
 
@@
   \includegraphics[width=\linewidth]{fig/casestudytext/casestudytext.pdf}
-  \caption{\label{fig:textvis}An example application developed and deployed in RCloud.}
+  \caption{\label{fig:textvis}The Topic Model Web Tool (TMWT), an example application developed
+    and deployed entirely in RCloud.  TMWT allows analysts to
+    fit and visually explore \emph{topic models}, statistical
+    representations of different topics in a document collection.
+  }
 \end{figure*}
 
-LDAVis was developed by two technical staff members at
+TMWT was developed by two technical staff members at
 AT\&T Labs, originally in RStudio Shiny~\cite{RStudio:2013:SWA},
@@
 RCloud's assumption that \emph{all developed notebooks are automatically
-deployed} simplifies this for LDAVis.
+deployed} simplifies this for TMWT.
 
-LDAVis combines text analysis, dimensionality reduction and
-interactive visualization. Text analysis is performed by
-combining a standard R library to fit LDA models using
-Gibbs sampling with custom developed R code.
+TMWT combines text preprocessing, topic modeling, and
+interactive visualization. Text preprocessing is done
+using standard string manipulation functions in R as well as
+some functions from the tm package \cite{R:tm} to perform
+such steps as global substitutions, tokenization, stop word removal, 
+and stemming. Each of these preprocessing steps is controlled by the user
+by allowing him/her to make selections in dropdown menu or, in 
+some cases (such as defining the stop word list) uploading 
+a custom file. Topic modeling is performed by
+combining a standard R library to fit Latent Dirichlet 
+Allocation (LDA) models using
+Gibbs sampling with custom-developed R code. Last, the interactive
+visualization of the fitted model is done using the LDAvis R package
+\cite{R:LDAvis}, which aids topic interpretation.
 The analysis module is a single R function
@@
 
-Each text topic is a probability distribution over the all the
-words of a document. To expose patterns in the relationships
+Each topic under LDA is a probability distribution over the all the unique
+words in the collection of documents. To expose patterns in the relationships
 between topics, LDAVis combines
@@
 
-The result of the dimensionality reduction process is a
+The result of the dimensionality reduction process in LDAvis is a
 two-dimensional plot of the topic space, an example of which is shown in
@@
 and Shiny are a subset of Vega's~\cite{vega}.
-Custom interactions in LDAVis (like hovering over a topic,
-topic cluster, or word) are not available yet in Vega~\cite{vega},
+Custom interactions in LDAVis (like hovering over a topic or word, or adjusting the 
+rankings of topic keywords) are not available yet in Vega~\cite{vega},
 although the required components were recently described by Satyanarayan et

### 2015-07-27 interviews.tex
s/coby/claude/g

@@
 
-Coby also works with text files and commands, rearranging source files
+Claude also works with text files and commands, rearranging source files
 so active code is at the top. RCloud does not readily support this
-workflow, so Coby often shares work by pasting code into an RCloud
+workflow, so Claude often shares work by pasting code into an RCloud
 notebook after he's done.
@@
 purpose, so having RCloud persist commands in cells can be annoying.
-Coby notes, ``It saves everything I do like everything is gold, but most
+Claude notes, ``It saves everything I do like everything is gold, but most
 of it is junk not meant to be saved.''
@@
 hours to run. In this situation, the Run Whole Notebook button is dangerous.
-Coby reports that when converting his work to notebooks, he
+Claude reports that when converting his work to notebooks, he
 ends up with a lot of comments saying ``This cell takes a long time to run.''

### 2015-08-17 casestudies.tex
back to 8 pages. do not peek into the patch lest you'd like a lesson in sausagemaking

@@
 \begin{figure*}
-  \includegraphics[width=\linewidth]{fig/casestudytext/casestudytext.pdf}
-  \caption{\label{fig:textvis}The Topic Model Web Tool (TMWT), an example application developed
+  \centering
+  \includegraphics[width=.85\linewidth]{fig/casestudytext/casestudytext.pdf}
+  \vspace{-1em}
+  \caption{\small \label{fig:textvis}The Topic Model Web Tool (TMWT), an example application developed
     and deployed entirely in RCloud.  TMWT allows analysts to
@@
   }
+  \vspace{-1.5em}
 \end{figure*}
@@
 
-Each topic under LDA is a probability distribution over the all the unique
+Each topic under LDA is a probability distribution over all the unique
 words in the collection of documents. To expose patterns in the relationships

### 2015-08-17 paper.tex
Fixed author locations
back to 8 pages. do not peek into the patch lest you'd like a lesson in sausagemaking

@@
-\documentclass[journal]{vgtc}                % final (journal style)
+\documentclass{vgtc}                % final (journal style)
+\let\ifpdf\relax
 %\documentclass[review,journal]{vgtc}         % review (journal style)
@@
 \usepackage{graphicx}
-\usepackage{times}
 \usepackage{xspace}
@@
 \newcommand{\comment}[1]{}
+
+\usepackage{times}
 
@@
 %% This turns references into clickable hyperlinks.
-\usepackage[bookmarks,backref=true,linkcolor=black]{hyperref} %,colorlinks
-\hypersetup{
-  pdfauthor = {},
-  pdftitle = {},
-  pdfsubject = {},
-  pdfkeywords = {},
-  colorlinks=true,
-  linkcolor= black,
-  citecolor= black,
-  pageanchor=true,
-  urlcolor = black,
-  plainpages = false,
-  linktocpage
-}
 
@@
 %% OnlineID. Otherwise, you may safely leave it at ``0''.
-\onlineid{341}
 
 %% declare the category of your paper, only shown in review mode
-\vgtccategory{Research}
 
 %% allow for this line if you want the electronic option to work properly
-\vgtcinsertpkg
 
@@
 %% Paper title.
+
+\linespread{0.9575} % I hate myself
 
@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{Stephen North, Carlos Scheidegger, Simon Urbanek and
-  Gordon Woodhull}
+\author{Stephen North \thanks{e-mail: s.c.n@ieee.org}\\ %
+  \scriptsize Infovisible %
+  \and Carlos Scheidegger \thanks{e-mail: cscheid@cs.arizona.edu}\\ %
+  \scriptsize University of Arizona %
+  \and Simon Urbanek\thanks{e-mail: urbanek@research.att.com}\\ %
+  \scriptsize AT\&T Labs %
+  \and Gordon Woodhull\thanks{e-mail: gordon@research.att.com}\\ %
+  \scriptsize AT\&T Labs}
 
-\authorfooter{
-%% insert punctuation at end of each item
-\item 
-Stephen North is with Infovisible LLC, Oldwick, USA, and graphviz.org.
-\\Email: s.c.n@ieee.org
-\item
-Carlos Scheidegger is with the University of Arizona, USA.
-\\Email: cscheid@email.arizona.edu
-\item
-Simon Urbanek is with AT\&T Labs in Bedminster, USA.
-\\Email: urbanek@research.att.com
-\item
-Gordon Woodhull is with AT\&T Labs in Bedminster, USA.
-\\Email: gordon@research.att.com
-}
+%% \authorfooter{
+%% %% insert punctuation at end of each item
+%% \item 
+%% Stephen North is with Infovisible, Oldwick, USA, and graphviz.org.
+%% \\Email: s.c.n@ieee.org
+%% \item
+%% Carlos Scheidegger is with the University of Arizona, USA.
+%% \\Email: cscheid@email.arizona.edu
+%% \item
+%% Simon Urbanek is with AT\&T Labs in Bedminster, USA.
+%% \\Email: urbanek@research.att.com
+%% \item
+%% Gordon Woodhull is with AT\&T Labs in New York, USA.
+%% \\Email: gordon@research.att.com
+%% }
 
@@
   \centering
-  \includegraphics[width=.80\linewidth]{fig/teaser/teaser.pdf}
+  \includegraphics[width=.70\linewidth]{fig/teaser/teaser.pdf}
   % \vspace{-.5em}
-  \caption{
+  \caption{ \small
 An overview of features in RCloud.
@@
   }
-% \vspace{-.5em}
+ \vspace{-1.5em}
 }
